### 2024-06-10 docs/verticality_future_mkd.Rmd
starting intro revisions

---
date: "`r Sys.Date()`"
author: "Your Name"
title: "officedown template"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)

fp <- fp_par(
  text.align = "center", 
  padding.bottom = 20, padding.top = 120, 
  border.bottom = fp_border())

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
```

\newpage

# Introduuction

Vertical stacking of species from beneath the ground to the top of the canopy is an emblematic signature of biologically rich regions and a major driver of biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. In MacArthur's [-@macarthurPopulationEcologyWarblers1958] seminal work on verticality in warblers, he notes distinct partitioning of vertical niche space among species in a community. Since then, vertical stratification at local scales has been widely observed across diverse taxonomic groups [@bashamVerticalStratificationPatterns2023a]. At regional scales, patterns of community verticality vary across environmental gradients, generally increasing in wetter habitats [@bashamVerticalStratificationPatterns2023a; @scheffersIncreasingArborealityAltitude2013]. For example, verticality of frogs increases at higher elevations where cooler and wetter climates increase habitat suitability in the canopy relative to low elevations where hot and dry conditions compress frog communities toward the ground [@scheffersIncreasingArborealityAltitude2013]. Globally, verticality increases toward the warmer and wetter tropics, where vertical partitioning may facilitate high biodiversity in structurally complex forests [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019]. Despite widespread agreement that verticality influences patterns of biodiversity across spatial scales, global patterns of verticality and its environmental drivers, including climatic and vegetative factors, have not been comprehensively described for vertebrates.

## possible paragraph describing hypotheses about climate and verticality
- Vertical stratification -> higher richness because species can partition the canopy
- Therefore expect higher richness to be associated with increasing verticality
- However, climate can limits verticality
- We propose that climate limits verticality
- Where climate is suitable, canopy height will positively impact verticality
- where climate is not suitable, canopy height will have no impact on verticality

As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted. Changes in taxonomic composition of assemblages changes, are often associated with changes in functional composition. Consequently, widespread patterns of deborealization, tropicalization, and thermophilization have occurred in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. Space-for-time substitutions can be used to predict impacts of climate change on the functional composition of species assemblages. For example, the thermal tolerance of assemblages displays predictable variation across elevation gradients. As the climate warms and species' ranges shift toward higher elevations, the thermal tolerance of assemblages increases as cool-adapted species become locally extirpated and warm-adapted species immigrate into communities (a process known as thermophilization) [@feeleyClimatedrivenChangesComposition2020].

```{r}
birds = read.csv("./../data/derivative_data/species_data/birds_breedresident_spdat_eltonvert_forestsOnly.csv")
mammals = read.csv("./../data/derivative_data/species_data/mammals_spdat_moura_forestsOnly.csv")
amph = read.csv("./../data/derivative_data/species_data/amph_spdat_moura_forestsOnly.csv")
repts = read.csv("./../data/derivative_data/species_data/rept_spdat_moura_forestsOnly.csv")

nbirds = nrow(birds)
nmammals = nrow(mammals)
namph = nrow(amph)
nrept = nrow(repts)

```

Relationships between climate and verticality across spatial and temporal environmental gradients have led to predictions that verticality may decline as climates become warmer and drier (cite). While forest understories are protected from extreme climatic conditions, the canopy experiences larger fluctuations in temperature and humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. Under climatically stressful conditions, arboreal species may move to lower vertical niche positions, reducing community verticality. Temporary vertical shifts have been observed for decades across a variety of biomes [@macarthurPopulationEcologyWarblers1958; @bashamVerticalStratificationCollapses2020; @kosheleffTemperatureInfluenceActivity2009; @eppleyFactorsInfluencingTerrestriality2022]. For example, arboreal frogs in a Panamanian rainforest move toward the understory during the dry season to escape the hot and dry conditions in the canopy [@bashamVerticalStratificationCollapses2020]. Similarly, arboreal primates from the Americas to Madagascar increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022]. Yet how vertical niche shifts may alter global patterns of verticality as the climate warms is not well understood. Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe. We then examine how climate change may impact environmental conditions that have led to present day patterns of verticality.

# Methods

*infographic for methods*

```{r}
mammals = fread("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel_forestsOnly.csv") %>% 
  dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)

amphibians = fread("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv") %>% 
  dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)

mammals.arb.tropicalmoist = round(mean(mammals$p.arb[mammals$biome == "Tropical & Subtropical Moist Broadleaf Forests"]) * 100, 2)
```

## Assemblage verticality

We mapped the verticality of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection. We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. [-@dinersteinEcoregionBasedApproachProtecting2017]. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we only used ranges that covered breeding and resident ranges, because seasonality impacts the verticality of assemblages. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. This dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature. For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

Verticality exhibited a strong positive correlation with species richness. To examine spatial patterns of verticality independently of species richness, we claculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool) . We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

 To minimize impacts of the positive relationship between verticality and species richness on global patterns, we calculated the standard effect size (SES) of mean assemblage verticality, which determines whether verticality is higher (positive) or lower (negative) than expected by random chance given the species richness in each cell and the vertical strategies of species in each geographic realm.

## Environmental data

We obtained present and future global climate data from CHELSA v2.1 [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Climate velocity represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011]. Canopy height was optained from [@langHighresolutionCanopyHeight2023] and vegetaion dnesiy Vegetative complexity was calculated as the product of vegetation density [@crowtherMappingTreeDensity2015] and canopy height [@langHighresolutionCanopyHeight2023]. Elevation data was obtained from *data source*. All environemtnal data was resmapled to the resolution and projection of the assemblage data.

## Analysis

We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of vegetative and climatic paramaters on mean and SES verticality. Spatial models were fit using R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and approximates spatial Gaussian random fields using stochastic partial differential equation (SPDE)-based approach [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015]. To implement the SPDE approach, we constructed a mesh using fmesher with a 300 km minimum allowable distance between vertices. We fit five alternative models for each taxonomic group, using the same fixed environmental predictors and varying the form of the random-effects structure. The general model is described as

$$
y_s ∼ 𝜇_{s}, \\
𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + ⍺_g + ⍵_s), \\
⍵_s ∼ MVNormal(0,𝚺_⍵)
$$

where $y_s$ is the mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality at all locations, and $𝑓^{-1}$ represpents an inverse link function. $𝜲^{main}_s$ is a vector of main effect covariates, 𝛽 represents a vector of estimated corresponding main effect coefficients. We used a Gaussian distribution for SES verticality, while we used a Beta distribution with a logit link function for mean verticality because values are constratined between 0 and 1. Environmental covariates included canopy height, vegetation density, minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, precipitation of the wettest month, precipitation seasonality, and climate change velocity since the LGM. Fitting models, we log10 transformed climate velocity and precipitation of the driest quarter to increase the linearity of the relationship between predictor and response and scaled all predictor variable to have a mean of zero and a standard deviation of 1. $⍺_g$ represents an optional IID random intercept defined by group $g$ ($⍺_g ∼ Normal(0,𝜎^2_⍺))$. The five alternative models each used a different variable to define the random intercept: biome, realm, biorealm, ecoregion, and no random intercept. $⍵_s$ represents a spatial random field at location $s$, which represents additional spatial variation in the data not explained by the environmental covariates or random intercept. The estimated covariance matrix ($𝚺_⍵$) was constrained by the Matérn covariance function, which defines the decay rate of correlation with distance.

We compared alternative models using five-fold cross validation. Grid cells were randomly assigned to one of five folds. Each model was the run five times, training the model using four of the folds (80% of the data) and testing the model on the withheld fold (20% of the data). After fitting the model using the training data, performance of each training set was calculated as the sum of log-likelihood values for the test data. Alternative models were then compared using the total log-likelihood summed across the five folds for each model structure. Across taxonomic groups, the best performing model contained a random intercept defined by ecoregion along with a spatial random field.

## Add spatially varying coefficient for canopy height
We then compared the best performing model to a model that additionally included a spatially varying coefficient for canopy height because we expect that canopy height may have a weaker impact on verticality in regions where verticality is homogeneously low than in regions were verticality is high.

Using the best performing random-effects structure, we compared three alternative models that differed in their fixed effects structure: the the full model with all fixed effects, the full model without the polynomial term for maximum temperature of the warmest month, and the full model without climate velocity. The base model included vegetative and current climatic predictors (canopy height, vegetation density, minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the wettest month. All these variables are assumed to directly impact the prevalence of different vertical strategies. The second alternative model included a polynomial term for maximum temperature of the warmest month because verticality increases in the tropics, where it is warmer, but extremely high maximum temperatures may limit arboreal strategies. The third alternative model a

To validate the best performing model, we refit it using all the data and calculated quantile residuals using the DHARMa R package.

\*we examined VIFs for predictor variables, sequentially removed variables until all VIFs were equal to or less than 4. We chose variables based on a combination of which ones had the highest VIF factors and which ones we thought had the most direct impact on physiological limits to arboreality. While, this is slightly higher than the standard VIF cutoff, we thought that it was important to keep both maximum temperature of the warmest month and minimum temperature of the coldest month, as the relative influence of these variables is likely to change across latitudes, with maximum temperatures having a larger impact on verticality in the tropics and minimum temperatures having a larger impact on verticality in temperate and boreal regions. We excluded precipitation seasonality because it was strongly correlated with dry season precipitation, has a less direct impact on physiological limits to verticality than dry season precipitation, and varies between current and future climate conditions than dry season precipitation.\*

To examine potential impacts of climate change on assemblage verticality, we predicted models to future climatic data while holding vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density. From the future predictions, we subtracted predictions to the data used to train the model and mapped the resulting differences.

We define a mesh, specifying a cutoff of 300 km as the minimum allowed distance between mesh verticies to maintain computation efficiencty while allowing suffcient complexity in the spatial field.

Vegetative and climatic parameters included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the driest quarter, precipitation of the wettest quarter, precipitation seasonality, and climate change velocity. All predictors had variance inflation factors below 4. We first identified the optimal random effects structure for the model using five-fold cross validation and the sum of the log-likelihood values across folds. Five random effects structures were compared, with each containing a spatial random field calculated from a 300 km mesh. Four of the five structures contained one additional random intercept: biome, realm, biorealm, or ecoregion. Models fit with a spatial random field and ecoregion as a random intercept consistently performed the best.

Using the optimal random effects structure, we then compared fixed effects using five-fold cross validation and the sum of the log-likelihood values for each fold. Our base model consisted of vegetation parameters (canopy height and vegetation density) and present-day climatic conditions (maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the driest quarter, precipitation of the wettest quarter, and precipitation seasonality). We compared this model to one that additionally included climate change velocity, which has previously been shown to correlate with assemblage verticality (cite).

Using the optimal fixed effects structure, we refit the model with all points and validated the model by examining the normality and spatial distribution of quantile residuals. We then projected the models for mean and SES verticality for each taxonomic group to future climatic conditions and mapped the difference between predictions of future and present verticality.

Full models were checked using sanity() function.

We used simultaneous autoregressive (SAR) models with a spatial error term to model the impact of environmental factors on the SES of mean verticality. Environmental variables included mean elevation, vegetative complexity, climate velocity, mean annual temperature, maximum temperature of the warmest quarter, minimum temperature of the coldest quarter, diurnal temperature range, temperature seasonality, annual precipitation, precipitation of the wettest quarter, precipitation of the driest quarter, and precipitation seasonality. To reduce multicollinearity among predictors, we initially applied an automatic selection procedure that eliminated the variable with the highest variance inflation factor (VIF) until all VIFs were below three. This resulted in the loss of both maximum temperature of the warmest month and minimum temperature of the coldest month, while temperature seasonality and diurnal temperature range were retained. Since maximum and minimum temperatures exhibit greater change from present to future climate scenarios and are more likely to directly impact verticality, we substituted these two variables for temperature seasonality and diurnal temperature range. Although this adjustment increased VIF, all VIF remained below five. All selected variables for use in the models included mean elevation, vegetative complexity, climate velocity, maximum temperature of the warmest quarter, minimum temperature of the coldest quarter, precipitation of the wettest quarter, precipitation of the driest quarter, and precipitation seasonality.

Prior to modeling, independent variables were checked for linear relationships with the response variable and log10 transformed if the relationship exhibited a strong right skew. Due to zero values in precipitation of the driest quarter, values were transformed using log10(precipitation + 1). If relationships appeared quadratic, a quadratic term was added to the model. Independent variables were then scaled to a mean of zero and standard deviation of one. To determine the optimal neighborhood weights matrix, we visually examined the distance to which spatial autocorrelation occurred in teh response based on correlograms. We then identified different neighborhood scenarios based on distance from the focal cell. The smallest distance included the eight neighbors immediates surrounding the focal cell and subsequently increased to include successive rings of cells up to the maximum distance at which spatial autocorrelation was present in the response variable. For each distance scenario, we produced three row-standardized neighborhood weights matrices based on distance, inverse distance, and inverse squared distance [@verhoefSpatialAutoregressiveModels2018].

We produced simultaneous autoregressive (SAR) models with a spatial error term using the additive effects of all scaled environmental variables for each neighborhood weight matrix and calculated Moran’s I for each model to determine if spatial autocorrelation remained in the residuals. Models in which spatial autocorrelation remained were discarded. Based on the remaining models, we selected the neighborhood weights matrix as the one used in the model with the lowest AICc. Initial model exploration indicated that spatial autocorrelation was present using larger spatial neighborhoods. To reduce computational demands, we therefore based subsequent model selection on spatial neighborhoods from a radius of one grid cell to seven grid cells around the focal cell. Using the optimal neighborhood weights matrix defined by the lowest AICc, we compared all possible combinations of independent variables using the ‘dredge’ function from the MuMIN R package. Optimal weight matrices used inverse distance squared weighting. We then used a model averaging procedure to estimate coefficients and standard errors for all models within 2 \u0394AICc of the best model [@symondsBriefGuideModel2011ods]. For each model within \u0394AICc of the best model, we calculated the AICc weight (wAICc), which represents the probability that the given model is the best (*cite Burnham and Anderson*). Because wAICc values were similar across candidate models, we used full-model averaging, where coefficients for parameters that do not appear in a model are set to zero. A weighted average of coefficients across all models is then calculated to estimate the impact of the parameter on the response variable. We additionally calculated relative variable importance by summing wAICc in the set of models where the variable occurs (*cite Burnham and Anderson*). To estimate model fit, we calculated pseudo-R^2^ as the squared Pearson correlation between predicted and observed values [@kisslingSpatialAutocorrelationSelection2008].

To estimate changes in mean and SES verticality under future climate scenarios, we predicted the averaged model to future environmental conditions, while climate velocity, elevation, and vegetative complexity remained at the values used in the models of present verticality. Predictions were made using a linear model equation with the coefficients from the averaged SAR model (i.e. the predicted trends without the spatial signal) because it is not possible to estimate the the spatial signal without knowing the values of the response variable. To estimate changes in verticality, we therefore used the same method to predict verticality under present environmental conditions and subtracted these values from predicted future verticality. Negative values indicate declines in verticality and positive values indicate increases in verticality.

# Results

## Global patterns of verticality

#Figures to include
# verticality map
# verticality by ecoregion - could be ciruclar bar plot for each realm or biome and taxonomic group
# coefficient plot
```{r}
load("./../results/sdmTMB_models/amphibians_meanvert.RData")
m.amph.meanvert = m.final
load("./../results/sdmTMB_models/amphibians_sesvert.RData")
m.amph.sesvert = m.final

load("./../results/sdmTMB_models/reptiles_meanvert.RData")
m.rept.meanvert = m.final
load("./../results/sdmTMB_models/reptiles_sesvert.RData")
m.rept.sesvert = m.final

load("./../results/sdmTMB_models/mammals_meanvert.RData")
m.mammals.meanvert = m.final
load("./../results/sdmTMB_models/mammals_sesvert.RData")
m.mammals.sesvert = m.final

load("./../results/sdmTMB_models/birds_meanvert.RData")
m.birds.meanvert = m.final
load("./../results/sdmTMB_models/birds_sesvert.RData")
m.birds.sesvert = m.final

```

# change in verticality map

# Supplementary figures/tables
# estimated coefficients and confidence intervals from the best model for each taxa - include all model parameters: fixed effects, spatial paramters (matern range, marginal spatial standard, deviation), distribution parameters

Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (\@ref(fig:vert-map-global)). This pattern was largely driven by declines in the proportion of arboreal species toward the poles. Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical broadleaf forests (\@ref(fig:vert-biome)).

Despite widespread coherence of global verticality patterns among taxa and within biomes, regional variation occurred. Within the tropics, verticality was highest in moist broadleaf forests and lowest in dry broadleaf forests (\@ref(fig:vert-biome)). Mammals exhibited the highest mean verticality in moist forests, particularly in the Amazon, the Congo Basin, Madagascar, and tropical and subtropical Asian forests where an average of `r mammals.arb.tropicalmoist`% of species per assemblage exhibited arboreal strategies while the proportion of terrestrial species declined. Verticality of birds closely mirrored this pattern except in Madagascar where SES of mean verticality declined. While reptiles and amphibians displayed similar verticality patterns, there were notable differences. Verticality of reptile assemblages peaked in the rainforests of Indonesia and Madagascar and declined in the neotropics where the proportion of fossorial species increased *map proportion arboreality and fossoriality*. Amphibian verticality additionally increased in the Brazilian Atlantic Forest while declining in SE Asia relative to other vertebrate assemblages.

In temperate biomes, verticality was similar across forests and declined in less dense wooded habitats (e.g., savannas) (\@ref(fig:vert-biome)). Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes. Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes, peaking in savannas and savanna-woodland transition zones (\@ref(fig:vert-map-global)L, \@ref(fig:vert-biome)). However, temperate rainforests (e.g., the Pacific-Northwest of North America) were notable exceptions, exhibiting higher verticality than than expected by chance. Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (\@ref(fig:vert-map-global)F). In contrast, the proportion of fossorial reptiles declined in temperate regions. Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species. Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal gradient in the SES of verticality.

# Environmental drivers of verticality

Models of mean and SES verticality indicated some similarities between the environmental drivers of mean and SES verticality across taxonomic groups, yet noticeable differences remained.
We hypothesized that canopy height would have an overall positive effect on verticality, but that its effect would vary over space, being stronger in tropical regions and lower in temperate regions, which generally have few arboreal vertebrates despite the presence of tall canopies. All of the best models included a spatially varying coefficient for canopy height. Overall, canopy height exhibited a positive effect on verticality of birds and mammals and weak positive effect or no effect on verticality of reptiles and amphibians (\@ref(fig:coef-plt). Over space the impact of canopy height was generally strongest in the Andes and Madagascar

If there is a positive relationship between canopy height and verticality, then taller canopies have more arboreal species
If there is a negative relationship between canopy height and verticality, then shorter canopies have more arboreal species
If there is not relationship between canopy height and verticality, then height of the canopy doesn't matter, potentially because climatic or evolutionary factors determine verticality.

Across taxa, precipitation of the driest and wettest months positively influenced the SES of verticality, though the impact of dry season precipitation was noticeably stronger. 
Impacts of climate on mean verticality generally mirrored those on the SES of verticality, though effects were smaller relative to the impact of canopy height.

variable effects of climate and vegetation across taxa. Mean verticality of birds and mammals was most strongly influenced by canopy height, though responses to canopy height varied across space. Though climatic variables had

Overall, the mean and SES of verticality exhibited similar responses to environmental variables across taxonomic groups. Verticality was positively associated with precipitation of the driest and wettest months and exhibited a quadratic relationship with maximum temperature of the warmest month for mean and SES verticality of birds and reptiles and SES verticality of mammals. The quadratic relationship was characterized by an initial increase in verticality followed by a decline in verticality at extremely high maximum temperatures. Mean verticality of mammals exhibited a relatively small positive relationship with maximum temperature and amphibians  except for amphibians, for which SES of verticality was negatively associated with maximum temperatures. Minimum cold temperatures positively impacted verticality of mammals, such that verticality increased as minimum temperatures increased, while having little impact on birds and mammals. Impacts of climate velocity also varied, negatively impacting birds, mammals, and amphibians, while having a small positive effect on reptiles.

*For difference section - current geographic niche underfilling and future potential to fill unused geographic space by filling currently unoccupied volume in the canopy as climate permits these habitats to become more suitable*

Though arboreal habitats remain in temperate biomes, the entire habitat volume is not occupied. This could be caused by arboreal habitats being climatically unsuitable in temperate regions and lead to a breakdown of the relationship between richness and volume. Overall, these patterns indicate a decline in the volume of habitable space toward the poles despite the structural presence of habitat.

-

    ## climate velocity represents climate instability across geographic timescales and negatively impacted verticality across vertebrate classes

-   Elevation

-   Vegetative complexity - selective logging could negatively impact verticality by reducing structural complexity

-   Decline in verticality in temperate regions despite the avialability of arboreal habitat space indicates physiological challenges of arboreal strategies in seaosnally cold and/or dry environmentals (Oliveira and Scheffers 2019)

-   for some vertebrate classes, fossoriality offers refuge from climate extremes

*plot richness by area x structural complexity and color by latitude* *then plot richness x latitude* *consider adding correlation of verticality with species richness and discuss here*

```{r}
load("./../results/amphibians_sar_vertmean2.RData")
amph.avgmod = vert.avg
amph.summ = summary(amph.avgmod)
amph.summ = round(amph.summ$coefmat.full, 2)
colnames(amph.summ) = c("Estimate", "SE", "z", "p")

load("./../results/mammals_sar_vertmean2.RData")
mammals.avgmod = vert.avg
mammals.summ = summary(mammals.avgmod)
mammals.summ = round(mammals.summ$coefmat.full, 2)
colnames(mammals.summ) = c("Estimate", "SE", "z", "p")

load("./../results/reptiles_sar_vertmean2.RData")
rept.avgmod = vert.avg
repts.summ = summary(rept.avgmod)
repts.summ = round(repts.summ$coefmat.full, 2)
colnames(repts.summ) = c("Estimate", "SE", "z", "p")

load("./../results/birdselton_sar_vertmean.RData")
birds.avgmod = vert.full
birds.summ = summary(birds.avgmod) # make sure spatialreg is loaded for summary to produce correct results
birds.summ = round(birds.summ$Coef, 2)
colnames(birds.summ) = c("Estimate", "SE", "z", "p")

```

# Environmental drivers of verticality

## Climate

Global patterns of verticality could largely be explained by environmental drivers. To examine the impacts of climate, topography, and vegetation on global patterns of verticality, we used simultaneous autoregressive (SAR) models with a spatial error term. The best models explained explained 88-98% of the global variance in the mean and SES of mean verticality across classes (\@ref(fig:model-coefs)). However, the spatial signal in all models was strong (lambda \> 0.9), which resulted in relatively small impacts of environmental variables as well as a lower pseudo-R^2^ value when calculated based only on the predicted trends without the spatial signal.

Precipitation of the driest quarter and minimum temperatures of the coldest quarter had the largest effects on the mean and SES of verticality across vertebrate classes, both exhibiting positive effects barring the impact of minimum temperatures on amphibians, which was insignificant ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)). Maximum temperatures of the warmest quarter had opposing effects on endotherms and ectotherms, negatively impacting verticality of reptiles and amphibians while generally having positive or insignificant effects on birds and mammals, except at the highest temperatures where verticality declined ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)). Arboreality is therefore primarily constrained by cold minimum temperatures and limited precipitation during the dry season, while high maximum temperatures additionally limit arboreal strategies in ectotherms. The vertical climate gradient from beneath the ground to the top of the canopy drives these global constraints on arboreality. Greater climatic fluctuations in the canopy coupled with water limitation pose physiological challenges to arboreal lifestyles. For example... In contrast, terrestrial and fossorial habitats are buffered from climate extremes, which facilitates species persistence where hot, cold, or dry conditions may prevent the occurrence of arboreal lifestyles.

Dry conditions negatively impact arboreal lizard, particularly under simultaneously hot environments. <https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1301>

```         
- Though arboreal habitats can exhibit average temperatures over 2C higher than terrestrial habitats (Scheffers et al. 2013) critical thermal maxima of arboreal species may not substantially exceed that of terrestrial species (Carilo Filho et al. 2021)
```

## Historical Climate - climate velocity

Verticality is also generally higher in regions that have experienced historically stable climatic conditions, as indicated by the significant negative relationships between climate velocity and the mean verticality of birds, mammals, and amphibians and SES of verticality for mammals, as well as the absence of positive relationships for any taxa. Low climate velocity (i.e., climate stability) generally increases speciation rates and decreases extinction rates (Colville et al. 2020). In tropical forests, complex vegetation structure characterized by multilayered forests (Terborh 1985) increases opportunities for diversification among arboreal niches (see Colville et al. 2020 11-13). For example, arboreal lineages of reptiles and frogs generally exhibit higher diversification rates [@liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017], while fossorial strategies have been associated with reduced speciation and increased extinction rates in snakes [@cyriacDiggingTheirOwn2018].

-   climate stability may therefore provide the conditions necessary for diversification in regions where climatic conditions are suitable for arboreal strategies, while increased dspersal capacity of arboreal species throughout 3D arboreal environments facilitates persistence in climatically unstable regions (Sandel et al. 2011; Scheffers et al. 2017).

```{r}
bird.meanvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_meanvert.RData")
bird.meanvert = pred.df
bird.meanvert$vertvar = "mean"
bird.sesvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_sesvert.RData")
bird.sesvert = pred.df
bird.sesvert$vertvar = "ses"
bird.pred = bind_rows(bird.meanvert, bird.sesvert) %>% mutate(class = "bird")

mammals.meanvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_meanvert.RData")
mammals.meanvert = pred.df
mammals.meanvert$vertvar = "mean"
mammals.sesvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_sesvert.RData")
mammals.sesvert = pred.df
mammals.sesvert$vertvar = "ses"
mammals.pred = bind_rows(mammals.meanvert, mammals.sesvert) %>% mutate(class = "mammals")

rept.meanvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_meanvert.RData")
rept.meanvert = pred.df
rept.meanvert$vertvar = "mean"
rept.sesvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_sesvert.RData")
rept.sesvert = pred.df
rept.sesvert$vertvar = "ses"
rept.pred = bind_rows(rept.meanvert, rept.sesvert) %>% mutate(class = "rept")

amph.meanvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_meanvert.RData")
amph.meanvert = pred.df
amph.meanvert$vertvar = "mean"
amph.sesvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_sesvert.RData")
amph.sesvert = pred.df
amph.sesvert$vertvar = "ses"
amph.pred = bind_rows(amph.meanvert, amph.sesvert) %>% mutate(class = "amph")

pred = bind_rows(bird.pred, mammals.pred, rept.pred, amph.pred)

```

# Impacts of climate change on verticality

To examine the potential impacts of climate change on the verticality of vertebrate assemblages, we used the coefficients from the SAR models to predict assemblage verticality for the time period 2071-2100 using the SSP585 future climate scenario, which represents the upper boundary of predicted warming under a scenario of fossil-fueled development. Though climate change will impact vegetation structure, we were unable to incorporate predictions of future vegetation into the models due to a lack of global future predictions of vegetation height and density. Because we could not predict the spatial signal in verticality, we compared predictions under future climate conditions to predictions under present climate conditions that do not incorporate the spatial signal. - We find that x% of communities could experience declines in verticality. - Declines in verticality are concentrated in the tropics. - Between 23 and -23 latitude, x% birds, x% mammals, x% reptiles, and x% amphibians exhibit declines in verticality - Declines in the Amazon were notably severe

## discussion of climate change

-   What causes changes in verticality - vertical niche shift, immigration, or local extinction?

-   in snakes, it is more common for species to exhibit shift from nonfossorial to fossorial than the other way around (Cyriac and Kodondaramiah 2018) - suggests increase in verticality for reptiles could be caused by extinction of fossorial species rather than vertical niche shift higher

    -   further evidence by Cyriac and Kodandoramaiah 2017 - high extinction rates of fossorial snakes during prolonged environmental fluctuations

-   but these patterns aren't consistent - e.g., in lizards, transitions from fossoriality to nonfossoriality is not less common that transisitons in the other direction

-   In temperate regions, we find...

-   Predicted changes in verticality led to homogenization of vertical strategies globally.

# Discussion

In the tropics, the use of arboreal habitat likely contributes to high species richness (cite Oliveira, also see Naikatini et al. 2022, MacArthur). The large volume of habitat in tropical biomes facilitates species packing within a given area as well as expansion across vertically stratified resources. However, habitat volume and structural diversity is generally not related to species richness in forested ecosystems in the US, except for birds (La Rue et al. 2023). Low arboreality toward higher latitudes may explain the breakdown in the volume-richness relationship and be caused by low climatic suitability of arboreal habitats in temperate climates, except for birds which continue to use arboreal habitats at northern latitudes.

-increasing arboreality at norther latitudes suggests that climate change could lead to arboreal habitats in boreal regions becoming more suitable for species, thereby releasing competitive pressure as species shift poleward by providing more habitat volume. -niche attrition - land use change eliminates the volume of arboreal habitats, which leads to declines in arboreality. Will climate change cause further attrition of arboreal habitats?? - elevator to extinction for mountaintop species = digging a grave

-   vertical niche shifts occur in response to novel habitat, global change, etc.
-   e.g., shift in birds to arboreal foraging on islands (Lapiedra et al. 2021 Niche shifts after island colonization...)

Despite the presence of biomes with vertical structure in temperate forests the proportion of arboreality steeply declines, suggesting that the geographic niche space is underfilled by the animal community.

Low climatic suitability may limit arboreality and thus diversity in temperate biomes.

## The influence of vegetation
Indeed, it is expected that the volume of habitat space explain variation in species richness among forested biomes. However, this relationship is not upheld for birds and mammals (LaRue et al. 2023). Declining arboreality in temperate forests that have the vegetation structure to support arboreal strategies may explain this pattern and be caused by environmental factors limiting the suitability of canopy habitats.

# Methods

# Figures

```{r fig.id="vert-map-global"}

```

```{r fig.id="model-coefs"}

```

```{r fig.id="vert-difs-global"}

```

```{r fig.id="vert-homogenization"}

```

# Extended Data Tables

```{r tab.id = "mean-vert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of mean verticality of vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}

```

```{r tab.id = "sesvert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of the standard effect size of mean verticality of vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}

```

# Extended Data Figures

```{r fig.id="vertses-hist-biome", fig.lp="ed", fig.cap="Histograms of verticality by biome for each vertebrate class. Red dotted line indicates the mean verticality for each biome/class."}
# plot SES vert by biome
("figures/supp_figs/vert_biome_forestOnly_forestSES.R")
plt_vertses_biome

```

```{r fig.id="vert-forest", fig.lp="ed", fig.cap="Mean verticality, standard effect size (SES) of mean verticality, and the average proportion of species exhibiting arboreal, terrestrial, and fossorial strategies across latitudinal bands. In maps of mean veticality, assemblages composed of only fossorial species are zero (red) and those composed of only arboreal species are quantified as one (purple). In maps of the SES of mean verticality, negative numbers (red) indicate assemblages have lower verticality than expected by random chance and positive numbers (purple) indicate assemblages have higher verticality than expected by random chance. Line graphs show the average proportion of species exhibiting arboreal (blue), terrestrial (yellow), and fossorial (red) strategies. Proportions for each assemblage can exceed one because species can exhibit multiple vertical strategies.}

```

```{r fig.id="forest-model-coefs"}

```

## Reference

## Acknowledgements

We would like to thank Roger Bivand for useful insights regarding predictions of SAR models.

### 2024-07-18 scripts/03_analyses/03_sar_amphibians.R
upload SAR models

@@ -1,0 +1,1051 @@
+source("scripts/00_functions/00_functions.R")
+source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
+library(tidyverse)
+library(vegan)
+library(pgirmess)
+library(ncf)
+library(spdep)
+library(sp)
+library(DHARMa)
+library(spatialreg)
+library(MuMIn)
+library(data.table)
+library(parallel)
+library(doParallel)
+library(terra)
+library(colorspace)
+library(patchwork)
+library(broom)
+
+options(na.action = "na.omit")
+
+
+# VIF variable selection
+env = read.csv("data/derivative_data/env_data.csv")
+
+#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
+
+# variable selection without hurs and vpd
+# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
+env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
+
+
+
+# Amphibians --------------------------------------------------------------
+
+amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel.csv")
+
+
+#  * - data check --------------------------------------------------------------
+
+## SES vert model
+sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
+  drop_na()
+
+hist(sesvert$vert.mean.ses)
+
+# check linear association between predictor and response
+as.data.frame(sesvert) %>% 
+  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses)) +
+  geom_point(col = "grey", alpha = 0.2, size = 1) +
+  geom_smooth(col = "blue", method = "lm") +
+  facet_wrap(facets = vars(vars), scales = "free_x") +
+  theme_classic()
+
+# transform clim_velocity
+# sesvert.t = as.data.frame(sesvert) %>% 
+#   mutate(log_clim_velocity = log10(clim_velocity),
+#          log_temp_sea = log10(temp_sea)) %>% 
+#   dplyr::select(!c(clim_velocity, temp_sea))
+
+sesvert.t = as.data.frame(sesvert) %>% 
[…]

### 2024-07-18 scripts/03_analyses/03_sar_birds.R

@@ -1,0 +1,1363 @@
+source("scripts/00_functions/00_functions.R")
+source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
+library(tidyverse)
+library(vegan)
+library(pgirmess)
+library(ncf)
+library(spdep)
+library(sp)
+library(DHARMa)
+library(spatialreg)
+library(MuMIn)
+library(data.table)
+library(parallel)
+library(foreach)
+library(doParallel)
+library(terra)
+library(colorspace)
+library(patchwork)
+library(broom)
+
+options(na.action = "na.omit")
+
+# VIF variable selection
+env = read.csv("data/derivative_data/env_data.csv")
+
+#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
+
+# variable selection without hurs and vpd
+#env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
+
+env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
+
+
+
+# Birds elton --------------------------------------------------------------
+
+birds = read.csv("data/derivative_data/gridcell_data/birds_comdat/birds_breedresident_comdat_elton.csv")
+
+#  * - data check --------------------------------------------------------------
+
+## SES vert model
+sesvert = birds[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
+  drop_na()
+
+hist(sesvert$vert.mean.ses)
+
+# check linear association between predictor and response
+as.data.frame(sesvert) %>% 
+  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses)) +
+  geom_point(col = "grey", alpha = 0.2, size = 1) +
+  geom_smooth(col = "blue", method = "lm") +
+  facet_wrap(facets = vars(vars), scales = "free_x") +
+  theme_classic()
+
+# transform clim_velocity and temp seasonality and elevation
+# sesvert.t = as.data.frame(sesvert) %>% 
+#   mutate(log_clim_velocity = log10(clim_velocity),
+#          log_precp_dry = log10(precip_dry),
+#          log_temp_sea = log10(temp_sea))
+
+sesvert.t = as.data.frame(sesvert) %>% 
[…]

### 2024-07-18 scripts/03_analyses/03_sar_mammals.R

@@ -1,0 +1,1017 @@
+source("scripts_lydia/00_functions.R")
+source("scripts_lydia/00_icorrelogram.R")
+source("scripts_lydia/00_functions_predict_sarlm_debug.R")
+source("scripts_lydia/00_icorrelogram.R")
+library(tidyverse)
+library(vegan)
+library(pgirmess)
+library(ncf)
+library(spdep)
+library(sp)
+library(DHARMa)
+library(spatialreg)
+library(MuMIn)
+library(data.table)
+library(parallel)
+library(foreach)
+library(doParallel)
+library(terra)
+library(colorspace)
+library(patchwork)
+library(broom)
+
+options(na.action = "na.omit")
+
+# VIF variable selection
+env = read.csv("data/derivative_data/env_data.csv")
+
+#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
+
+# variable selection without hurs and vpd
+# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
+
+env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
+
+
+
+# Mammals --------------------------------------------------------------
+
+mammals = read.csv("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel.csv")
+
+#  * - data check --------------------------------------------------------------
+
+## SES vert model
+sesvert = mammals[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
+  drop_na()
+
+hist(sesvert$vert.mean.ses)
+
+# check linear association between predictor and response
+as.data.frame(sesvert) %>% 
+  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses)) +
+  geom_point(col = "grey", alpha = 0.2, size = 1) +
+  geom_smooth(col = "blue", method = "lm") +
+  facet_wrap(facets = vars(vars), scales = "free_x") +
+  theme_classic()
+
+# transform clim_velocity and temp seasonality and elevation
+# sesvert.t = as.data.frame(sesvert) %>% 
+#   mutate(log_clim_velocity = log10(clim_velocity),
+#          log_elev = log10(elev),
[…]

### 2024-07-18 scripts/03_analyses/03_sar_reptiles.R

@@ -1,0 +1,919 @@
+source("scripts_lydia/00_functions.R")
+source("scripts_lydia/00_icorrelogram.R")
+source("scripts_lydia/00_functions_predict_sarlm_debug.R")
+source("scripts_lydia/00_icorrelogram.R")
+library(tidyverse)
+library(vegan)
+library(pgirmess)
+library(ncf)
+library(spdep)
+library(sp)
+library(DHARMa)
+library(spatialreg)
+library(MuMIn)
+library(data.table)
+library(parallel)
+library(foreach)
+library(doParallel)
+library(terra)
+library(colorspace)
+library(patchwork)
+library(broom)
+
+options(na.action = "na.omit")
+
+# VIF variable selection
+env = read.csv("data/derivative_data/env_data.csv")
+
+#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
+
+# variable selection without hurs and vpd
+# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
+env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
+
+
+
+# Reptiles --------------------------------------------------------------
+
+repts = read.csv("data/derivative_data/gridcell_data/reptiles_comdat/rept_comdat_parallel.csv")
+
+#  * - data check --------------------------------------------------------------
+
+## SES vert model
+sesvert = repts[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
+  drop_na()
+
+hist(sesvert$vert.mean.ses)
+
+# check linear association between predictor and response
+as.data.frame(sesvert) %>% 
+  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses)) +
+  geom_point(col = "grey", alpha = 0.2, size = 1) +
+  geom_smooth(col = "blue", method = "lm") +
+  facet_wrap(facets = vars(vars), scales = "free_x") +
+  theme_classic()
+
+# transform clim_velocity and temp seasonality and elevation
+# sesvert.t = as.data.frame(sesvert) %>% 
+#   mutate(log_clim_velocity = log10(clim_velocity),
+#          log_elev = log10(elev),
+#          log_temp_sea = log10(temp_sea))
+
+
[…]

### 2024-07-18 scripts/03_analyses/03_glms_amphibians.R
delete glms

@@ -1,1052 +1,0 @@
-source("scripts_lydia/00_functions.R")
-source("scripts_lydia/00_icorrelogram.R")
-source("scripts_lydia/00_functions_predict_sarlm_debug.R")
-source("scripts_lydia/00_icorrelogram.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
-
-# variable selection without hurs and vpd
-# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Amphibians --------------------------------------------------------------
-
-amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel.csv")
-
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-# transform clim_velocity
-# sesvert.t = as.data.frame(sesvert) %>% 
-#   mutate(log_clim_velocity = log10(clim_velocity),
-#          log_temp_sea = log10(temp_sea)) %>% 
-#   dplyr::select(!c(clim_velocity, temp_sea))
-
[…]

### 2024-07-18 scripts/03_analyses/03_glms_birds.R

@@ -1,1363 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(foreach)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
-
-# variable selection without hurs and vpd
-#env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
-
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Birds elton --------------------------------------------------------------
-
-birds = read.csv("data/derivative_data/gridcell_data/birds_comdat/birds_breedresident_comdat_elton.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = birds[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-# transform clim_velocity and temp seasonality and elevation
-# sesvert.t = as.data.frame(sesvert) %>% 
-#   mutate(log_clim_velocity = log10(clim_velocity),
-#          log_precp_dry = log10(precip_dry),
-#          log_temp_sea = log10(temp_sea))
-
-sesvert.t = as.data.frame(sesvert) %>% 
[…]

### 2024-07-18 scripts/03_analyses/03_glms_mammals.R

@@ -1,1017 +1,0 @@
-source("scripts_lydia/00_functions.R")
-source("scripts_lydia/00_icorrelogram.R")
-source("scripts_lydia/00_functions_predict_sarlm_debug.R")
-source("scripts_lydia/00_icorrelogram.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(foreach)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
-
-# variable selection without hurs and vpd
-# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
-
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Mammals --------------------------------------------------------------
-
-mammals = read.csv("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = mammals[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-# transform clim_velocity and temp seasonality and elevation
-# sesvert.t = as.data.frame(sesvert) %>% 
-#   mutate(log_clim_velocity = log10(clim_velocity),
-#          log_elev = log10(elev),
[…]

### 2024-07-18 scripts/01_get_data/01_get_env_data.R
update get data

@@ -1,84 +1,101 @@
 library(terra)
 library(sf)
 library(fasterize)
+library(doParallel)
+library(cluster)
+library(foreach)
 
 # change tempdir so that I can easily delete temp files that take up too much space
 tempdir(tmpdir = "tempfiles/")
 
 # template raster used to get grid level pres/abs data at 111km
 mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
-res(mapa) = 111000
+#res(mapa) = 111000
+res(mapa) = 50000
 
 ## Current climate conditions (CHELSA v2.1)
 
 ## Chelsa climate data
 ## Downloaded from https://chelsa-climate.org/downloads/
 bio = list.files(path = "data/original/env_data/chelsa/", pattern = ".tif", full.names = T)
-bio = lapply(bio, rast)
-bio = rast(bio)
-bio = crop(bio, ext(-180, 180, -64.7937, 90))
+#bio = lapply(bio, rast)
+bio = bio[c(1:5,9:12)]
 
-tempsea = rast("data/original/env_data/chelsa/CHELSA_bio4_1981-2010_V.2.1.tif")
+cl = makeCluster(7)
+registerDoParallel(cl)
+bio = foreach(i = 1:length(bio), .packages = c("terra")) %dopar% {
+  mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
+  res(mapa) = 50000
+  r = rast(bio[[i]])
+  r = project(r, "+proj=cea +datum=WGS84")
+  r = resample(r, mapa, threads = T)
+  writeRaster(r, paste0("data/derivative_data/resampled_env_rasters_50km/", names(r), ".tif"))
+}
+stopCluster(cl)
 
-mat = project(bio$`CHELSA_bio1_1981-2010_V.2.1`, "+proj=cea +datum=WGS84")
-mat = resample(mat, mapa, threads = T)
-writeRaster(mat, "data/derivative_data/resampled_env_rasters/chelsa_bio_01.tif")
-
-tic()
-max_temp_warm = project(bio$`CHELSA_bio5_1981-2010_V.2.1`, "+proj=cea +datum=WGS84")
-max_temp_warm = resample(max_temp_warm, mapa, threads = T)
[…]

### 2024-07-18 scripts/01_get_data/01_get_pres_abs_data.R

@@ -32,7 +32,7 @@ library(tidyterra)
 vert = read.csv("data/original/Moura_arboreality_vertebrates.csv")
 
 # mammal data
-data = vect("./../data/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
+data = vect("data/original/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
 
 vert_mammal = vert %>% dplyr::filter(Class == "Mammalia") %>% 
   rename(sci_name = "Scientific.Name")
@@ -54,49 +54,103 @@ which(valid_check == FALSE)
 rm(data, data_reproj)
 gc()
 
+library(numform)
+for (i in 1:ceiling(nrow(data_reproj_valid) / 1000)) {
+  st = 1 + 1000*(i-1)
+  en = st + 999
+  en = ifelse(en > nrow(data_reproj_valid), nrow(data_reproj_valid), en)
+  
+  d = data_reproj_valid[st:en]
+  
+  writeVector(d, paste0("tempfiles/mammals_", f_pad_zero(i,2), ".shp"))
+}
+
+# split range file for processing
+library(doParallel)
+library(foreach)
+cl = makeCluster(5)
+registerDoParallel(cl)
+foreach(i = 1:ceiling(nrow(data_reproj_valid) / 1000), .packages = c("letsR", "numform")) %dopar% {
+  
+  f = list.files(path = "tempfiles/", pattern = ".shp", full.names = T)
+  f = f[[i]]
+  f = vect(f)
+  
+  # st = 1 + 2000*(i-1)
+  # en = st + 1999
+  # en = ifelse(en > nrow(data_reproj_valid), nrow(data_reproj_valid), en)
+  # 
+  # d = data_reproj_valid[st:en]
+  
+  # presence: extant or probably extant
+  # origin: native only
+  # seasonal: all resident, breeding, non-breeding
+  # keep cells in matrix where richness = 0
+  presab = lets.presab(f, resol = 50000, 
+                            crs = "+proj=cea +datum=WGS84", crs.grid = "+proj=cea +datum=WGS84",
+                            count = T, cover = 0,
+                            xmn = -20592508, xmx = 20588492, ymn = -5743602, ymx = 6573398,
+                            remove.cells = FALSE)
+  
[…]

### 2024-07-18 scripts/01_get_data/02_get_bird_data.R

@@ -1,5 +1,5 @@
 # Lydia Soifer
-# December 23, 2023
+# May 1, 2024
 # Get arboreality and environmental data and add to the matrix from get_pres_abs_data
 
 library(terra)
@@ -18,7 +18,7 @@ mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
              crs = "+proj=cea +datum=WGS84")
 res(mapa) <- 111000
 
-mundi = vect("./../data/Mundi_contour/Mundi_contour.shp")
+mundi = vect("data/Mundi_contour/Mundi_contour.shp")
 mundi = project(mundi, "+proj=cea +datum=WGS84")
 mundi.rast = rasterize(mundi, mapa)
 mundi.df = as.data.frame(mundi.rast, xy = T) %>% 

### 2024-07-18 scripts/01_get_data/02_get_sp_and_cell_data.R

@@ -1,7 +1,7 @@
 # Make species level and gridcell level dataframes for each class
 library(tidyverse)
 library(rnaturalearth)
-source("scripts_lydia/00_functions.R")
+source("scripts/00_functions/00_functions.R")
 library(terra)
 library(data.table)
 library(tictoc)
@@ -36,7 +36,15 @@ world.df = as.data.frame(world.rast, xy = T) %>%
 
 # load env data
 env = read.csv("data/derivative_data/env_data.csv")
+ggplot() +
+  geom_raster(data =env, aes(x = x, y = y, fill = biome)) +
+  coord_sf(crs = "+proj=cea +datum=WGS84") +
+  theme_classic()
 
+ggplot() +
+  geom_raster(data =env, aes(x = x, y = y, fill = canopy_height>5)) +
+  coord_sf(crs = "+proj=cea +datum=WGS84") +
+  theme_classic()
 
 
 # Moura traits
@@ -130,31 +138,33 @@ traits_birds = traits_birds %>%
 sum(traits_birds$sciname != colnames(occr[3:ncol(occr)]))
 
 # * - species level data ------------------------------------------------------
-birds_breed_spdat = get_sp_dat(trait_dat = traits_birds, occ = occr, env = env)
-write.csv(birds_breed_spdat, "data/derivative_data/species_data/birds_breedresident_spdat_moura.csv", row.names = F)
+# Using Elton data for birds
 
-test = occr[1:10,]
-get_gridcell_dat_parallel(trait_dat = birds_breed_spdat, occ = test, env = env, rich_min = 5, ncore = 7, nsim = 3)
-
-birds_spdat = read.csv("data/derivative_data/species_data/birds_breedresident_spdat_moura.csv")
-
-# takes 19 hours
-tic()
-birds_breed_comdat = get_gridcell_dat_parallel(trait_dat = birds_breed_spdat, occ = occr, env = env, rich_min = 5, ncore = 7, nsim = 100)
-write.csv(birds_breed_comdat, "data/derivative_data/gridcell_data/birds_comdat/birds_breedresident_comdat_parallel.csv", row.names = F)
-toc()
-
-
-# minutes
-tic()
-birds_trait_volume = get_trait_volume(trait_dat = birds_spdat, occ = occr, env = env, rich_min = 5, ncore = 7, nsim = 100)
[…]

### 2024-07-18 scripts/01_get_data/02_get_sp_and_cell_data_forestsOnly.R

@@ -2,11 +2,14 @@
 library(tidyverse)
 library(rnaturalearth)
 source("scripts/00_functions/00_functions.R")
+source("scripts/00_functions/00_functions2.R")
+source("scripts/00_functions/manuscript_functions.R")
 library(terra)
 library(data.table)
 library(tictoc)
 library(foreach)
 library(doParallel)
+library(picante)
 
 ### HARD CODES ###
 # phylogenetic imputation threshold for binary diel activity and verticality variables
@@ -17,7 +20,8 @@ impute_thresh = 0.7
 # extent of occurrence layer
 mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
              crs = "+proj=cea +datum=WGS84")
-res(mapa) = 111000
+#res(mapa) = 111000
+res(mapa) = 50000
 
 # terrestrial outline and gridcells
 # world = rnaturalearth::ne_countries()
@@ -35,11 +39,12 @@ world.df = as.data.frame(world.rast, xy = T) %>%
   dplyr::select(x,y)
 
 # load env data
-env = read.csv("data/derivative_data/env_data.csv")
+env = read.csv("data/derivative_data/env_data_50km.csv")
 
 # clip env data to wooded habitats only
 env.forest = env %>% 
   filter(grepl("forest|woodland|taiga|Forest|várzea|Yungas|savanna|thicket|mallee", ecoregion))
+fwrite(env.forest, "data/derivative_data/env_data_50km_forest.csv", row.names = F)
 
 # Moura traits
 traits = read.csv("data/original/trait_data/TetrapodTraits_1.0.0/TetrapodTraits_1.0.0.csv")
@@ -97,7 +102,8 @@ traits = traits %>%
 # Amphibians --------------------------------------------------------------
 
 # load amphibian occurrence data (cover0 includes all species, even if ranges cover less than 1/3 grid cell)
-occr <- fread("data/derivative_data/pres_abs_cover0/pres_abs_amph.csv")
+#occr <- fread("data/derivative_data/pres_abs_cover0/pres_abs_amph.csv")
+occr = fread("data/derivative_data/pres_abs_cover0/resolution_50km/pres_abs_amph.csv")
 colnames(occr)[1:2] = c("x", "y")
 colnames(occr) = gsub(" ", "_", colnames(occr))
 
[…]

### 2024-07-18 scripts/01_get_data/03_get_bbs_data.R

@@ -1,0 +1,70 @@
+# download breeding bird survey data for select regions and examine changes in arboreality over time
+
+library(bbsAssistant)
+library(data.table)
+
+sb_items
+View(region_codes)
+
+grab_bbs_data(bbs_dir = "data/original/bbs_data")
+
+dat = fread("data/original/bbs_data/50-StopData/50-StopData/1997ToPresent_SurveyWide/Fifty3/fifty3.csv")
+head(dat)
+
+dat = fread("data/original/bbs_data/50-StopData/50-StopData/1997ToPresent_SurveyWide/Fifty6/fifty6.csv")
+
+
+# FL state num
+fl = region_codes %>% filter(State == "florida")
+nc = region_codes %>% filter(State == "north carolina")
+
+dat.fl = dat %>% filter(StateNum == nc$StateNum)
+
+# calculate presab on route
+dat.fl[, pa := apply(.SD, 1, sum), .SDcols = Stop1:Stop50]
+dat.fl[, pa := ifelse(pa>0, 1, 0)]
+
+# reomve stop columns
+dat.fl[, 8:57 := NULL]
+
+# add species names
+head(species_list)
+splist = species_list %>% dplyr::select(AOU, Scientific_Name)
+
+dat.fl = left_join(dat.fl, splist, by = "AOU")
+dat.fl = dat.fl %>% rename(sciname = Scientific_Name)
+
+# verticality data for birds
+vert =  read.csv("data/derivative_data/species_data/birds_breedresident_spdat_eltonvert.csv")
+vert = vert %>% 
+  dplyr::select(sciname, Verticality) %>% 
+  mutate(sciname = gsub("_", " ", sciname))
+
+# combine vert data with bbs data for FL
+dat.fl = left_join(dat.fl, vert, by = "sciname")
+
+dat.fl.sum = dat.fl %>% 
+  filter(pa == 1) %>% 
+  group_by(RouteDataID, CountryNum, StateNum, Route, Year) %>% 
+  summarise(vert = mean(Verticality, na.rm = T))
+
+ggplot(dat.fl.sum, aes(x = Year, y = vert)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  theme_classic()
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/50km/sar_amphibians.R
update SAR models

@@ -1,0 +1,360 @@
+source("scripts/00_functions/00_functions.R")
+library(nlme)
+library(car)
+library(ncf)
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+
+
+options(na.action = "na.omit")
+
+
+# VIF variable selection
+#env = read.csv("data/derivative_data/env_data_50km_forest.csv")
+
+# amphibian forest only 50km resolution
+dat = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+
+# make key value dataframe for categorical rasters
+key.ecoregion = data.frame(ecoregion = unique(dat$ecoregion))
+key.ecoregion$key.ecoregion = 1:nrow(key.ecoregion)
+
+key.biome = data.frame(biome = unique(dat$biome))
+key.biome$key.biome = 1:nrow(key.biome)
+
+key.realm = data.frame(realm = unique(dat$realm))
+key.realm$key.realm = 1:nrow(key.realm)
+
+datr = dat  %>% relocate(x,y, .before = rich) %>% 
+  left_join(key.biome, by = "biome") %>% 
+  left_join(key.realm, by = "realm") %>% 
+  left_join(key.ecoregion, by = "ecoregion") %>% 
+  rast(type = "xyz", crs = "+proj=cea +datum=WGS84")
+plot(datr$vert.mean)
+plot(datr$vert.mean.ses)
+
+# variable selection 
+#env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
+
+# randomly select 5000
+set.seed(12345)
+d = dat %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na() %>% 
+  slice_sample(n=5000) %>% 
+  mutate_at(.vars = vars(canopy_height:clim_velocity, elev, veg_den, veg_complexity), .funs = scale)
+
+# look at the data
+# plot vert.mean.ses
+d %>% 
+  dplyr::select(vert.mean.ses, biome:clim_velocity, elev, veg_den, veg_complexity) %>%
+  pivot_longer(cols = 3:16, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+    geom_point(pch = ".") +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/mammals.R

@@ -1,840 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Mammals --------------------------------------------------------------
-
-mammals = read.csv("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel_forestsOnly.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = mammals[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry))
-# add quadratic to tmaxwarm
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_birds.R
update SAR models

@@ -365,7 +365,7 @@ sesvert.avg.full %>%
   geom_vline(xintercept = 0, linetype = "dashed") +
   scale_color_manual(values = c("black", "red3")) +
   scale_y_discrete("") +
-  ggtitle("SES Mean Verticality: Amphibians") +
+  ggtitle("SES Mean Verticality: Birds") +
   theme_classic() +
   theme(plot.title = element_text(hjust = 0.5))
 
@@ -733,3 +733,4 @@ vert.avg.full %>%
 pred.r = rast(pred.df, type = "xyz", crs = "+proj=cea +datum=WGS84")
 plot(pred.r$pred.future)
 plot(pred.r$pred.future - pred.r$pred.pres.trend)
+

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_reptiles.R

@@ -742,4 +742,7 @@ vert.avg.full %>%
   theme_classic() +
   theme(plot.title = element_text(hjust = 0.5))
 
-
+pred.r = rast(pred.df, type = "xyz", crs = "+proj=cea +datum=WGS84")
+plot(pred.r$pred.pres.trend)
+plot(pred.r$pred.future)
+plot(pred.r$pred.future - pred.r$pred.pres.trend)

### 2024-07-18 scripts/03_analyses/04_glmms/sdmTMB_amphibians.R
update sdmTMB models

@@ -14,6 +14,9 @@ library(DHARMa)
 
 # amphibian forest only 50km resolution
 raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
+
 
 # remove cells with fewer than five species
 raw = raw %>% 
@@ -28,7 +31,7 @@ env.f = rast(list.files(path = "data/derivative_data/resampled_env_rasters_50km/
 env.f = env.f[[2:6]]
 
 dat.f = raw %>% 
-  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, ecoregion)
+  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, realm)
 
 env.f = terra::extract(env.f, dat.f[,c('x', "y")], ID = F)
 
@@ -112,542 +115,735 @@ vif(lm(vert.mean.ses ~ canopy_height + veg_den + tmax_warm + tmin_cold  + precip
 
 f1 = formula(vert.mean.ses ~ canopy_height + veg_den + I(tmax_warm^2) + tmax_warm + tmin_cold + precip_wet + log_precip_dry + log_clim_velocity)
 
-
-
-
-# Cross validation --------------------------------------------------------
-
-# compare random effects structure using cross validation
-# preliminary assessment indicated that residuals displayed strong spatial autocorrelation when
-# spatial effects were not accounted for in any way and when biome or biorealm were included as random intercepts in the model
-
-# set up spatial mesh
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
 dat$biome = factor(dat$biome)
 dat$biorealm = factor(paste(dat$biome, dat$realm, sep = "_"))
+dat$realm = factor(dat$realm)
 
-# use 300km mesh - relatively fine with reasonable fitting speed
-mesh300km = make_mesh(data = dat, xy_cols = c("x", "y"), cutoff = 3)
-plot(mesh300km)
-
-# set up five random folds for cross validation
-folds = sample(1:5, size = nrow(dat), replace = T)
-
-# 1: fit random intercept of ecoregion without spatial random field
-library(future)
-plan(multisession)
-
-mod.ecoregion.cv = sdmTMB_cv(update(f1, ~ . + (1|ecoregion)), 
-                             data = dat,
-                             mesh = mesh300km,
[…]

### 2024-07-18 scripts/03_analyses/04_glmms/sdmTMB_birds.R

@@ -11,6 +11,8 @@ library(terra)
 library(car)
 library(data.table)
 library(DHARMa)
+source("scripts/00_functions/00_plot_functions.R")
+source("scripts/00_functions/manuscript_functions.R")
 
 # amphibian forest only 50km resolution
 raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
@@ -28,7 +30,7 @@ env.f = rast(list.files(path = "data/derivative_data/resampled_env_rasters_50km/
 env.f = env.f[[2:6]]
 
 dat.f = raw %>% 
-  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, ecoregion)
+  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, realm)
 
 env.f = terra::extract(env.f, dat.f[,c('x', "y")], ID = F)
 
@@ -112,545 +114,744 @@ vif(lm(vert.mean.ses ~ canopy_height + veg_den + tmax_warm + tmin_cold  + precip
 
 f1 = formula(vert.mean.ses ~ canopy_height + veg_den + I(tmax_warm^2) + tmax_warm + tmin_cold + precip_wet + log_precip_dry + log_clim_velocity)
 
-
-
-
-# Cross validation --------------------------------------------------------
-
-# compare random effects structure using cross validation
-# preliminary assessment indicated that residuals displayed strong spatial autocorrelation when
-# spatial effects were not accounted for in any way and when biome or biorealm were included as random intercepts in the model
-
-# set up spatial mesh
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
 dat$biome = factor(dat$biome)
 dat$biorealm = factor(paste(dat$biome, dat$realm, sep = "_"))
+dat$realm = factor(dat$realm)
 
-# use 300km mesh - relatively fine with reasonable fitting speed
-mesh300km = make_mesh(data = dat, xy_cols = c("x", "y"), cutoff = 3)
-plot(mesh300km)
-
-# set up five random folds for cross validation
-folds = sample(1:5, size = nrow(dat), replace = T)
-
-# 1: fit random intercept of ecoregion without spatial random field
-library(future)
-plan(multisession)
-
-mod.ecoregion.cv = sdmTMB_cv(update(f1, ~ . + (1|ecoregion)), 
-                             data = dat,
-                             mesh = mesh300km,
[…]

### 2024-07-18 scripts/03_analyses/04_glmms/sdmTMB_mammals.R

@@ -11,9 +11,12 @@ library(terra)
 library(car)
 library(data.table)
 library(DHARMa)
+library(fmesher)
 
 # mammals forest only 50km resolution
 raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
 
 # remove cells with fewer than five species
 raw = raw %>% 
@@ -28,7 +31,7 @@ env.f = rast(list.files(path = "data/derivative_data/resampled_env_rasters_50km/
 env.f = env.f[[2:6]]
 
 dat.f = raw %>% 
-  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, ecoregion)
+  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, realm)
 
 env.f = terra::extract(env.f, dat.f[,c('x', "y")], ID = F)
 
@@ -111,136 +114,355 @@ vif(lm(vert.mean.ses ~ canopy_height + veg_den + tmax_warm + tmin_cold  + precip
 vif(lm(vert.mean.ses ~ canopy_height + veg_den + tmax_warm + tmin_cold  + precip_wet + log_precip_dry + log_clim_velocity, data = dat))
 
 
-# VIF all under 4
+# VIF all under 5
 
 f1 = formula(vert.mean.ses ~ canopy_height + veg_den + I(tmax_warm^2) + tmax_warm + tmin_cold + precip_wet + log_precip_dry + log_clim_velocity)
 
-
-# Cross validation --------------------------------------------------------
-
-# compare random effects structure using cross validation
-# preliminary assessment indicated that residuals displayed strong spatial autocorrelation when
-# spatial effects were not accounted for in any way and when biome or biorealm were included as random intercepts in the model
-
-# set up spatial mesh
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
 dat$biome = factor(dat$biome)
 dat$biorealm = factor(paste(dat$biome, dat$realm, sep = "_"))
+dat$realm = factor(dat$realm)
+
+dat.f$realm = factor(dat.f$realm)
+dat.f$x = dat.f$x/1e5
+dat.f$y = dat.f$y/1e5
 
 
-# use 300km mesh - relatively fine with reasonable fitting speed
[…]

### 2024-07-18 scripts/03_analyses/04_glmms/sdmTMB_reptiles.R

@@ -28,7 +28,7 @@ env.f = rast(list.files(path = "data/derivative_data/resampled_env_rasters_50km/
 env.f = env.f[[2:6]]
 
 dat.f = raw %>% 
-  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, ecoregion)
+  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, realm)
 
 env.f = terra::extract(env.f, dat.f[,c('x', "y")], ID = F)
 
@@ -110,559 +110,755 @@ vif(lm(vert.mean.ses ~ canopy_height + veg_den + tmax_warm + tmin_cold  + precip
 
 # VIF all under 4
 
-f1 = formula(vert.mean.ses ~ canopy_height + veg_den + I(tmax_warm^2) + tmax_warm + tmin_cold + precip_wet + log_precip_dry + log_clim_velocity)
 
-
-
-
-# Cross validation --------------------------------------------------------
-
-# compare random effects structure using cross validation
-# preliminary assessment indicated that residuals displayed strong spatial autocorrelation when
-# spatial effects were not accounted for in any way and when biome or biorealm were included as random intercepts in the model
-
-# set up spatial mesh
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
 dat$biome = factor(dat$biome)
 dat$biorealm = factor(paste(dat$biome, dat$realm, sep = "_"))
+dat$realm = as.factor(dat$realm)
 
-# use 300km mesh - relatively fine with reasonable fitting speed
-mesh300km = make_mesh(data = dat, xy_cols = c("x", "y"), cutoff = 3)
-plot(mesh300km)
-
-# set up five random folds for cross validation
-folds = sample(1:5, size = nrow(dat), replace = T)
-
-# 1: fit random intercept of ecoregion without spatial random field
-library(future)
-plan(multisession)
-
-mod.ecoregion.cv = sdmTMB_cv(update(f1, ~ . + (1|ecoregion)), 
-                             data = dat,
-                             mesh = mesh300km,
-                             spatial = "off",
-                             reml = T, 
-                             fold_ids = folds,
-                             k_folds = length(unique(folds)))
-mod.ecoregion.cv$converged
-lapply(mod.ecoregion.cv$models, sanity)
-# gradients not okay for folds 2 and 5
-
[…]

### 2024-07-18 scripts/03_analyses/04_glmms/sdmTMB_squamates.R

@@ -1,0 +1,673 @@
+# Fit models of verticality using sdmTMB
+# Step 1: compare random effects structure using cross validation, reml = T, include all fixed effects
+# Step 2: compare fixed effects structure for different hypotheses regarding environmental predictors using cross validation, reml = F
+# Step 3: refit the best model using all data
+# Step 4: predict the best model to future env data
+
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+library(car)
+library(data.table)
+library(DHARMa)
+
+# squamate forest only 50km resolution
+raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/squamates_comdat.csv")
+
+# remove cells with fewer than five species
+raw = raw %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na()
+
+# future env data
+env.f = rast(list.files(path = "data/derivative_data/resampled_env_rasters_50km/chelsa_future/2071_2100/ensemble/",
+                        pattern = ".tif", full.names = T))
+env.f = env.f[[2:6]]
+
+dat.f = raw %>% 
+  dplyr::select(x,y,canopy_height, veg_den, clim_velocity, ecoregion)
+
+env.f = terra::extract(env.f, dat.f[,c('x', "y")], ID = F)
+
+dat.f = cbind(dat.f, env.f)
+dat.f = dat.f %>% rename(tmax_warm = max_temp_warm, tmin_cold = min_temp_cold)
+
+# remove NA values from future and present data
+raw = raw[which(!is.na(dat.f$precip_dry)),]
+dat.f = dat.f[which(!is.na(dat.f$precip_dry)),]
+
+
+# scale the data
+dat = raw %>%
+  mutate_at(.vars = vars(canopy_height:clim_velocity, elev, veg_den, veg_complexity), .funs = scale)
+
+# plot relationship between SES verticality and env predictors
+dat %>% 
+  dplyr::select(vert.mean.ses, biome:clim_velocity, elev, veg_den, veg_complexity) %>%
+  pivot_longer(cols = 3:16, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
[…]

### 2024-07-18 scripts/00_functions/00_functions.R
update manuscript functions

@@ -68,7 +68,7 @@ library(mFD)
 
 # Get species level data --------------------------------------------------
 
-get_sp_dat = function(trait_dat, filter_col=NA, filter_val=NA, occ, env){
+get_sp_dat = function(trait_dat, filter_col=NA, filter_val=NA, occ, env, traitbase = NA){
   # trait_dat: dataframe; species trait information. Columns related to diet must include the word "Diet"
   # occ: dataframe with presence absence data and xy values for cells
   # env: env data with values for xy in occ dataframe. xy values in occ and env MUST match
@@ -89,7 +89,13 @@ get_sp_dat = function(trait_dat, filter_col=NA, filter_val=NA, occ, env){
   }
   
   # vertical plasticity: number of different vertical strata in which a species occurs
-  trait_dat$vert.plast = apply(subset(trait_dat, select = c("Fos", "Ter", "Aqu", "Arb", "Aer")), 1, function(x) length(which(x>0)))
+  if(traitbase == "elton") {
+    trait_dat$vert.plast = apply(subset(trait_dat, select = ForStrat.watbelowsurf:ForStrat.aerial), 
+                                 1, function(x) length(which(x>0)))
+  } else {
+    trait_dat$vert.plast = apply(subset(trait_dat, select = c("Fos", "Ter", "Aqu", "Arb", "Aer")), 1, function(x) length(which(x>0)))
+  }
+  
   
   # mean temperature within species range
   trait_dat$mat = apply(subset(occ, select = trait_dat$sciname), 2, function(x) mean(env[which(x==1), "mat"], na.rm = T))
@@ -1261,4 +1267,328 @@ get_gridcell_dat_parallel = function(trait_dat, filter_col=NA, filter_val=NA, oc
 }
 
 
+
+# Get gridcell vert parallel ----------------------------------------------
+get_gridcell_vert_parallel = function(trait_dat, filter_col=NA, filter_val=NA, occ, env, rich_min = 5, ncore, nsim, eltonbirds = FALSE) {
+  # trait_dat: datatable or dataframe with taxonomy, vertical position, and traits
+  ##     species names should be in column named "sciname"  
+  ##     This should be the whole taxonomic group so that we can get taxonomic subset with respect to whole group
[…]

### 2024-07-18 scripts/00_functions/00_functions2.R

@@ -1,0 +1,76 @@
+mat = matrix(c(0,1,1,0,0,
+              0,1,0,1,0,
+              1,1,1,1,1,
+              0,0,0,1,1), byrow = T, nrow = 4)
+
+vert = c(1,0.5,0,1,1)
+p.arb = c(1,0,0,1,1)
+names(vert) = names(mat)
+
+colnames(mat) = paste0("sp", 1:ncol(mat))
+rownames(mat) = paste0("site", 1:nrow(mat))
+
+mat = as.data.table(mat)
+
+
+
+trait_summ = function(m, t, fun) {
+  # summarise trait value per site
+  # replace zeros with NAs
+  # fun: function to summarize - if fun = "prop" calculates proportions of ones for a binary variable
+  
+  if(sum(!colnames(m) %in% names(t)) > 0) {
+    stop("species differ between community matrix and trait vector")
+  }
+  if(sum(!colnames(t) %in% names(m)) > 0) {
+    stop("species differ between community matrix and trait vector")
+  }
+  
+  if(sum(colnames(m) != names(t)) > 0){
+    t[order(match(names(t), colnames(m)))]
+  }
+  
+  m[m == 0] = NA
+  
+  # replace vert scores in matrix and calculate summary trait value for each site
+  if(fun == "prop") {
+    apply(t(m)*t, 2, function(x) sum(x, na.rm = T)/sum(!is.na(x)))
+  } else {
+    apply(t(m)*t, 2, fun, na.rm = T)
+  }
+}
+
+sestrait = function(m, t, fun, reps, quantiles = c(0.025, 0.975)) {
+  # m = site x species matrix
+  # t = vector of traits (names of trait vector must correspond to column names of matrix)
+  # fun = function to summarize trait values per site (e.g. mean will provide mean and SES of mean trait)
+  # number of replicates to generate random communities
+  # quantiles = vector of quantiles to summarize replicates
+  
+  if(sum(!colnames(m) %in% names(t)) > 0) {
+    stop("species differ between community matrix and trait vector")
+  }
+  if(sum(!colnames(t) %in% names(m)) > 0) {
+    stop("species differ between community matrix and trait vector")
+  }
+  
+  if(sum(colnames(m) != names(t)) > 0){
+    t[order(match(names(t), colnames(m)))]
+  }
+  
[…]

### 2024-07-18 scripts/00_functions/00_plot_functions.R

@@ -1,0 +1,87 @@
+
+# PLOT COEFFICIENTS FOR MODEL COMPARISONS ---------------------------------
+
+plot_compMods_coefs = function(mods, fname) {
+  # mods: list of named models to plot coefs
+  # fname: filename to save plot to
+  coefs = foreach(i = 1:length(mods), .combine = "rbind") %do% {
+    df = tidy(mods[[i]], conf.int = T, conf.level = 0.95) %>% 
+      mutate(model = names(mods)[i],
+             term = factor(term, levels = c("log_clim_velocity",
+                                            "precip_wet", "log_precip_dry", 
+                                            "I(tmax_warm^2)", "tmax_warm","tmin_cold",
+                                            "veg_den", "canopy_height")))
+  }
+  
+  p = coefs %>% 
+    filter(term != "(Intercept)") %>% 
+    ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, color = model)) +
+    geom_vline(xintercept = 0, linetype = "dashed") +
+    geom_pointrange(position = position_dodge2(width = 0.3)) +
+    scale_x_continuous("Estimate") +
+    theme_bw() + 
+    theme(axis.title.y = element_blank())
+  
+  png(fname, height = 180, width = 180, units = "mm", res = 300) 
+  p
+  dev.off()
+}
+
+
+
+# CHECK RESIDUAL PLOTS ----------------------------------------------------
+
+plot_resids = function(mod, response_var, fpath) {
+  # mod: sdmTMB model to check residuals on
+  # response variable used in the model
+  # file path without file extension
+  
+  pred = predict(mod, type = "response")
+  s = simulate(mod, nsim = 500, seed = 12345, type = "mle-mvn")
+  
+  # simulate quantile residuals
+  r <- DHARMa::createDHARMa(
+    simulatedResponse = s,
+    observedResponse = pred[,response_var],
+    fittedPredictedResponse = pred[,"est"]
+  )
+  
+  png(paste0(fpath, "01.png"), height = 105, width = 210, res = 300, units = "mm")
+  plot(r)
+  dev.off()
+  
+  png(paste0(fpath, "02.png"), height = 100, width = 150, res = 300, units = "mm")
+  testQuantiles(r)
+  dev.off()
+  
+  png(paste0(fpath, "03.png"), height = 100, width = 300, res = 300, units = "mm")
[…]

### 2024-07-18 scripts/00_functions/manuscript_functions.R

@@ -1,0 +1,908 @@
+library(foreach)
+library(tidyterra)
+library(ggplot2)
+library(terra)
+library(colorspace)
+library(patchwork)
+library(fmesher)
+library(sdmTMB)
+
+# testing data
+# occ = matrix(c(0,1,1,0,0,0,
+#              0,1,0,1,0,0,
+#              1,1,1,1,1,0,
+#              0,0,0,1,1,0,
+#              1,0,1,1,0,0,
+#              0,1,1,1,1,0,
+#              0,0,0,1,1,1,
+#              0,0,0,1,0,1,
+#              0,1,0,0,0,1), byrow = T, nrow = 9)
+# 
+# tr = data.frame(Verticality = c(1,0.5,0,1,1,0.5),
+#            Arb = c(1,0,0,1,1,0),
+#            Fos = c(0,0,1,0,0,0),
+#            Ter = c(0,1,0,0,0,1),
+#            sciname = paste0("sp", 1:ncol(occ)))
+# colnames(occ) = paste0("sp", 1:ncol(occ))
+# rownames(occ) = paste0("site", 1:nrow(occ))
+# 
+# env = data.frame(realm = c(rep("a",4), NA, rep("b", 4)), biome = c(rep("b", 4), NA, rep("b2", 4)))
+
+gridcell_dat = function(occ, tr, env, reps = 100) {
+  # test if species names in trait_dat are in same order as occ
+  if(sum(colnames(occ)[3:ncol(occ)] != tr$sciname) > 0) {
+    stop("Species names in occurrence dataframe and trait dataframe do not match")
+  }
+  
+  # test match between occ and env
+  if(all.equal(env[,c("x", "y")], occ[,c("x", "y")]) != TRUE){
+    stop("env and occ coordinates do not match")
+  }
+  
+  # name occ rows and remove xy columns
+  rownames(occ) = paste0(occ$x, "_", occ$y)
+  rownames(env) = paste0(env$x, "_", env$y)
+  occ = occ[,-c(1,2)]
+  
+  realms = unique(env$realm)
+  realms = realms[which(!is.na(realms))]
+  
+  cl = makeCluster(7)
+  registerDoParallel(cl)
+  
+  dat = foreach(i = realms,
+          .packages = "picante") %do% {
+    # filter to realm
+    occ.r = occ[which(env$realm==i),]
+    colnames(occ.r) = colnames(occ)
+    #occ.r = as.matrix(occ.r)
[…]

### 2024-07-18 scripts/03_analyses/05_glms/glm_amphibians.Rmd
update glms

---
title: "glm_amphibians"
output: html_document
date: "2024-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ncf)
library(DHARMa)
source("./../../00_functions/00_functions.R")

options(na.action = "na.omit")

```

# SES verticality

VIF variable selection

```{r}
env = read.csv("./../../../data/derivative_data/env_data.csv")

# variable selection 
env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)

```

```{r}
amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv")

sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
  drop_na()

hist(sesvert$vert.mean.ses)

```

Look at the correlation between variables
The strongest correlation coefficient is between tmax_warm and tmin_cold at 0.62
```{r}
pairs(sesvert[,6:13])
cor(sesvert[,6:13], method = "pearson")

```

Check linear association between predictor and response
```{r}
as.data.frame(sesvert) %>% 
  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
  ggplot(aes(x = val, y = vert.mean.ses)) +
  geom_point(col = "grey", alpha = 0.2, size = 1) +
  geom_smooth(col = "blue", method = "lm") +
  facet_wrap(facets = vars(vars), scales = "free_x") +
  theme_classic()

```

There is not apparent nonlinearity between response and predictors for any variables except for tmin_cold. However, there are non-homogeneous variances.

Next, we need to examine spatial autocorrelation in the response variable
```{r}
correlog = ncf::correlog(x = sesvert$x, y = sesvert$y, z = sesvert$vert.mean.ses, increment = 111000, resamp = 100, latlon = F)
plot(correlog)
save(correlog, file = "glm_amphibian_objects.RData")
correlog.df = data.frame(distclass = correlog$mean.of.class, moranI = correlog$correlation, p = correlog$p)

# this takes a long time to run
# do we get the same result if we subsample the points

sesvert.sub = sesvert %>% 
  slice_sample(n=1000)

correlog.sub = ncf::correlog(x = sesvert.sub$x, y = sesvert.sub$y, z = sesvert.sub$vert.mean.ses, increment = 111000, resamp = 100, latlon = F)
plot(correlog.sub)
save.image(file = "glm_amphibian_objects.RData")

# zoom in on plot
plot(correlog.sub, xlim = c(0,5e6))
correlog.sub$x.intercept
correlog.sub.df = data.frame(dist = correlog.sub$mean.of.class, moranI = correlog.sub$correlation, p = correlog.sub$p)
```

We know that there is spatial autocorrelation in the response variable and that there are nonhomogeneous variances and potential non-linearities with response variables. But before we start adjusting anything, we will run a simple linear regression to start.

```{r}
# first set the formula

f = formula(vert.mean.ses ~ tmax_warm + tmin_cold + precip_wet + precip_dry + precip_sea + clim_velocity + elev + veg_complexity)

# before running the model, we will scale the predictor variables so that coefficients are comparable
sesvert.scale = sesvert %>% 
  mutate_at(.vars = 6:ncol(sesvert), .funs = function(x){scale(x)})
colnames(sesvert.scale) = colnames(sesvert)

mod.lm1 = lm(f, data = sesvert.scale)

# look at validation plots
plot(mod.lm1)
hist(mod.lm1$residuals)

# not bad residual plots - some slightly high residuals, which I suspect are caused by clim_velocity

# check spatial autocorrelation
# subset residuals to look at spatial autocorrelation
resid.sub = data.frame(x = sesvert$x, y = sesvert$y, resid = mod.lm1$residuals) %>% slice_sample(n = 1000)
correlog.m1.resid = ncf::correlog(x = resid.sub$x, y = resid.sub$y, z = resid.sub$resid, increment = 111000, resamp = 100, latlon = F)

plot(correlog.m1.resid)

# there is clear spatial autocorrelation in the residuals until 1294 km

correlog.m1.resid$x.intercept

# save model coefficients
mod.lm1.summ = summary(mod.lm1)
#coefs.df = data.frame(estimate = mod.lm1.summ$coefficients[,1], se = mod.lm1.summ$coefficients[,2], p = mod.lm1.summ$coefficients[,4]) %>% 
#  mutate(mod = "mod.lm1")

# run the same model using glm instead of lm
mod.glm1 = glm(f, data = sesvert.scale)
plot(mod.glm1)

# get quantile residuals with DHARMA package
mod.glm1.resid = simulateResiduals(mod.glm1)

# look at spatial autocorrelation in resids
resid.sub = data.frame(x = sesvert.scale$x, y = sesvert.scale$y, resid = residuals(mod.glm1.resid)) %>% slice_sample(n = 1000)
correlog.glm1.resid = ncf::correlog(x = resid.sub$x, y = resid.sub$y, z = resid.sub$resid, increment = 111000, resamp = 100, latlon = F)
plot(correlog.glm1.resid) # clear spatial autocorrelation

# also test overall autocorrelation using Dharma
testSpatialAutocorrelation(mod.glm1.resid, x = sesvert.scale$x, y = sesvert.scale$y, plot = F)

# add coefficients to comparison dataframe
glm1.summ = summary(mod.glm1)
glm1.summ = data.frame(estimate = glm1.summ$coefficients[,1], se = glm1.summ$coefficients[,2], p = glm1.summ$coefficients[,4])
glm1.summ$mod = "mod.glm1"
glm1.summ$var = rownames(glm1.summ)

#coefs.df = bind_rows(coefs.df, glm1.summ)
```

We know that there is spatial autocorrelation in the residuals and that the residuals are not quite normally distributed.
First we address the issue of spatial autocorrelation by subsetting the data to the point x-intercept of the correlogram
```{r}
library(enmSdmX)

sesvert.scale.sf = st_as_sf(sesvert.scale, coords = c("x", "y"), remove = F)
st_crs(sesvert.scale.sf) = "+proj=cea +datum=WGS84"

# thin points by x-intercept of correlogram from glm residuals
sesvert.scale.thin = geoThin(sesvert.scale.sf, minDist = correlog.glm1.resid$x.intercept, method = "complete")
sesvert.scale.thin = as.data.frame(sesvert.scale.thin)

# run glm on thinned points (only 122 points left after thinning)
mod.glm.thin = glm(f, data = sesvert.scale.thin)
plot(mod.glm.thin) # some patterning in residuals

# get quantile residuals with DHARMA package
mod.glmthin.resid = simulateResiduals(mod.glm.thin)

# look at spatial autocorrelation in resids
correlog.glmthin.resid = ncf::correlog(x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, z = residuals(mod.glmthin.resid), increment = correlog.glm1.resid$x.intercept, resamp = 100, latlon = F)
plot(correlog.glmthin.resid) # clear spatial autocorrelation

# also test overall autocorrelation using Dharma
testSpatialAutocorrelation(mod.glmthin.resid, x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, plot = F)

# add coefficients to comparison dataframe
glmthin.summ = summary(mod.glm.thin)
glmthin.summ = data.frame(estimate = glmthin.summ$coefficients[,1], se = glmthin.summ$coefficients[,2], p = glmthin.summ$coefficients[,4])
glmthin.summ$mod = "mod.glm.thin"
glmthin.summ$var = rownames(glmthin.summ)

coefs.df = bind_rows(glm1.summ, glmthin.summ)

```

How do the coefficients compare between glm and thinned glm
```{r}
ggplot(coefs.df) +
  geom_pointrange(aes(x = estimate, xmin = estimate-se, xmax = estimate+se, y = var, color = mod))
save.image(file = "glm_amphibian_objects.RData")
```

Load the SAR models for amphibians
```{r}
load("./../../../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_sesvert.RData")
sar.summ = summary(sesvert.avg)$coefmat.full
sar.summ = data.frame(estimate = sar.summ[,1], se = sar.summ[,2], p = sar.summ[,4], mod = "sar")
sar.summ$var = rownames(sar.summ)
coefs.df = bind_rows(coefs.df, sar.summ)

ggplot(coefs.df) +
  geom_pointrange(aes(y = estimate, ymin = estimate-se, ymax = estimate+se, x = var, color = mod),
                  position = position_dodge()) +
  coord_flip()

```

There are some clear differences, but we have not transformed the predictor variables for the glm.
Try transforming the same as for sar models and then compare.

```{r}
as.data.frame(sesvert) %>% 
  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
  ggplot(aes(x = val, y = vert.mean.ses)) +
  geom_point(col = "grey", alpha = 0.2, size = 1) +
  geom_smooth(col = "blue", method = "lm") +
  facet_wrap(facets = vars(vars), scales = "free_x") +
  theme_classic()

sesvert.t = as.data.frame(sesvert) %>% 
  mutate(log_clim_velocity = log10(clim_velocity),
         log_elev = log10(elev),
         log_precip_sea = log10(precip_sea),
         log_precip_dry = log10(precip_dry+1))

sesvert.t %>% 
  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
  ggplot(aes(x = val, y = vert.mean.ses)) +
  geom_point(col = "grey", alpha = 0.2, size = 1) +
  geom_smooth(col = "blue", method = "lm") +
  facet_wrap(facets = vars(vars), scales = "free_x") +
  theme_classic()

# keep log transformation for climate velocity and elevation
sesvert.t = sesvert.t %>% 
  dplyr::select(!c(clim_velocity, elev, log_precip_sea, precip_dry))
# add quadratic term to model for tmin_cold

# scale predictor variables
sesvert.scale = sesvert.t %>% 
  mutate_at(.vars = 6:ncol(sesvert.t), .funs = function(x){scale(x)})
colnames(sesvert.scale) = colnames(sesvert.t)

f = formula(vert.mean.ses ~ log_elev + log_clim_velocity + tmax_warm + tmin_cold + log_precip_dry + precip_sea + precip_wet + veg_complexity)

```

Run the glm with and withoug thinning the points
```{r}
# run the same model using glm instead of lm
mod.glm1 = glm(f, data = sesvert.scale)
plot(mod.glm1)

# get quantile residuals with DHARMA package
mod.glm1.resid = simulateResiduals(mod.glm1)

# look at spatial autocorrelation in resids
resid.sub = data.frame(x = sesvert.scale$x, y = sesvert.scale$y, resid = residuals(mod.glm1.resid)) %>% slice_sample(n = 1000)
correlog.glm1.resid = ncf::correlog(x = resid.sub$x, y = resid.sub$y, z = resid.sub$resid, increment = 111000, resamp = 100, latlon = F)
plot(correlog.glm1.resid, xlim = c(0,5e6), ylim = c(-1,1)) # clear spatial autocorrelation

# also test overall autocorrelation using Dharma
testSpatialAutocorrelation(mod.glm1.resid, x = sesvert.scale$x, y = sesvert.scale$y, plot = F)

# add coefficients to comparison dataframe
glm1.summ = summary(mod.glm1)
glm1.summ = data.frame(estimate = glm1.summ$coefficients[,1], se = glm1.summ$coefficients[,2], p = glm1.summ$coefficients[,4])
glm1.summ$mod = "mod.glm1"
glm1.summ$var = rownames(glm1.summ)

```

Now for the thinned points
```{r}
sesvert.scale.sf = st_as_sf(sesvert.scale, coords = c("x", "y"), remove = F)
st_crs(sesvert.scale.sf) = "+proj=cea +datum=WGS84"

# thin points by x-intercept of correlogram from glm residuals
sesvert.scale.thin = geoThin(sesvert.scale.sf, minDist = correlog.glm1.resid$x.intercept, method = "complete")
sesvert.scale.thin = as.data.frame(sesvert.scale.thin)

# run glm on thinned points (only 122 points left after thinning)
mod.glm.thin = glm(f, data = sesvert.scale.thin)
plot(mod.glm.thin) # some patterning in residuals

# get quantile residuals with DHARMA package
mod.glmthin.resid = simulateResiduals(mod.glm.thin)

# look at spatial autocorrelation in resids
correlog.glmthin.resid = ncf::correlog(x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, z = residuals(mod.glmthin.resid), increment = correlog.glm1.resid$x.intercept, resamp = 100, latlon = F)
plot(correlog.glmthin.resid) # clear spatial autocorrelation

# also test overall autocorrelation using Dharma
testSpatialAutocorrelation(mod.glmthin.resid, x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, plot = F)

# add coefficients to comparison dataframe
glmthin.summ = summary(mod.glm.thin)
glmthin.summ = data.frame(estimate = glmthin.summ$coefficients[,1], se = glmthin.summ$coefficients[,2], p = glmthin.summ$coefficients[,4])
glmthin.summ$mod = "mod.glm.thin"
glmthin.summ$var = rownames(glmthin.summ)

coefs.df = bind_rows(glm1.summ, glmthin.summ)
```

Compare coefs to SAR model
```{r}
coefs.df = bind_rows(coefs.df, sar.summ)

ggplot(coefs.df) +
  geom_pointrange(aes(y = estimate, ymin = estimate-se, ymax = estimate+se, x = var, color = mod),
                  position = position_dodge()) +
  coord_flip()
```

What do predictions from the different models look like?
```{r}
predict.glm1 = predict(mod.glm1, newdata = sesvert.scale, type = "response")
hist(predict.glm1)
hist(sesvert.scale$vert.mean.ses)

glm1.df = data.frame(x = sesvert.scale$x, y = sesvert.scale$y, resid = mod.glm1$residuals, pred = predict.glm1)
plot(glm1.df$pred, glm1.df$resid)
abline(a = 0, b = 0, col = "red3")

plot(sesvert.scale$vert.mean.ses, glm1.df$pred, ylim = c(-4,6))
abline(a = 0, b = 1, col = "red3")

ggplot(glm1.df) +
  geom_tile(aes(x=x, y=y, fill = resid)) +
  scale_fill_continuous_divergingx("spectral") +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  theme_classic()

predict.glmthin = predict(mod.glm.thin, newdata = sesvert.scale.thin, type = "response")
hist(predict.glmthin)

glmthin.df = data.frame(x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, pred = predict.glmthin, resid = mod.glm.thin$residuals)
plot(glmthin.df$pred, glmthin.df$resid)
abline(a = 0, b = 0, col = "red3")

plot(sesvert.scale.thin$vert.mean.ses, glmthin.df$pred, ylim = c(-4,3))
abline(a = 0, b = 1, col = "red3")

ggplot(glmthin.df) +
  geom_point(aes(x=x, y=y, color = resid), size = 2) +
  scale_color_continuous_divergingx("spectral") +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  theme_classic()
```

Looks like high values are still compressed as in SAR model
```{r}
plot(predict.glm1, mod.glm1$residuals)
plot(sesvert.scale$tmax_warm, mod.glm1$residuals)
plot(sesvert.scale$tmin_cold, mod.glm1$residuals)
plot(sesvert.scale$precip_wet, mod.glm1$residuals)
plot(sesvert.scale$precip_sea, mod.glm1$residuals)
plot(sesvert.scale$veg_complexity, mod.glm1$residuals)
plot(sesvert.scale$log_clim_velocity, mod.glm1$residuals)
plot(sesvert.scale$log_elev, mod.glm1$residuals)
plot(sesvert.scale$log_precip_dry, mod.glm1$residuals)

glm1.resids = data.frame(x = sesvert.scale$x, y = sesvert.scale$y, resid = mod.glm1$residuals, 
                         vert.mean.ses = sesvert.scale$vert.mean.ses, 
                         vert.mean.ses.pred = predict.glm1)

ggplot(glm1.resids) +
  geom_tile(aes(x = x, y = y, fill = resid)) +
  coord_sf(crs = "+proj=cea +datum=WGS84")

ggplot(glm1.resids) +
  geom_tile(aes(x = x, y = y, fill = vert.mean.ses - vert.mean.ses.pred)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral")

glmthin.resids = data.frame(x = sesvert.scale.thin$x, y = sesvert.scale.thin$y, resid = mod.glm.thin$residuals, 
                         vert.mean.ses = sesvert.scale.thin$vert.mean.ses, 
                         vert.mean.ses.pred = predict.glmthin)

ggplot(glmthin.resids) +
  geom_point(aes(x = x, y = y, color = vert.mean.ses - vert.mean.ses.pred)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_color_continuous_divergingx("spectral")

# now plot from SAR models
ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = vert.mean.ses - pred.pres.trend)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral") +
  theme_classic()

ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = vert.mean.ses - pred.signal)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral") +
  theme_classic()

pred.signal = predict(sesvert.avg)
pred.df$pred.signal = pred.signal

# plot predictions with signal
ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = pred.signal)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral", rev = T, limits = c(-5,6.8)) +
  theme_classic()

ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = vert.mean.ses)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral", rev = T, limits = c(-5,6.8)) +
  theme_classic()

ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = pred.pres.trend)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral", rev = T, limits = c(-5,6.8)) +
  theme_classic()

ggplot(pred.df) +
  geom_tile(aes(x = x, y = y, fill = pred.future)) +
  coord_sf(crs = "+proj=cea +datum=WGS84") +
  scale_fill_continuous_divergingx("spectral", rev = T, limits = c(-5,6.8)) +
  theme_classic()

plot(pred.df$vert.mean.ses, pred.df$pred.signal)
abline(a = 0, b = 1, col = "red3")

plot(pred.df$vert.mean.ses, pred.df$pred.pres.trend, ylim = c(-4,6))
abline(a = 0, b = 1, col = "red3")

plot(pred.df$vert.mean.ses, pred.df$vert.mean.ses - pred.df$pred.pres.trend, ylim = c(-4,6))
abline(a = 0, b = 0, col = "red3")

plot(pred.df$vert.mean.ses, pred.df$vert.mean.ses - pred.df$pred.signal, ylim = c(-4,6))
abline(a = 0, b = 0, col = "red3")
```

I suspect the model is misspecified, but I don't know what else to include in the model.
Something other than environmental variables is causing high verticality in the tropics

Try with phylogenetic distance in the models
```{r}
amph = read.csv("./../../../data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv")

sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", "mpd.obs.z", env_vars)] %>% 
  drop_na()

as.data.frame(sesvert) %>% 
  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
  ggplot(aes(x = val, y = vert.mean.ses)) +
  geom_point(col = "grey", alpha = 0.2, size = 1) +
  geom_smooth(col = "blue", method = "lm") +
  facet_wrap(facets = vars(vars), scales = "free_x") +
  theme_classic()

sesvert.t = as.data.frame(sesvert) %>% 
  mutate(log_clim_velocity = log10(clim_velocity),
         log_elev = log10(elev),
         log_precip_sea = log10(precip_sea),
         log_precip_dry = log10(precip_dry+1))

sesvert.t %>% 
  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
  ggplot(aes(x = val, y = vert.mean.ses)) +
  geom_point(col = "grey", alpha = 0.2, size = 1) +
  geom_smooth(col = "blue", method = "lm") +
  facet_wrap(facets = vars(vars), scales = "free_x") +
  theme_classic()

# keep log transformation for climate velocity and elevation
sesvert.t = sesvert.t %>% 
  dplyr::select(!c(clim_velocity, elev, log_precip_sea, precip_dry))
# add quadratic term to model for tmin_cold

# scale predictor variables
sesvert.scale = sesvert.t %>% 
  mutate_at(.vars = 6:ncol(sesvert.t), .funs = function(x){scale(x)})
colnames(sesvert.scale) = colnames(sesvert.t)

f = formula(vert.mean.ses ~ log_elev + log_clim_velocity + tmax_warm + tmin_cold + log_precip_dry + precip_sea + precip_wet + veg_complexity + mpd.obs.z)

```

Run a linear model including phylodistance
```{r}
mod.glm2 = glm(f, data = sesvert.scale)
plot(mod.glm2)

plot(sesvert.scale$vert.mean.ses, mod.glm2$fitted.values, ylim = c(-4,6))
abline(a = 0, b = 1, col = "red3")

glm2.summ = summary(mod.glm2)
glm2.summ = glm2.summ$coefficients
colnames(glm2.summ)[2] = "se"
colnames(glm2.summ)[4] = "p"
glm2.summ = as.data.frame(glm2.summ)
glm2.summ$var = rownames(glm2.summ)

ggplot(glm2.summ) +
  geom_pointrange(aes(x = Estimate, y = var, xmin = Estimate-se, xmax = Estimate+se, color = p<0.05)) +
  geom_vline(xintercept = 0)

```

```{r}
ggplot(glm1.df)
```

### 2024-07-18 scripts/03_analyses/glms_reptiles.R

@@ -1,919 +1,0 @@
-source("scripts_lydia/00_functions.R")
-source("scripts_lydia/00_icorrelogram.R")
-source("scripts_lydia/00_functions_predict_sarlm_debug.R")
-source("scripts_lydia/00_icorrelogram.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(foreach)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
-
-# variable selection without hurs and vpd
-# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Reptiles --------------------------------------------------------------
-
-repts = read.csv("data/derivative_data/gridcell_data/reptiles_comdat/rept_comdat_parallel.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = repts[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-# transform clim_velocity and temp seasonality and elevation
-# sesvert.t = as.data.frame(sesvert) %>% 
-#   mutate(log_clim_velocity = log10(clim_velocity),
-#          log_elev = log10(elev),
-#          log_temp_sea = log10(temp_sea))
-
-
[…]

### 2024-07-18 scripts/04_plots/env_maps.R
map environmental variables

@@ -1,0 +1,11 @@
+library(terra)
+library(tidyverse)
+library(tidyterra)
+library(data.table)
+
+env = fread("data/derivative_data/env_data.csv")
+
+ggplot(data = env, aes(x = x, y = y, fill = biome)) +
+  geom_raster() +
+  coord_sf(crs = "+proj=cea +datum=WGS84")
+

### 2024-07-18 scripts/04_plots/env_scatter_plots.R
env scatter plots present vs. future

@@ -1,0 +1,33 @@
+library(terra)
+library(tidyverse)
+
+env = read.csv("data/derivative_data/env_data.csv")
+
+env.future = list.files(path = "data/derivative_data/resampled_env_rasters/chelsa_future/2071_2100/ensemble/", pattern = ".tif", full.names = T)
+env.future = lapply(env.future, rast)
+env.future = rast(env.future)
+names(env.future)[6] = "precip_ann"
+env.future$temp_sea = env.future$temp_sea/100
+env.future = as.data.frame(env.future, xy = T)
+env.future = env.future %>% 
+  rename_at(.vars = 3:11, .funs = function(x) paste0(x, "_future"))
+
+load("results/amphibians_sar_vertmean2.RData")
+amph.vert = vert.scale.sub %>% 
+  dplyr::select(x,y,vert.mean, vert.mean.future) %>% 
+  left_join(env, by = c("x", "y")) %>% 
+  left_join(env.future, by = c("x", "y"))
+
+ggplot(amph.vert) +
+  geom_point(aes(x = precip_sea, y = precip_sea_future, color = vert.mean.future - vert.mean), size = 0.5) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  #facet_wrap(~realm) +
+  scale_color_continuous_divergingx("spectral") +
+  theme_classic()
+
+ggplot(amph.vert) +
+  geom_point(aes(x = tmin_cold, y = min_temp_cold_future, color = vert.mean.future - vert.mean), size = 0.5) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  facet_wrap(~realm) +
+  scale_color_continuous_divergingx("spectral") +
+  theme_classic()

### 2024-07-18 scripts/04_plots/plot_spatial_varying.R
map spatially varying coefficients

@@ -1,0 +1,168 @@
+library(terra)
+library(ggplot2)
+library(colorspace)
+library(tidyverse)
+library(patchwork)
+library(cowplot)
+
+wd = vect("data/original/rnaturalearth_world.shp") %>% 
+  project("epsg:4326") %>% 
+  crop(ext(-180,180,-60,83.64))
+
+plot_spatial_varying = function(mod, var, v, legend_title) {
+  coefs = tidy(mod)
+  coef.var = as.numeric(coefs[which(coefs$term == var), "estimate"])
+  pred = predict(mod, type = "response")
+  dat = pred %>% 
+    mutate(x = x*1e5, y = y*1e5,
+           canopy_height_est_slope = zeta_s_canopy_height + coef.var) %>% 
+    relocate(x, y, .before = rich) %>% 
+    rast(crs = "+proj=cea + datum=WGS84") %>% 
+    project("epsg:4326")
+    
+    
+  ggplot() +
+    geom_spatvector(data = v, color = "black", fill = "gray70") +
+    geom_spatraster(data = dat, aes(fill = canopy_height_est_slope)) +
+    geom_spatvector(data = v, color = "black", fill = NA) +
+    scale_fill_continuous_divergingx("BrBG", na.value = NA, guide = guide_colorbar(legend_title)) +
+    theme(axis.text = element_blank(),
+          axis.title = element_blank(),
+          axis.ticks = element_blank(),
+          panel.background = element_rect(color = "black", fill = NA),
+          panel.grid = element_blank())
+
+}
+
+# plot canopy height coefficienby canopy height standard deviation
+plot_coef_height = function(pred) {
+  height = pred %>% 
+    dplyr::select(x,y,canopy_height) %>% 
+    mutate(x = x*1e5, y = y*1e5) %>% 
+    rast(crs = "+proj=cea +datum=WGS84") %>% 
+    focal(w = 5, fun = sd, na.rm = T) %>% 
+    as.data.frame(xy = T) %>% 
+    mutate(x = round(x/1e5,5), y = round(y/1e5,5)) %>% 
+    inner_join(pred, by = c("x", "y")) %>% 
+    dplyr::select(x,y,biome, ecoregion, biorealm, zeta_s_canopy_height, focal_sd) %>% 
+    rename(canopy_height_sd = focal_sd)
+  
+  p = height %>% 
[…]

### 2024-07-18 scripts/04_plots/richness_vert_correlations.R
plot richness by verticality based on canopy height coefficient

@@ -1,0 +1,188 @@
+# plot verticality x richness for regions with canopy_coef > 1 and canopy_coef < 1
+
+library(ggplot2)
+library(tidyverse)
+library(patchwork)
+
+amph = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+birds = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
+rept = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv")
+mammals = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
+
+amph$taxa = "Amphibians"
+birds$taxa = "Birds"
+rept$taxa = "Reptiles"
+mammals$taxa = "Mammals"
+
+dat = rbind(amph, birds, rept, mammals)
+
+dat %>% 
+  filter(rich >= 5) %>% 
+  ggplot(aes(x = rich, y = vert.mean.ses, color = biome)) +
+    geom_point(pch = ".") +
+    geom_smooth(method = "lm") +
+    facet_grid(rows = vars(biome), cols = vars(taxa), scales = "free_x") +
+    theme_classic()
+
+dat %>% 
+  filter(rich >= 5) %>% 
+  ggplot(aes(x = rich, y = vert.mean, color = biome)) +
+  geom_point(pch = ".") +
+  geom_smooth(method = "lm") +
+  facet_wrap(~taxa, scales = "free_x") +
+  theme_classic()
+
+
+plot_cor = function(mod, pred, taxa) {
+  
+  coefs = tidy(mod)
+  coef.var = as.numeric(coefs[which(coefs$term == "canopy_height"), "estimate"])
+  
+  pred.canopy.pos = pred %>% 
+    filter(zeta_s_canopy_height + coef.var > 1)
+  
+  pred.canopy.n = pred %>% 
+    filter(zeta_s_canopy_height + coef.var <= 1)
+  
+  cor.p = cor.test(pred.canopy.pos$vert.mean.ses, pred.canopy.pos$rich)
+  cor.n = cor.test(pred.canopy.n$vert.mean.ses, pred.canopy.n$rich)
+
+  sesvert = pred %>% 
+    ggplot(aes(x = vert.mean.ses, y = rich, color = (zeta_s_canopy_height + coef.var) > 1, 
+               fill = (zeta_s_canopy_height + coef.var) > 1,
[…]

### 2024-07-18 scripts/04_plots/sar_coef_plot.R
plot SAR model coefficients

@@ -122,7 +122,7 @@ coefs = bind_rows(sesvert.coefs, meanvert.coefs) %>%
 library(ggh4x)
 coef.plt = coefs %>%
   ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
-  geom_pointrange() +
+  geom_pointrange(size = 0.15) +
   geom_vline(xintercept = 0, linetype = "dashed") +
   #geom_text(data = r2, aes(label = paste0("R\u00b2 = ", r2), x = Estimate), y = 2, inherit.aes = F, hjust = 1, vjust = 0) +
   #geom_text(data = r2, aes(label = paste0("R\u00b2 trend = ", r2.trend), x = Estimate), y = 0.9, inherit.aes = F, hjust = 1, vjust = 0) +

### 2024-07-18 scripts/04_plots/sdmTMB_conditional effects.R
plot conditional effects for sdmTMB models

@@ -1,0 +1,123 @@
+library(sdmTMB)
+library(foreach)
+library(doParallel)
+
+load("results/sdmTMB_models/amphibians_meanvert.RData")
+visreg(m.final, "tmax_warm")
+
+dat = m.final$data
+dat %>% 
+  mutate(tt = ifelse(abs(y) < 23.5, "Tropical", "Temperate")) %>% 
+  ggplot(aes(x = vert.mean.ses, y = rich, color = tt)) +
+  geom_point(, size = 0.05) +
+  geom_smooth()
+
+conditional_effects = function(m, random_int, ncore = 7, fpath) {
+  # m: sdmTMB model object
+  # random_int: name of variable used as a random intercept in the model
+  # ncore: number of cores for parallelization
+  # fpath: file path to save predictions to
+  
+  dat = m$data
+  terms = as.character(m$formula[[1]])[3]
+  terms = strsplit(terms, split = "\\+")[[1]]
+  terms = grep("\\(", terms, invert = T, value = T)
+  terms = str_trim(terms, side = "both")
+  
+  dat = dat[,which(colnames(dat) %in% terms)]
+  
+  # make matrix with median values
+  newdat = apply(dat, 2, median)
+  newdat = matrix(newdat, nrow = 100, ncol = length(newdat), byrow = T)
+  colnames(newdat) = colnames(dat)
+  
+  if(!is.null(random_int)){
+    randint = unique(m$data[,random_int])
+  }
+  
+  # replace variables one by one to get conditional predictions
+  cl = makeCluster(ncore)
+  registerDoParallel(cl)
+  cond_resp = foreach(i = 1:ncol(dat)) %dopar% {
+    var = seq(min(dat[,i]), max(dat[,i]), length.out = 100)
+    n = newdat
+    n[,i] = var
+    n = as.data.frame(n)
+    p = foreach(r = randint, .combine = "rbind") %do% {
+      n[,random_int] = rep(r, 100)
+      
+      # predict to newdata
+      pred = sdmTMB:::predict.sdmTMB(m, newdata = n, type = "response", re_form = NA, se_fit = F)
+    }
+    
+  }
+  stopCluster(cl)
+  names(cond_resp) = colnames(dat)
+  save(cond_resp, file = fpath)
+}
+
+
+# Unscale the data --------------------------------------------------------
+
[…]

### 2024-07-18 scripts/04_plots/vert_biome_hist.R
plot histograms of verticality by biome

@@ -1,0 +1,301 @@
+library(terra)
+library(tidyverse)
+library(data.table)
+library(colorspace)
+library(foreach)
+
+env = read.csv("data/derivative_data/env_data.csv")
+
+amph = fread("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, biome) %>% 
+  mutate(class = "Amphibians")
+
+bird = fread("data/derivative_data/gridcell_data/birds_comdat/birds_comdat_parallel_elton_forestsOnly.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, biome) %>% 
+  mutate(class = "Birds")
+
+mammals = fread("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel_forestsOnly.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, biome) %>% 
+  mutate(class = "Mammals")
+
+rept = fread("data/derivative_data/gridcell_data/reptiles_comdat/rept_comdat_parallel_forestsOnly.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, biome) %>% 
+  mutate(class = "Reptiles")
+
+df = bind_rows(amph, bird, mammals, rept) %>% 
+  mutate(class = factor(class, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")))
+
+df.mean = df %>% 
+  group_by(biome, class) %>% 
+  summarise(vert.mean.ses.mean = mean(vert.mean.ses, na.rm = T),
+            vert.mean.mean = mean(vert.mean, na.rm = T)) %>% 
+  mutate(class = factor(class, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")))
+
+plt_vertses_biome = ggplot(df, aes(x = vert.mean.ses)) +
+  geom_histogram() +
+  geom_vline(data = df.mean, aes(xintercept = vert.mean.ses.mean), linetype = "dashed", color = "red3", linewidth = 1) +
+  facet_grid(biome~class, scales = "free", labeller = label_wrap_gen(width = 5)) +
+  scale_x_continuous("SES Mean Verticality") +
+  theme(panel.grid.major = element_line(color = "gray80"),
+        panel.grid.minor = element_blank(),
+        panel.background = element_rect(color = "black", fill = NA),
[…]

### 2024-07-18 scripts/04_plots/vertdif_by_biorealm.R
plot change in verticality by biome

@@ -1,0 +1,147 @@
+library(terra)
+library(tidyterra)
+library(tidyverse)
+
+
+# Load SES vert and Mean vert models ----------------------------------------------------
+load("results/sdmTMB_models/amphibians_sesvert.RData")
+amph.sesvert = pred.f
+amph.sesvert$vertvar = "SES Mean Verticality"
+
+load("results/sdmTMB_models/amphibians_meanvert.RData")
+amph.meanvert = pred.f
+amph.meanvert$vertvar = "Mean Verticality"
+
+amph.pdat = amph.sesvert %>% 
+  bind_rows(amph.meanvert) %>%
+  dplyr::select(x,y,est.dif, vertvar) %>% 
+  mutate(class = "Amphibians")
+
+
+load("results/sdmTMB_models/mammals_sesvert.RData")
+mammals.sesvert = pred.f
+mammals.sesvert$vertvar = "SES Mean Verticality"
+
+load("results/sdmTMB_models/mammals_meanvert.RData")
+mammals.meanvert = pred.f
+mammals.meanvert$vertvar = "Mean Verticality"
+
+mammals.pdat = mammals.sesvert %>% 
+  bind_rows(mammals.meanvert) %>%
+  dplyr::select(x,y,est.dif, vertvar) %>% 
+  mutate(class = "Mammals")
+
+
+load("results/sdmTMB_models/birds_sesvert.RData")
+birds.sesvert = pred.f
+birds.sesvert$vertvar = "SES Mean Verticality"
+
+load("results/sdmTMB_models/birds_meanvert.RData")
+birds.meanvert = pred.f
+birds.meanvert$vertvar = "Mean Verticality"
+
+birds.pdat = birds.sesvert %>% 
+  bind_rows(birds.meanvert) %>%
+  dplyr::select(x,y,est.dif, vertvar) %>% 
+  mutate(class = "Birds")
+
+
+load("results/sdmTMB_models/reptiles_sesvert.RData")
+reptiles.sesvert = pred.f
+reptiles.sesvert$vertvar = "SES Mean Verticality"
+
+load("results/sdmTMB_models/reptiles_meanvert.RData")
+reptiles.meanvert = pred.f
+reptiles.meanvert$vertvar = "Mean Verticality"
+
+repts.pdat = reptiles.sesvert %>% 
+  bind_rows(reptiles.meanvert) %>%
+  dplyr::select(x,y,est.dif, vertvar) %>% 
+  mutate(class = "Reptiles")
+
[…]

### 2024-07-18 scripts/04_plots/vertdif_by_envdif.R
plot change in verticality by change in environment

@@ -1,0 +1,126 @@
+library(terra)
+library(tidyterra)
+library(tidyverse)
+
+load("results/amphibians_sar_sesvert2.RData")
+amph = sesvert.scale.sub
+
+amph.pdat = amph %>% 
+  rast(type = "xyz", crs = "+proj=cea +datum=WGS84") %>% 
+  #project("epsg:4326") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(sesvert.dif = vert.mean.ses.future-vert.mean.ses) %>% 
+  dplyr::select(x,y,sesvert.dif) %>% 
+  rename(sesvert.dif.amph = sesvert.dif)
+
+
+load("results/mammals_sar_sesvert2.RData")
+mammals = sesvert.scale.sub
+
+mammals.pdat = mammals %>% 
+  rast(type = "xyz", crs = "+proj=cea +datum=WGS84") %>% 
+  #project("epsg:4326") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(sesvert.dif = vert.mean.ses.future-vert.mean.ses) %>% 
+  dplyr::select(x,y,sesvert.dif) %>% 
+  rename(sesvert.dif.mammals = sesvert.dif)
+
+
+
+load("results/birdselton_sar_sesvert2.RData")
+birds = sesvert.scale.sub
+
+birds.pdat = birds %>% 
+  rast(type = "xyz", crs = "+proj=cea +datum=WGS84") %>% 
+  #project("epsg:4326") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(sesvert.dif = vert.mean.ses.future-vert.mean.ses) %>% 
+  dplyr::select(x,y,sesvert.dif) %>% 
+  rename(sesvert.dif.birds = sesvert.dif)
+
+
+load("results/reptiles_sar_sesvert2.RData")
+repts = sesvert.scale.sub
+
+repts.pdat = repts %>% 
+  rast(type = "xyz", crs = "+proj=cea +datum=WGS84") %>% 
+  #project("epsg:4326") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(sesvert.dif = vert.mean.ses.future-vert.mean.ses) %>% 
+  dplyr::select(x,y,sesvert.dif) %>% 
+  rename(sesvert.dif.repts = sesvert.dif)
+
+
+# env data ----------------------------------------------------------------
+
[…]

### 2024-07-18 scripts/04_plots/vert_maps.R
update verticality maps

@@ -95,7 +95,7 @@ library(colorspace)
   
  
   return(list(mp.pres, plt.final))
-}
+
 
 
 # plot ses vert mean ------------------------------------------------------
@@ -422,26 +422,30 @@ plt_vert_margin = function(v, df, plot.title, tags = list(c("A", "B", "C"))) {
   
     # create main plot
   plot_main = ggplot() +
-    geom_spatraster(data = r$vert.mean.ses, maxcell = Inf) +
     # Overlay world
-    geom_spatvector(data = wd, color = alpha("black", 0.7), fill = NA, linewidth = .1) +
-    #geom_segment(aes(x = xmax, xend = xmax, y = -60, yend = ymax)) +
+    geom_spatvector(data = wd, color = alpha("black", 0.7), fill = "gray70", linewidth = .1) +
+    geom_spatraster(data = r$vert.mean.ses, maxcell = Inf) +
+    geom_spatvector(data = wd, color = "black", fill = NA, linewidth = .1) +
     scale_fill_continuous_divergingx("spectral", na.value = NA, limits = col.lims, rev = T) +
-    guides(fill = guide_colorbar(title = "SES Mean Verticality")) +
+    guides(fill = guide_colorbar(title = "")) +
    # theme_classic() +
-    theme(legend.position = "bottom",
-          legend.margin = margin(0,0,0,0),
+    theme(legend.position = "inside",
+          legend.position.inside = c(0.6,0.45),
+          legend.justification = c(0,1),
+          legend.margin = margin(0,0,0.2,0.3),
           legend.key.height = unit(3, units = "mm"),
           legend.background = element_blank(),
-          legend.key.width = unit(10, units = "mm"),
+          legend.key.width = unit(2, units = "mm"),
           legend.box.margin = margin(0,0,0,0),
           legend.box.spacing = unit(0,"mm"),
           legend.box.background = element_blank(),
           plot.background = element_blank(),
+          plot.margin = unit(c(0,0,0,0), units = "mm"),
           panel.spacing = unit(0,"mm"),
+          #panel.background = element_rect(color = "black", fill = NA),
           panel.background = element_blank(),
           #legend.title = element_blank(),
-          legend.title.position = "top",
+          #legend.title.position = "top",
[…]

### 2024-07-18 scripts/03_analyses/01_sar_models/03_sar_amphibians.R
delete global SAR models

@@ -1,1051 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-#env_vars = vif_func(in_frame = env[,c(4:18,22)], thresh = 3, trace = T)
-
-# variable selection without hurs and vpd
-# env_vars = vif_func(in_frame = env[c(4:12,17:18,22)], thresh = 3, trace = T)
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Amphibians --------------------------------------------------------------
-
-amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel.csv")
-
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-# transform clim_velocity
-# sesvert.t = as.data.frame(sesvert) %>% 
-#   mutate(log_clim_velocity = log10(clim_velocity),
-#          log_temp_sea = log10(temp_sea)) %>% 
-#   dplyr::select(!c(clim_velocity, temp_sea))
-
-sesvert.t = as.data.frame(sesvert) %>% 
[…]

### 2024-07-18 scripts/03_analyses/02_sar_models_forest_only/sar_amphibians_forest_only.R
Delete forest only SAR models

@@ -1,870 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Amphibians --------------------------------------------------------------
-
-amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel.csv")
-
-
-# subset data to forested areas
-
-# option 1: subset by biome
-unique(amph$biome)
-amph.forests = amph %>% 
-  filter(grepl("Forests", biome))
-
-ggplot(amph.forests) +
-  geom_boxplot(aes(x = biome, y = veg_den))
-
-ggplot(amph) +
-  geom_boxplot(aes(x = biome, y = veg_den)) +
-  theme(axis.text = element_text(angle = 270))
-
-ggplot(amph) +
-  geom_boxplot(aes(x = biome, y = veg_complexity)) +
-  theme(axis.text = element_text(angle = 270))
-
-# urbanization might be impacting veg density
-veg = amph %>% 
-  dplyr::select(x,y,veg_den, veg_complexity, canopy_height, biome) %>% 
-  rast(type = "xyz", crs = "+proj=cea +datum=WGS84")
-
-plot(veg$veg_den)
-plot(veg$biome)
-
-# biorealm = vect("data/original/env_data/ecoregions/Ecoregions2017.shp")
-
-# ggplot(biorealm) +
-#   geom_spatvector(aes(fill = BIOME_NAME)) +
-#   coord_sf(crs=4326)
-
-# subset by biome to avoid threshold value
-
-amph = amph.forests
-rm(amph.forests)
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
[…]

### 2024-07-18 scripts/03_analyses/02_sar_models_forest_only/sar_birds_forest_only.R

@@ -1,802 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Birds elton --------------------------------------------------------------
-
-birds = read.csv("data/derivative_data/gridcell_data/birds_comdat/birds_breedresident_comdat_elton.csv")
-
-# subset to only forests
-birds = birds %>% 
-  filter(grepl("Forests", biome))
-
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = birds[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_precip_dry = log10(precip_dry),
-         log_elev = log10(elev),
-         log_precip_wet = log10(precip_wet),
-         log_precip_sea = log10(precip_sea))
-
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
[…]

### 2024-07-18 scripts/03_analyses/02_sar_models_forest_only/sar_mammals_forest_only.R

@@ -1,821 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(foreach)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Mammals --------------------------------------------------------------
-
-mammals = read.csv("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel.csv")
-
-mammals = mammals %>% 
-  filter(grepl("Forests", biome))
-
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = mammals[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea))
-
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
[…]

### 2024-07-18 scripts/03_analyses/02_sar_models_forest_only/sar_reptiles_forest_only.R

@@ -1,803 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(foreach)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Reptiles --------------------------------------------------------------
-
-repts = read.csv("data/derivative_data/gridcell_data/reptiles_comdat/rept_comdat_parallel.csv")
-
-repts = repts %>% 
-  filter(grepl("Forests", biome))
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = repts[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry))
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/50km/sar_amphibians.R
Delete forest only SAR models where SES of verticality was calculated using only forest species

@@ -1,360 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-library(nlme)
-library(car)
-library(ncf)
-library(sdmTMB)
-library(lme4)
-library(tidyverse)
-library(terra)
-
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-#env = read.csv("data/derivative_data/env_data_50km_forest.csv")
-
-# amphibian forest only 50km resolution
-dat = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
-
-# make key value dataframe for categorical rasters
-key.ecoregion = data.frame(ecoregion = unique(dat$ecoregion))
-key.ecoregion$key.ecoregion = 1:nrow(key.ecoregion)
-
-key.biome = data.frame(biome = unique(dat$biome))
-key.biome$key.biome = 1:nrow(key.biome)
-
-key.realm = data.frame(realm = unique(dat$realm))
-key.realm$key.realm = 1:nrow(key.realm)
-
-datr = dat  %>% relocate(x,y, .before = rich) %>% 
-  left_join(key.biome, by = "biome") %>% 
-  left_join(key.realm, by = "realm") %>% 
-  left_join(key.ecoregion, by = "ecoregion") %>% 
-  rast(type = "xyz", crs = "+proj=cea +datum=WGS84")
-plot(datr$vert.mean)
-plot(datr$vert.mean.ses)
-
-# variable selection 
-#env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-# randomly select 5000
-set.seed(12345)
-d = dat %>% 
-  # subset to richness >= 5
-  filter(rich >= 5) %>% 
-  filter(!is.na(vert.mean.ses)) %>% 
-  drop_na() %>% 
-  slice_sample(n=5000) %>% 
-  mutate_at(.vars = vars(canopy_height:clim_velocity, elev, veg_den, veg_complexity), .funs = scale)
-
-# look at the data
-# plot vert.mean.ses
-d %>% 
-  dplyr::select(vert.mean.ses, biome:clim_velocity, elev, veg_den, veg_complexity) %>%
-  pivot_longer(cols = 3:16, names_to = "var", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
-    geom_point(pch = ".") +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_amphibians.R

@@ -1,733 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Amphibians --------------------------------------------------------------
-
-amph = read.csv("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = amph[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry+1))
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_birds.R

@@ -1,736 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Birds --------------------------------------------------------------
-
-birds = read.csv("data/derivative_data/gridcell_data/birds_comdat/birds_comdat_parallel_elton_forestsOnly.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = birds[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry+1))
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_mammals.R

@@ -1,755 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Mammals --------------------------------------------------------------
-
-mammals = read.csv("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel_forestsOnly.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = mammals[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry+1))
-# add quadratic to tmaxwarm
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
[…]

### 2024-07-18 scripts/03_analyses/03_sar_models_forest_only_ses/sar_reptiles.R

@@ -1,748 +1,0 @@
-source("scripts/00_functions/00_functions.R")
-source("scripts/00_functions/00_functions_predict_sarlm_debug.R")
-library(tidyverse)
-library(vegan)
-library(pgirmess)
-library(ncf)
-library(spdep)
-library(sp)
-library(DHARMa)
-library(spatialreg)
-library(MuMIn)
-library(data.table)
-library(parallel)
-library(doParallel)
-library(terra)
-library(colorspace)
-library(patchwork)
-library(broom)
-
-options(na.action = "na.omit")
-
-
-# VIF variable selection
-env = read.csv("data/derivative_data/env_data.csv")
-
-# variable selection 
-env_vars = vif_func(in_frame = env[c(4, 7:12,17:18,22)], thresh = 5, trace = T)
-
-
-
-# Reptiles --------------------------------------------------------------
-
-reptiles = read.csv("data/derivative_data/gridcell_data/reptiles_comdat/rept_comdat_parallel_forestsOnly.csv")
-
-#  * - data check --------------------------------------------------------------
-
-## SES vert model
-sesvert = reptiles[,c("x", "y", "vert.mean.ses", "biome", "realm", env_vars)] %>% 
-  drop_na()
-
-hist(sesvert$vert.mean.ses)
-
-# check linear association between predictor and response
-as.data.frame(sesvert) %>% 
-  pivot_longer(cols = 6:ncol(sesvert), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
-  geom_smooth(col = "blue", method = "lm") +
-  facet_wrap(facets = vars(vars), scales = "free_x") +
-  theme_classic()
-
-sesvert.t = as.data.frame(sesvert) %>% 
-  mutate(log_clim_velocity = log10(clim_velocity),
-         log_elev = log10(elev),
-         log_precip_sea = log10(precip_sea),
-         log_precip_dry = log10(precip_dry+1))
-
-sesvert.t %>% 
-  pivot_longer(cols = 6:ncol(sesvert.t), names_to = "vars", values_to = "val") %>% 
-  ggplot(aes(x = val, y = vert.mean.ses)) +
-  geom_point(col = "grey", alpha = 0.2, size = 1) +
[…]

### 2024-07-18 scripts/01_get_data/03_get_bbs_data.R
Delete get bbs data

@@ -1,70 +1,0 @@
-# download breeding bird survey data for select regions and examine changes in arboreality over time
-
-library(bbsAssistant)
-library(data.table)
-
-sb_items
-View(region_codes)
-
-grab_bbs_data(bbs_dir = "data/original/bbs_data")
-
-dat = fread("data/original/bbs_data/50-StopData/50-StopData/1997ToPresent_SurveyWide/Fifty3/fifty3.csv")
-head(dat)
-
-dat = fread("data/original/bbs_data/50-StopData/50-StopData/1997ToPresent_SurveyWide/Fifty6/fifty6.csv")
-
-
-# FL state num
-fl = region_codes %>% filter(State == "florida")
-nc = region_codes %>% filter(State == "north carolina")
-
-dat.fl = dat %>% filter(StateNum == nc$StateNum)
-
-# calculate presab on route
-dat.fl[, pa := apply(.SD, 1, sum), .SDcols = Stop1:Stop50]
-dat.fl[, pa := ifelse(pa>0, 1, 0)]
-
-# reomve stop columns
-dat.fl[, 8:57 := NULL]
-
-# add species names
-head(species_list)
-splist = species_list %>% dplyr::select(AOU, Scientific_Name)
-
-dat.fl = left_join(dat.fl, splist, by = "AOU")
-dat.fl = dat.fl %>% rename(sciname = Scientific_Name)
-
-# verticality data for birds
-vert =  read.csv("data/derivative_data/species_data/birds_breedresident_spdat_eltonvert.csv")
-vert = vert %>% 
-  dplyr::select(sciname, Verticality) %>% 
-  mutate(sciname = gsub("_", " ", sciname))
-
-# combine vert data with bbs data for FL
-dat.fl = left_join(dat.fl, vert, by = "sciname")
-
-dat.fl.sum = dat.fl %>% 
-  filter(pa == 1) %>% 
-  group_by(RouteDataID, CountryNum, StateNum, Route, Year) %>% 
-  summarise(vert = mean(Verticality, na.rm = T))
-
-ggplot(dat.fl.sum, aes(x = Year, y = vert)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  theme_classic()
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-

### 2024-07-18 scripts/01_get_data/01_data_download.R
Document, clean, and rename data download and get env data

@@ -1,3 +1,6 @@
+# This script downloads CHELSA climate data for present and future time periods, canopy height, and elevation
+# For future climate data, the five ssp585 scenarios were downloaded (I later average these to obtain a future climate layer)
+
 # download climate data
 library(terra)
 library(numform)

### 2024-07-18 scripts/01_get_data/01_get_env_data.R

@@ -1,310 +1,0 @@
-library(terra)
-library(sf)
-library(fasterize)
-library(doParallel)
-library(cluster)
-library(foreach)
-
-# change tempdir so that I can easily delete temp files that take up too much space
-tempdir(tmpdir = "tempfiles/")
-
-# template raster used to get grid level pres/abs data at 111km
-mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
-#res(mapa) = 111000
-res(mapa) = 50000
-
-## Current climate conditions (CHELSA v2.1)
-
-## Chelsa climate data
-## Downloaded from https://chelsa-climate.org/downloads/
-bio = list.files(path = "data/original/env_data/chelsa/", pattern = ".tif", full.names = T)
-#bio = lapply(bio, rast)
-bio = bio[c(1:5,9:12)]
-
-cl = makeCluster(7)
-registerDoParallel(cl)
-bio = foreach(i = 1:length(bio), .packages = c("terra")) %dopar% {
-  mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
-  res(mapa) = 50000
-  r = rast(bio[[i]])
-  r = project(r, "+proj=cea +datum=WGS84")
-  r = resample(r, mapa, threads = T)
-  writeRaster(r, paste0("data/derivative_data/resampled_env_rasters_50km/", names(r), ".tif"))
-}
-stopCluster(cl)
-
-# bio = crop(bio, ext(-180, 180, -64.7937, 90))
-# 
-# tempsea = rast("data/original/env_data/chelsa/CHELSA_bio4_1981-2010_V.2.1.tif")
-# 
-# mat = project(bio$`CHELSA_bio1_1981-2010_V.2.1`, "+proj=cea +datum=WGS84")
-# mat = resample(mat, mapa, threads = T)
-# writeRaster(mat, "data/derivative_data/resampled_env_rasters/chelsa_bio_01.tif")
-# 
-# tic()
-# max_temp_warm = project(bio$`CHELSA_bio5_1981-2010_V.2.1`, "+proj=cea +datum=WGS84")
-# max_temp_warm = resample(max_temp_warm, mapa, threads = T)
[…]

### 2024-07-18 scripts/01_get_data/02_get_env_data.R

@@ -1,0 +1,248 @@
+# This code loads all environmental variables, projects to cylindrical equal area and 50 km resolution
+# Saves to a csv file (each row = 1 cell represented by x,y coords)
+
+library(terra)
+library(sf)
+library(fasterize)
+library(doParallel)
+library(cluster)
+library(foreach)
+
+# change tempdir so that I can easily delete temp files that take up too much space
+tempdir(tmpdir = "tempfiles/")
+
+# template raster used to get grid level pres/abs data at 111km
+mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
+#res(mapa) = 111000
+res(mapa) = 50000
+
+## Current climate conditions (CHELSA v2.1)
+
+## Chelsa climate data
+## Downloaded from https://chelsa-climate.org/downloads/
+bio = list.files(path = "data/original/env_data/chelsa/", pattern = ".tif", full.names = T)
+#bio = lapply(bio, rast)
+bio = bio[c(1:5,9:12)]
+
+# Resmaple chelsa climate data to 50 km resolution and project to cylindrical equal area projection
+cl = makeCluster(7)
+registerDoParallel(cl)
+bio = foreach(i = 1:length(bio), .packages = c("terra")) %dopar% {
+  mapa = rast(extent = c(-20592508, 20588492, -5743602, 6573398), crs = "+proj=cea +datum=WGS84")
+  res(mapa) = 50000
+  r = rast(bio[[i]])
+  r = project(r, "+proj=cea +datum=WGS84")
+  r = resample(r, mapa, threads = T)
+  writeRaster(r, paste0("data/derivative_data/resampled_env_rasters_50km/", names(r), ".tif"))
+}
+stopCluster(cl)
+
+
+# future climate ----------------------------------------------------------
+# Load future climate data files and make raster stacks for each scenario
+# gfdl-esm4-ssp585
+bio.f.gfdl = list.files(path = "data/original/env_data/chelsa_future/2071-2100/gfdl-esm4-ssp585/", pattern = ".tif", full.names = T)
+bio.f.gfdl = lapply(bio.f.gfdl, rast)
[…]

### 2024-07-19 scripts/01_get_data/02_get_sp_and_cell_data.R
Delete get sp and cell dat (111 km resolution)

@@ -1,351 +1,0 @@
-# Make species level and gridcell level dataframes for each class
-library(tidyverse)
-library(rnaturalearth)
-source("scripts/00_functions/00_functions.R")
-library(terra)
-library(data.table)
-library(tictoc)
-library(foreach)
-library(doParallel)
-
-### HARD CODES ###
-# phylogenetic imputation threshold for binary diel activity and verticality variables
-# greater than threshold = 1
-# less than or equal to threshold = 0
-impute_thresh = 0.7
-
-# extent of occurrence layer
-mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
-             crs = "+proj=cea +datum=WGS84")
-res(mapa) = 111000
-
-# terrestrial outline and gridcells
-# world = rnaturalearth::ne_countries()
-# world = vect(world)
-# world = as.polygons(world)
-# world = aggregate(world, dissolve = T)
-# writeVector(world, "data/original/rnaturalearth_world.shp")
-world = vect("data/original/rnaturalearth_world.shp")
-values(world) <- 1
-world.cea = project(world, "+proj=cea +datum=WGS84")
-world.rast = rasterize(world, mapa)
-values(world.rast) <- 1
-world.rast = mask(world.rast, world.cea)
-world.df = as.data.frame(world.rast, xy = T) %>% 
-  dplyr::select(x,y)
-
-# load env data
-env = read.csv("data/derivative_data/env_data.csv")
-ggplot() +
-  geom_raster(data =env, aes(x = x, y = y, fill = biome)) +
-  coord_sf(crs = "+proj=cea +datum=WGS84") +
-  theme_classic()
-
-ggplot() +
-  geom_raster(data =env, aes(x = x, y = y, fill = canopy_height>5)) +
-  coord_sf(crs = "+proj=cea +datum=WGS84") +
-  theme_classic()
-
-
-# Moura traits
-traits = read.csv("data/original/trait_data/TetrapodTraits_1.0.0/TetrapodTraits_1.0.0.csv")
-
-traits %>% 
-  filter(Palearctic == 1) %>% 
-  ggplot(aes(x = factor(Verticality), y = RangeSize)) +
-  geom_point()+
-  geom_boxplot() +
-  coord_cartesian(ylim = c(0,500))
-
-plot(traits$Verticality, traits$RangeSize)
-
-colnames(traits)
-
-traits = traits %>%
[…]

### 2024-07-19 scripts/01_get_data/02_get_bird_data.R
Delete get bird data.R

@@ -1,413 +1,0 @@
-# Lydia Soifer
-# May 1, 2024
-# Get arboreality and environmental data and add to the matrix from get_pres_abs_data
-
-library(terra)
-library(data.table)
-library(tictoc)
-library(chelsaDL)
-library(tidyverse)
-library(doParallel)
-library(parallel)
-library(foreach)
-library(mFD)
-source("scripts_lydia/00_functions.R")
-
-# REFERENCE PROJECTION MAP
-mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
-             crs = "+proj=cea +datum=WGS84")
-res(mapa) <- 111000
-
-mundi = vect("data/Mundi_contour/Mundi_contour.shp")
-mundi = project(mundi, "+proj=cea +datum=WGS84")
-mundi.rast = rasterize(mundi, mapa)
-mundi.df = as.data.frame(mundi.rast, xy = T) %>% 
-  dplyr::select(x,y)
-
-# load breeding bird occurrence data (cover0 includes all species, even if ranges cover less than 1/3 grid cell)
-occr <- fread("data/derivative_data/pres_abs_cover0/pres_abs_birds_breeding_resident.csv")
-colnames(occr)[1:2] = c("x", "y")
-colnames(occr) = gsub(" ", "_", colnames(occr))
-
-# remove ocean pixels
-occr = right_join(occr, mundi.df, by = c("x", "y"))
-
-# list of species from IUCN range maps
-spp <- colnames(occr)[-1:-2]
-spp <- data.frame(sciname = gsub(" ","_",spp))
-
-# load env data
-env = read.csv("data/derivative_data/env_data.csv")
-
-# Trait and Verticality data --------------------------------------------------------------
-
-# Using Elton traits  
-# Hamish Wilman, Jonathan Belmaker, Jennifer Simpson, Carolina de la Rosa, 
-# Marcelo M. Rivadeneira, and Walter Jetz. 2014. EltonTraits 1.0: Species-level 
-# foraging attributes of the world's birds and mammals. 
-# Ecology 95:2027. http://dx.doi.org/10.1890/13-1917.1
-
-### OPEN LARGE DATASET
-trait <- read.csv("data/original/trait_data/elton_traits/BirdFuncDat.txt", sep = "\t")
-
-names(trait)
-
-# Get traits of interest - ForStrat-Value, BodyMass-Value
-trait = trait %>% 
[…]

### 2024-07-19 scripts/01_get_data/01_get_pres_abs_data.R
file rename get_pres_abs_data

@@ -1,669 +1,0 @@
-# get gridded presence absence data from polygon range maps
-# Lydia Soifer
-
-library(terra)
-library(sf)
-library(letsR)
-library(data.table)
-library(tidyverse)
-library(tidyterra)
-
-# mammals
-# data = st_read("./../data/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
-# 
-# # test behaviour of letsR when there are multiple polygons for one species
-# u_arctos = data %>% filter(sci_name == "Ursus arctos")
-# u_arctos_pa = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80)
-# View(u_arctos_pa$Presence_and_Absence_Matrix) # combines all polygons into 1 for the species
-# plot(u_arctos_pa)
-# 
-# unique(u_arctos$seasonal)
-# 
-# u_arctos_nonbreed = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80, seasonal = 3)
-# plot(u_arctos_nonbreed)
-# 
-# u_arctos_breed = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80, seasonal = 1)
-# plot(u_arctos_breed)
-# 
-# rm(list = ls())
-# gc()
-
-# verticality data
-vert = read.csv("data/original/Moura_arboreality_vertebrates.csv")
-
-# mammal data
-data = vect("data/original/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
-
-vert_mammal = vert %>% dplyr::filter(Class == "Mammalia") %>% 
-  rename(sci_name = "Scientific.Name")
-
-# reduce data to only species for which we have verticality data
-data = data %>% inner_join(vert_mammal, by = "sci_name")
-
-# check if all polygons are valid
-valid = is.valid(data)
-for (n in which(valid == FALSE)) {print(n)}
-
-# reproject first because reprojecting after making valid undoes valid polygons
-data_reproj = project(data, "+proj=cea +datum=WGS84")
-data_reproj_valid = makeValid(data_reproj)
-
-# recheck polygon validity
-valid_check = is.valid(data_reproj_valid)
-which(valid_check == FALSE)
[…]

### 2024-07-19 scripts/01_get_data/03_get_pres_abs_data.R

@@ -1,0 +1,663 @@
+# get gridded presence absence data at 50 km resolution from polygon range maps
+# Lydia Soifer
+
+library(terra)
+library(sf)
+library(letsR)
+library(data.table)
+library(tidyverse)
+library(tidyterra)
+
+
+# Mammals -----------------------------------------------------------------
+
+# data = st_read("./../data/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
+# 
+# # test behaviour of letsR when there are multiple polygons for one species
+# u_arctos = data %>% filter(sci_name == "Ursus arctos")
+# u_arctos_pa = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80)
+# View(u_arctos_pa$Presence_and_Absence_Matrix) # combines all polygons into 1 for the species
+# plot(u_arctos_pa)
+# 
+# unique(u_arctos$seasonal)
+# 
+# u_arctos_nonbreed = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80, seasonal = 3)
+# plot(u_arctos_nonbreed)
+# 
+# u_arctos_breed = lets.presab(u_arctos, resol = 1, xmn = -190, xmx = 190, ymn = 18, ymx = 80, seasonal = 1)
+# plot(u_arctos_breed)
+# 
+# rm(list = ls())
+# gc()
+
+# verticality data
+vert = read.csv("data/original/Moura_arboreality_vertebrates.csv")
+
+# mammal data
+data = vect("data/original/ranges/MAMMALS_TERRESTRIAL_ONLY/MAMMALS_TERRESTRIAL_ONLY.shp")
+
+vert_mammal = vert %>% dplyr::filter(Class == "Mammalia") %>% 
+  rename(sci_name = "Scientific.Name")
+
+# reduce data to only species for which we have verticality data
+data = data %>% inner_join(vert_mammal, by = "sci_name")
+
+# check if all polygons are valid
+valid = is.valid(data)
+for (n in which(valid == FALSE)) {print(n)}
+
+# reproject first because reprojecting after making valid undoes valid polygons
+data_reproj = project(data, "+proj=cea +datum=WGS84")
+data_reproj_valid = makeValid(data_reproj)
+
+# recheck polygon validity
[…]

### 2024-08-02 docs/verticality_future_mkd.Rmd
intro revision

@@
 
-# Introduuction
+# Introduction
 
-Vertical stacking of species from beneath the ground to the top of the canopy is an emblematic signature of biologically rich regions and a major driver of biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
-In MacArthur's [-@macarthurPopulationEcologyWarblers1958] seminal work on verticality in warblers, he notes distinct partitioning of vertical niche space among species in a community.
-Since then, vertical stratification at local scales has been widely observed across diverse taxonomic groups [@bashamVerticalStratificationPatterns2023a].
-At regional scales, patterns of community verticality vary across environmental gradients, generally increasing in wetter habitats [@bashamVerticalStratificationPatterns2023a; @scheffersIncreasingArborealityAltitude2013].
-For example, verticality of frogs increases at higher elevations where cooler and wetter climates increase habitat suitability in the canopy relative to low elevations where hot and dry conditions compress frog communities toward the ground [@scheffersIncreasingArborealityAltitude2013].
-Globally, verticality increases toward the warmer and wetter tropics, where vertical partitioning may facilitate high biodiversity in structurally complex forests [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019].
-Despite widespread agreement that verticality influences patterns of biodiversity across spatial scales, global patterns of verticality and its environmental drivers, including climatic and vegetative factors, have not been comprehensively described for vertebrates.
+Vertical stacking of species is an emblematic signature of biologically rich regions and influences biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
+At local scales, species assemblages may be stratified across vertical climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
+These climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
+Stratification across this vertical climate gradient may reduce competition and increase species richness (*cite*).
+Tall, multi-layered forests are thus expected to support both highly diverse and highly arboreal species assemblages by increasing the volume of habitat and enabling species to finely partition the vertical forest structure [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+These conditions occur in tropical forests where high proportions of arboreal species may contribute to their biologically rich assemblages [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019].
+However, vegetation structure often fails to predict patterns of species richness at continental to global scales [@rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].
 
-## possible paragraph describing hypotheses about climate and verticality
-- Vertical stratification -> higher richness because species can partition the canopy
-- Therefore expect higher richness to be associated with increasing verticality
-- However, climate can limits verticality
-- We propose that climate limits verticality
-- Where climate is suitable, canopy height will positively impact verticality
-- where climate is not suitable, canopy height will have no impact on verticality
+At global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests.
+For example, low minimum temperatures in the canopies of high latitude forests may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
+Similarly, exposure to hot and dry conditions in the canopies of tropical lowland forests may reduce the suitability of arboreal habitats for heat or desiccation sensitive species, such as frogs [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020].
+Release of these climatic constraints at higher elevations, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013].
+We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological constraints.
+
+Similarly, spatial variation may occur in the relationship between species richness and assemblage verticality, defined here as the average vertical position of species in an assemblage [@oliveiraVerticalStratificationInfluences2019].
+While verticality and richness are often assumed to exhibit a positive relationship, the strength of this relationship is variable [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*.
+In regions where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weka because species are unable to vertically partition habitat.
+In contrast, when an increase in canopy height allows the addition of arboreal species to the assemblage, increasing assemblage verticality, we expect a strong positive relationship between verticality and species richness.
+
+As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness.
+In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and result in either shifts toward greater use of lower vertical positions or local extirpation [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
+Both scenarios would reduce assemblage verticality, reflecting climate-induced shifts in assemblage thermal tolerance caused by local extirpation or range shifts [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
+At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
+However, the extent to which climate change may alter global patterns of verticality is unknown.
+
+Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates.
+To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes.
+We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness.
+Finally, we predict how climate change may impact global patterns of verticality.
+
+Long-term climate instability may additionally impact global patterns of verticality [@scheffersVerticalArborealityHorizontal2017].
+Physiological adaptations to climate variability in the canopy may confer tolerance to extreme hot and dry conditions [@tracyNotJustSmall2010; @leahyArborealityDrivesHeat2022].
+Alternatively, vertical dispersal between the ground and canopy increases opportunities for behavioral adaptations to climate fluctuations [@scheffersIncreasingArborealityAltitude2013; @scheffersVerticalArborealityHorizontal2017; @bashamVerticalStratificationCollapses2020].
+However, observed impacts of historical climate instability on assemblage verticality are inconsistent [@scheffersVerticalArborealityHorizontal2017; @oliveiraVerticalStratificationInfluences2019].
 
 As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted.
-Changes in taxonomic composition of assemblages changes, are often associated with changes in functional composition.
-Consequently, widespread patterns of deborealization, tropicalization, and thermophilization have occurred in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
-Space-for-time substitutions can be used to predict impacts of climate change on the functional composition of species assemblages.
-For example, the thermal tolerance of assemblages displays predictable variation across elevation gradients.
-As the climate warms and species' ranges shift toward higher elevations, the thermal tolerance of assemblages increases as cool-adapted species become locally extirpated and warm-adapted species immigrate into communities (a process known as thermophilization) [@feeleyClimatedrivenChangesComposition2020].
+Functional changes to community composition are often detected in response to climate change.
+For example, deborealization, tropicalization, and thermophilization have occurred as community thermal tolerances increase in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
+Warming and drying climates may cause arboreal species to seek refuge at lower positions within the vertical forest structure, disperse to cooler climates toward higher latitudes or elevations, or become locally extinct.
+These processes will alter assemblage verticality.
+For example, verticality may decline in tropical lowlands where species are increasingly exposed to extreme hot and dry conditions [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
+In contrast, verticality may increase toward higher latitudes or elevations where warming temperatures may release constraints on arboreality.
+While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.
 
@@
 
-Relationships between climate and verticality across spatial and temporal environmental gradients have led to predictions that verticality may decline as climates become warmer and drier (cite).
-While forest understories are protected from extreme climatic conditions, the canopy experiences larger fluctuations in temperature and humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-Under climatically stressful conditions, arboreal species may move to lower vertical niche positions, reducing community verticality.
-Temporary vertical shifts have been observed for decades across a variety of biomes [@macarthurPopulationEcologyWarblers1958; @bashamVerticalStratificationCollapses2020; @kosheleffTemperatureInfluenceActivity2009; @eppleyFactorsInfluencingTerrestriality2022].
-For example, arboreal frogs in a Panamanian rainforest move toward the understory during the dry season to escape the hot and dry conditions in the canopy [@bashamVerticalStratificationCollapses2020].
-Similarly, arboreal primates from the Americas to Madagascar increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022].
-Yet how vertical niche shifts may alter global patterns of verticality as the climate warms is not well understood.
+Despite widespread agreement that verticality influences patterns of biodiversity and vulnerability to climate change, global patterns of verticality and its environmental drivers have not been comprehensively described for vertebrates.
 Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe.
-We then examine how climate change may impact environmental conditions that have led to present day patterns of verticality.
+We then examine how climate change may impact assemblage verticality.
 
@@
 To validate the best performing model, we refit it using all the data and calculated quantile residuals using the DHARMa R package.
+
+To validate the model, we examined quantile residuals to ensure normality and lack of spatial pattern.
 
@@
 
-# Environmental drivers of verticality
+## Environmental drivers of verticality
 
-Models of mean and SES verticality indicated some similarities between the environmental drivers of mean and SES verticality across taxonomic groups, yet noticeable differences remained.
-We hypothesized that canopy height would have an overall positive effect on verticality, but that its effect would vary over space, being stronger in tropical regions and lower in temperate regions, which generally have few arboreal vertebrates despite the presence of tall canopies.
-All of the best models included a spatially varying coefficient for canopy height.
-Overall, canopy height exhibited a positive effect on verticality of birds and mammals and weak positive effect or no effect on verticality of reptiles and amphibians (\@ref(fig:coef-plt).
-Over space the impact of canopy height was generally strongest in the Andes and Madagascar
+We expected verticality to increase with increasing canopy height, vegetation density, precipitation, and minimum cold temperatures.
+While the models were generally consistent with these hypotheses, there were some notable differences.
 
-If there is a positive relationship between canopy height and verticality, then taller canopies have more arboreal species
-If there is a negative relationship between canopy height and verticality, then shorter canopies have more arboreal species
-If there is not relationship between canopy height and verticality, then height of the canopy doesn't matter, potentially because climatic or evolutionary factors determine verticality.
+### need a section on model fit ###
+Models predicted verticality well.
+However, much of the predictive ability was due to the spatial random field.
+When the spatial random field was excluded from model predictions, models generally underestimated verticality in regions with high verticality and overestimated verticality in regions with low verticality.
 
-Across taxa, precipitation of the driest and wettest months positively influenced the SES of verticality, though the impact of dry season precipitation was noticeably stronger. 
-Impacts of climate on mean verticality generally mirrored those on the SES of verticality, though effects were smaller relative to the impact of canopy height.
+### Climate
 
-variable effects of climate and vegetation across taxa.
-Mean verticality of birds and mammals was most strongly influenced by canopy height, though responses to canopy height varied across space.
-Though climatic variables had
+The models generally indicated that verticality was limited by low precipitation, high maximum temperatures, and low minimum temperatures, though some variation occurred between taxa.
+Additionally, climatic impacts relative to the impact of canopy height was stronger for SES verticality than mean verticality.
 
-Overall, the mean and SES of verticality exhibited similar responses to environmental variables across taxonomic groups.
-Verticality was positively associated with precipitation of the driest and wettest months and exhibited a quadratic relationship with maximum temperature of the warmest month for mean and SES verticality of birds and reptiles and SES verticality of mammals.
-The quadratic relationship was characterized by an initial increase in verticality followed by a decline in verticality at extremely high maximum temperatures.
-Mean verticality of mammals exhibited a relatively small positive relationship with maximum temperature and amphibians  except for amphibians, for which SES of verticality was negatively associated with maximum temperatures.
-Minimum cold temperatures positively impacted verticality of mammals, such that verticality increased as minimum temperatures increased, while having little impact on birds and mammals.
-Impacts of climate velocity also varied, negatively impacting birds, mammals, and amphibians, while having a small positive effect on reptiles.
+Across taxa, precipitation of the driest and wettest months positively influences mean and SES verticality, though the impact of dry season precipitation on SES verticality was noticeable stronger (\@ref(fig:mod-coefs)).
+Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality.
+Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile verticality, as well as SES verticality of mammals.
+This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further.
+Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship.
+Climate velocity was not included in the best models for mammals verticality or for SES verticality of amphibians.
+For the remaining models, climate velocity (i.e., climate instability) negatively impacted bird verticality and mean verticality of amphibians and positively impacted reptile verticality.
 
-*For difference section - current geographic niche underfilling and future potential to fill unused geographic space by filling currently unoccupied volume in the canopy as climate permits these habitats to become more suitable*
+### Vegetation
 
-Though arboreal habitats remain in temperate biomes, the entire habitat volume is not occupied.
-This could be caused by arboreal habitats being climatically unsuitable in temperate regions and lead to a breakdown of the relationship between richness and volume.
-Overall, these patterns indicate a decline in the volume of habitable space toward the poles despite the structural presence of habitat.
+We hypothesized that canopy height and vegetation density would positively impact verticality, but that the impact of canopy height would vary across space.
+Specifically, we expected canopy height to have a strong positive effect on verticality in warmer and wetter regions that support arboreal assemblages and a negligible effect in regions with climate conditions that support fewer arboreal species.
 
--
+Globally, canopy height positively impacted verticality for mammals and birds and had little impact on verticality of reptiles and amphibians, while vegetation density had a negligible impact on verticality of all taxa (\@ref(fig:coef-plt).
+The best models for all taxa included a spatially varying coefficient for canopy height, which caused the wide confidence interval in the global estimate its coefficient.
 
-    ## climate velocity represents climate instability across geographic timescales and negatively impacted verticality across vertebrate classes
+Spatial variation in the effect of canopy height was greater for models of SES verticality than for mean verticality, though both exhibited similar patterns (\@ref(fig:svc-map)).
+For birds, canopy height exhibited the strongest positive effect on SES verticality in the Andes, central America, Indomalayan moist deciduous forests, the forest-savanna woodlands and thickets of eastern Africa, and northern taiga forests (\@ref(fig:svc-map)A,E).
+Spatial variability of the coefficient of canopy height for models of SES verticality exhibited similar pattern in mammals.
+However, strong positive effects additionally occurred in the south-eastern US, subpolar forests of South America, Guinean forest-savanna, and Australian temperate savanna and woodlands (\@ref(fig:svc-map)B,F).
+For reptiles, similar regions showed weaker positive relationships between canopy height and SES verticality.
+However, Madagascar exhibited a strong positive coefficient (\@ref(fig:svc-map)C,G).
+The coefficient of canopy height was also strongly positive for Amphibians in Madagascar as well as Sri Lanka relative to other regions (\@ref(fig:svc-map)D,H).
+Notably, the coefficient of canopy height was near zero in tropical broadleaf forests across the globe and was not negligible in cooler regions, contrary to our predictions.
 
--   Elevation
-
--   Vegetative complexity - selective logging could negatively impact verticality by reducing structural complexity
-
--   Decline in verticality in temperate regions despite the avialability of arboreal habitat space indicates physiological challenges of arboreal strategies in seaosnally cold and/or dry environmentals (Oliveira and Scheffers 2019)
-
--   for some vertebrate classes, fossoriality offers refuge from climate extremes
-
-*plot richness by area x structural complexity and color by latitude* *then plot richness x latitude* *consider adding correlation of verticality with species richness and discuss here*
+## Climate change impacts
 
 ```{r}
-load("./../results/amphibians_sar_vertmean2.RData")
-amph.avgmod = vert.avg
-amph.summ = summary(amph.avgmod)
-amph.summ = round(amph.summ$coefmat.full, 2)
-colnames(amph.summ) = c("Estimate", "SE", "z", "p")
+# calculate percent change in verticality
+calc_change = function(pred.f, taxa, response) {
+  pred.f = pred.f %>% 
+    dplyr::select(x,y,est.dif) %>% 
+    mutate(taxa = taxa,
+         response = response)
+return(pred.f)
+}
 
-load("./../results/mammals_sar_vertmean2.RData")
-mammals.avgmod = vert.avg
-mammals.summ = summary(mammals.avgmod)
-mammals.summ = round(mammals.summ$coefmat.full, 2)
-colnames(mammals.summ) = c("Estimate", "SE", "z", "p")
+load("./../results/sdmTMB_models/amphibians_meanvert.RData")
+amph.meanvert = calc_change(pred.f, taxa = "Amphibians", response = "Mean Verticality")
 
-load("./../results/reptiles_sar_vertmean2.RData")
-rept.avgmod = vert.avg
-repts.summ = summary(rept.avgmod)
-repts.summ = round(repts.summ$coefmat.full, 2)
-colnames(repts.summ) = c("Estimate", "SE", "z", "p")
+load("./../results/sdmTMB_models/amphibians_sesvert.RData")
+amph.sesvert = calc_change(pred.f, taxa = "Amphibians", response = "SES Verticality")
 
-load("./../results/birdselton_sar_vertmean.RData")
-birds.avgmod = vert.full
-birds.summ = summary(birds.avgmod) # make sure spatialreg is loaded for summary to produce correct results
-birds.summ = round(birds.summ$Coef, 2)
-colnames(birds.summ) = c("Estimate", "SE", "z", "p")
+load("./../results/sdmTMB_models/reptiles_meanvert.RData")
+rept.meanvert = calc_change(pred.f, taxa = "Reptiles", response = "Mean Verticality")
+
+load("./../results/sdmTMB_models/reptiles_sesvert.RData")
+rept.sesvert = calc_change(pred.f, taxa = "Reptiles", response = "SES Verticality")
+
+load("./../results/sdmTMB_models/mammals_meanvert.RData")
+mammals.meanvert = calc_change(pred.f, taxa = "Mammals", response = "Mean Verticality")
+
+load("./../results/sdmTMB_models/mammals_sesvert.RData")
+mammals.sesvert = calc_change(pred.f, taxa = "Mammals", response = "SES Verticality")
+
+load("./../results/sdmTMB_models/birds_meanvert.RData")
+birds.meanvert = calc_change(pred.f, taxa = "Birds", response = "Mean Verticality")
+
+load("./../results/sdmTMB_models/birds_sesvert.RData")
+birds.sesvert = calc_change(pred.f, taxa = "Birds", response = "SES Verticality")
+
+dat = rbind(amph.meanvert, amph.sesvert, rept.meanvert, rept.sesvert, mammals.meanvert, mammals.sesvert, birds.meanvert, birds.sesvert)
+
+dat.change = dat %>% 
+  group_by(taxa, response) %>% 
+  summarise(max = max(est.dif), min = min(est.dif))
+
+dat.change.ses = dat.change %>% filter(response == "SES Verticality")
+dat.change.mean = dat.change %>% filter(response == "Mean Verticality")
+
+ses.poschange = dat.change.ses[which(dat.change.ses$max == max(dat.change.ses$max)),]
+ses.negchange = dat.change.ses[which(dat.change.ses$min == min(dat.change.ses$min)),]
+
+mean.poschange = dat.change.mean[which(dat.change.mean$max == max(dat.change.mean$max)),]
+mean.negchange = dat.change.mean[which(dat.change.mean$min == min(dat.change.mean$min)),]
 
@@
 
-# Environmental drivers of verticality
-
-## Climate
-
-Global patterns of verticality could largely be explained by environmental drivers.
-To examine the impacts of climate, topography, and vegetation on global patterns of verticality, we used simultaneous autoregressive (SAR) models with a spatial error term.
-The best models explained explained 88-98% of the global variance in the mean and SES of mean verticality across classes (\@ref(fig:model-coefs)).
-However, the spatial signal in all models was strong (lambda \> 0.9), which resulted in relatively small impacts of environmental variables as well as a lower pseudo-R^2^ value when calculated based only on the predicted trends without the spatial signal.
-
-Precipitation of the driest quarter and minimum temperatures of the coldest quarter had the largest effects on the mean and SES of verticality across vertebrate classes, both exhibiting positive effects barring the impact of minimum temperatures on amphibians, which was insignificant ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)).
-Maximum temperatures of the warmest quarter had opposing effects on endotherms and ectotherms, negatively impacting verticality of reptiles and amphibians while generally having positive or insignificant effects on birds and mammals, except at the highest temperatures where verticality declined ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)).
-Arboreality is therefore primarily constrained by cold minimum temperatures and limited precipitation during the dry season, while high maximum temperatures additionally limit arboreal strategies in ectotherms.
-The vertical climate gradient from beneath the ground to the top of the canopy drives these global constraints on arboreality.
-Greater climatic fluctuations in the canopy coupled with water limitation pose physiological challenges to arboreal lifestyles.
-For example...
-In contrast, terrestrial and fossorial habitats are buffered from climate extremes, which facilitates species persistence where hot, cold, or dry conditions may prevent the occurrence of arboreal lifestyles.
-
-Dry conditions negatively impact arboreal lizard, particularly under simultaneously hot environments. <https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1301>
-
-```         
-- Though arboreal habitats can exhibit average temperatures over 2C higher than terrestrial habitats (Scheffers et al.
-2013) critical thermal maxima of arboreal species may not substantially exceed that of terrestrial species (Carilo Filho et al.
-2021)
-```
-
-## Historical Climate - climate velocity
-
-Verticality is also generally higher in regions that have experienced historically stable climatic conditions, as indicated by the significant negative relationships between climate velocity and the mean verticality of birds, mammals, and amphibians and SES of verticality for mammals, as well as the absence of positive relationships for any taxa.
-Low climate velocity (i.e., climate stability) generally increases speciation rates and decreases extinction rates (Colville et al.
-2020).
-In tropical forests, complex vegetation structure characterized by multilayered forests (Terborh 1985) increases opportunities for diversification among arboreal niches (see Colville et al.
-2020 11-13).
-For example, arboreal lineages of reptiles and frogs generally exhibit higher diversification rates [@liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017], while fossorial strategies have been associated with reduced speciation and increased extinction rates in snakes [@cyriacDiggingTheirOwn2018].
-
--   climate stability may therefore provide the conditions necessary for diversification in regions where climatic conditions are suitable for arboreal strategies, while increased dspersal capacity of arboreal species throughout 3D arboreal environments facilitates persistence in climatically unstable regions (Sandel et al.
-2011; Scheffers et al.
-2017).
-
-```{r}
-bird.meanvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_meanvert.RData")
-bird.meanvert = pred.df
-bird.meanvert$vertvar = "mean"
-bird.sesvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_sesvert.RData")
-bird.sesvert = pred.df
-bird.sesvert$vertvar = "ses"
-bird.pred = bind_rows(bird.meanvert, bird.sesvert) %>% mutate(class = "bird")
-
-mammals.meanvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_meanvert.RData")
-mammals.meanvert = pred.df
-mammals.meanvert$vertvar = "mean"
-mammals.sesvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_sesvert.RData")
-mammals.sesvert = pred.df
-mammals.sesvert$vertvar = "ses"
-mammals.pred = bind_rows(mammals.meanvert, mammals.sesvert) %>% mutate(class = "mammals")
-
-rept.meanvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_meanvert.RData")
-rept.meanvert = pred.df
-rept.meanvert$vertvar = "mean"
-rept.sesvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_sesvert.RData")
-rept.sesvert = pred.df
-rept.sesvert$vertvar = "ses"
-rept.pred = bind_rows(rept.meanvert, rept.sesvert) %>% mutate(class = "rept")
-
-amph.meanvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_meanvert.RData")
-amph.meanvert = pred.df
-amph.meanvert$vertvar = "mean"
-amph.sesvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_sesvert.RData")
-amph.sesvert = pred.df
-amph.sesvert$vertvar = "ses"
-amph.pred = bind_rows(amph.meanvert, amph.sesvert) %>% mutate(class = "amph")
-
-pred = bind_rows(bird.pred, mammals.pred, rept.pred, amph.pred)
-
-```
-
-# Impacts of climate change on verticality
-
-To examine the potential impacts of climate change on the verticality of vertebrate assemblages, we used the coefficients from the SAR models to predict assemblage verticality for the time period 2071-2100 using the SSP585 future climate scenario, which represents the upper boundary of predicted warming under a scenario of fossil-fueled development.
-Though climate change will impact vegetation structure, we were unable to incorporate predictions of future vegetation into the models due to a lack of global future predictions of vegetation height and density.
-Because we could not predict the spatial signal in verticality, we compared predictions under future climate conditions to predictions under present climate conditions that do not incorporate the spatial signal.
-- We find that x% of communities could experience declines in verticality.
-- Declines in verticality are concentrated in the tropics.
-- Between 23 and -23 latitude, x% birds, x% mammals, x% reptiles, and x% amphibians exhibit declines in verticality - Declines in the Amazon were notably severe
+Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality.
+The largest changes in SES verticality were observed for birds, which exhibited a maximum decline of `r round(as.numeric(ses.negchange[1, "min"]),2)` and a maximum increase of `r round(as.numeric(ses.poschange[1, "max"]),2)`.
+The larges changes in mean verticality were observed for amphibians, which exhibited a maximum decline of `r round(as.numeric(mean.negchange[1, "min"]), 2)` and a maximum increase of `r round(as.numeric(mean.poschange[1, "max"]), 2)`.
+For birds and mammals, mean and SES verticality generally declined in tropical moist broadleaf forests globally in addition to the temperate forests, savannas, and woodlands of southern Australia and the Brazilian Atlantic forests.
+In contrast, verticality primarily increased in the boreal forests of Canada and northern Eurasia, tropical and subtropical savannas and woodlands in Africa, and tropical savannas in northern Australia (\@ref(fig:vert-dif-biome)).
+Changes in verticality of Reptiles and Amphibians were characterized by declines in most regions.
+However, amphibian verticality was expected to experience minor increases in African tropical and subtropical savannas and woodlands (\@ref(fig:vert-dif-biome)).
 
 ## discussion of climate change
-
--   What causes changes in verticality - vertical niche shift, immigration, or local extinction?
 
@@
 
--   In temperate regions, we find...
-
--   Predicted changes in verticality led to homogenization of vertical strategies globally.
-
 # Discussion
-
-In the tropics, the use of arboreal habitat likely contributes to high species richness (cite Oliveira, also see Naikatini et al.
-2022, MacArthur).
-The large volume of habitat in tropical biomes facilitates species packing within a given area as well as expansion across vertically stratified resources.
-However, habitat volume and structural diversity is generally not related to species richness in forested ecosystems in the US, except for birds (La Rue et al.
-2023).
-Low arboreality toward higher latitudes may explain the breakdown in the volume-richness relationship and be caused by low climatic suitability of arboreal habitats in temperate climates, except for birds which continue to use arboreal habitats at northern latitudes.
-
--increasing arboreality at norther latitudes suggests that climate change could lead to arboreal habitats in boreal regions becoming more suitable for species, thereby releasing competitive pressure as species shift poleward by providing more habitat volume.
--niche attrition - land use change eliminates the volume of arboreal habitats, which leads to declines in arboreality.
-Will climate change cause further attrition of arboreal habitats??
-- elevator to extinction for mountaintop species = digging a grave
-
--   vertical niche shifts occur in response to novel habitat, global change, etc.
--   e.g., shift in birds to arboreal foraging on islands (Lapiedra et al. 2021 Niche shifts after island colonization...)
-
-Despite the presence of biomes with vertical structure in temperate forests the proportion of arboreality steeply declines, suggesting that the geographic niche space is underfilled by the animal community.
-
-Low climatic suitability may limit arboreality and thus diversity in temperate biomes.
-
-## The influence of vegetation
-Indeed, it is expected that the volume of habitat space explain variation in species richness among forested biomes.
-However, this relationship is not upheld for birds and mammals (LaRue et al.
-2023).
-Declining arboreality in temperate forests that have the vegetation structure to support arboreal strategies may explain this pattern and be caused by environmental factors limiting the suitability of canopy habitats.
-
-# Methods
 

### 2024-08-02 docs/verticality_future_mkd.Rmd
methods revision

@@
 As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness.
-In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and result in either shifts toward greater use of lower vertical positions or local extirpation [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
-Both scenarios would reduce assemblage verticality, reflecting climate-induced shifts in assemblage thermal tolerance caused by local extirpation or range shifts [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
+In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
+These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
 At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
-However, the extent to which climate change may alter global patterns of verticality is unknown.
-
-Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates.
-To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes.
-We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness.
-Finally, we predict how climate change may impact global patterns of verticality.
-
-Long-term climate instability may additionally impact global patterns of verticality [@scheffersVerticalArborealityHorizontal2017].
-Physiological adaptations to climate variability in the canopy may confer tolerance to extreme hot and dry conditions [@tracyNotJustSmall2010; @leahyArborealityDrivesHeat2022].
-Alternatively, vertical dispersal between the ground and canopy increases opportunities for behavioral adaptations to climate fluctuations [@scheffersIncreasingArborealityAltitude2013; @scheffersVerticalArborealityHorizontal2017; @bashamVerticalStratificationCollapses2020].
-However, observed impacts of historical climate instability on assemblage verticality are inconsistent [@scheffersVerticalArborealityHorizontal2017; @oliveiraVerticalStratificationInfluences2019].
-
-As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted.
-Functional changes to community composition are often detected in response to climate change.
-For example, deborealization, tropicalization, and thermophilization have occurred as community thermal tolerances increase in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
-Warming and drying climates may cause arboreal species to seek refuge at lower positions within the vertical forest structure, disperse to cooler climates toward higher latitudes or elevations, or become locally extinct.
-These processes will alter assemblage verticality.
-For example, verticality may decline in tropical lowlands where species are increasingly exposed to extreme hot and dry conditions [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
-In contrast, verticality may increase toward higher latitudes or elevations where warming temperatures may release constraints on arboreality.
 While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.
@@
 nrept = nrow(repts)
-
 ```
 
-Despite widespread agreement that verticality influences patterns of biodiversity and vulnerability to climate change, global patterns of verticality and its environmental drivers have not been comprehensively described for vertebrates.
-Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe.
-We then examine how climate change may impact assemblage verticality.
+Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates.
+To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes.
+We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness.
+Finally, we predict how climate change may impact global patterns of verticality.
 
@@
 
-## Assemblage verticality
+## Species data
 
-We mapped the verticality of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection.
+We mapped the verticality and richness of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection.
+Grid cells with fewer than five species were excluded.
 We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al.
@@
 Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
-For birds, we only used ranges that covered breeding and resident ranges, because seasonality impacts the verticality of assemblages.
+For birds, we used breeding plus resident ranges, because seasonality impacts the verticality of assemblages.
 To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell.
@@
 Verticality exhibited a strong positive correlation with species richness.
-To examine spatial patterns of verticality independently of species richness, we claculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool) .
+To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool).
 We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.
-
- To minimize impacts of the positive relationship between verticality and species richness on global patterns, we calculated the standard effect size (SES) of mean assemblage verticality, which determines whether verticality is higher (positive) or lower (negative) than expected by random chance given the species richness in each cell and the vertical strategies of species in each geographic realm.
 
@@
 
-We obtained present and future global climate data from CHELSA v2.1 [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
+We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
 Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100.
-Climate velocity represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011].
-Canopy height was optained from [@langHighresolutionCanopyHeight2023] and vegetaion dnesiy Vegetative complexity was calculated as the product of vegetation density [@crowtherMappingTreeDensity2015] and canopy height [@langHighresolutionCanopyHeight2023].
-Elevation data was obtained from *data source*.
-All environemtnal data was resmapled to the resolution and projection of the assemblage data.
+Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density).
+Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality (Oliveira and Scheffers 2019).
+All environmental data was resampled to the resolution and projection of the assemblage data.
 
 ## Analysis
+
+### Variable selection
+We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity.
+To linearize the relationships between predictor and response variables, we log-transformed climate velocity and precipitation of the driest month.
+We then scaled all predictor variables to mean zero and variance one.
+To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than 5 [@wartonEcostatsDataAnalysis2022].
+We chose variables to remove based on a combination of which had the highest VIF, which were most likely to directly impact physiological limits to arboreality, and which were expected to change the most in response to climate change.
+For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation, has a less direct impact on physiological limits to verticality than dry season precipitation, and varies less between current and future climate conditions than dry season precipitation.
+This procedure resulted in a final set of seven variables that included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the wettest month, precipitation of the driest month, and climate velocity.
+The additive effects of these variables were included as fixed effects in the model in addition to a quadratic term for maximum temperature of the warmest month, as we expected that verticality may increase with temperature up to a certain point but then decline under extremely high temperatures.
+
+### Model description
+
+We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of environmental variables on mean and SES verticality.
+Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields.
+[@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015].
+sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent, and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].
+
+The general model can be described as:
+
+$$
+y_s ∼ 𝜇_{s}, \\
+𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + α_g + 𝜲^{svc}_sζ_s + ⍵_s), \\
+⍵_s ∼ MVNormal(0,𝚺_⍵)
+$$
+
+where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function.
+We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1.
+$𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients.
+$α_g$ represents an optional random intercept by group g.
+$⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept.
+$𝜲^{svc}_sζ_s$ represents an optional spatially varying coefficient, where $ 𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$).
+The spatial range for the covariance function is shared with the range for the spatial random field.
+Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with the terms following a multivariate normal distribution with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
+The model is specified such that both spatial terms share the same range parameter.
+
+### Mesh development
+
+The Gaussian random field is approximated using a spatial mesh.
+We constructed separate meshes for all taxa and for mean and SES verticality using the ‘fmesher’ R package.
+The mesh used to approximate the spatial Gaussian random field (GRF) impacts model results as well as computational time.
+To assess sensitivity of the model to the mesh and select the optimal mesh, we used an iterative procedure to update minimum and maximum edge lengths of triangles in the mesh based on the spatial range of the data (i.e., the distance at which points exhibit a correlation of approximately 0.1) [@thorsonSpatioTemporalModelsEcologists2024].
+We first estimated the spatial range by visually inspecting spatial correlations based on a subset of 1000 points.
+To avoid over-smoothing the GRF while or over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024].
+To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range.
+Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range (*cite*).
+To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
+We then fit a model using the resulting mesh.
+If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model.
+We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of spatial domain used to construct the mesh.
+We selected the mesh used in the final model for all following analysis.
+
+### Model comparisons
 

### 2024-08-04 docs/verticality_future_mkd.Rmd
revise results

@@
 library(spatialreg)
+library(data.table)
 
@@
 ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
+```
+
+```{r}
+word_comment <- function(
+  comment, 
+  highlight = "", 
+  author = "Lydia", 
+  time = Sys.time(),
+  id = "0"
+) 
+{
+  if (isTRUE(knitr:::pandoc_to() == "docx")) {
+    return(
+      sprintf(
+        '[%s]{.comment-start id="%s" author="%s" date="%s"} %s []{.comment-end id="%s"}',
+        comment,
+        id,
+        author,
+        time,
+        highlight,
+        id
+      )
+    )
+  } else {
+    return(
+      sprintf(
+        "*%s* **[Comment id %s by %s at time %s: %s]**", 
+        highlight,
+        id,
+        author, 
+        time, 
+        comment
+      )
+    )
+  }
+}
 ```
@@
 ```{r}
-birds = read.csv("./../data/derivative_data/species_data/birds_breedresident_spdat_eltonvert_forestsOnly.csv")
+birds = read.csv("./../data/derivative_data/species_data/birds_spdat_elton_forestsOnly.csv")
 mammals = read.csv("./../data/derivative_data/species_data/mammals_spdat_moura_forestsOnly.csv")
@@
 ```{r}
-mammals = fread("data/derivative_data/gridcell_data/mammals_comdat/mammals_comdat_parallel_forestsOnly.csv") %>% 
+mammals = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
 
-amphibians = fread("data/derivative_data/gridcell_data/amphibians_comdat/amph_comdat_parallel_forestsOnly.csv") %>% 
+birds = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
 
-mammals.arb.tropicalmoist = round(mean(mammals$p.arb[mammals$biome == "Tropical & Subtropical Moist Broadleaf Forests"]) * 100, 2)
+amphibians = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
+
+rept = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv") %>% 
+  dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
+
+mammals.arb.tropicalmoist = round(mean(mammals$p.arb[mammals$biome == "Tropical & Subtropical Moist Broadleaf Forests"], na.rm = T) * 100, 2)
+birds.arb.tropicalmoist = round(mean(birds$p.arb[birds$biome == "Tropical & Subtropical Moist Broadleaf Forests"], na.rm = T) * 100, 2)
+rept.arb.tropicalmoist = round(mean(rept$p.arb[rept$biome == "Tropical & Subtropical Moist Broadleaf Forests"], na.rm = T) * 100, 2)
+amphibians.arb.tropicalmoist = round(mean(amphibians$p.arb[amphibians$biome == "Tropical & Subtropical Moist Broadleaf Forests"], na.rm = T) * 100, 2)
+
 ```
@@
 
-We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of vegetative and climatic paramaters on mean and SES verticality.
-Spatial models were fit using R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and approximates spatial Gaussian random fields using stochastic partial differential equation (SPDE)-based approach [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015].
-To implement the SPDE approach, we constructed a mesh using fmesher with a 300 km minimum allowable distance between vertices.
-We fit five alternative models for each taxonomic group, using the same fixed environmental predictors and varying the form of the random-effects structure.
-The general model is described as
+For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structures [@zuurMixedEffectsModels2009].
+The base model only included a spatial random field.
+Subsequent models additionally included biogeographic realm [sensu @dinersteinEcoregionBasedApproachProtecting2017] as a random intercept, a spatially varying coefficient for canopy height, and the combination of both.
+Including realm as a random intercept accounts for evolutionary differences among biogeographic realms that could lead to differences in verticality.
+The spatially varying coefficient allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.
+We compared models using AIC and cross-validation.
+AIC is frequently used to identify the most parsimonious model [e.g., @barnettImprovingEstimatesSpecies2021]; however, the metric is not completely appropriate for hierarchical models (see Commander et al.
+2022 `r word_comment(comment = "check citation", highlight = "Commander et al.
+2022")`).
+Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability (*again check citation Commander et al.
+2022*).
+We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing.
+The sum of the log-likelihoods for each left-out fold provides a measure of predictive ability that is compared across models.
+Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package.
+Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Table S1 `r word_comment(comment = "add supp table", highlight = "Table S1")`).
+Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.
 
-$$
-y_s ∼ 𝜇_{s}, \\
-𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + ⍺_g + ⍵_s), \\
-⍵_s ∼ MVNormal(0,𝚺_⍵)
-$$
+### Model validation
 
-where $y_s$ is the mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality at all locations, and $𝑓^{-1}$ represpents an inverse link function.
-$𝜲^{main}_s$ is a vector of main effect covariates, 𝛽 represents a vector of estimated corresponding main effect coefficients.
-We used a Gaussian distribution for SES verticality, while we used a Beta distribution with a logit link function for mean verticality because values are constratined between 0 and 1.
-Environmental covariates included canopy height, vegetation density, minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, precipitation of the wettest month, precipitation seasonality, and climate change velocity since the LGM.
-Fitting models, we log10 transformed climate velocity and precipitation of the driest quarter to increase the linearity of the relationship between predictor and response and scaled all predictor variable to have a mean of zero and a standard deviation of 1.
-$⍺_g$ represents an optional IID random intercept defined by group $g$ ($⍺_g ∼ Normal(0,𝜎^2_⍺))$.
-The five alternative models each used a different variable to define the random intercept: biome, realm, biorealm, ecoregion, and no random intercept.
-$⍵_s$ represents a spatial random field at location $s$, which represents additional spatial variation in the data not explained by the environmental covariates or random intercept.
-The estimated covariance matrix ($𝚺_⍵$) was constrained by the Matérn covariance function, which defines the decay rate of correlation with distance.
+We validated models first using the ‘sanity’ function from the sdmTMB package.
+We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space, rank transformed model predictions, raw model predictions.
 
-We compared alternative models using five-fold cross validation.
-Grid cells were randomly assigned to one of five folds.
-Each model was the run five times, training the model using four of the folds (80% of the data) and testing the model on the withheld fold (20% of the data).
-After fitting the model using the training data, performance of each training set was calculated as the sum of log-likelihood values for the test data.
-Alternative models were then compared using the total log-likelihood summed across the five folds for each model structure.
-Across taxonomic groups, the best performing model contained a random intercept defined by ecoregion along with a spatial random field.
+To estimate uncertainty of SVCs for canopy height, we added draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations.
+We then calculated the median and 0.025 and 0.975 quantiles to obtain a 95% confidence interval for each SVC estimate.
+Confidence intervals that did not overlap zero were considered significant [@andersonSdmTMBPackageFast2024].
 
-## Add spatially varying coefficient for canopy height
-We then compared the best performing model to a model that additionally included a spatially varying coefficient for canopy height because we expect that canopy height may have a weaker impact on verticality in regions where verticality is homogeneously low than in regions were verticality is high.
+## Relationship between verticality and richness
 
-Using the best performing random-effects structure, we compared three alternative models that differed in their fixed effects structure: the the full model with all fixed effects, the full model without the polynomial term for maximum temperature of the warmest month, and the full model without climate velocity.
-The base model included vegetative and current climatic predictors (canopy height, vegetation density, minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the wettest month.
-All these variables are assumed to directly impact the prevalence of different vertical strategies.
-The second alternative model included a polynomial term for maximum temperature of the warmest month because verticality increases in the tropics, where it is warmer, but extremely high maximum temperatures may limit arboreal strategies.
-The third alternative model a
+To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points, only points where canopy height significantly impacted verticality, and only points where canopy height had a positive significant effect on verticality.
 
-To validate the best performing model, we refit it using all the data and calculated quantile residuals using the DHARMa R package.
+## Future predictions
 
-To validate the model, we examined quantile residuals to ensure normality and lack of spatial pattern.
-
-\*we examined VIFs for predictor variables, sequentially removed variables until all VIFs were equal to or less than 4.
-We chose variables based on a combination of which ones had the highest VIF factors and which ones we thought had the most direct impact on physiological limits to arboreality.
-While, this is slightly higher than the standard VIF cutoff, we thought that it was important to keep both maximum temperature of the warmest month and minimum temperature of the coldest month, as the relative influence of these variables is likely to change across latitudes, with maximum temperatures having a larger impact on verticality in the tropics and minimum temperatures having a larger impact on verticality in temperate and boreal regions.
-We excluded precipitation seasonality because it was strongly correlated with dry season precipitation, has a less direct impact on physiological limits to verticality than dry season precipitation, and varies between current and future climate conditions than dry season precipitation.\*
-
-To examine potential impacts of climate change on assemblage verticality, we predicted models to future climatic data while holding vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density.
-From the future predictions, we subtracted predictions to the data used to train the model and mapped the resulting differences.
-
-We define a mesh, specifying a cutoff of 300 km as the minimum allowed distance between mesh verticies to maintain computation efficiencty while allowing suffcient complexity in the spatial field.
-
-Vegetative and climatic parameters included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the driest quarter, precipitation of the wettest quarter, precipitation seasonality, and climate change velocity.
-All predictors had variance inflation factors below 4.
-We first identified the optimal random effects structure for the model using five-fold cross validation and the sum of the log-likelihood values across folds.
-Five random effects structures were compared, with each containing a spatial random field calculated from a 300 km mesh.
-Four of the five structures contained one additional random intercept: biome, realm, biorealm, or ecoregion.
-Models fit with a spatial random field and ecoregion as a random intercept consistently performed the best.
-
-Using the optimal random effects structure, we then compared fixed effects using five-fold cross validation and the sum of the log-likelihood values for each fold.
-Our base model consisted of vegetation parameters (canopy height and vegetation density) and present-day climatic conditions (maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the driest quarter, precipitation of the wettest quarter, and precipitation seasonality).
-We compared this model to one that additionally included climate change velocity, which has previously been shown to correlate with assemblage verticality (cite).
-
-Using the optimal fixed effects structure, we refit the model with all points and validated the model by examining the normality and spatial distribution of quantile residuals.
-We then projected the models for mean and SES verticality for each taxonomic group to future climatic conditions and mapped the difference between predictions of future and present verticality.
-
-Full models were checked using sanity() function.
-
-We used simultaneous autoregressive (SAR) models with a spatial error term to model the impact of environmental factors on the SES of mean verticality.
-Environmental variables included mean elevation, vegetative complexity, climate velocity, mean annual temperature, maximum temperature of the warmest quarter, minimum temperature of the coldest quarter, diurnal temperature range, temperature seasonality, annual precipitation, precipitation of the wettest quarter, precipitation of the driest quarter, and precipitation seasonality.
-To reduce multicollinearity among predictors, we initially applied an automatic selection procedure that eliminated the variable with the highest variance inflation factor (VIF) until all VIFs were below three.
-This resulted in the loss of both maximum temperature of the warmest month and minimum temperature of the coldest month, while temperature seasonality and diurnal temperature range were retained.
-Since maximum and minimum temperatures exhibit greater change from present to future climate scenarios and are more likely to directly impact verticality, we substituted these two variables for temperature seasonality and diurnal temperature range.
-Although this adjustment increased VIF, all VIF remained below five.
-All selected variables for use in the models included mean elevation, vegetative complexity, climate velocity, maximum temperature of the warmest quarter, minimum temperature of the coldest quarter, precipitation of the wettest quarter, precipitation of the driest quarter, and precipitation seasonality.
-
-Prior to modeling, independent variables were checked for linear relationships with the response variable and log10 transformed if the relationship exhibited a strong right skew.
-Due to zero values in precipitation of the driest quarter, values were transformed using log10(precipitation + 1).
-If relationships appeared quadratic, a quadratic term was added to the model.
-Independent variables were then scaled to a mean of zero and standard deviation of one.
-To determine the optimal neighborhood weights matrix, we visually examined the distance to which spatial autocorrelation occurred in teh response based on correlograms.
-We then identified different neighborhood scenarios based on distance from the focal cell.
-The smallest distance included the eight neighbors immediates surrounding the focal cell and subsequently increased to include successive rings of cells up to the maximum distance at which spatial autocorrelation was present in the response variable.
-For each distance scenario, we produced three row-standardized neighborhood weights matrices based on distance, inverse distance, and inverse squared distance [@verhoefSpatialAutoregressiveModels2018].
-
-We produced simultaneous autoregressive (SAR) models with a spatial error term using the additive effects of all scaled environmental variables for each neighborhood weight matrix and calculated Moran’s I for each model to determine if spatial autocorrelation remained in the residuals.
-Models in which spatial autocorrelation remained were discarded.
-Based on the remaining models, we selected the neighborhood weights matrix as the one used in the model with the lowest AICc.
-Initial model exploration indicated that spatial autocorrelation was present using larger spatial neighborhoods.
-To reduce computational demands, we therefore based subsequent model selection on spatial neighborhoods from a radius of one grid cell to seven grid cells around the focal cell.
-Using the optimal neighborhood weights matrix defined by the lowest AICc, we compared all possible combinations of independent variables using the ‘dredge’ function from the MuMIN R package.
-Optimal weight matrices used inverse distance squared weighting.
-We then used a model averaging procedure to estimate coefficients and standard errors for all models within 2 \u0394AICc of the best model [@symondsBriefGuideModel2011ods].
-For each model within \u0394AICc of the best model, we calculated the AICc weight (wAICc), which represents the probability that the given model is the best (*cite Burnham and Anderson*).
-Because wAICc values were similar across candidate models, we used full-model averaging, where coefficients for parameters that do not appear in a model are set to zero.
-A weighted average of coefficients across all models is then calculated to estimate the impact of the parameter on the response variable.
-We additionally calculated relative variable importance by summing wAICc in the set of models where the variable occurs (*cite Burnham and Anderson*).
-To estimate model fit, we calculated pseudo-R^2^ as the squared Pearson correlation between predicted and observed values [@kisslingSpatialAutocorrelationSelection2008].
-
-To estimate changes in mean and SES verticality under future climate scenarios, we predicted the averaged model to future environmental conditions, while climate velocity, elevation, and vegetative complexity remained at the values used in the models of present verticality.
-Predictions were made using a linear model equation with the coefficients from the averaged SAR model (i.e.
-the predicted trends without the spatial signal) because it is not possible to estimate the the spatial signal without knowing the values of the response variable.
-To estimate changes in verticality, we therefore used the same method to predict verticality under present environmental conditions and subtracted these values from predicted future verticality.
-Negative values indicate declines in verticality and positive values indicate increases in verticality.
+To examine potential impacts of climate change on assemblage verticality, we predicted mean and SES verticality under a future climate scenario using the models.
+However, we held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density.
+From the future predictions, we subtracted predictions to the data used to train the model (present predictions) and mapped the resulting differences.
+To standardize measures of change across taxa and between mean and SES verticality, we additionally calculated the percent relative change between future and present predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
+Thus, negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
+*map relative change as well*
 
@@
 ## Global patterns of verticality
+
+Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (\@ref(fig:vert-map-global)).
+This pattern was largely driven by declines in the proportion of arboreal species toward the poles.
+Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical moist broadleaf forests (\@ref(fig:vert-biome)).
+
+Despite widespread coherence of global verticality patterns among taxa and within biomes, regional variation occurred.
+Within tropical and subtropical biomes, verticality was highest in moist broadleaf forests and lowest in dry broadleaf forests (\@ref(fig:vert-biome)).
+In moist broadleaf forests, verticality of bird and mammal assemblages was consistently high, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies, while the proportion of terrestrial species declined.
+In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Malaysian and Indonesian forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar.
+Additionally, the proportion of reptiles and amphibians that exhibit terrestrial strategies did not decline to the same extent as birds and mammals.
+*include proportion maps in supplement*
+
+In temperate biomes, verticality was similar across forest types but declined in less dense wooded habitats (e.g., savannas and shrublands) (\@ref(fig:vert-biome)).
+Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes.
+Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (\@ref(fig:vert-map-global)L, though temperate rainforests (e.g., the Pacific-Northwest of North America) were a notable exception.
+Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (\@ref(fig:vert-map-global)F).
+In contrast, the proportion of fossorial reptiles declined in temperate regions.
+Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species.
+Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal gradient in SES verticality.
+
+## Environmental drivers of verticality
+
+### Climate
+
+The spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (\@ref(fig:mod-coefs), *supp table for model results*).
+Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality, though the impact of dry season precipitation on SES verticality was stronger (\@ref(fig:mod-coefs)).
+Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality.
+Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile mean and SES verticality, as well as SES verticality of mammals.
+This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further (*conditional effects plot*).
+Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship.
+Climate velocity negatively impacted bird and amphibian verticality and positively impacted mammal and reptile verticality.
+However, environmental factors were unable to explain extremely high and extremely low verticality as indicated by larger residuals at the ends of the verticality spectrum when spatial random effects were not included in estimates of assemblage verticality (*supp figure*).
+
+### Vegetation
+
+We hypothesized that canopy height and vegetation density would positively impact verticality, but that the impact of canopy height would vary across space.
+Specifically, we expected canopy height to have a strong positive effect on verticality in warmer and wetter regions that support arboreal assemblages and a negligible effect in regions with climate conditions that support fewer arboreal species.
+Results of the GLMMs did not entirely support our predictions.
+Globally, canopy height positively impacted SES verticality of reptiles and amphibians and mean verticality of amphibians, while vegetation density had no significant impact on mean or SES verticality of any taxa (\@ref(fig:mod-coefs)).
+In contrast to expectations, canopy height did not have a positive impact on verticality across much of the tropics where climate conditions support arboreality (\ref(fig:svc)).
+Significant positive impacts of canopy height generally occurred in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (\@ref(fig:svc-sig)).
+
+## Verticality-species richness relationships
+
+When considering all points, Pearson correlations indicated strong positive relationships between SES or mean verticality and species richness for birds and mammals, a weak positive relationship for amphibians, and nearly no relationship for reptiles (*add R2 and p-values*) (\@ref(fig:corplot)).
+When we restricted the correlation to only include points where canopy height had a significant positive effect on verticality, the strength of the relationship between SES verticality and richness increased for amphibians and SES verticality of birds, but declined for mammals, reptiles, and mean verticality of birds.
+
+## Climate change impacts
+
+*Calculate relative change as (future-present)/(presentmax-presentmin)*
+
+```{r}
+# calculate percent change in verticality
+source("./../scripts/04_plots/vertdif_by_biorealm.R")
+
+sessumm = dat %>% 
+  filter(vertvar == "SES Mean Verticality") %>% 
+  group_by(class, biome) %>% 
+  summarize(mean.reldif = mean(est.reldif)) %>% 
+  arrange(mean.reldif) %>% 
+  as.data.frame()
+
+meansumm = dat %>% 
+  filter(vertvar == "Mean Verticality") %>% 
+  group_by(class, biome) %>% 
+  summarize(mean.reldif = mean(est.reldif)) %>% 
+  arrange(mean.reldif) %>% 
+  as.data.frame()
+
+```
+
+Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality (\@ref(fig:vert-dif-map)).
+On average verticality was predicted to decline the most in tropical and subtropical moist broadleaf forests and increase the most at high latitudes in boreal forests.
+Within these biomes, the models predicted the greatest relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (\@ref(fig:vert-dif-biome)).
 
@@
 ```{r}
-load("./../results/sdmTMB_models/amphibians_meanvert.RData")
-m.amph.meanvert = m.final
-load("./../results/sdmTMB_models/amphibians_sesvert.RData")
-m.amph.sesvert = m.final
-
-load("./../results/sdmTMB_models/reptiles_meanvert.RData")
-m.rept.meanvert = m.final
-load("./../results/sdmTMB_models/reptiles_sesvert.RData")
-m.rept.sesvert = m.final
-
-load("./../results/sdmTMB_models/mammals_meanvert.RData")
-m.mammals.meanvert = m.final
-load("./../results/sdmTMB_models/mammals_sesvert.RData")
-m.mammals.sesvert = m.final
-
-load("./../results/sdmTMB_models/birds_meanvert.RData")
-m.birds.meanvert = m.final
-load("./../results/sdmTMB_models/birds_sesvert.RData")
-m.birds.sesvert = m.final
+# load("./../results/sdmTMB_models/amphibians_meanvert.RData")
+# m.amph.meanvert = m.final
+# load("./../results/sdmTMB_models/amphibians_sesvert.RData")
+# m.amph.sesvert = m.final
+# 
+# load("./../results/sdmTMB_models/reptiles_meanvert.RData")
+# m.rept.meanvert = m.final
+# load("./../results/sdmTMB_models/reptiles_sesvert.RData")
+# m.rept.sesvert = m.final
+# 
+# load("./../results/sdmTMB_models/mammals_meanvert.RData")
+# m.mammals.meanvert = m.final
+# load("./../results/sdmTMB_models/mammals_sesvert.RData")
+# m.mammals.sesvert = m.final
+# 
+# load("./../results/sdmTMB_models/birds_meanvert.RData")
+# m.birds.meanvert = m.final
+# load("./../results/sdmTMB_models/birds_sesvert.RData")
+# m.birds.sesvert = m.final
 
@@
 # estimated coefficients and confidence intervals from the best model for each taxa - include all model parameters: fixed effects, spatial paramters (matern range, marginal spatial standard, deviation), distribution parameters
-
-Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (\@ref(fig:vert-map-global)).
-This pattern was largely driven by declines in the proportion of arboreal species toward the poles.
-Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical broadleaf forests (\@ref(fig:vert-biome)).
-
-Despite widespread coherence of global verticality patterns among taxa and within biomes, regional variation occurred.
-Within the tropics, verticality was highest in moist broadleaf forests and lowest in dry broadleaf forests (\@ref(fig:vert-biome)).
-Mammals exhibited the highest mean verticality in moist forests, particularly in the Amazon, the Congo Basin, Madagascar, and tropical and subtropical Asian forests where an average of `r mammals.arb.tropicalmoist`% of species per assemblage exhibited arboreal strategies while the proportion of terrestrial species declined.
-Verticality of birds closely mirrored this pattern except in Madagascar where SES of mean verticality declined.
-While reptiles and amphibians displayed similar verticality patterns, there were notable differences.
-Verticality of reptile assemblages peaked in the rainforests of Indonesia and Madagascar and declined in the neotropics where the proportion of fossorial species increased *map proportion arboreality and fossoriality*.
-Amphibian verticality additionally increased in the Brazilian Atlantic Forest while declining in SE Asia relative to other vertebrate assemblages.
-
-In temperate biomes, verticality was similar across forests and declined in less dense wooded habitats (e.g., savannas) (\@ref(fig:vert-biome)).
-Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes.
-Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes, peaking in savannas and savanna-woodland transition zones (\@ref(fig:vert-map-global)L, \@ref(fig:vert-biome)).
-However, temperate rainforests (e.g., the Pacific-Northwest of North America) were notable exceptions, exhibiting higher verticality than than expected by chance.
-Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (\@ref(fig:vert-map-global)F).
-In contrast, the proportion of fossorial reptiles declined in temperate regions.
-Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species.
-Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal gradient in the SES of verticality.
-
-## Environmental drivers of verticality
-
-We expected verticality to increase with increasing canopy height, vegetation density, precipitation, and minimum cold temperatures.
-While the models were generally consistent with these hypotheses, there were some notable differences.
-
-### need a section on model fit ###
-Models predicted verticality well.
-However, much of the predictive ability was due to the spatial random field.
-When the spatial random field was excluded from model predictions, models generally underestimated verticality in regions with high verticality and overestimated verticality in regions with low verticality.
-
-### Climate
-
-The models generally indicated that verticality was limited by low precipitation, high maximum temperatures, and low minimum temperatures, though some variation occurred between taxa.
-Additionally, climatic impacts relative to the impact of canopy height was stronger for SES verticality than mean verticality.
-
-Across taxa, precipitation of the driest and wettest months positively influences mean and SES verticality, though the impact of dry season precipitation on SES verticality was noticeable stronger (\@ref(fig:mod-coefs)).
-Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality.
-Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile verticality, as well as SES verticality of mammals.
-This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further.
-Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship.
-Climate velocity was not included in the best models for mammals verticality or for SES verticality of amphibians.
-For the remaining models, climate velocity (i.e., climate instability) negatively impacted bird verticality and mean verticality of amphibians and positively impacted reptile verticality.
-
-### Vegetation
-
-We hypothesized that canopy height and vegetation density would positively impact verticality, but that the impact of canopy height would vary across space.
-Specifically, we expected canopy height to have a strong positive effect on verticality in warmer and wetter regions that support arboreal assemblages and a negligible effect in regions with climate conditions that support fewer arboreal species.
-
-Globally, canopy height positively impacted verticality for mammals and birds and had little impact on verticality of reptiles and amphibians, while vegetation density had a negligible impact on verticality of all taxa (\@ref(fig:coef-plt).
-The best models for all taxa included a spatially varying coefficient for canopy height, which caused the wide confidence interval in the global estimate its coefficient.
-
-Spatial variation in the effect of canopy height was greater for models of SES verticality than for mean verticality, though both exhibited similar patterns (\@ref(fig:svc-map)).
-For birds, canopy height exhibited the strongest positive effect on SES verticality in the Andes, central America, Indomalayan moist deciduous forests, the forest-savanna woodlands and thickets of eastern Africa, and northern taiga forests (\@ref(fig:svc-map)A,E).
-Spatial variability of the coefficient of canopy height for models of SES verticality exhibited similar pattern in mammals.
-However, strong positive effects additionally occurred in the south-eastern US, subpolar forests of South America, Guinean forest-savanna, and Australian temperate savanna and woodlands (\@ref(fig:svc-map)B,F).
-For reptiles, similar regions showed weaker positive relationships between canopy height and SES verticality.
-However, Madagascar exhibited a strong positive coefficient (\@ref(fig:svc-map)C,G).
-The coefficient of canopy height was also strongly positive for Amphibians in Madagascar as well as Sri Lanka relative to other regions (\@ref(fig:svc-map)D,H).
-Notably, the coefficient of canopy height was near zero in tropical broadleaf forests across the globe and was not negligible in cooler regions, contrary to our predictions.
-
-## Climate change impacts
-
-```{r}
-# calculate percent change in verticality
-calc_change = function(pred.f, taxa, response) {
-  pred.f = pred.f %>% 
-    dplyr::select(x,y,est.dif) %>% 
-    mutate(taxa = taxa,
-         response = response)
-return(pred.f)
-}
-
-load("./../results/sdmTMB_models/amphibians_meanvert.RData")
-amph.meanvert = calc_change(pred.f, taxa = "Amphibians", response = "Mean Verticality")
-
-load("./../results/sdmTMB_models/amphibians_sesvert.RData")
-amph.sesvert = calc_change(pred.f, taxa = "Amphibians", response = "SES Verticality")
-
-load("./../results/sdmTMB_models/reptiles_meanvert.RData")
-rept.meanvert = calc_change(pred.f, taxa = "Reptiles", response = "Mean Verticality")
-
-load("./../results/sdmTMB_models/reptiles_sesvert.RData")
-rept.sesvert = calc_change(pred.f, taxa = "Reptiles", response = "SES Verticality")
-
-load("./../results/sdmTMB_models/mammals_meanvert.RData")
-mammals.meanvert = calc_change(pred.f, taxa = "Mammals", response = "Mean Verticality")
-
-load("./../results/sdmTMB_models/mammals_sesvert.RData")
-mammals.sesvert = calc_change(pred.f, taxa = "Mammals", response = "SES Verticality")
-
-load("./../results/sdmTMB_models/birds_meanvert.RData")
-birds.meanvert = calc_change(pred.f, taxa = "Birds", response = "Mean Verticality")
-
-load("./../results/sdmTMB_models/birds_sesvert.RData")
-birds.sesvert = calc_change(pred.f, taxa = "Birds", response = "SES Verticality")
-
-dat = rbind(amph.meanvert, amph.sesvert, rept.meanvert, rept.sesvert, mammals.meanvert, mammals.sesvert, birds.meanvert, birds.sesvert)
-
-dat.change = dat %>% 
-  group_by(taxa, response) %>% 
-  summarise(max = max(est.dif), min = min(est.dif))
-
-dat.change.ses = dat.change %>% filter(response == "SES Verticality")
-dat.change.mean = dat.change %>% filter(response == "Mean Verticality")
-
-ses.poschange = dat.change.ses[which(dat.change.ses$max == max(dat.change.ses$max)),]
-ses.negchange = dat.change.ses[which(dat.change.ses$min == min(dat.change.ses$min)),]
-
-mean.poschange = dat.change.mean[which(dat.change.mean$max == max(dat.change.mean$max)),]
-mean.negchange = dat.change.mean[which(dat.change.mean$min == min(dat.change.mean$min)),]
-
-```
 
@@
 
-```{r fig.id="vert-map-global"}
-
+```{r fig.id="vert-map-global", fig.width=160, fig.height=96}
+include_graphics("./../figures/main_figs/vert_maps.png", error = F)
 ```
 
-```{r fig.id="model-coefs"}
-
+```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3}
+include_graphics("./../figures/main_figs/sdmTMB_coefs.png", error = F)
 ```
 
-```{r fig.id="vert-difs-global"}
-
+```{r fig.id="svc", fig.width=160, fig.height=143.95}
+include_graphics("./../figures/main_figs/canopy_height_coefs.png", error = F)
 ```
 
-```{r fig.id="vert-homogenization"}
+```{r fig.id="corplot"}
+include_graphics("./../figures/main_figs/vert_rich_cor.png", error = F)
+```
 
+```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3}
+include_graphics("./../figures/main_figs/vert_difs.png", error = F)
 ```
@@
 
-```{r tab.id = "mean-vert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of mean verticality of vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}
+#```{r tab.id = "mean-vert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of mean verticality of vertebrate assemblages as a #function of climatic, topographic, and vegetative factors."}
 
-```
+#```
 
-```{r tab.id = "sesvert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of the standard effect size of mean verticality of vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}
+#```{r tab.id = "sesvert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of the standard effect size of mean verticality of #vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}
 
-```
+#```
 
@@
 
-```{r fig.id="vertses-hist-biome", fig.lp="ed", fig.cap="Histograms of verticality by biome for each vertebrate class. Red dotted line indicates the mean verticality for each biome/class."}
-# plot SES vert by biome
-("figures/supp_figs/vert_biome_forestOnly_forestSES.R")
-plt_vertses_biome
-
+```{r fig.id="vert-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
+include_graphics("./../figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
 
-```{r fig.id="vert-forest", fig.lp="ed", fig.cap="Mean verticality, standard effect size (SES) of mean verticality, and the average proportion of species exhibiting arboreal, terrestrial, and fossorial strategies across latitudinal bands.
-In maps of mean veticality, assemblages composed of only fossorial species are zero (red) and those composed of only arboreal species are quantified as one (purple).
-In maps of the SES of mean verticality, negative numbers (red) indicate assemblages have lower verticality than expected by random chance and positive numbers (purple) indicate assemblages have higher verticality than expected by random chance.
-Line graphs show the average proportion of species exhibiting arboreal (blue), terrestrial (yellow), and fossorial (red) strategies.
-Proportions for each assemblage can exceed one because species can exhibit multiple vertical strategies.}
-
+```{r fig.id="vert-dif-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
+include_graphics("./../figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
 
-```{r fig.id="forest-model-coefs"}
+#```{r fig.id="vertses-hist-biome", fig.lp="ed", fig.cap="Histograms of verticality by biome for each vertebrate class. Red dotted line indicates the mean #verticality for each biome/class."}
+## plot SES vert by biome
+#("figures/supp_figs/vert_biome_forestOnly_forestSES.R")
+#plt_vertses_biome
 
-```
+#```
+
+#```{r fig.id="vert-forest", fig.lp="ed", fig.cap="Mean verticality, standard effect size (SES) of mean verticality, and the average proportion of species #exhibiting arboreal, terrestrial, and fossorial strategies across latitudinal bands.
+In maps of mean veticality, assemblages composed of only fossorial species #are zero (red) and those composed of only arboreal species are quantified as one (purple).
+In maps of the SES of mean verticality, negative numbers (red) #indicate assemblages have lower verticality than expected by random chance and positive numbers (purple) indicate assemblages have higher verticality than #expected by random chance.
+Line graphs show the average proportion of species exhibiting arboreal (blue), terrestrial (yellow), and fossorial (red) strategies.
+#Proportions for each assemblage can exceed one because species can exhibit multiple vertical strategies.}
+
+#```
+
+#```{r fig.id="forest-model-coefs"}
+
+#```
 

### 2024-08-04 scripts/04_plots/richness_vert_correlations.R
update plot

@@ -24,15 +24,15 @@ dat %>%
   ggplot(aes(x = rich, y = vert.mean.ses, color = biome)) +
     geom_point(pch = ".") +
     geom_smooth(method = "lm") +
-    facet_grid(rows = vars(biome), cols = vars(taxa), scales = "free_x") +
+    facet_grid(cols = vars(biome),rows = vars(taxa), scales = "free") +
     theme_classic()
 
 dat %>% 
   filter(rich >= 5) %>% 
-  ggplot(aes(x = rich, y = vert.mean, color = biome)) +
+  ggplot(aes(x = vert.mean.ses, y = rich, color = biome)) +
   geom_point(pch = ".") +
   geom_smooth(method = "lm") +
-  facet_wrap(~taxa, scales = "free_x") +
+  facet_wrap(~taxa, scales = "free") +
   theme_classic()
 
 

### 2024-08-04 scripts/04_plots/vertdif_by_biorealm.R
change to percent relative difference

@@ -2,86 +2,108 @@ library(terra)
 library(tidyterra)
 library(tidyverse)
 
+# Supplementary figure: Plot the percent relative change ((vert.future - vert.present)/(vertmax - vertmin)*100).
+# biomes are ordered along the y-axis from the greatest average positive change at the top to the greatest
+# average negative change at the bottom. Averages for ordering the y-axis are calculate for mean or ses verticality across all taxa.
+
 
 # Load SES vert and Mean vert models ----------------------------------------------------
 load("results/sdmTMB_models/amphibians_sesvert.RData")
 amph.sesvert = pred.f
 amph.sesvert$vertvar = "SES Mean Verticality"
+amph.sesvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean.ses) - min(pred$vert.mean.ses))*100
+amph.sesvert$biome = pred$biome
 
 load("results/sdmTMB_models/amphibians_meanvert.RData")
 amph.meanvert = pred.f
 amph.meanvert$vertvar = "Mean Verticality"
+amph.meanvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean) - min(pred$vert.mean))*100
+amph.meanvert$biome = pred$biome
 
 amph.pdat = amph.sesvert %>% 
   bind_rows(amph.meanvert) %>%
-  dplyr::select(x,y,est.dif, vertvar) %>% 
+  dplyr::select(x,y,est.dif, est.reldif, vertvar, biome) %>% 
   mutate(class = "Amphibians")
 
 
 load("results/sdmTMB_models/mammals_sesvert.RData")
 mammals.sesvert = pred.f
 mammals.sesvert$vertvar = "SES Mean Verticality"
+mammals.sesvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean.ses) - min(pred$vert.mean.ses))*100
+mammals.sesvert$biome = pred$biome
+
 
 load("results/sdmTMB_models/mammals_meanvert.RData")
 mammals.meanvert = pred.f
 mammals.meanvert$vertvar = "Mean Verticality"
+mammals.meanvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean) - min(pred$vert.mean))*100
+mammals.meanvert$biome = pred$biome
 
 mammals.pdat = mammals.sesvert %>% 
[…]

### 2024-08-07 docs/verticality_future_mkd.Rmd
intro revisions

@@
 
-Vertical stacking of species is an emblematic signature of biologically rich regions and influences biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
-At local scales, species assemblages may be stratified across vertical climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-These climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-Stratification across this vertical climate gradient may reduce competition and increase species richness (*cite*).
-Tall, multi-layered forests are thus expected to support both highly diverse and highly arboreal species assemblages by increasing the volume of habitat and enabling species to finely partition the vertical forest structure [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-These conditions occur in tropical forests where high proportions of arboreal species may contribute to their biologically rich assemblages [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019].
-However, vegetation structure often fails to predict patterns of species richness at continental to global scales [@rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].
+Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
+At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
+Tall forests generally have greater foliage height diversity.
+These vegetation layers may act as a multi-floored apartment building that enables species to finely partition habitat [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness (*cite positive relationships between canopy height and species richness at local scales*).
+*concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019].
+However, canopy height and other metrics of vegetation structure do not consisently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
+Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.
 
-At global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests.
-For example, low minimum temperatures in the canopies of high latitude forests may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
+At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests.
+Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
+In high latitude forests, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
 Similarly, exposure to hot and dry conditions in the canopies of tropical lowland forests may reduce the suitability of arboreal habitats for heat or desiccation sensitive species, such as frogs [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020].
-Release of these climatic constraints at higher elevations, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013].
-We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological constraints.
+Release of these climatic constraints at higher elevations in the tropics, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013].
+We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.
 
-Similarly, spatial variation may occur in the relationship between species richness and assemblage verticality, defined here as the average vertical position of species in an assemblage [@oliveiraVerticalStratificationInfluences2019].
-While verticality and richness are often assumed to exhibit a positive relationship, the strength of this relationship is variable [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*.
-In regions where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weka because species are unable to vertically partition habitat.
-In contrast, when an increase in canopy height allows the addition of arboreal species to the assemblage, increasing assemblage verticality, we expect a strong positive relationship between verticality and species richness.
+Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019].
+Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*.
+Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage.
+In contrast, where taller canopies do not facilitate higher verticality, the relationship bewteen verticality and richness may be weak (*concept fig*).
 
@@
 
-#Figures to include
-# verticality map
-# verticality by ecoregion - could be ciruclar bar plot for each realm or biome and taxonomic group
-# coefficient plot
-```{r}
-# load("./../results/sdmTMB_models/amphibians_meanvert.RData")
-# m.amph.meanvert = m.final
-# load("./../results/sdmTMB_models/amphibians_sesvert.RData")
-# m.amph.sesvert = m.final
-# 
-# load("./../results/sdmTMB_models/reptiles_meanvert.RData")
-# m.rept.meanvert = m.final
-# load("./../results/sdmTMB_models/reptiles_sesvert.RData")
-# m.rept.sesvert = m.final
-# 
-# load("./../results/sdmTMB_models/mammals_meanvert.RData")
-# m.mammals.meanvert = m.final
-# load("./../results/sdmTMB_models/mammals_sesvert.RData")
-# m.mammals.sesvert = m.final
-# 
-# load("./../results/sdmTMB_models/birds_meanvert.RData")
-# m.birds.meanvert = m.final
-# load("./../results/sdmTMB_models/birds_sesvert.RData")
-# m.birds.sesvert = m.final
+# Discussion
 
-```
+## Environmental drivers of verticality
 
-# change in verticality map
+### Climate
 
-# Supplementary figures/tables
-# estimated coefficients and confidence intervals from the best model for each taxa - include all model parameters: fixed effects, spatial paramters (matern range, marginal spatial standard, deviation), distribution parameters
+### Vegetation
 
-Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality.
-The largest changes in SES verticality were observed for birds, which exhibited a maximum decline of `r round(as.numeric(ses.negchange[1, "min"]),2)` and a maximum increase of `r round(as.numeric(ses.poschange[1, "max"]),2)`.
-The larges changes in mean verticality were observed for amphibians, which exhibited a maximum decline of `r round(as.numeric(mean.negchange[1, "min"]), 2)` and a maximum increase of `r round(as.numeric(mean.poschange[1, "max"]), 2)`.
-For birds and mammals, mean and SES verticality generally declined in tropical moist broadleaf forests globally in addition to the temperate forests, savannas, and woodlands of southern Australia and the Brazilian Atlantic forests.
-In contrast, verticality primarily increased in the boreal forests of Canada and northern Eurasia, tropical and subtropical savannas and woodlands in Africa, and tropical savannas in northern Australia (\@ref(fig:vert-dif-biome)).
-Changes in verticality of Reptiles and Amphibians were characterized by declines in most regions.
-However, amphibian verticality was expected to experience minor increases in African tropical and subtropical savannas and woodlands (\@ref(fig:vert-dif-biome)).
+## Verticality species-richness relationships
 
-## discussion of climate change
-
--   in snakes, it is more common for species to exhibit shift from nonfossorial to fossorial than the other way around (Cyriac and Kodondaramiah 2018) - suggests increase in verticality for reptiles could be caused by extinction of fossorial species rather than vertical niche shift higher
-
-    -   further evidence by Cyriac and Kodandoramaiah 2017 - high extinction rates of fossorial snakes during prolonged environmental fluctuations
-
--   but these patterns aren't consistent - e.g., in lizards, transitions from fossoriality to nonfossoriality is not less common that transisitons in the other direction
-
-# Discussion
+## Impact of climate change on verticality
 
@@
 
-# Extended Data Tables
+\newpage
 
-#```{r tab.id = "mean-vert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of mean verticality of vertebrate assemblages as a #function of climatic, topographic, and vegetative factors."}
+# Supplementary tables
 
-#```
+# estimated coefficients and confidence intervals from the best model for each taxa - include all model parameters: fixed effects, spatial paramters (matern range, marginal spatial standard, deviation), distribution parameters
 
-#```{r tab.id = "sesvert-mods", tab.lp="ed", tab.cap="Simultaneous autoregressive model results for models of the standard effect size of mean verticality of #vertebrate assemblages as a function of climatic, topographic, and vegetative factors."}
+\newpage
 
-#```
-
-# Extended Data Figures
+# Supplementary figures
 
@@
 
-```{r fig.id="vert-dif-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
+```{r fig.id="vert-dif-biome", fig.lp="ed", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
+Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
+Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}
+
 include_graphics("./../figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)

### 2024-08-13 docs/verticality_future_mkd.Rmd
spatial varying relationships discussion

@@
 At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall forests generally have greater foliage height diversity.
-These vegetation layers may act as a multi-floored apartment building that enables species to finely partition habitat [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness (*cite positive relationships between canopy height and species richness at local scales*).
+Tall forests with high foliage hight diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@wagnerRemotelySensedTree2023] (*cite positive relationships between canopy height and species richness at local scales*).
 *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019].
@@
 Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density).
-Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality (Oliveira and Scheffers 2019).
+Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
 All environmental data was resampled to the resolution and projection of the assemblage data.
@@
 The spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (\@ref(fig:mod-coefs), *supp table for model results*).
-Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality, though the impact of dry season precipitation on SES verticality was stronger (\@ref(fig:mod-coefs)).
+Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (\@ref(fig:mod-coefs)).
 Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality.
 Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile mean and SES verticality, as well as SES verticality of mammals.
-This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further (*conditional effects plot*).
 Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship.
+These patterns resulted in declines in verticality at both extremely high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (*see conditional effect plots*).
 Climate velocity negatively impacted bird and amphibian verticality and positively impacted mammal and reptile verticality.
@@
 # Discussion
+
+Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studeis that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019].
+Though climate conditions influenced verticality, they were neither able to predict extremely high or low verticality nor spatial variation in the influence of canopy height on verticality.
+Furthermore, `r word_comment(highlight = "Furthermore", comment = "Try bootstrapping method to compare slopes")` regions exhibitin positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between verticality and species richness.
+Our results thus do not support our hypothesis that climtatic conditions that support high assemblage verticality lead to positive impacts of canopy hiehg on verticality and positive associations between verticality and species richness.
+Vertical stratification across increasingly tall and multi-layered forests may therefore not be a primary mehcanism driving positive relationships between canopy height, verticality, and species richness at global scales.
+Rather, we propose that species packing and evolutionary patterns may explain our unexpected results.
+The ways in which climate change will impact verticality and its relationship with species richness is thus difficult to predict.
+
+## Environmental drivers of verticality
+
+Climate influences global patterns of verticality because more extreme climates in teh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
+Water availabitliy therefore facilitates verticality, while temperatures at both hot and cold extremes and awter scaracity generally limit verticality.
+For exmaple, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life straties outside moist forests in tropical regions.
+However, our models indicated nuances in the relative influences of different drivers across taxa.
+
+Impacts of precipitation were particularly pronounced for amphibian verticality.
+In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consitently available (*cite*).
+However, in regions that experience pronounced dry seasons, amphibians behaviourally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
+Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desccation witout adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
+*Birds were also influenced by precipitation, likely because they are dependent on the availability of standing water in teh canopy and water content of leaves.* `r word_comment(highlith = "leaves", comment = "need to look more into impacts of drought on birds - see https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.10781")`
+
+For mammals and reptiles, our modesl indicated that minimum cold temperatures had the greatest impact on verticality.
+In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014].
+Species are therefore largely constratined to terrestrial or fossorial habiats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019].
+Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors contribute to unexplained variation.
+
+## Spatially varying relationships between canopy height, verticality, and richness
+
+Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015].
+These results align with Basham et al.
+[-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics.
+Interestingly, previously reported spatial variation in the impact of canopy height on taxonomic and functional richness mirror, to some extent, spatial variation in the impacts of canopy height on verticality [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018].
+For example, canopy height positively impacts verticality and species richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea [@rollLinkingVertebrateSpecies2015].
+Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015].
+Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality.
+Yet, our results did not consistently support this hypothesis.
+
+A niche expansion model of community assembly characterizes the hypothesis that taller canopies with greater foliage height diversity allow greater vertical stratification and subsequent increases in verticality and species richness [@macarthurBirdSpeciesDiversity1961; @macarthurPatternsSpeciesDiversity1965].
+The niche expansion model posits that species richness increases when additional species occupy novel habitat or resource space [@pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016].
+Thus, arboreal species may be added to the assemblage by occupying the novel habitat space that becomes available as canopy height increases, producing a stronger positive relationship between verticality and richness in regions where canopy height has a positive influence on verticality.
+This pattern was evident for birds and amphibians.
+However, this three-way positive relationship did not occur in the most diverse regions, including much of the tropics.
+Instead, it was largely restricted to areas where relatively rapid increases in vegetation structure from high to low elevations or dry to wet forests (e.g., Madagascar) may facilitate the addition of species outside the habitat space occupied by species in assemblages of less productive habitats [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018].
+Yet in regions where canopy height did not have significant positive impacts on verticality, positive correlations between verticality and richness remained for birds and amphibians and increased in strength for mammals and reptiles, suggesting vertical stratification (i.e., niche expansion) may not be the dominant mechanism causing positive relationships between canopy height, verticality, and species richness at large spatial scales.
+
+Rather than expansion across tall forests, positive associations between verticality and species richness could be driven by niche packing.
+Niche packing occurs when smaller niche spaces or greater niche overlap allow more species to co-exist within a habitat [@pigotFunctionalTraitsReveal2016; @pellissierNichePackingExpansion2018].
+Specialization on specific resources within the canopy over evolutionary time could thus increase arboreal species diversity without increasing canopy height [@fengForestCanopyHeight2020; @pellissierNichePackingExpansion2018].
+Furthermore, evolutionary rates of arboreal lineages are generally faster than those of terrestrial and fossorial lineages [@cyriacDiggingTheirOwn2018; @liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017].
+In the tropics where climate conditions support arboreality, the combination of niche packing and rapid evolutionary rates of arboreal species could lead to a disproportionately high number of arboreal species, increasing verticality relative to what would be expected by climate conditions alone and causing positive relationships between verticality and richness in regions where canopy height has insignificant or even negative effects on verticality.
 
@@
 
+```{r fig.id="svc-sig", fig.lp="ed", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}
+
+include_graphics("./../figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
+```
+
 #```{r fig.id="vertses-hist-biome", fig.lp="ed", fig.cap="Histograms of verticality by biome for each vertebrate class. Red dotted line indicates the mean #verticality for each biome/class."}

### 2024-08-16 docs/verticality_future_mkd.Rmd
full first draft

@@
 title: "officedown template"
+bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
 output: 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE)
+knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE)
 library(officedown)
@@
 library(data.table)
+library(flextable)
+library(knitr)
 
@@
 At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall forests with high foliage hight diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
 By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@wagnerRemotelySensedTree2023] (*cite positive relationships between canopy height and species richness at local scales*).
 *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019].
-However, canopy height and other metrics of vegetation structure do not consisently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
+However, canopy height and other metrics of vegetation structure do not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
 Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.
@@
 Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage.
-In contrast, where taller canopies do not facilitate higher verticality, the relationship bewteen verticality and richness may be weak (*concept fig*).
+In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak (*concept fig*).
 
@@
 ```{r}
-birds = read.csv("./../data/derivative_data/species_data/birds_spdat_elton_forestsOnly.csv")
-mammals = read.csv("./../data/derivative_data/species_data/mammals_spdat_moura_forestsOnly.csv")
-amph = read.csv("./../data/derivative_data/species_data/amph_spdat_moura_forestsOnly.csv")
-repts = read.csv("./../data/derivative_data/species_data/rept_spdat_moura_forestsOnly.csv")
+birds = read.csv("data/derivative_data/species_data/birds_spdat_elton_forestsOnly.csv")
+mammals = read.csv("data/derivative_data/species_data/mammals_spdat_moura_forestsOnly.csv")
+amph = read.csv("data/derivative_data/species_data/amph_spdat_moura_forestsOnly.csv")
+repts = read.csv("data/derivative_data/species_data/rept_spdat_moura_forestsOnly.csv")
 
@@
 
-*infographic for methods*
-
 ```{r}
-mammals = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
+mammals = fread("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
 
-birds = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
+birds = fread("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
 
-amphibians = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
+amphibians = fread("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
 
-rept = fread("./../data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv") %>% 
+rept = fread("data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv") %>% 
   dplyr::select(vert.mean, vert.mean.ses, p.arb, p.ter, p.fos, biome, ecoregion, realm)
@@
 To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than 5 [@wartonEcostatsDataAnalysis2022].
-We chose variables to remove based on a combination of which had the highest VIF, which were most likely to directly impact physiological limits to arboreality, and which were expected to change the most in response to climate change.
-For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation, has a less direct impact on physiological limits to verticality than dry season precipitation, and varies less between current and future climate conditions than dry season precipitation.
+We chose variables to remove based on a combination of which had the highest VIF and which were most likely to directly impact physiological limits to arboreality.
+For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation and has a less direct impact on physiological limits to verticality than dry season precipitation.
 This procedure resulted in a final set of seven variables that included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the wettest month, precipitation of the driest month, and climate velocity.
@@
 Thus, negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
-*map relative change as well*
 
@@
 # calculate percent change in verticality
-source("./../scripts/04_plots/vertdif_by_biorealm.R")
+#source("./../scripts/04_plots/vertdif_by_biorealm.R")
+source("scripts/04_plots/vertdif_by_biorealm.R")
 
@@
 
+load("results/sdmTMB_models/amphibians_sesvert.RData")
+amph.sesvert = pred.f
+amph.sesvert$vertvar = "SES Mean Verticality"
+amph.sesvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean.ses) - min(pred$vert.mean.ses))*100
+
+amphses.minreldif = round(min(amph.sesvert$est.reldif),2)
+amphses.maxreldif = round(max(amph.sesvert$est.reldif),2)
+
+load("results/sdmTMB_models/amphibians_meanvert.RData")
+amph.meanvert = pred.f
+amph.meanvert$vertvar = "Mean Verticality"
+amph.meanvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean) - min(pred$vert.mean))*100
+
+amphmean.minreldif = round(min(amph.meanvert$est.reldif),2)
+amphmean.maxreldif = round(max(amph.meanvert$est.reldif),2)
+
 ```
 
-Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality (\@ref(fig:vert-dif-map)).
-On average verticality was predicted to decline the most in tropical and subtropical moist broadleaf forests and increase the most at high latitudes in boreal forests.
-Within these biomes, the models predicted the greatest relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (\@ref(fig:vert-dif-biome)).
+Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality and varied across taxa (\@ref(fig:vert-dif-map)).
+Amphibians exhibited the highest predicted relative percent change, with predicted declines reaching up to `r amphses.minreldif*-1`% for SES verticality and `r amphmean.minreldif*-1`% for mean verticality in parts of the tropics and predicted increases reaching up to `r amphses.maxreldif`% for SES verticality and `r amphmean.maxreldif`% for mean verticality in the African forest-savanna matrix.
+Across biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist broadleaf forests and the highest relative increases at high latitudes in boreal forests.
+Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (\@ref(fig:vert-dif-biome)).
 
@@
 
-Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studeis that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019].
+Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studies that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019].
 Though climate conditions influenced verticality, they were neither able to predict extremely high or low verticality nor spatial variation in the influence of canopy height on verticality.
-Furthermore, `r word_comment(highlight = "Furthermore", comment = "Try bootstrapping method to compare slopes")` regions exhibitin positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between verticality and species richness.
-Our results thus do not support our hypothesis that climtatic conditions that support high assemblage verticality lead to positive impacts of canopy hiehg on verticality and positive associations between verticality and species richness.
-Vertical stratification across increasingly tall and multi-layered forests may therefore not be a primary mehcanism driving positive relationships between canopy height, verticality, and species richness at global scales.
-Rather, we propose that species packing and evolutionary patterns may explain our unexpected results.
-The ways in which climate change will impact verticality and its relationship with species richness is thus difficult to predict.
+Furthermore, regions exhibiting positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between canopy height and species richness.
+We propose that the rate of niche packing across evolutionary time spans rather than expansion across tall forests may explain both the limited capacity for climate alone to predict high levels of verticality and the lack of the three-way positive relationship between canopy height, verticality, and species richness for some taxa.
+Given the potentially large role of evolutionary patterns in global distributions of verticality, our models may underpredict the impacts of climate change on vertical assemblage structure.
 
@@
 
-Climate influences global patterns of verticality because more extreme climates in teh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
-Water availabitliy therefore facilitates verticality, while temperatures at both hot and cold extremes and awter scaracity generally limit verticality.
-For exmaple, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life straties outside moist forests in tropical regions.
+Climate influences global patterns of verticality because more extreme climates in thh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
+Water availability therefore facilitates verticality, while temperatures at both hot and cold extremes and water scarcity generally limit verticality.
+For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist forests in tropical regions.
 However, our models indicated nuances in the relative influences of different drivers across taxa.
@@
 Impacts of precipitation were particularly pronounced for amphibian verticality.
-In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consitently available (*cite*).
-However, in regions that experience pronounced dry seasons, amphibians behaviourally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
-Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desccation witout adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
-*Birds were also influenced by precipitation, likely because they are dependent on the availability of standing water in teh canopy and water content of leaves.* `r word_comment(highlith = "leaves", comment = "need to look more into impacts of drought on birds - see https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.10781")`
+In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consistently available [@sundayThermalsafetyMarginsNecessity2014].
+However, in regions that experience pronounced dry seasons, amphibians behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
+Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
 
-For mammals and reptiles, our modesl indicated that minimum cold temperatures had the greatest impact on verticality.
+For mammals and reptiles, our models indicated that minimum cold temperatures had the greatest impact on verticality.
 In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014].
-Species are therefore largely constratined to terrestrial or fossorial habiats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019].
+Species are therefore largely constrained to terrestrial or fossorial habitats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019].
 Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors contribute to unexplained variation.
@@
 
-## Environmental drivers of verticality
+## Impacts of climate change on verticality
 
-### Climate
+Increasing temperature and declining precipitation could alter global patterns of assemblage verticality via vertical niche shifts, range shifts, or local extinction.
+Our models indicated relatively small declines of verticality in tropical latitudes in response to climate change at a global scale.
+In the absence of climatic buffering experienced in the understory, species living in arboreal habitats may be threatened by the increasing magnitude and frequency of extreme heat and drought [@ipccClimateChange20232023].
+The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small themral safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014].
+To avoid unsuitable conditions that exceed physiological tolerances, species could behaviorally regulate climate exposure by spending more time in the understory, thus reducing assemblage verticality [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020].
+Indeed, vertical niche shifts have been observed at smaller spatiotemporal scales across tropical forests.
+For example, birds and arboreal frogs increase their use of understory resources during the dry season [@delgado-martinezDifferentialUtilizationSurface2023; @bashamVerticalStratificationCollapses2020].
+Arboreal primates from the Americas to Madagascar similarly increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022].
+Alternatively, arboreal species could become locally extinct if they are not able to adjust microhabitat use to avoid extreme climate conditions.
 
-### Vegetation
+Increasing minimum cold temperatures could release constraints on arboreality in temperate and boreal regions for birds and mammals.
+Our models predicted the highest increases in verticality in boreal forests, which could occur, for example, as optimal latitudinal ranges of birds in low latitude assemblages with higher verticality shift poleward [@laughlinWinterRangeShifts2023; @martinsSignificantShiftsLatitudinal2024].
+Increases in verticality caused by range shifts will likely be restricted to extra-tropical assemblages, because the ranges of tropical species typically move across elevational rather than latitudinal gradients in response to climate change [@colwellStillLittleEvidence2024].
+Vertical niche shifts are less likely to contribute to increases in verticality due to morphological constraints on arboreal strategies are likely to prevent vertical niche shifts causing increases in vertical strategies [@camargoContrastingRealizedFundamental2016; @dealencarArborealityConstrainsMorphological2017].
+Yet, use of arboreal habitats by terrestrial species has occasionally been observed in high latitude temperate climates [@petrovanWhyLinkDiverse2022].
 
-## Verticality species-richness relationships
-
-## Impact of climate change on verticality
+Despite the small changes in verticality predicted by our models under future climate scenarios, realized change, particularly declines may be larger.
+First, underprediction of declines may arse because warming and drying climates could impact a disproportionate number of arboreal species in the tropics where climate alone failed to predict high verticality.
+Second, habitat loss caused by logging and deforestation may have more immediate effects on the vertical structure of species assemblages than climate change [@lindenmayerResponseArborealMarsupials2021; @murrayClimateLandUse2021].
+Understanding the extent to which increases or decreases in verticality are realized and attributable to climate change will require long-term monitoring of taxa and and vertical microclimates across latitudes.
 
@@
 
-```{r fig.id="vert-map-global", fig.width=160, fig.height=96}
-include_graphics("./../figures/main_figs/vert_maps.png", error = F)
+```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians.
+Line plots inidicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
+Gray areas indicate areas that are not wooded or that have fewer than five species."}
+
+read_docx() |>
+  body_add_img(src = "figures/main_figs/vert_maps.png", style="centered", width=160, height=96)
+#include_graphics("figures/main_figs/vert_maps.png")
 ```
 
-```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3}
-include_graphics("./../figures/main_figs/sdmTMB_coefs.png", error = F)
+```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
+tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}
+
+read_docx() |>
+  body_add_img(src = "figures/main_figs/sdmTMB_coefs.png", style="centered", width=160, height=87.3)
+
+#include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
 ```
 
-```{r fig.id="svc", fig.width=160, fig.height=143.95}
-include_graphics("./../figures/main_figs/canopy_height_coefs.png", error = F)
+```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians.
+Gray areas indicate areas that are not wooded or that have fewer than five species."}
+
+#include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
 ```
 
-```{r fig.id="corplot"}
-include_graphics("./../figures/main_figs/vert_rich_cor.png", error = F)
+```{r fig.id="corplot", fig.width = 160, fig.height=120, fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive.
+"}
+
+#include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
 ```
 
-```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3}
-include_graphics("./../figures/main_figs/vert_difs.png", error = F)
+```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians.
+Bar plots inidcate average relative percent change per latitude.
+Gray areas indicate areas that are not wooded or that have fewer than five species."}
+
+#include_graphics("figures/main_figs/vert_difs.png", error = F)
 ```
@@
 # Supplementary tables
+*Remove NA values because gradient issue is okay Re: question posted to sdmTMB forum*
+
+```{r tab.id="mod-selection", tab.lp="stab", tab.cap="Model selection criteria for models of mean and SES verticality for all four vertebrate taxa.
+The log-likelihood, AIC, and delta AIC are for models trained with all points.
+The average log-likelihood is the average sum of log-likelihood values for left out test data across five folds.
+NA values are caused by not meeting all model assumptions across all five cross-validation folds."}
+tab = read.csv("tables/supplement/model_selection_compare.csv")
+
+tab = tab %>% 
+ # filter(response_var == "Mean verticality") %>% 
+  relocate(taxon) %>% 
+  relocate(response_var, .after = taxon) %>% 
+  arrange(taxon, response_var, deltaAIC) %>% 
+  mutate(taxon = factor(taxon, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
+  mutate_at(.vars = c("logLik", "AIC", "deltaAIC", "avg_logLik"), .funs = function(x)round(x, 2))
+
+flextable(tab) %>% 
+  merge_v(j=~ taxon + response_var) %>%
+  hline(i=seq(4,32,4), part = "body",
+        border = fp_border(color = "black", width = 3)) %>% 
+  vline(j = c(1, 2), part = "body",
+        border = fp_border(color="black", width = 3)) %>% 
+  colformat_double(i = ~response_var, j = "taxon") %>% 
+  fix_border_issues()
+```
 
@@
 ```{r fig.id="vert-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
-include_graphics("./../figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
+#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
@@
 
-include_graphics("./../figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
+#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
@@
 
-include_graphics("./../figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
+#include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
 ```
@@
 
-## Reference
+# References
 
-## Acknowledgements
-
-We would like to thank Roger Bivand for useful insights regarding predictions of SAR models.
+# Acknowledgements

### 2024-08-16 verticality_future_mkd.Rmd
full first draft

--- 2024-08-16 docs/verticality_future_mkd.Rmd
@@
 ---
-date: "`r Sys.Date()`"
-author: "Your Name"
-title: "officedown template"
+author: "Lydia Soifer, David Edwards, Brett Scheffers"
+title: "Global patterns and drivers of vertebrate verticality"
 bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
@@
   officedown::rdocx_document:
+    reference_docx: "style-guide.docx"
     mapstyles:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE)
+knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
 library(officedown)
@@
 library(knitr)
+library(chunkhooks)
 
@@
 ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
+
+hook_figure_unit("mm")
 ```
@@
 Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@wagnerRemotelySensedTree2023] (*cite positive relationships between canopy height and species richness at local scales*).
+By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @wagnerRemotelySensedTree2023].
 *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019].
@@
 Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019].
-Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*.
+Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019].
 Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage.
@@
 $$
+\begin{align}
 y_s ∼ 𝜇_{s}, \\
@@
 ⍵_s ∼ MVNormal(0,𝚺_⍵)
+\end{align}
 $$
@@
 
-*Calculate relative change as (future-present)/(presentmax-presentmin)*
-
-```{r}
+```{r cache=TRUE, cache.path="cache/"}
 # calculate percent change in verticality
@@
 
-Climate influences global patterns of verticality because more extreme climates in thh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
+Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
 Water availability therefore facilitates verticality, while temperatures at both hot and cold extremes and water scarcity generally limit verticality.
@@
 In the absence of climatic buffering experienced in the understory, species living in arboreal habitats may be threatened by the increasing magnitude and frequency of extreme heat and drought [@ipccClimateChange20232023].
-The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small themral safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014].
+The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small thermal safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014].
 To avoid unsuitable conditions that exceed physiological tolerances, species could behaviorally regulate climate exposure by spending more time in the understory, thus reducing assemblage verticality [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020].
@@
 
-```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians.
+```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap.pre="Fig.
+", fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians.
 Line plots inidicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
+Gray areas indicate areas that are not wooded or that have fewer than five species.", include=TRUE}
+
+knitr::include_graphics("figures/main_figs/vert_maps.png", error = F)
+```
+
+```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap.pre="Fig.
+", fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
+tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}
+
+include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
+```
+
+```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap.pre="Fig.
+", fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians.
 Gray areas indicate areas that are not wooded or that have fewer than five species."}
 
-read_docx() |>
-  body_add_img(src = "figures/main_figs/vert_maps.png", style="centered", width=160, height=96)
-#include_graphics("figures/main_figs/vert_maps.png")
+include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
 ```
 
-```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
-tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}
+```{r fig.id="corplot", fig.width = 160, fig.height=213.4, fig.cap.pre="Fig.
+", fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive.
+"}
 
-read_docx() |>
-  body_add_img(src = "figures/main_figs/sdmTMB_coefs.png", style="centered", width=160, height=87.3)
-
-#include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
+include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
 ```
 
-```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians.
-Gray areas indicate areas that are not wooded or that have fewer than five species."}
-
-#include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
-```
-
-```{r fig.id="corplot", fig.width = 160, fig.height=120, fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive.
-"}
-
-#include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
-```
-
-```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians.
+```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap.pre="Fig.
+", fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians.
 Bar plots inidcate average relative percent change per latitude.
@@
 
-#include_graphics("figures/main_figs/vert_difs.png", error = F)
+include_graphics("figures/main_figs/vert_reldifs.png", error = F)
 ```
@@
 \newpage
+
+# Supplementary methods
+## Species data
+
+We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. (2017).
+
+The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
+For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7.
+
+## Mesh development
 
@@
 
-```{r fig.id="vert-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
-#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
+```{r fig.id="vert-biome", fig.lp="suppfig", fig.cap="", fig.cap.pre="Fig. S", fig.autonum.start_at=1, fig.height=80, fig.width=160}
+include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
 
-```{r fig.id="vert-dif-biome", fig.lp="ed", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
+```{r fig.id="vert-dif-biome", fig.lp="suppfig", fig.cap.pre="Fig.
+S", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
 Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
@@
 
-#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
+include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
 ```
 
-```{r fig.id="svc-sig", fig.lp="ed", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}
+```{r fig.id="svc-sig", fig.lp="suppfig", fig.cap.pre="Fig.
+S", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}
 
-#include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
+include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
 ```

### 2024-08-16 scripts/04_plots/coef_plot.R
add significance color

@@ -4,6 +4,7 @@ library(tidyverse)
 library(lemon)
 library(ggh4x)
 library(spatialreg)
+library(sdmTMB)
 
 # VERTMEAN ----------------------------------------------------------------
 
@@ -37,7 +38,8 @@ birds.coefs$class = "Birds"
 
 meanvert.coefs = bind_rows(amph.coefs, mammals.coefs, reptiles.coefs, birds.coefs)%>% 
   mutate(class = factor(class, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")),
-         type = "Mean Verticality")
+         type = "Mean Verticality",
+         sig = ifelse(conf.low*conf.high > 0, "sig", "notsig"))
 
 
 # * - plot --------------------------------------------------------------------
@@ -91,7 +93,8 @@ birds.coefs$class = "Birds"
 
 sesvert.coefs = bind_rows(amph.coefs, mammals.coefs, reptiles.coefs, birds.coefs)%>% 
   mutate(class = factor(class, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")),
-         type = "SES Mean Verticality")
+         type = "SES Mean Verticality",
+         sig = ifelse(conf.low*conf.high > 0, "sig", "notsig"))
 
 
 
@@ -123,7 +126,7 @@ coefs = bind_rows(sesvert.coefs, meanvert.coefs) %>%
 
 library(ggh4x)
 coef.plt = coefs %>%
-  ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
+  ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, color = sig)) +
   geom_pointrange(size = 0.15) +
   geom_vline(xintercept = 0, linetype = "dashed") +
   #geom_text(data = r2, aes(label = paste0("R\u00b2 = ", r2), x = Estimate), y = 2, inherit.aes = F, hjust = 1, vjust = 0) +
@@ -132,6 +135,7 @@ coef.plt = coefs %>%
   #scale_size_continuous(range = c(0.02,0.5)) +
   scale_y_discrete("") +
   scale_x_continuous("Estimate") +
+  scale_color_manual(values = c("black", "red3")) +
   facet_grid2(rows = vars(type), cols = vars(class), scales = "free_x", independent = "x") +
   scale_x_facet(ROW == 1, limits = c(-0.08, 0.25)) +
   scale_x_facet(ROW == 2, limits = c(-0.2, 0.8)) +
[…]

### 2024-08-16 verticality_future_mkd.Rmd
results revisions

@@
 
-### Mesh development
-
-The Gaussian random field is approximated using a spatial mesh.
-We constructed separate meshes for all taxa and for mean and SES verticality using the ‘fmesher’ R package.
-The mesh used to approximate the spatial Gaussian random field (GRF) impacts model results as well as computational time.
-To assess sensitivity of the model to the mesh and select the optimal mesh, we used an iterative procedure to update minimum and maximum edge lengths of triangles in the mesh based on the spatial range of the data (i.e., the distance at which points exhibit a correlation of approximately 0.1) [@thorsonSpatioTemporalModelsEcologists2024].
-We first estimated the spatial range by visually inspecting spatial correlations based on a subset of 1000 points.
-To avoid over-smoothing the GRF while or over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024].
-To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range.
-Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range (*cite*).
-To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
-We then fit a model using the resulting mesh.
-If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model.
-We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of spatial domain used to construct the mesh.
-We selected the mesh used in the final model for all following analysis.
-
 ### Model comparisons
@@
 
-Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (\@ref(fig:vert-map-global)).
+Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (Fig.
+\@ref(fig:vert-map-global)).
 This pattern was largely driven by declines in the proportion of arboreal species toward the poles.
-Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical moist broadleaf forests (\@ref(fig:vert-biome)).
+Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical moist forests (Fig.
+S\@ref(fig:vert-biome)).
+Within tropical and subtropical biomes, verticality was highest in moist forests and lowest in dry forests (Fig.
+\@ref(fig:vert-biome)).
+In moist forests, verticality of bird and mammal assemblages was consistently high, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies.
+In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Malaysian and Indonesian forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar.
 
-Despite widespread coherence of global verticality patterns among taxa and within biomes, regional variation occurred.
-Within tropical and subtropical biomes, verticality was highest in moist broadleaf forests and lowest in dry broadleaf forests (\@ref(fig:vert-biome)).
-In moist broadleaf forests, verticality of bird and mammal assemblages was consistently high, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies, while the proportion of terrestrial species declined.
-In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Malaysian and Indonesian forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar.
-Additionally, the proportion of reptiles and amphibians that exhibit terrestrial strategies did not decline to the same extent as birds and mammals.
-*include proportion maps in supplement*
-
-In temperate biomes, verticality was similar across forest types but declined in less dense wooded habitats (e.g., savannas and shrublands) (\@ref(fig:vert-biome)).
+In temperate biomes, verticality was similar across forest types but declined in less dense wooded habitats (e.g., savannas and shrublands) (Fig.
+S \@ref(fig:vert-biome)).
 Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes.
-Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (\@ref(fig:vert-map-global)L, though temperate rainforests (e.g., the Pacific-Northwest of North America) were a notable exception.
-Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (\@ref(fig:vert-map-global)F).
+Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (Fig.
+\@ref(fig:vert-map-global)L), though temperate rainforests (e.g., the Pacific-Northwest of North America) were a notable exception.
+Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (Fig.
+\@ref(fig:vert-map-global)F).
 In contrast, the proportion of fossorial reptiles declined in temperate regions.
 Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species.
-Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal gradient in SES verticality.
+Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal verticality gradient.
 
@@
 
-The spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (\@ref(fig:mod-coefs), *supp table for model results*).
-Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (\@ref(fig:mod-coefs)).
-Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality.
-Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile mean and SES verticality, as well as SES verticality of mammals.
-Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship.
-These patterns resulted in declines in verticality at both extremely high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (*see conditional effect plots*).
-Climate velocity negatively impacted bird and amphibian verticality and positively impacted mammal and reptile verticality.
-However, environmental factors were unable to explain extremely high and extremely low verticality as indicated by larger residuals at the ends of the verticality spectrum when spatial random effects were not included in estimates of assemblage verticality (*supp figure*).
-
-### Vegetation
-
-We hypothesized that canopy height and vegetation density would positively impact verticality, but that the impact of canopy height would vary across space.
-Specifically, we expected canopy height to have a strong positive effect on verticality in warmer and wetter regions that support arboreal assemblages and a negligible effect in regions with climate conditions that support fewer arboreal species.
-Results of the GLMMs did not entirely support our predictions.
-Globally, canopy height positively impacted SES verticality of reptiles and amphibians and mean verticality of amphibians, while vegetation density had no significant impact on mean or SES verticality of any taxa (\@ref(fig:mod-coefs)).
-In contrast to expectations, canopy height did not have a positive impact on verticality across much of the tropics where climate conditions support arboreality (\ref(fig:svc)).
-Significant positive impacts of canopy height generally occurred in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (\@ref(fig:svc-sig)).
+Spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (Fig.
+\@ref(fig:mod-coefs)).
+Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (Fig.
+\@ref(fig:mod-coefs)).
+Minimum temperature of the coldest month positively impacted mammal and reptile verticality and did not significantly impact bird and amphibian verticality.
+Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammals verticality.
+These patterns resulted in lower verticality at both high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (Fig.
+S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)).
+Canopy height had globally positive impacts on SES verticality of reptiles and amphibians and mean verticality of amphibians, and vegetation density had no significant impact on mean or SES verticality of any taxon Fig.
+(\@ref(fig:mod-coefs)).
+However, impacts of canopy height varied spatially, exhibiting neutral impacts on verticality across much of the tropics and positive impacts in regions in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (Fig.
+\@ref(fig:svc), Fig.
+S\@ref(fig:svc-sig) ).
+Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals when spatial random effects were excluded from model predictions (*supp figure*).
 
@@
 
-When considering all points, Pearson correlations indicated strong positive relationships between SES or mean verticality and species richness for birds and mammals, a weak positive relationship for amphibians, and nearly no relationship for reptiles (*add R2 and p-values*) (\@ref(fig:corplot)).
+When considering all points, Pearson correlations indicated strong positive relationships between SES or mean verticality and species richness for birds and mammals, a weak positive relationship for amphibians, and nearly no relationship for reptiles (Fig.
+\@ref(fig:corplot)).
 When we restricted the correlation to only include points where canopy height had a significant positive effect on verticality, the strength of the relationship between SES verticality and richness increased for amphibians and SES verticality of birds, but declined for mammals, reptiles, and mean verticality of birds.
@@
 
-Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality and varied across taxa (\@ref(fig:vert-dif-map)).
+Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality and varied across taxa (Fig.
+\@ref(fig:vert-dif-map)).
 Amphibians exhibited the highest predicted relative percent change, with predicted declines reaching up to `r amphses.minreldif*-1`% for SES verticality and `r amphmean.minreldif*-1`% for mean verticality in parts of the tropics and predicted increases reaching up to `r amphses.maxreldif`% for SES verticality and `r amphmean.maxreldif`% for mean verticality in the African forest-savanna matrix.
-Across biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist broadleaf forests and the highest relative increases at high latitudes in boreal forests.
-Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (\@ref(fig:vert-dif-biome)).
+Across biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist forests and the highest relative increases at high latitudes in boreal forests.
+Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (Fig.
+S\@ref(fig:vert-dif-biome)).
 
@@
 
+The Gaussian random field is approximated using a spatial mesh.
+We constructed separate meshes for all taxa and for mean and SES verticality using the ‘fmesher’ R package.
+The mesh used to approximate the spatial Gaussian random field (GRF) impacts model results as well as computational time.
+To assess sensitivity of the model to the mesh and select the optimal mesh, we used an iterative procedure to update minimum and maximum edge lengths of triangles in the mesh based on the spatial range of the data (i.e., the distance at which points exhibit a correlation of approximately 0.1) [@thorsonSpatioTemporalModelsEcologists2024].
+We first estimated the spatial range by visually inspecting spatial correlations based on a subset of 1000 points.
+To avoid over-smoothing the GRF while or over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024].
+To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range.
+Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range (*cite*).
+To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
+We then fit a model using the resulting mesh.
+If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model.
+We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of spatial domain used to construct the mesh.
+We selected the mesh used in the final model for all following analysis.
+
 # Supplementary tables
@@
 
-```{r tab.id="mod-selection", tab.lp="stab", tab.cap="Model selection criteria for models of mean and SES verticality for all four vertebrate taxa.
+```{r tab.id="mod-tab", tab.lp="stab", tab.cap="Model selection criteria for models of mean and SES verticality for all four vertebrate taxa.
 The log-likelihood, AIC, and delta AIC are for models trained with all points.
@@
 NA values are caused by not meeting all model assumptions across all five cross-validation folds."}
-tab = read.csv("tables/supplement/model_selection_compare.csv")
-
-tab = tab %>% 
- # filter(response_var == "Mean verticality") %>% 
-  relocate(taxon) %>% 
-  relocate(response_var, .after = taxon) %>% 
-  arrange(taxon, response_var, deltaAIC) %>% 
-  mutate(taxon = factor(taxon, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
-  mutate_at(.vars = c("logLik", "AIC", "deltaAIC", "avg_logLik"), .funs = function(x)round(x, 2))
-
-flextable(tab) %>% 
-  merge_v(j=~ taxon + response_var) %>%
-  hline(i=seq(4,32,4), part = "body",
-        border = fp_border(color = "black", width = 3)) %>% 
-  vline(j = c(1, 2), part = "body",
-        border = fp_border(color="black", width = 3)) %>% 
-  colformat_double(i = ~response_var, j = "taxon") %>% 
-  fix_border_issues()
+# tab = read.csv("tables/supplement/model_selection_compare.csv")
+# 
+# tab = tab %>% 
+#  # filter(response_var == "Mean verticality") %>% 
+#   relocate(taxon) %>% 
+#   relocate(response_var, .after = taxon) %>% 
+#   arrange(taxon, response_var, deltaAIC) %>% 
+#   mutate(taxon = factor(taxon, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
+#   mutate_at(.vars = c("logLik", "AIC", "deltaAIC", "avg_logLik"), .funs = function(x)round(x, 2))
+# 
+# flextable(tab) %>% 
+#   merge_v(j=~ taxon + response_var) %>%
+#   hline(i=seq(4,32,4), part = "body",
+#         border = fp_border(color = "black", width = 3)) %>% 
+#   vline(j = c(1, 2), part = "body",
+#         border = fp_border(color="black", width = 3)) %>% 
+#   colformat_double(i = ~response_var, j = "taxon") %>% 
+#   fix_border_issues()
 ```
@@
 
-```{r fig.id="vert-dif-biome", fig.lp="suppfig", fig.cap.pre="Fig.
-S", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
-Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
-Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}
+```{r fig.id="condeff-mean", fig.lp="suppfig", fig.cap="Conditional effect plots for mean verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
+S"}
+include_graphics("figures/supp_figs/cond_effects_meanvert.png", error = F)
+```
 
-include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
+```{r fig.id="condeff-ses", fig.lp="suppfig", fig.cap="Conditional effect plots for SES verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
+S"}
+include_graphics("figures/supp_figs/cond_effects_sesvert.png", error = F)
 ```
@@
 include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
+```
+
+```{r fig.id="vert-dif-biome", fig.lp="suppfig", fig.cap.pre="Fig.
+S", fig.cap="Average percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
+Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
+Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}
+
+include_graphics("figures/supp_figs/dif_biome_forestOnly_forestSES.png", error = F)
 ```

### 2024-08-21 verticality_future_mkd.Rmd
revision

@@
 ---
-author: "Lydia Soifer, David Edwards, Brett Scheffers"
+author: ""
 title: "Global patterns and drivers of vertebrate verticality"
@@
 
-\newpage
-
 # Introduction
@@
 At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @wagnerRemotelySensedTree2023].
-*concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019].
-However, canopy height and other metrics of vegetation structure do not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
-Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.
+Tall, multi-layered forests increase both habitat volume and niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+By reducing competition, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023].
+However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
+Examining the impacts of environmental drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.
 
 At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests.
-Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-In high latitude forests, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
-Similarly, exposure to hot and dry conditions in the canopies of tropical lowland forests may reduce the suitability of arboreal habitats for heat or desiccation sensitive species, such as frogs [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020].
-Release of these climatic constraints at higher elevations in the tropics, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013].
+Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
+At high latitudes, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
+In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient and contribute to the high diversity of arboreal species found in these habitats [@oliveiraVerticalStratificationInfluences2019].
 We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.
@@
 Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage.
-In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak (*concept fig*).
+In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak.
 
@@
 At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
-While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.
 
@@
 
-Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates.
+Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, its relationship with species richness, and its vulnerability to climate change have not been comprehensively described for vertebrates at a global scale.
 To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes.
@@
 
-We mapped the verticality and richness of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection.
+We mapped verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] using 50 km grid cells and a cylindrical equal area projection (Supplementary methods).
 Grid cells with fewer than five species were excluded.
-We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al.
-[-@dinersteinEcoregionBasedApproachProtecting2017].
 Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
-For birds, we used breeding plus resident ranges, because seasonality impacts the verticality of assemblages.
+For birds, we used breeding plus resident ranges.
 To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell.
 For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024].
-This dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
-For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7.
 For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014].
@@
 
-Verticality exhibited a strong positive correlation with species richness.
-To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool).
+To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality ($SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool).
 We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.
@@
 Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
-All environmental data was resampled to the resolution and projection of the assemblage data.
+All environmental data were resampled to the resolution and projection of the species data.
 
@@
 $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients.
-$α_g$ represents an optional random intercept by group g.
+$α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality drive by global evolutionary patterns.
 $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept.
-$𝜲^{svc}_sζ_s$ represents an optional spatially varying coefficient, where $ 𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$).
-The spatial range for the covariance function is shared with the range for the spatial random field.
-Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with the terms following a multivariate normal distribution with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
+$𝜲^{svc}_sζ_s$ represents a spatially varying coefficient, where $𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$).
+Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
+Gaussian random fields were approximated for models of mean and SES verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019] (Supplementary methods).
 The model is specified such that both spatial terms share the same range parameter.
 
-### Model comparisons
+### Model comparison and evaluation
 
@@
 Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package.
-Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Table S1 `r word_comment(comment = "add supp table", highlight = "Table S1")`).
+Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabe S\@ref(tab:mod-compare-tab)).
 Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.
 
-### Model validation
+We validated models first using the ‘sanity’ function from the sdmTMB package.
+We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space and to examine any patterns across rank transformed and raw model predictions.
 
-We validated models first using the ‘sanity’ function from the sdmTMB package.
-We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space, rank transformed model predictions, raw model predictions.
-
-To estimate uncertainty of SVCs for canopy height, we added draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations.
+We estimated uncertainty of SVCs for canopy height by adding draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations.
 We then calculated the median and 0.025 and 0.975 quantiles to obtain a 95% confidence interval for each SVC estimate.
@@
 
-To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points, only points where canopy height significantly impacted verticality, and only points where canopy height had a positive significant effect on verticality.
+To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points and only points where canopy height had a positive significant effect on verticality.
 
@@
 
-To examine potential impacts of climate change on assemblage verticality, we predicted mean and SES verticality under a future climate scenario using the models.
-However, we held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density.
-From the future predictions, we subtracted predictions to the data used to train the model (present predictions) and mapped the resulting differences.
-To standardize measures of change across taxa and between mean and SES verticality, we additionally calculated the percent relative change between future and present predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
-Thus, negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
+To examine potential impacts of climate change on assemblage verticality, we used the models to predict mean and SES verticality under a future climate scenario.
+We held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density.
+We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences.
+To standardize measures of change across taxa and between mean and SES verticality, we calculated the percent relative change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
+Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
 
@@
 
-### Climate
-
 Spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (Fig.
@@
 Minimum temperature of the coldest month positively impacted mammal and reptile verticality and did not significantly impact bird and amphibian verticality.
-Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammals verticality.
+Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammal verticality.
 These patterns resulted in lower verticality at both high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (Fig.
@@
 S\@ref(fig:svc-sig) ).
-Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals when spatial random effects were excluded from model predictions (*supp figure*).
+Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals (Fig.
+S\@ref(fig:resid-plt)).
 
@@
 Impacts of precipitation were particularly pronounced for amphibian verticality.
-In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consistently available [@sundayThermalsafetyMarginsNecessity2014].
-However, in regions that experience pronounced dry seasons, amphibians behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
+In tropical wet forests, consistent water availability allows amphibians to maintain body temperatures below critical thermal maxima via evapotranspiration [@sundayThermalsafetyMarginsNecessity2014].
+However, in regions that experience pronounced dry seasons, amphibians may behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
 Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
@@
 Species are therefore largely constrained to terrestrial or fossorial habitats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019].
-Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors contribute to unexplained variation.
+Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors influence global patterns of verticality.
 
@@
 Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015].
-These results align with Basham et al.
+These results correspond to Basham et al.
 [-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics.
@@
 Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015].
-Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (Fig.
+S\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
 Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality.
@@
 However, this three-way positive relationship did not occur in the most diverse regions, including much of the tropics.
-Instead, it was largely restricted to areas where relatively rapid increases in vegetation structure from high to low elevations or dry to wet forests (e.g., Madagascar) may facilitate the addition of species outside the habitat space occupied by species in assemblages of less productive habitats [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018].
+Instead, it was largely restricted to areas where relatively rapid increases in vegetation structure from high to low elevations or dry to wet forests (e.g., Madagascar) may facilitate the addition of species outside the habitat space occupied by species with lower richness in less productive habitats [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018].
 Yet in regions where canopy height did not have significant positive impacts on verticality, positive correlations between verticality and richness remained for birds and amphibians and increased in strength for mammals and reptiles, suggesting vertical stratification (i.e., niche expansion) may not be the dominant mechanism causing positive relationships between canopy height, verticality, and species richness at large spatial scales.
@@
 Niche packing occurs when smaller niche spaces or greater niche overlap allow more species to co-exist within a habitat [@pigotFunctionalTraitsReveal2016; @pellissierNichePackingExpansion2018].
-Specialization on specific resources within the canopy over evolutionary time could thus increase arboreal species diversity without increasing canopy height [@fengForestCanopyHeight2020; @pellissierNichePackingExpansion2018].
+Specialization on specific resources within the canopy over evolutionary time could thus increase arboreal species diversity in the absence of increasing canopy height [@fengForestCanopyHeight2020; @pellissierNichePackingExpansion2018].
 Furthermore, evolutionary rates of arboreal lineages are generally faster than those of terrestrial and fossorial lineages [@cyriacDiggingTheirOwn2018; @liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017].
-In the tropics where climate conditions support arboreality, the combination of niche packing and rapid evolutionary rates of arboreal species could lead to a disproportionately high number of arboreal species, increasing verticality relative to what would be expected by climate conditions alone and causing positive relationships between verticality and richness in regions where canopy height has insignificant or even negative effects on verticality.
+In the tropics where climate conditions support arboreality, the combination of niche packing and rapid evolutionary rates of arboreal species could thus lead to a disproportionately high number of arboreal species, increasing verticality relative to what would be expected by climate conditions alone and causing positive relationships between verticality and richness in regions where canopy height has insignificant or even negative effects on verticality.
 
@@
 Increases in verticality caused by range shifts will likely be restricted to extra-tropical assemblages, because the ranges of tropical species typically move across elevational rather than latitudinal gradients in response to climate change [@colwellStillLittleEvidence2024].
-Vertical niche shifts are less likely to contribute to increases in verticality due to morphological constraints on arboreal strategies are likely to prevent vertical niche shifts causing increases in vertical strategies [@camargoContrastingRealizedFundamental2016; @dealencarArborealityConstrainsMorphological2017].
+Vertical niche shifts are less likely to contribute to increases in verticality due to morphological constraints on arboreal strategies [@camargoContrastingRealizedFundamental2016; @dealencarArborealityConstrainsMorphological2017].
 Yet, use of arboreal habitats by terrestrial species has occasionally been observed in high latitude temperate climates [@petrovanWhyLinkDiverse2022].
 
-Despite the small changes in verticality predicted by our models under future climate scenarios, realized change, particularly declines may be larger.
-First, underprediction of declines may arse because warming and drying climates could impact a disproportionate number of arboreal species in the tropics where climate alone failed to predict high verticality.
+Despite the small changes in verticality predicted by our models under future climate scenarios, realized change, particularly declines in verticality, may be larger.
+First, underprediction of declines may arise because warming and drying climates could impact a disproportionate number of arboreal species in the tropics where climate alone failed to predict high verticality.
 Second, habitat loss caused by logging and deforestation may have more immediate effects on the vertical structure of species assemblages than climate change [@lindenmayerResponseArborealMarsupials2021; @murrayClimateLandUse2021].
-Understanding the extent to which increases or decreases in verticality are realized and attributable to climate change will require long-term monitoring of taxa and and vertical microclimates across latitudes.
+Understanding the extent to which increases or decreases in verticality are realized and attributable to climate change will require long-term monitoring of taxa and vertical microclimates across latitudes.
 
@@
 ", fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians.
-Line plots inidicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
+Line plots indicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
 Gray areas indicate areas that are not wooded or that have fewer than five species.", include=TRUE}
@@
 
+\newpage
+
 ```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap.pre="Fig.
 ", fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
+Red points indicate coefficients with 95% CIs that do not overlap zero.
 tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}
@@
 ```
+
+\newpage
 
@@
 
+\newpage
+
 ```{r fig.id="corplot", fig.width = 160, fig.height=213.4, fig.cap.pre="Fig.
@@
 
+\newpage
+
 ```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap.pre="Fig.
 ", fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians.
-Bar plots inidcate average relative percent change per latitude.
+Bar plots indicate average relative percent change per latitude.
 Gray areas indicate areas that are not wooded or that have fewer than five species."}
@@
 # Supplementary methods
+
 ## Species data
 
-We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. (2017).
-
+We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al.
+(2017).
 The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
@@
 To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range.
-Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range (*cite*).
+Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range.
 To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
@@
 # Supplementary tables
-*Remove NA values because gradient issue is okay Re: question posted to sdmTMB forum*
 
-```{r tab.id="mod-tab", tab.lp="stab", tab.cap="Model selection criteria for models of mean and SES verticality for all four vertebrate taxa.
+```{r tab.id="mod-compare-tab", tab.lp="stab", tab.cap="Model selection criteria for models of mean and SES verticality for all four vertebrate taxa.
 The log-likelihood, AIC, and delta AIC are for models trained with all points.
@@
 NA values are caused by not meeting all model assumptions across all five cross-validation folds."}
-# tab = read.csv("tables/supplement/model_selection_compare.csv")
-# 
-# tab = tab %>% 
-#  # filter(response_var == "Mean verticality") %>% 
-#   relocate(taxon) %>% 
-#   relocate(response_var, .after = taxon) %>% 
-#   arrange(taxon, response_var, deltaAIC) %>% 
-#   mutate(taxon = factor(taxon, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
-#   mutate_at(.vars = c("logLik", "AIC", "deltaAIC", "avg_logLik"), .funs = function(x)round(x, 2))
-# 
-# flextable(tab) %>% 
-#   merge_v(j=~ taxon + response_var) %>%
-#   hline(i=seq(4,32,4), part = "body",
-#         border = fp_border(color = "black", width = 3)) %>% 
-#   vline(j = c(1, 2), part = "body",
-#         border = fp_border(color="black", width = 3)) %>% 
-#   colformat_double(i = ~response_var, j = "taxon") %>% 
-#   fix_border_issues()
+tab = read.csv("tables/supplement/model_selection_compare.csv")
+
+tab = tab %>%
+ # filter(response_var == "Mean verticality") %>%
+  relocate(taxon) %>%
+  relocate(response_var, .after = taxon) %>%
+  arrange(taxon, response_var, deltaAIC) %>%
+  mutate(taxon = factor(taxon, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>%
+  mutate_at(.vars = c("logLik", "AIC", "deltaAIC", "avg_logLik"), .funs = function(x)round(x, 2))
+
+flextable(tab) %>%
+  merge_v(j=~ taxon + response_var) %>%
+  hline(i=seq(4,32,4), part = "body",
+        border = fp_border(color = "black", width = 3)) %>%
+  vline(j = c(1, 2), part = "body",
+        border = fp_border(color="black", width = 3)) %>%
+  colformat_double(i = ~response_var, j = "taxon") %>%
+  fix_border_issues()
 ```
@@
 
-```{r fig.id="vert-biome", fig.lp="suppfig", fig.cap="", fig.cap.pre="Fig. S", fig.autonum.start_at=1, fig.height=80, fig.width=160}
+```{r fig.id="vert-biome", fig.lp="suppfig", fig.cap="Average mean and SES verticality for each taxonomic group in global biomes.
+Error bars indicate standard deviations.", fig.cap.pre="Fig.
+S", fig.autonum.start_at=1, fig.height=80, fig.width=160}
 include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
@@
 
+\newpage
+
 ```{r fig.id="condeff-mean", fig.lp="suppfig", fig.cap="Conditional effect plots for mean verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
-S"}
+S", height=213.3, width=160}
 include_graphics("figures/supp_figs/cond_effects_meanvert.png", error = F)
@@
 
+\newpage
+
 ```{r fig.id="condeff-ses", fig.lp="suppfig", fig.cap="Conditional effect plots for SES verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
-S"}
+S", height=213.3, width=160}
 include_graphics("figures/supp_figs/cond_effects_sesvert.png", error = F)
 ```
+
+\newpage
 
@@
 ```
+
+\newpage
+
+```{r fig.id="resid-plt", fig.lp="suppfig", fig.cap.pre="Fig.
+S", fig.cap="A smoothed scatter plot of quantile residuals as a function of SES or mean verticality, plotted using the DHARMa R package.
+Red line indicates a spline regression and red dots indicate outliers.", fig.height=213.3, fig.width=160}
+
+include_graphics("figures/supp_figs/resid_plots.png", error = F)
+
+```
+
+\newpage
 
@@
 
-#```{r fig.id="vertses-hist-biome", fig.lp="ed", fig.cap="Histograms of verticality by biome for each vertebrate class. Red dotted line indicates the mean #verticality for each biome/class."}
-## plot SES vert by biome
-#("figures/supp_figs/vert_biome_forestOnly_forestSES.R")
-#plt_vertses_biome
-
-#```
-
-#```{r fig.id="vert-forest", fig.lp="ed", fig.cap="Mean verticality, standard effect size (SES) of mean verticality, and the average proportion of species #exhibiting arboreal, terrestrial, and fossorial strategies across latitudinal bands.
-In maps of mean veticality, assemblages composed of only fossorial species #are zero (red) and those composed of only arboreal species are quantified as one (purple).
-In maps of the SES of mean verticality, negative numbers (red) #indicate assemblages have lower verticality than expected by random chance and positive numbers (purple) indicate assemblages have higher verticality than #expected by random chance.
-Line graphs show the average proportion of species exhibiting arboreal (blue), terrestrial (yellow), and fossorial (red) strategies.
-#Proportions for each assemblage can exceed one because species can exhibit multiple vertical strategies.}
-
-#```
-
-#```{r fig.id="forest-model-coefs"}
-
-#```
-
 # References
-
-# Acknowledgements

### 2024-09-05 verticality_future_mkd.Rmd
intro revisions

@@
 
-Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
-At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall, multi-layered forests increase both habitat volume and niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-By reducing competition, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023].
-However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
-Examining the impacts of environmental drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.
+In forest ecosystems, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy, i.e., vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
+Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+This vertical stacking of species has become an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
+As such, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023].
+However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at poleward latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
 
-At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests.
-Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-At high latitudes, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
-In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient and contribute to the high diversity of arboreal species found in these habitats [@oliveiraVerticalStratificationInfluences2019].
-We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.
+Tall forests contain steep vertical climate gradients characterized by lower thermal variability and greater moisture toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
+At temperate and boreal latitudes, low temperature and moisture in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
+In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient [@oliveiraVerticalStratificationInfluences2019].
+We therefore expect a globally varying relationship between canopy height and verticality pending baseline regional conditions from positive in tropical to neutral in temperate and boreal systems.
 
-Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019].
-Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019].
-Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage.
-In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak.
+Spatial variation may also occur in the relationship between verticality and species richness [@oliveiraVerticalStratificationInfluences2019].
+Though prior studies have shown positive relationships between verticality and richness for select taxa, the strength of this relationship is variable and generally more positive in tropical regions where climatic conditions pose fewer physiological constraints on arboreal strategies [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019].
+Niche expansion, characterized by vertical partitioning of the foerst, may allow this pattern to occur.
+In contrast, the relationship between verticality and richness may be weak where taller canopies do not facilitate higher verticality.
 
@@
 In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
-These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
-At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
+These scenarios would reduce assemblage verticality in the lowlands, akin to studies that show shifts in ecological processes such as climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
+At poleward latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
 
@@
 
-Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, its relationship with species richness, and its vulnerability to climate change have not been comprehensively described for vertebrates at a global scale.
-To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes.
-We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness.
+Despite decatdes of ecological inquiry into the relationship between forest height and species richness, little is known as to whether tall forests ubiquitously support niche filling and thus promote greater species coexistence per unit of area (high alpha or gamma diversity) To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes.
+We then examine whether the dynamic of filling vertical niche space and its effects of species richness is informed by the relationship between verticality, canopy height, and climate.
 Finally, we predict how climate change may impact global patterns of verticality.

### 2025-01-22 verticality_future_mkd.Rmd
update manuscript

@@
 
+# Abstract
+_Aim_: Vertical stratification of species from the ground to canopy is posited as a mechanistic link explaining positive relationships between canopy height and species richness.
+A key question is understanding how climate and vegetation impact verticality and spatial variation in verticality-diversity patterns.
+We examine these relationships and then project models of verticality under future climate conditions to examine how climate change may impact vertical community structure.
+
+_Location_: Global
+
+_Time period_: 1981-2010 to 2071-2100
+
+_Major taxa studied_: Vertebrates
+
+_Methods_: We modeled impacts of climate and vegetation on vertebrate verticality globally using spatial generalized linear mixed models with a spatially varying coefficient for canopy height.
+We projected models to future climate conditions to examine impacts of climate change on vertical community structure.
+We also examined global and biome-specific relationships between verticality and species richness.
+
+_Results_: Verticality patterns are globally consistent across vertebrate taxa, declining from low to high latitudes.
+Low precipitation and temperature constrain verticality outside moist tropical regions.
+Globally, positive relationships occurred between canopy height and verticality and between verticality and richness (except for reptiles).
+Regionally, canopy height exhibited largely neutral impacts on verticality, while verticality-richness relationships varied from negative or weakly positive in temperate and boreal biomes to generally positive in tropical biomes.
+Model projections indicated relatively small changes in verticality under future climate conditions.
+
+_Main conclusions_: Physiological constraints to arboreal lifestyles limit verticality in cold and dry macroclimate conditions.
+Globally, vertical stratification across increasingly tall forests (niche expansion) contributes to high richness in tropical regions.
+Regionally, climatic constraints limit positive verticality-richness relationships in higher latitudes, while niche packing largely contributes to positive verticality-richness relationships in tropical regions.
+Small changes to vertical community structure at a global scale could be caused by arboreal species increasing their use of understory habitat, becoming extirpated, or shifting geographic ranges to higher latitudes.
+
 # Introduction
 
-In forest ecosystems, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy, i.e., vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-This vertical stacking of species has become an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
-As such, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023].
-However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at poleward latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
+In forest ecosystems, species assemblages stratify vertically along resource and climate gradients that extend from beneath the ground to the top of the canopy, a phenomenon called vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
+Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments (Fig.
+\@ref(fig:concept-fig)a;[@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]).
+This vertical stacking of species has been described as a mechanistic link between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023], and is thought to contribute to the exceptionally high biodiversity in the tropics via niche expansion processes [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
+Thus, as the amount of habitat space increases with forest height, species may be added to the assemblage, increasing richness relative to shorter forests (Fig.
+\@ref(fig:concept-fig); @pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016).
+However, canopy height does not consistently impact species richness at regional to global scales, with weaker relationships observed at high latitudes where verticality is lower [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
+Whether verticality plays a central role in driving richness gradients globally remains an open question, and surprisingly little progress has been made in assessing whether this relationship holds across multiple taxonomic groups with differing sensitivities to climate variability.
 
-Tall forests contain steep vertical climate gradients characterized by lower thermal variability and greater moisture toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-At temperate and boreal latitudes, low temperature and moisture in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
-In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient [@oliveiraVerticalStratificationInfluences2019].
-We therefore expect a globally varying relationship between canopy height and verticality pending baseline regional conditions from positive in tropical to neutral in temperate and boreal systems.
+Tall forests exhibit steep vertical climate gradients, characterized by high maximum temperatures, low minimum temperatures, and reduced moisture availability in the canopy and lower thermal variability and higher moisture availability toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
+In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to stratify across the entire vertical habitat gradient, partitioning niche space including nesting and food resources [@oliveiraVerticalStratificationInfluences2019].
+Taller canopies in these regions could thus lead to higher verticality and richness (Fig.
+\@ref(fig:concept-fig)).
+In contrast, at temperate and boreal latitudes, macroclimate conditions, including low temperature and precipitation, may cause canopy microclimates to exceed physiological thresholds for many species, often constraining them to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
+As a result, taller canopies in these regions are not necessarily associated with higher verticality and richness as harsh macroclimate conditions limit the availability of vertical habitat space (Fig.
+\@ref(fig:concept-fig)).
+We therefore expect a globally varying relationship between canopy height and verticality as well as between verticality and species richness, pending baseline regional conditions from positive in tropical to neutral in temperate and boreal ecosystems.
 
-Spatial variation may also occur in the relationship between verticality and species richness [@oliveiraVerticalStratificationInfluences2019].
-Though prior studies have shown positive relationships between verticality and richness for select taxa, the strength of this relationship is variable and generally more positive in tropical regions where climatic conditions pose fewer physiological constraints on arboreal strategies [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019].
-Niche expansion, characterized by vertical partitioning of the foerst, may allow this pattern to occur.
-In contrast, the relationship between verticality and richness may be weak where taller canopies do not facilitate higher verticality.
+Since verticality is influenced by both climate and vegetation structure, changing climatic conditions may alter constraints on arboreal life strategies, potentially reshaping assemblage verticality and its association with species richness.
+In tropical lowlands, novel high temperatures [@trewNovelTemperaturesAre2024] may impose physiological stress on arboreal animals, leading them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
+These scenarios would reduce assemblage verticality in the lowlands, mirroring documented climate-driven shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
+In contrast, at higher latitudes, warming minimum temperatures may alleviate climatic constraints on arboreal species, promoting range expansions that could increase assemblage verticality.
+These contrasting dynamics underscore the need for further investigation into how climate-driven changes influence the vertical structure of assemblages and its impacts on biodiversity gradients.
 
-As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness.
-In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
-These scenarios would reduce assemblage verticality in the lowlands, akin to studies that show shifts in ecological processes such as climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
-At poleward latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.
+Despite decades of ecological inquiry into the relationships between verticality and species richness, global patterns of verticality, the environmental factors driving verticality, and the mechanistic link between verticality and species richness have not been comprehensively described for terrestrial vertebrates.
+To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes, as well as relationships between verticality and species richness.
+We ask 1) how do climate and vegetation structure impact verticality and does the impact of canopy height vary spatially, 2) how do relationships between verticality and richness vary across biomes, and 3) how might climate change impact global patterns of verticality?
 
@@
 ```
-
-Despite decatdes of ecological inquiry into the relationship between forest height and species richness, little is known as to whether tall forests ubiquitously support niche filling and thus promote greater species coexistence per unit of area (high alpha or gamma diversity) To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes.
-We then examine whether the dynamic of filling vertical niche space and its effects of species richness is informed by the relationship between verticality, canopy height, and climate.
-Finally, we predict how climate change may impact global patterns of verticality.
 
@@
 
-We mapped verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] using 50 km grid cells and a cylindrical equal area projection (Supplementary methods).
-Grid cells with fewer than five species were excluded.
-Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
-For birds, we used breeding plus resident ranges.
+We mapped global verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells and using a cylindrical equal area projection (Supplementary Methods).
+Grid cells with fewer than five species were excluded, because they may reduce the performance of null models.
+Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022; using breeding plus resident ranges], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
+
 To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell.
 For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024].
-For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014].
+For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale vertical classification than the dataset from Moura et al.
+[-@mouraPhylogenyinformedCharacterisationGlobal2024].
 We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1.
@@
 
-To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality ($SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool).
-We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.
+To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool.
+We defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness.
+Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.
 
@@
 
-We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
+We obtained present and future global climate data from CHELSA v2.1 (30 arc-second resolution) for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
 Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100.
 Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density).
-Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
+Additionally, we included historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
 All environmental data were resampled to the resolution and projection of the species data.
@@
 We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity.
-To linearize the relationships between predictor and response variables, we log-transformed climate velocity and precipitation of the driest month.
+To linearize the relationships between predictor and response variables, we log10-transformed climate velocity and precipitation of the driest month.
 We then scaled all predictor variables to mean zero and variance one.
-To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than 5 [@wartonEcostatsDataAnalysis2022].
+To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than five [@wartonEcostatsDataAnalysis2022].
 We chose variables to remove based on a combination of which had the highest VIF and which were most likely to directly impact physiological limits to arboreality.
@@
 We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of environmental variables on mean and SES verticality.
-Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields.
-[@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015].
-sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent, and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].
+Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015].
+sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].
 
@@
 $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients.
-$α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality drive by global evolutionary patterns.
+$α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality driven by global evolutionary patterns.
 $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept.
 $𝜲^{svc}_sζ_s$ represents a spatially varying coefficient, where $𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$).
+
 Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
@@
 
-For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structures [@zuurMixedEffectsModels2009].
+For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structure [@zuurMixedEffectsModels2009].
 The base model only included a spatial random field.
-Subsequent models additionally included biogeographic realm [sensu @dinersteinEcoregionBasedApproachProtecting2017] as a random intercept, a spatially varying coefficient for canopy height, and the combination of both.
-Including realm as a random intercept accounts for evolutionary differences among biogeographic realms that could lead to differences in verticality.
-The spatially varying coefficient allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.
+Subsequent models additionally included biogeographic realm (sensu @dinersteinEcoregionBasedApproachProtecting2017) as a random intercept, a spatially varying coefficient for canopy height, and the combination of both.
+The SVC allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.
+
 We compared models using AIC and cross-validation.
-AIC is frequently used to identify the most parsimonious model [e.g., @barnettImprovingEstimatesSpecies2021]; however, the metric is not completely appropriate for hierarchical models (see Commander et al.
-2022 `r word_comment(comment = "check citation", highlight = "Commander et al.
-2022")`).
-Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability (*again check citation Commander et al.
-2022*).
+AIC is frequently used to identify the most parsimonious model (e.g., @barnettImprovingEstimatesSpecies2021); however, the metric is not completely appropriate for hierarchical models [@commanderShadowModelHow2022].
+Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability [@commanderShadowModelHow2022].
 We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing.
@@
 Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package.
-Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabe S\@ref(tab:mod-compare-tab)).
+Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabel S\@ref(tab:mod-compare-tab)).
 Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.
@@
 
-## Relationship between verticality and richness
+## Relationships between verticality and richness
 
-To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points and only points where canopy height had a positive significant effect on verticality.
+We fit linear models of richness as a function of SES or mean verticality for all assemblages as well as for assemblages in each biome [@dinersteinEcoregionBasedApproachProtecting2017].
+Biomes with fewer than 30 assemblages were not included.
 
@@
 We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences.
-To standardize measures of change across taxa and between mean and SES verticality, we calculated the percent relative change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
+To standardize measures of change across taxa and between mean and SES verticality, we calculated the relative percent change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
 Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
@@
 
-Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (Fig.
+The standard effect size (SES) of verticality and mean verticality exhibited latitudinal gradients across taxa characterized by high verticality in the tropics and lower verticality toward poleward latitudes (Fig.
 \@ref(fig:vert-map-global)).
-This pattern was largely driven by declines in the proportion of arboreal species toward the poles.
-Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical moist forests (Fig.
+These patterns were largely driven by declines in the proportion of arboreal species toward the poles (Fig.
+\@ref(fig:vert-map-global)).
+
+Within tropical and subtropical biomes, verticality across taxa was highest in moist forests and lowest in dry forests (Fig.
 S\@ref(fig:vert-biome)).
-Within tropical and subtropical biomes, verticality was highest in moist forests and lowest in dry forests (Fig.
-\@ref(fig:vert-biome)).
-In moist forests, verticality of bird and mammal assemblages was consistently high, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies.
-In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Malaysian and Indonesian forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar.
+In moist forests, verticality of bird and mammal assemblages was consistently high globally, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies.
+In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Indo-Malay forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar (Fig.
+\@ref(fig:vert-map-global)).
 
-In temperate biomes, verticality was similar across forest types but declined in less dense wooded habitats (e.g., savannas and shrublands) (Fig.
-S \@ref(fig:vert-biome)).
-Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes.
+In temperate biomes, verticality was lower in less-dense wooded habitats (e.g., savannas and shrublands) (Fig.
+S\@ref(fig:vert-biome)).
+Additionally, the proportion of species exhibiting fossorial strategies largely influenced differences in verticality among vertebrate classes in temperate regions.
 Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (Fig.
-\@ref(fig:vert-map-global)L), though temperate rainforests (e.g., the Pacific-Northwest of North America) were a notable exception.
-Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (Fig.
-\@ref(fig:vert-map-global)F).
-In contrast, the proportion of fossorial reptiles declined in temperate regions.
-Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species.
-Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal verticality gradient.
+\@ref(fig:vert-map-global)L), followed by mammals (Fig.
+\@ref(fig:vert-map-global)F), although amphibians in temperate rainforests (e.g., the Pacific-Northwest of North America) had relatively high verticality.
+In contrast, the proportion of fossorial reptiles declined in temperate regions, with low verticality driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species.
+Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal verticality gradient (Fig.
+\@ref(fig:vert-map-global)).
 
@@
 
-Spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (Fig.
+Spatial GLMMs indicated that mean and SES verticality were limited by dry and wet season precipitation, while impacts of maximum temperatures, minimum temperatures, and canopy height varied across taxa (Fig.
 \@ref(fig:mod-coefs)).
-Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (Fig.
+Specifically, the minimum temperature of the coldest month positively impacted mean and SES verticality of mammals and reptiles, but did not significantly impact birds and amphibians.
+Maximum temperatures of the warmest month exhibited a quadratic relationship with mean and SES verticality of birds, reptiles, and amphibians, while exhibiting a positive linear relationship with mammal verticality.
+The negative effects of high and low maximum temperatures were most pronounced under dry conditions (Fig.
+S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)).
+These patterns led to lower mean and SES verticality at both high and low maximum temperatures in regions with low dry season precipitation for reptiles and amphibians, while verticality increased with maximum temperature across varying levels of dry season precipitation for birds and mammals (Fig.
+S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)).
+
+Canopy height had globally significant positive impacts on SES verticality of reptiles and amphibians and mean verticality of amphibians, and trended toward a positive impact for mammals, birds, and mean verticality of reptiles.
+Vegetation density had no significant impact on mean or SES verticality for any taxon (Fig.
 \@ref(fig:mod-coefs)).
-Minimum temperature of the coldest month positively impacted mammal and reptile verticality and did not significantly impact bird and amphibian verticality.
-Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammal verticality.
-These patterns resulted in lower verticality at both high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (Fig.
-S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)).
-Canopy height had globally positive impacts on SES verticality of reptiles and amphibians and mean verticality of amphibians, and vegetation density had no significant impact on mean or SES verticality of any taxon Fig.
-(\@ref(fig:mod-coefs)).
-However, impacts of canopy height varied spatially, exhibiting neutral impacts on verticality across much of the tropics and positive impacts in regions in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (Fig.
-\@ref(fig:svc), Fig.
-S\@ref(fig:svc-sig) ).
-Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals (Fig.
-S\@ref(fig:resid-plt)).
+Notably, the impact of canopy height varied spatially.
+Despite exhibiting neutral impacts on verticality across most regions globally, canopy height positively impacted verticality in tropical regions with greater topographic relief, such as the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (Fig.
+\@ref(fig:svc)).
 
@@
 
-When considering all points, Pearson correlations indicated strong positive relationships between SES or mean verticality and species richness for birds and mammals, a weak positive relationship for amphibians, and nearly no relationship for reptiles (Fig.
-\@ref(fig:corplot)).
-When we restricted the correlation to only include points where canopy height had a significant positive effect on verticality, the strength of the relationship between SES verticality and richness increased for amphibians and SES verticality of birds, but declined for mammals, reptiles, and mean verticality of birds.
+Linear models indicated globally positive relationships between richness and mean or SES verticality across all taxa (Fig.
+\@ref(fig:richness-mods)).
+However, R^2^ values indicated generally weaker global relationships for reptiles and amphibians than for birds and mammals.
+When restricted to individual biomes, boreal and temperate regions generally exhibited negative or weak positive slopes, while tropical regions generally exhibited stronger positive slopes (Fig.
+\@ref(fig:richness-mods), S\@ref(fig:cor-biome)).
+However, reptiles deviated from this pattern, exhibiting weak positive or negative slopes between richness and verticality in the tropics (Fig.
+\@ref(fig:richness-mods), S\@ref(fig:cor-biome)).
 
@@
 \@ref(fig:vert-dif-map)).
-Amphibians exhibited the highest predicted relative percent change, with predicted declines reaching up to `r amphses.minreldif*-1`% for SES verticality and `r amphmean.minreldif*-1`% for mean verticality in parts of the tropics and predicted increases reaching up to `r amphses.maxreldif`% for SES verticality and `r amphmean.maxreldif`% for mean verticality in the African forest-savanna matrix.
-Across biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist forests and the highest relative increases at high latitudes in boreal forests.
-Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (Fig.
+However, models predicted that tropical regions will experience small declines in verticality across all taxonomic groups.
+Amphibians exhibited the highest predicted relative percent change, with predicted declines reaching 5% for SES and mean verticality in parts of the tropics, and predicted increases reaching up to 4.4% for SES verticality and 5.1% for mean verticality in the African forest-savanna matrix.
+For all other taxa, changes in verticality are expected to vary less than 1.5%.
+Within biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist forests and the highest relative increases at high latitudes in boreal forests.
+Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%)) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (Fig.
 S\@ref(fig:vert-dif-biome)).
@@
 
-Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studies that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019].
-Though climate conditions influenced verticality, they were neither able to predict extremely high or low verticality nor spatial variation in the influence of canopy height on verticality.
-Furthermore, regions exhibiting positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between canopy height and species richness.
-We propose that the rate of niche packing across evolutionary time spans rather than expansion across tall forests may explain both the limited capacity for climate alone to predict high levels of verticality and the lack of the three-way positive relationship between canopy height, verticality, and species richness for some taxa.
-Given the potentially large role of evolutionary patterns in global distributions of verticality, our models may underpredict the impacts of climate change on vertical assemblage structure.
+Understanding global patterns and drivers of verticality is key to understanding how verticality mediates geographic patterns of species richness.
+Our results reveal globally consistent macroscale patterns of verticality across taxa in wooded biomes, with epicenters of verticality in the wet tropics and declining verticality toward the poles (Fig.
+\@ref(fig:vert-map-global)).
 
-## Environmental drivers of verticality
+## Climatic drivers of verticality
 
-Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
-Water availability therefore facilitates verticality, while temperatures at both hot and cold extremes and water scarcity generally limit verticality.
-For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist forests in tropical regions.
-However, our models indicated nuances in the relative influences of different drivers across taxa.
+Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats in regions where macroclimate conditions are also more extreme [@oliveiraVerticalStratificationInfluences2019].
+Consistently high precipitation facilitates verticality, while temperatures at both hot and cold extremes and low or highly seasonal precipitation generally limit verticality.
+For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist tropical forests.
+However, our models indicated nuanced influences of drivers across taxa.
 
-Impacts of precipitation were particularly pronounced for amphibian verticality.
-In tropical wet forests, consistent water availability allows amphibians to maintain body temperatures below critical thermal maxima via evapotranspiration [@sundayThermalsafetyMarginsNecessity2014].
-However, in regions that experience pronounced dry seasons, amphibians may behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020].
-Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
+Precipitation strongly influenced amphibian verticality, while a combination of precipitation and maximum temperatures influenced bird verticality.
+In tropical wet forests, consistent water availability from precipitation and/or fog allows amphibians and birds to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014].
+In tropical regions with pronounced dry seasons, amphibians and birds may increase their use of understory habitats to avoid desiccation, reducing assemblage verticality [@bashamVerticalStratificationCollapses2020; @rajaonariveloSeasonalVariationVertical2021].
+In many temperate or boreal biomes, low precipitation similarly constrains amphibian arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].
 
-For mammals and reptiles, our models indicated that minimum cold temperatures had the greatest impact on verticality.
-In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014].
-Species are therefore largely constrained to terrestrial or fossorial habitats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019].
-Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors influence global patterns of verticality.
+For mammals and reptiles, our models indicated that cold temperatures had the greatest impact on verticality.
+In boreal forests, cold and windy conditions in the canopy can cause body temperatures to drop below critical thermal minima in ectotherms and pose large energetic costs to endotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014].
+Limited food availability in the canopy may exacerbate high energy demands for arboreal strategies in these ecosystems.
+Species are therefore largely constrained to thermally buffered terrestrial or fossorial microhabitats [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014].
 
-## Spatially varying relationships between canopy height, verticality, and richness
+## Canopy height, verticality, and richness
 
-Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015].
-These results correspond to Basham et al.
-[-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics.
-Interestingly, previously reported spatial variation in the impact of canopy height on taxonomic and functional richness mirror, to some extent, spatial variation in the impacts of canopy height on verticality [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018].
-For example, canopy height positively impacts verticality and species richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea [@rollLinkingVertebrateSpecies2015].
-Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015].
-Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (Fig.
-S\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
-Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality.
-Yet, our results did not consistently support this hypothesis.
+At a global scale, our models indicated positive relationships between both canopy height and verticality and canopy height and species richness (barring reptiles; Figs.
+\@ref(fig:mod-coefs), \@ref(fig:richness-mods).
+The co-occurrence of these patterns suggests that verticality may mediate positive relationships between canopy height and richness via niche expansion, which have been observed for several taxonomic groups, including birds, mammals, and amphibians [@fengForestCanopyHeight2020; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015].
+Therefore, from high latitude boreal forests to the equatorial tropics, increasing tree height [@langHighresolutionCanopyHeight2023] combined with macroclimate conditions that facilitate arboreality allow the addition of arboreal species to assemblages, increasing both verticality and species richness (Fig.
+\@ref(fig:concept-fig)) [@oliveiraVerticalStratificationInfluences2019].
 
-A niche expansion model of community assembly characterizes the hypothesis that taller canopies with greater foliage height diversity allow greater vertical stratification and subsequent increases in verticality and species richness [@macarthurBirdSpeciesDiversity1961; @macarthurPatternsSpeciesDiversity1965].
-The niche expansion model posits that species richness increases when additional species occupy novel habitat or resource space [@pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016].
-Thus, arboreal species may be added to the assemblage by occupying the novel habitat space that becomes available as canopy height increases, producing a stronger positive relationship between verticality and richness in regions where canopy height has a positive influence on verticality.
-This pattern was evident for birds and amphibians.
-However, this three-way positive relationship did not occur in the most diverse regions, including much of the tropics.
-Instead, it was largely restricted to areas where relatively rapid increases in vegetation structure from high to low elevations or dry to wet forests (e.g., Madagascar) may facilitate the addition of species outside the habitat space occupied by species with lower richness in less productive habitats [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018].
-Yet in regions where canopy height did not have significant positive impacts on verticality, positive correlations between verticality and richness remained for birds and amphibians and increased in strength for mammals and reptiles, suggesting vertical stratification (i.e., niche expansion) may not be the dominant mechanism causing positive relationships between canopy height, verticality, and species richness at large spatial scales.
+At a regional scale, positive relationships between both verticality and richness and verticality and canopy height were limited to small regions within the tropics where relatively rapid increases in vegetation height from high to low elevations (e.g., the Andes) or dry to wet forests (e.g., Madagascar) may facilitate the addition of arboreal species to novel habitat space [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018].
+These regions generally correspond to previously reported spatially varying impacts of canopy height on taxonomic and functional richness [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018].
+For example, canopy height positively affects verticality and richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea, birds in the Himalayas, New Guinea, and southeastern Australia, and mammals in northwestern Mexico and eastern Madagascar, while negative effects occur for mammals in parts of the Amazon basin (Fig.
+\@ref(fig:svc)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018].
+The co-occurrence of these patterns further supports the roll of verticality in mediating relationships between canopy height and richness in select regions.
 
-Rather than expansion across tall forests, positive associations between verticality and species richness could be driven by niche packing.
-Niche packing occurs when smaller niche spaces or greater niche overlap allow more species to co-exist within a habitat [@pigotFunctionalTraitsReveal2016; @pellissierNichePackingExpansion2018].
-Specialization on specific resources within the canopy over evolutionary time could thus increase arboreal species diversity in the absence of increasing canopy height [@fengForestCanopyHeight2020; @pellissierNichePackingExpansion2018].
-Furthermore, evolutionary rates of arboreal lineages are generally faster than those of terrestrial and fossorial lineages [@cyriacDiggingTheirOwn2018; @liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017].
-In the tropics where climate conditions support arboreality, the combination of niche packing and rapid evolutionary rates of arboreal species could thus lead to a disproportionately high number of arboreal species, increasing verticality relative to what would be expected by climate conditions alone and causing positive relationships between verticality and richness in regions where canopy height has insignificant or even negative effects on verticality.
+Yet, positive relationships between verticality and richness still occurred across most tropical biomes for all taxa except reptiles despite neutral relationships between canopy height and verticality.
+These results are consistent with Basham et al.
+[-@bashamVerticalStratificationPatterns2023a], who found non-significant impacts of canopy height on verticality in the tropics using meta-analysis.
+In these regions, verticality may facilitate increasing species richness without coincident increases in canopy height via niche packing, which occurs when smaller niche spaces or greater niche overlap allow more species to coexist in a habitat (Fig.
+\@ref(fig:concept-fig)); @pigotFunctionalTraitsReveal2016; @pellissierNichePackingExpansion2018).
+Recent analyses of tropical forest structure support this mechanism, with large swaths of tropical forest exhibiting limited vertical stratification of vegetation, which was identified as a single (rather than multiple) peaks in foliage density from the ground to the canopy [@doughtyTropicalForestsAre2023].
+In the absence of increasing canopy height, specialization on specific resources within the canopy over evolutionary time combined with generally faster diversification rates of arboreal lineages could increase arboreal species diversity, assemblage verticality, and total richness [@fengForestCanopyHeight2020; @cyriacDiggingTheirOwn2018; @liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017; @pellissierNichePackingExpansion2018].
+
+In temperate and boreal regions, where canopy height had a neutral impact on verticality (Fig.
+\@ref(fig:svc)) and verticality and richness exhibited negative or shallow positive slopes (Fig.
+\@ref(fig:richness-mods)), climatic constraints on arboreal lifestyles may prevent expansion across vertical forest strata or packing within the canopy [@oliveiraVerticalStratificationInfluences2019], thus limiting positive relationships between canopy height, verticality, and species richness [@larueDiversityVolumeRelationships2023].
+For taxa that exhibit fossorial strategies, negative correlations between verticality and richness could, in contrast, be caused by expansion vertically downward rather than upward.
+More nuanced data on the degree of arboreality (opposed to categorical vertical strategy data) would be required to verify patterns of niche expansion and packing.
 
@@
 
-Increasing temperature and declining precipitation could alter global patterns of assemblage verticality via vertical niche shifts, range shifts, or local extinction.
-Our models indicated relatively small declines of verticality in tropical latitudes in response to climate change at a global scale.
-In the absence of climatic buffering experienced in the understory, species living in arboreal habitats may be threatened by the increasing magnitude and frequency of extreme heat and drought [@ipccClimateChange20232023].
-The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small thermal safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014].
-To avoid unsuitable conditions that exceed physiological tolerances, species could behaviorally regulate climate exposure by spending more time in the understory, thus reducing assemblage verticality [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020].
-Indeed, vertical niche shifts have been observed at smaller spatiotemporal scales across tropical forests.
-For example, birds and arboreal frogs increase their use of understory resources during the dry season [@delgado-martinezDifferentialUtilizationSurface2023; @bashamVerticalStratificationCollapses2020].
-Arboreal primates from the Americas to Madagascar similarly increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022].
-Alternatively, arboreal species could become locally extinct if they are not able to adjust microhabitat use to avoid extreme climate conditions.
+Increasing temperature and altered precipitation patterns could have minor impacts on global patterns of assemblage verticality based on our models via vertical niche shifts, range shifts, or local extinction (Fig.
+\@ref(fig:vert-dif-map)).
+For example, arboreal species could spend more time in the understory to avoid unsuitable climate conditions in the canopy that may increase in frequency with the rising magnitude and frequency of extreme heat and drought events [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020; @ipccClimateChange20232023].
+Indeed, vertical niche shifts have been observed at smaller spatiotemporal scales across tropical forests and taxa (e.g.
+birds, @delgado-martinezDifferentialUtilizationSurface2023; frogs, @bashamVerticalStratificationCollapses2020; primates, @eppleyFactorsInfluencingTerrestriality2022).
+However, logging and deforestation may have more immediate effects by reducing the availability of arboreal habitats and increasing local temperatures, potentially generating more extreme shifts in verticality [@lindenmayerResponseArborealMarsupials2021; @nowakowskiChangingThermalLandscapes2018; @seniorPantropicalAnalysisImpacts2017].
 
 Increasing minimum cold temperatures could release constraints on arboreality in temperate and boreal regions for birds and mammals.
-Our models predicted the highest increases in verticality in boreal forests, which could occur, for example, as optimal latitudinal ranges of birds in low latitude assemblages with higher verticality shift poleward [@laughlinWinterRangeShifts2023; @martinsSignificantShiftsLatitudinal2024].
-Increases in verticality caused by range shifts will likely be restricted to extra-tropical assemblages, because the ranges of tropical species typically move across elevational rather than latitudinal gradients in response to climate change [@colwellStillLittleEvidence2024].
-Vertical niche shifts are less likely to contribute to increases in verticality due to morphological constraints on arboreal strategies [@camargoContrastingRealizedFundamental2016; @dealencarArborealityConstrainsMorphological2017].
+Verticality could subsequently increase as optimal latitudinal ranges of birds living in the upper canopy shift poleward [@laughlinWinterRangeShifts2023; @martinsSignificantShiftsLatitudinal2024].
+Range shift induced increases in verticality will likely be restricted to extra-tropical assemblages, because ranges of tropical species typically shift across elevations rather than latitudes [@colwellStillLittleEvidence2024].
+Vertical niche shifts are less likely to contribute to increased verticality, because morphology constrains arboreal strategies [@camargoContrastingRealizedFundamental2016; @dealencarArborealityConstrainsMorphological2017].
 Yet, use of arboreal habitats by terrestrial species has occasionally been observed in high latitude temperate climates [@petrovanWhyLinkDiverse2022].
 
-Despite the small changes in verticality predicted by our models under future climate scenarios, realized change, particularly declines in verticality, may be larger.
-First, underprediction of declines may arise because warming and drying climates could impact a disproportionate number of arboreal species in the tropics where climate alone failed to predict high verticality.
-Second, habitat loss caused by logging and deforestation may have more immediate effects on the vertical structure of species assemblages than climate change [@lindenmayerResponseArborealMarsupials2021; @murrayClimateLandUse2021].
-Understanding the extent to which increases or decreases in verticality are realized and attributable to climate change will require long-term monitoring of taxa and vertical microclimates across latitudes.
+## Conclusions
+
+We show that verticality exhibits globally consistent patterns across taxa and is constrained by minimum temperatures and low precipitation.
+Global positive impacts of canopy height on verticality suggest vertical stratification across increasingly tall forests contributes to high species richness in tropical regions.
+However, neutral regional impacts of canopy height suggest that climatic constraints limit relationships between verticality and richness in boreal and temperate regions, while niche packing contributes to positive relationships between verticality and richness in tropical regions.
+While predictions of verticality under future climate conditions indicate small changes in global patterns of verticality, understanding the extent to which increases or decreases in verticality are realized and attributable to climate change will require long-term monitoring of taxa and vertical microclimates across latitudes.
 
@@
 
+```{r fig.id="concept-fig", fig.width=160, fig.height=90, fig.cap.pre="Fig.
+", fig.cap="a) Niche expansion: Canopy height positively impacts verticality by allowing species to occupy novel habitat space as canopy height increases.
+As assemblage verticality increases, species richness also increases.
+Thus verticality mediates positive relationships that have been observed between canopy height and species richness.
+b) Niche packing: Canopy height has no impact on verticality or richness, yet verticality and richness are still positively correlated.
+This pattern could be caused by niche packing within arboreal habitats.", include=TRUE5}
+
+knitr::include_graphics("figures/main_figs/vert_maps.png", error = F)
+```
+
 ```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap.pre="Fig.
-", fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians.
+", fig.cap="Standard effect size (SES) and mean verticality of birds, mammals, reptiles, and amphibians.
 Line plots indicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
@@
 ```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap.pre="Fig.
-", fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
-Red points indicate coefficients with 95% CIs that do not overlap zero.
-tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}
+", fig.cap=" Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
+Red points indicate coefficients with 95% CIs that do not overlap zero."}
 
@@
 ```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap.pre="Fig.
-", fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians.
-Gray areas indicate areas that are not wooded or that have fewer than five species."}
+", fig.cap="Spatially varying effect of canopy height on the mean and standard effect size (SES) of verticality for birds, mammals, reptiles, and amphibians.
+The spatially varying effect is the sum of the main effect of canopy and height and the spatially varying random effect at any given point.
+Larger positive numbers indicate a stronger positive effect of canopy height on verticality and larger negative numbers indicate a stronger negative effect of canopy height on verticality.
+Dark gray areas indicate locations where the effect of canopy height was not significant.
+Light gray areas indicate areas that are not wooded or that have fewer than five species."}
 
@@
 
-```{r fig.id="corplot", fig.width = 160, fig.height=213.4, fig.cap.pre="Fig.
-", fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive.
-"}
+```{r fig.id="richness-mods", fig.width = 160, fig.height=115.6, fig.cap.pre="Fig.
+", fig.cap="Slopes for linear models of richness as a function of SES or mean verticality across all biomes and for individual biomes.
+Hollow circles indicate non-significant relationships (p ≥ 0.05) and filled circles indicate significant relationships.
+R^2^ values are represented by the size of the circle.
+Purple = all biomes, gray = boreal biomes, orange = temperate biomes, green = tropical biomes.
+Flooded GS = Flooded grassland and savannas, Mediterranean FWS = Mediterranean forests, woodlands, and scrub, Montane GS = Montane grasslands and shrublands, Temperate BMF = Temperate broadleaf and mixed forests, Temperate CF = Temperate conifer forests, Temperate GSS = Temperate grasslands, savannas, and shrublands, Tropical CF = Tropical and subtropical conifer forests, Tropical DBF = Tropical and subtropical dry broadleaf forests, Tropical GSS = Tropical and subtropical grasslands, savannas, and shrublands, Tropical MBF = Tropical and subtropical moist broadleaf forests"}
 
-include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
+include_graphics("figures/main_figs/vert_rich_lm_slopes.png", error = F)
 ```
@@
 
-include_graphics("figures/main_figs/vert_reldifs.png", error = F)
+include_graphics("figures/main_figs/fig5_vert_reldifs.png", error = F)
 ```
@@
 
-We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al.
-(2017).
+We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinerstein et al.
+[-@dinersteinEcoregionBasedApproachProtecting2017].
 The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
@@
 We first estimated the spatial range by visually inspecting spatial correlations based on a subset of 1000 points.
-To avoid over-smoothing the GRF while or over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024].
+To avoid over-smoothing the GRF while simultaneously not over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024].
 To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range.
-Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range.
-To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
+Additionally, we specified an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range.
+To avoid small triangles, we also specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
 We then fit a model using the resulting mesh.
 If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model.
-We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of spatial domain used to construct the mesh.
+We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of the spatial domain used to construct the mesh.
 We selected the mesh used in the final model for all following analysis.
@@
 The log-likelihood, AIC, and delta AIC are for models trained with all points.
-The average log-likelihood is the average sum of log-likelihood values for left out test data across five folds.
-NA values are caused by not meeting all model assumptions across all five cross-validation folds."}
+The average log-likelihood is the average sum of log-likelihood values for left out test data across five folds."}
 tab = read.csv("tables/supplement/model_selection_compare.csv")
@@
 
-```{r fig.id="condeff-mean", fig.lp="suppfig", fig.cap="Conditional effect plots for mean verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
+```{r fig.id="condeff-mean", fig.lp="suppfig", fig.cap="Conditional effect plots for mean verticality (indicated by blue-red color scale) in response to precipitation of the driest month and maximum temperature of the warmest month.
+The first column shows mean verticality across gradients of maximum temperature of the warmest month and log(dry season precipitation).
+The second column shows mean verticality across gradients of minimum temperature of the coldest month and log(dry season precipitation).", fig.cap.pre="Fig.
 S", height=213.3, width=160}
@@
 
-```{r fig.id="svc-sig", fig.lp="suppfig", fig.cap.pre="Fig.
-S", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}
-
-include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
-```
-
-\newpage
-
-```{r fig.id="resid-plt", fig.lp="suppfig", fig.cap.pre="Fig.
-S", fig.cap="A smoothed scatter plot of quantile residuals as a function of SES or mean verticality, plotted using the DHARMa R package.
-Red line indicates a spline regression and red dots indicate outliers.", fig.height=213.3, fig.width=160}
-
-include_graphics("figures/supp_figs/resid_plots.png", error = F)
-
+```{r fig.id="cor-biome", fig.lp="suppfig", fig.cap="Relationships between richness and mean or SES Verticality for birds, mammals, reptiles, and amphibians in each biome.
+Lines indicate slopes of linear models for richness as a function of mean or SES verticality.", fig.cap.pre="Fig.
+S", height=124.5, width=160}
+include_graphics("figures/supp_figs/vert_richness_correlations.png", error = F)
 ```

### 2025-01-28 verticality_future_mkd.Rmd
manuscript revisions

@@
 ---
-author: ""
-title: "Global patterns and drivers of vertebrate verticality"
 bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
+csl: global-ecology-and-biogeography.csl
 output: 
@@
     sep: ': '
+header-includes:
+- \usepackage{fancyhdr}
+- \pagestyle{fancy}
+- \rhead{Global Verticality}
+- \begin{document}
 ---
@@
 
-# Abstract
-_Aim_: Vertical stratification of species from the ground to canopy is posited as a mechanistic link explaining positive relationships between canopy height and species richness.
+**Global patterns and drivers of vertebrate verticality**
+
+Lydia G. Soifer^1\*^, David P. Edwards^2,3^, Brett R. Scheffers^4^
+
+1 School of Natural Resource and Environment, University of Florida, USA
+
+2 Department of Plant Sciences and Centre for Global Wood Security, University of Cambridge, Cambridge, UK
+
+3 Conservation Research Institue, University of Cambridge, Cambridge, UK
+
+4 Department of Wildlife Ecology and Conservation, University of Florida, USA
+
+Email addresses:
+
+^\*^ lysoifer@gmail.com (corresponding author)
+
+dpe29@cam.ac.uk
+
+brett.scheffers@ufl.edu
+
+**Funding Statement**
+
+LGS was funded by the National Science Foundation Graduate Research Fellowship.
+This material is based upon work supported by the National Science Foundation Graduate Research Fellowship under Grant No.
+DGE1842473.
+Any opinions, findings, and conclusions or recommendations expressed in this material are those of the authors and do not  necessarily reflect the views of the National Science Foundation.
+
+**Conflicts of Interest**
+
+The authors have no conflicts of interest.
+
+\newpage
+
+**Global patterns and drivers of vertebrate verticality**
+
+**Abstract**
+
+**Aim**: Vertical stratification of species from the ground to canopy is posited as a mechanistic link explaining positive relationships between canopy height and species richness.
 A key question is understanding how climate and vegetation impact verticality and spatial variation in verticality-diversity patterns.
@@
 
-_Location_: Global
+**Location**: Global
 
-_Time period_: 1981-2010 to 2071-2100
+**Time period**: 1981-2010 to 2071-2100
 
-_Major taxa studied_: Vertebrates
+**Major taxa studied**: Vertebrates
 
-_Methods_: We modeled impacts of climate and vegetation on vertebrate verticality globally using spatial generalized linear mixed models with a spatially varying coefficient for canopy height.
+**Methods**: We modeled impacts of climate and vegetation on vertebrate verticality globally using spatial generalized linear mixed models with a spatially varying coefficient for canopy height.
 We projected models to future climate conditions to examine impacts of climate change on vertical community structure.
@@
 
-_Results_: Verticality patterns are globally consistent across vertebrate taxa, declining from low to high latitudes.
+**Results**: Verticality patterns are globally consistent across vertebrate taxa, declining from low to high latitudes.
 Low precipitation and temperature constrain verticality outside moist tropical regions.
@@
 
-_Main conclusions_: Physiological constraints to arboreal lifestyles limit verticality in cold and dry macroclimate conditions.
+**Main conclusions**: Physiological constraints to arboreal lifestyles limit verticality in cold and dry macroclimate conditions.
 Globally, vertical stratification across increasingly tall forests (niche expansion) contributes to high richness in tropical regions.
@@
 
+\newpage
+
 # Introduction
 
-In forest ecosystems, species assemblages stratify vertically along resource and climate gradients that extend from beneath the ground to the top of the canopy, a phenomenon called vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023].
-Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments (Fig.
-\@ref(fig:concept-fig)a;[@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]).
+In forest ecosystems, species assemblages stratify vertically along resource and climate gradients that extend from beneath the ground to the top of the canopy, a phenomenon called vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
+Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments [Fig.
+\@ref(fig:concept-fig)a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
 This vertical stacking of species has been described as a mechanistic link between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023], and is thought to contribute to the exceptionally high biodiversity in the tropics via niche expansion processes [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a].
-Thus, as the amount of habitat space increases with forest height, species may be added to the assemblage, increasing richness relative to shorter forests (Fig.
-\@ref(fig:concept-fig); @pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016).
+Thus, as the amount of habitat space increases with forest height, species may be added to the assemblage, increasing richness relative to shorter forests [Fig.
+\@ref(fig:concept-fig); @pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016].
 However, canopy height does not consistently impact species richness at regional to global scales, with weaker relationships observed at high latitudes where verticality is lower [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].
-Whether verticality plays a central role in driving richness gradients globally remains an open question, and surprisingly little progress has been made in assessing whether this relationship holds across multiple taxonomic groups with differing sensitivities to climate variability.
+Whether animal verticality plays a central role in driving richness gradients globally remains an open question, and surprisingly little progress has been made in assessing whether this relationship holds across multiple taxonomic groups with differing sensitivities to climate variability.
 
 Tall forests exhibit steep vertical climate gradients, characterized by high maximum temperatures, low minimum temperatures, and reduced moisture availability in the canopy and lower thermal variability and higher moisture availability toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019].
-In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to stratify across the entire vertical habitat gradient, partitioning niche space including nesting and food resources [@oliveiraVerticalStratificationInfluences2019].
-Taller canopies in these regions could thus lead to higher verticality and richness (Fig.
+In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to stratify across the entire vertical habitat gradient, partitioning niche space for nesting and food resources [@oliveiraVerticalStratificationInfluences2019].
+Taller canopies in tropical regions could thus support higher animal verticality and richness (Fig.
 \@ref(fig:concept-fig)).
 In contrast, at temperate and boreal latitudes, macroclimate conditions, including low temperature and precipitation, may cause canopy microclimates to exceed physiological thresholds for many species, often constraining them to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021].
-As a result, taller canopies in these regions are not necessarily associated with higher verticality and richness as harsh macroclimate conditions limit the availability of vertical habitat space (Fig.
+As a result, taller canopies in these regions do not necessarily provide additional vertical habitat opportunities or support higher verticality and richness, as harsh macroclimate conditions limit the suitability of vertical habitat space (Fig.
 \@ref(fig:concept-fig)).
-We therefore expect a globally varying relationship between canopy height and verticality as well as between verticality and species richness, pending baseline regional conditions from positive in tropical to neutral in temperate and boreal ecosystems.
+We therefore expect a globally varying relationship between animal verticality and two key factors: canopy height and species richness.
+These relationships are likely to depend on baseline regional conditions, ranging from positive in tropical ecosystems to neutral in temperate and boreal regions.
 
-Since verticality is influenced by both climate and vegetation structure, changing climatic conditions may alter constraints on arboreal life strategies, potentially reshaping assemblage verticality and its association with species richness.
-In tropical lowlands, novel high temperatures [@trewNovelTemperaturesAre2024] may impose physiological stress on arboreal animals, leading them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
+Since animal verticality is influenced by both climate and vegetation structure, changing climatic conditions may alter constraints on arboreal life strategies, potentially reshaping assemblage verticality and its association with species richness.
+In tropical lowlands, novel high temperatures [@trewNovelTemperaturesAre2024] may impose physiological stress on arboreal animals, leading them to disperse toward higher elevations, lower vertical habitats, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019].
 These scenarios would reduce assemblage verticality in the lowlands, mirroring documented climate-driven shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024].
@@
 These contrasting dynamics underscore the need for further investigation into how climate-driven changes influence the vertical structure of assemblages and its impacts on biodiversity gradients.
-
-Despite decades of ecological inquiry into the relationships between verticality and species richness, global patterns of verticality, the environmental factors driving verticality, and the mechanistic link between verticality and species richness have not been comprehensively described for terrestrial vertebrates.
-To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes, as well as relationships between verticality and species richness.
-We ask 1) how do climate and vegetation structure impact verticality and does the impact of canopy height vary spatially, 2) how do relationships between verticality and richness vary across biomes, and 3) how might climate change impact global patterns of verticality?
 
@@
 ```
+
+Despite decades of ecological inquiry into the relationships between verticality and species richness, key aspects remain poorly understood for terrestrial vertebrates.
+These include global patterns of verticality, their environmental drivers, and the mechanisms linking verticality and species richness.
+To improve our understanding of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of nearly 24,000 vertebrate species with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes, as well as relationships between verticality and species richness.
+We ask 1) how do climate and vegetation structure impact verticality and does the impact of canopy height vary spatially, 2) how do relationships between verticality and richness vary across biomes, and 3) how might climate change impact global patterns of verticality?
 
@@
 
-We mapped global verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells and using a cylindrical equal area projection (Supplementary Methods).
+We mapped global verticality and richness of terrestrial vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells and using a cylindrical equal area projection (Supplementary Methods).
 Grid cells with fewer than five species were excluded, because they may reduce the performance of null models.
-Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022; using breeding plus resident ranges], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
+Range maps were obtained from the IUCN Red List for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022; using breeding plus resident ranges], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
+Overall, we obtained range maps for `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians.
 
-To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell.
+To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score from zero to one and then took the average verticality of all species whose ranges overlapped with the grid cell.
 For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024].
@@
 To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool.
-We defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness.
+For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness.
 Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.
@@
 
-We obtained present and future global climate data from CHELSA v2.1 (30 arc-second resolution) for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
-Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100.
-Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density).
-Additionally, we included historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
-All environmental data were resampled to the resolution and projection of the species data.
+We used 10 variables to represent variation in environmental conditions.
+Climate data were represented with six predictors: (i) mean daily maximum air temperature of the warmest month, (ii) mean daily minimum air temperature of the coldest month, (iii) temperature seasonality, (iv) precipitation of the wettest month, (v) precipitation of the driest month, and (vi)  precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
+All climatic variables were extracted from CHELSA v2.1 (30 arc-second resolution) for both current (1981-2010) and future (2071-2100) periods, the later represented by an ensemble of the five general circulation models in CHELSA CMIP6 ISIMIP3 data projected under the Shared Socioeconomic Pathway (SSP) 585.
+Vegetation data included (vii) canopy height [@langHighresolutionCanopyHeight2023], (viii) vegetation density [@crowtherMappingTreeDensity2015], and (ix) vegetation complexity (the product of canopy height and vegetation density).
+Additionally, we included (x) historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
+All environmental data were resampled to the spatial resolution and projection of the species data.
 
@@
 ### Variable selection
-We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity.
+We examined relationships between mean verticality and SES verticality and the ten environmental variables.
 To linearize the relationships between predictor and response variables, we log10-transformed climate velocity and precipitation of the driest month.
-We then scaled all predictor variables to mean zero and variance one.
-To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than five [@wartonEcostatsDataAnalysis2022].
+We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using the Variation Inflation Factor (VIF).
+We iteratively removed the predictor variables until all VIFs were less than five [@wartonEcostatsDataAnalysis2022].
 We chose variables to remove based on a combination of which had the highest VIF and which were most likely to directly impact physiological limits to arboreality.
@@
 
-Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
+Both the spatial random field and the spatially varying coefficients were modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
 Gaussian random fields were approximated for models of mean and SES verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019] (Supplementary methods).
@@
 
-For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structure [@zuurMixedEffectsModels2009].
+For each taxonomic group and verticality metric (i.e., mean or SES verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structure [@zuurMixedEffectsModels2009].
 The base model only included a spatial random field.
-Subsequent models additionally included biogeographic realm (sensu @dinersteinEcoregionBasedApproachProtecting2017) as a random intercept, a spatially varying coefficient for canopy height, and the combination of both.
-The SVC allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.
+Subsequent models additionally included biogeographic realm (sensu @dinersteinEcoregionBasedApproachProtecting2017) as a random intercept, a spatially varying coefficient (SVC) for canopy height, and the combination of both.
+The SVC allows the impact of canopy height on verticality to vary globally, enabling us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.
 
@@
 Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package.
-Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabel S\@ref(tab:mod-compare-tab)).
-Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.
+Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Table S\@ref(tab:mod-compare-tab)).
+Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model including an SVC for canopy height and a random effect of realm.
 
 We validated models first using the ‘sanity’ function from the sdmTMB package.
-We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space and to examine any patterns across rank transformed and raw model predictions.
+We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to assess residual normality and lack of patterning across space.
 
-We estimated uncertainty of SVCs for canopy height by adding draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations.
-We then calculated the median and 0.025 and 0.975 quantiles to obtain a 95% confidence interval for each SVC estimate.
+We estimated uncertainty of SVCs for canopy height by adding draws of the SVC random field to draws of the fixed effect parameter for canopy height across 300 simulations.
+From these draws, we calculated the median and a 95% confidence interval bounded by 2.5% and 97.5% quantiles.
 Confidence intervals that did not overlap zero were considered significant [@andersonSdmTMBPackageFast2024].
@@
 
-To examine potential impacts of climate change on assemblage verticality, we used the models to predict mean and SES verticality under a future climate scenario.
+To examine potential impacts of climate change on assemblage verticality, we used the models to project mean and SES verticality under a future climate scenario.
 We held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density.
-We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences.
-To standardize measures of change across taxa and between mean and SES verticality, we calculated the relative percent change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
+We calculated the verticality change by subtracting present-day predictions, based on current climate, from the future projections.
+These differences were then standardized to reflect the  relative change in mean and SES verticality:
+
+$relative change = (verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell.
 Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
@@
 
+\newpage
+
+# References
+
+\newpage
+
+**Data Availability Statement**
+
+All data used are publicly available. Code is available at https://github.com/lysoifer/verticality_future.
+
+\newpage
+
 # Figures
 
-```{r fig.id="concept-fig", fig.width=160, fig.height=90, fig.cap.pre="Fig.
+```{r fig.id="concept-fig", fig.width=160, fig.height=100.65, fig.cap.pre="Fig.
 ", fig.cap="a) Niche expansion: Canopy height positively impacts verticality by allowing species to occupy novel habitat space as canopy height increases.
@@
 b) Niche packing: Canopy height has no impact on verticality or richness, yet verticality and richness are still positively correlated.
-This pattern could be caused by niche packing within arboreal habitats.", include=TRUE5}
+This pattern could be caused by niche packing within arboreal habitats.", include=TRUE}
 
-knitr::include_graphics("figures/main_figs/vert_maps.png", error = F)
+knitr::include_graphics("figures/main_figs/fig1_concept_fig.png", error = F)
 ```
 
-```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap.pre="Fig.
+```{r fig.id="vert-map-global", fig.width=180, fig.height=140.02, fig.cap.pre="Fig.
 ", fig.cap="Standard effect size (SES) and mean verticality of birds, mammals, reptiles, and amphibians.
@@
 
-knitr::include_graphics("figures/main_figs/vert_maps.png", error = F)
+knitr::include_graphics("figures/main_figs/fig2_vert_maps.png", error = F)
 ```
@@
 
-include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
+include_graphics("figures/main_figs/fig3_sdmTMB_coefs.png", error = F)
 ```
@@
 
-```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap.pre="Fig.
+```{r fig.id="svc", fig.width=180, fig.height=120, fig.cap.pre="Fig.
 ", fig.cap="Spatially varying effect of canopy height on the mean and standard effect size (SES) of verticality for birds, mammals, reptiles, and amphibians.
@@
 
-include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
+include_graphics("figures/main_figs/fig4_canopy_height_coefs.png", error = F)
 ```
@@
 
-include_graphics("figures/main_figs/vert_rich_lm_slopes.png", error = F)
+include_graphics("figures/main_figs/fig5_vert_rich_lm_slopes.png", error = F)
 ```
@@
 
-include_graphics("figures/main_figs/fig5_vert_reldifs.png", error = F)
+include_graphics("figures/main_figs/fig6_vert_reldifs.png", error = F)
 ```
@@
 
-# estimated coefficients and confidence intervals from the best model for each taxa - include all model parameters: fixed effects, spatial paramters (matern range, marginal spatial standard, deviation), distribution parameters
-
 \newpage
@@
 
-```{r fig.id="condeff-ses", fig.lp="suppfig", fig.cap="Conditional effect plots for SES verticality in response to precipitation of the driest month and maximum temperature of the warmest month.", fig.cap.pre="Fig.
+```{r fig.id="condeff-ses", fig.lp="suppfig", fig.cap="Conditional effect plots for SES verticality (indicated by blue-red color scale) in response to precipitation of the driest month and maximum temperature of the warmest month.
+The first column shows SES verticality across gradients of maximum temperature of the warmest month and log(dry season precipitation).
+The second column shows SES verticality across gradients of minimum temperature of the coldest month and log(dry season precipitation).", fig.cap.pre="Fig.
 S", height=213.3, width=160}
@@
 ```{r fig.id="vert-dif-biome", fig.lp="suppfig", fig.cap.pre="Fig.
-S", fig.cap="Average percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
+S", fig.cap="Average relative percent difference between SES and mean verticality predicted under present and future climate conditions for global biomes.
 Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.
-Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}
+Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average invrease in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}
 
@@
 ```
-
-# References

### 2025-02-09 scripts/04_plots/fig1_vert_maps.R
update maps with dry forest

@@ -7,182 +7,174 @@ library(data.table)
 library(tidyterra)
 library(colorspace)
 library(grid)
+library(tricolore)
+library(lemon)
 
 
 # add proportion plots ----------------------------------------------------
-
-plt_vert_margin = function(v, df, plot.title, tags = list(c("A", "B", "C"))) {
-  # r: spatraster for plotting
-  # v: spatvector for outline
-  # plot.title = plot title
-  # margin.titl = y-axis margin title
-  
-  # get minmax values from vector
-  ymin = terra::ext(v)[3]
-  ymax = terra::ext(v)[4]
-  xmax = terra::ext(v)[2]
-  
-  # get color scale limits
-  # col.lims = round(minmax(r),2)
-  col.lims = c(round(min(df$vert.mean.ses),2), round(max(df$vert.mean.ses),2))
-  
-  r = df %>% rast(crs = "+proj=cea +datum=WGS84") %>% 
-    project("epsg:4326") 
-  
-    # create main plot
-  plot_main = ggplot() +
-    # Overlay world
-    geom_spatvector(data = wd, color = alpha("black", 0.7), fill = "gray70", linewidth = .1) +
-    geom_spatraster(data = r$vert.mean.ses, maxcell = Inf) +
-    geom_spatvector(data = wd, color = "black", fill = NA, linewidth = .1) +
-    scale_fill_continuous_divergingx("spectral", na.value = NA, limits = col.lims, rev = T) +
-    guides(fill = guide_colorbar(title = "")) +
-   # theme_classic() +
-    theme(legend.position = "inside",
-          legend.position.inside = c(0.6,0.45),
-          legend.justification = c(0,1),
-          legend.margin = margin(0,0,0.2,0.3),
-          #legend.key.height = unit(3, units = "mm"),
-          legend.key.height = unit(2, units = "mm"),
-          legend.background = element_blank(),
-          legend.key.width = unit(2, units = "mm"),
-          legend.box.margin = margin(0,0,0,0),
-          legend.box.spacing = unit(0,"mm"),
-          legend.box.background = element_blank(),
-          legend.text = element_text(size = 6),
-          plot.background = element_blank(),
-          plot.margin = unit(c(0,0,0,0), units = "mm"),
[…]

### 2025-02-09 verticality_future_mkd.Rmd
update fig 2

@@
 
-Lydia G. Soifer^1\*^, David P. Edwards^2,3^, Brett R. Scheffers^4^
+Lydia G. Soifer^1\*^, David P. Edwards^2,3^, Brunno F Oliveira^4,5^, Mario R Moura^6^, Brett R. Scheffers^4^
 
@@
 
-4 Department of Wildlife Ecology and Conservation, University of Florida, USA
+4 Centre de Synthèse et d’Analyse sur la Biodiversité, Montpellier, France
+
+5 Department of Wildlife Ecology and Conservation, University of Florida, USA
+
+6 Departamento de Sistemática e Ecologia, Universidade Federal da Paraíba, João Pessoa, Brazil
 
@@
 dpe29@cam.ac.uk
+
+brunno.oliveira@me.com
+
+mariormoura@gmail.com
 
@@
 
-Spatial GLMMs indicated that mean and SES verticality were limited by dry and wet season precipitation, while impacts of maximum temperatures, minimum temperatures, and canopy height varied across taxa (Fig.
+Spatial GLMMs indicated that mean and SES verticality were limited by dry and, to some extent, wet season precipitation, while impacts of maximum temperatures, minimum temperatures, and canopy height varied across taxa (Fig.
 \@ref(fig:mod-coefs)).
@@
 
-Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality and varied across taxa (Fig.
+Changes in verticality in response to projected climate change were small relative to the total variation in mean and SES verticality and varied across taxa (Fig.
 \@ref(fig:vert-dif-map)).
-However, models predicted that tropical regions will experience small declines in verticality across all taxonomic groups.
-Amphibians exhibited the highest predicted relative percent change, with predicted declines reaching 5% for SES and mean verticality in parts of the tropics, and predicted increases reaching up to 4.4% for SES verticality and 5.1% for mean verticality in the African forest-savanna matrix.
+However, our future projections show small declines in verticality across tropical regions for all taxonomic groups.
+Amphibians exhibited the highest predicted relative percent change, with the SES and mean of verticality expected to decrease by up to 5% in parts of the tropics, whereas a 4.4-5.1% increase is projected along the African forest-savanna matrix.
 For all other taxa, changes in verticality are expected to vary less than 1.5%.
-Within biomes, the models predicted the highest relative declines in verticality in tropical and subtropical moist forests and the highest relative increases at high latitudes in boreal forests.
-Within these biomes, the models predicted the greatest average relative decline for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%)) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (Fig.
+Within biomes, the models predicted the highest relative decreases in verticality in tropical and subtropical moist forests and the highest relative increases at high latitudes in boreal forests.
+Within these biomes, the models predicted the greatest average relative decrease for amphibians (SES verticality: `r round(sessumm[1, "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[1, "mean.reldif"], 2)`%)) and the greatest average relative increase for birds (SES verticality: `r round(sessumm[which(sessumm$class == "Birds" & sessumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%, mean verticality: `r round(meansumm[which(meansumm$class == "Birds" & meansumm$biome == "Boreal Forests/Taiga"), "mean.reldif"], 2)`%) (Fig.
 S\@ref(fig:vert-dif-biome)).
@@
 
-Understanding global patterns and drivers of verticality is key to understanding how verticality mediates geographic patterns of species richness.
-Our results reveal globally consistent macroscale patterns of verticality across taxa in wooded biomes, with epicenters of verticality in the wet tropics and declining verticality toward the poles (Fig.
+Understanding global patterns and drivers of verticality is key to understanding how vertical habitat use mediates species richness gradients.
+Our results reveal globally consistent patterns of animal verticality in wooded biomes, with epicenters of verticality in the wet tropics and declining verticality toward the poles (Fig.
 \@ref(fig:vert-map-global)).
@@
 Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats in regions where macroclimate conditions are also more extreme [@oliveiraVerticalStratificationInfluences2019].
-Consistently high precipitation facilitates verticality, while temperatures at both hot and cold extremes and low or highly seasonal precipitation generally limit verticality.
+Consistently high precipitation facilitates verticality, while temperatures at both hot and cold extremes and low precipitation generally limit verticality.
 For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist tropical forests.
@@
 
-At a global scale, our models indicated positive relationships between both canopy height and verticality and canopy height and species richness (barring reptiles; Figs.
+At a global scale, our findings indicate positive relationships between both canopy height and verticality and verticality and species richness (barring reptiles; Figs.
 \@ref(fig:mod-coefs), \@ref(fig:richness-mods).
@@
 
-```{r fig.id="vert-map-global", fig.width=180, fig.height=140.02, fig.cap.pre="Fig.
-", fig.cap="Standard effect size (SES) and mean verticality of birds, mammals, reptiles, and amphibians.
-Line plots indicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes.
+```{r fig.id="vert-map-global", fig.width=180, fig.height=175.52, fig.cap.pre="Fig.
+", fig.cap="a) Line plots and b) ternary maps showing the proportion of birds, mammals, reptiles, and amphibians exhibiting arboreal, terrestrial, and fossorial life strategies.
+In line plots, lines represent the average proportion for a given latitude ± standard deviation.
+The sum of pProportions may exceed one, because a single species can exhibit multiple strategies.
+c) The standard effect size (SES) of verticality in wooded biomes.
 Gray areas indicate areas that are not wooded or that have fewer than five species.", include=TRUE}
 
-knitr::include_graphics("figures/main_figs/fig2_vert_maps.png", error = F)
+knitr::include_graphics("figures/main_figs/fig2_vert_maps/fig2_vert_maps.png", error = F)
 ```
@@
 ", fig.cap="Spatially varying effect of canopy height on the mean and standard effect size (SES) of verticality for birds, mammals, reptiles, and amphibians.
-The spatially varying effect is the sum of the main effect of canopy and height and the spatially varying random effect at any given point.
+The spatially varying effect is the sum of the main effect of canopy height and the spatially varying random effect at any given point.
 Larger positive numbers indicate a stronger positive effect of canopy height on verticality and larger negative numbers indicate a stronger negative effect of canopy height on verticality.
@@
 The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
-For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7.
+For species where imputation was necessary, we considered a species to use a given vertical stratum if over 70% of the 10,000 imputations predicted use of the stratum.
 

### 2025-02-09 scripts/04_plots/fig2_vert_maps.R
add phylopic

@@ -178,3 +178,18 @@ grid.draw(pfinal)
 dev.off()
 
 # figure finished in inkscape
+
+library(rphylopic)
+squirrel = get_uuid("Sciurus vulgaris")
+squirrel = get_phylopic(squirrel)
+save_phylopic(squirrel, path = "figures/main_figs/fig2_vert_maps/squirrel.svg")
+
+lizard = pick_phylopic("Coleonyx fasciatus")
+save_phylopic(lizard, path = "figures/main_figs/fig2_vert_maps/lizard.svg")
+
+
+frog = pick_phylopic("Plectrohyla avia")
+save_phylopic(frog, path = "figures/main_figs/fig2_vert_maps/frog.svg")
+
+bird = pick_phylopic("Cuculidae")
+save_phylopic(bird, path = "figures/main_figs/fig2_vert_maps/bird.svg")

### 2025-02-09 scripts/04_plots/fig3_coef_plot.R
update fig3 coef plot

@@ -6,159 +6,91 @@ library(ggh4x)
 library(spatialreg)
 library(sdmTMB)
 
-# VERTMEAN ----------------------------------------------------------------
 
-# ADD variable importance weights to the graphs ##
+# SES Verticality Model Coefficients --------------------------------------
 
-load("results/sdmTMB_models/amphibians_meanvert.RData")
-amph.mod = mod
+amph = readRDS("results/sdmTMB_models2/amphibians_sesvert.rds")
+birds = readRDS("results/sdmTMB_models2/birds_sesvert.rds")
+mammals = readRDS("results/sdmTMB_models2/mammals_sesvert.rds")
+repts = readRDS("results/sdmTMB_models2/reptiles_sesvert.rds")
 
-load("results/sdmTMB_models/mammals_meanvert.RData")
-mammals.mod = mod
+mods = list(amph$bestmod, birds$bestmod, mammals$bestmod, repts$bestmod)
+mods = lapply(mods, tidy, effects = "fixed", conf.int = T)
+mods[[1]]$taxa = "Amphibians"
+mods[[2]]$taxa = "Birds"
+mods[[3]]$taxa = "Mammals"
+mods[[4]]$taxa = "Reptiles"
 
-load("results/sdmTMB_models/reptiles_meanvert.RData")
-rept.mod = mod
-
-load("results/sdmTMB_models/birds_meanvert.RData")
-birds.mod = mod
-
-# * - get coefficients with 95% CI for each class -----------------------------------------
-amph.coefs = tidy(amph.mod, conf.int = T, conf.level = 0.95, effects = "fixed")
-amph.coefs$class = "Amphibians"
-
-mammals.coefs = tidy(mammals.mod, conf.int = T, conf.level = 0.95, effects = "fixed")
-mammals.coefs$class = "Mammals"
-
-reptiles.coefs = tidy(rept.mod, conf.int = T, conf.level = 0.95, effects = "fixed")
-reptiles.coefs$class = "Reptiles"
-
-birds.coefs = tidy(birds.mod, conf.int = T, conf.level = 0.95, effects = "fixed")
-birds.coefs$class = "Birds"
-
-
-meanvert.coefs = bind_rows(amph.coefs, mammals.coefs, reptiles.coefs, birds.coefs)%>% 
-  mutate(class = factor(class, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")),
-         type = "Mean Verticality",
[…]

### 2025-02-10 scripts/04_plots/vert_maps_dif.R
update with new models

@@ -123,20 +123,24 @@ wd = vect("data/original/rnaturalearth_world.shp") %>%
 
 # * amphibians SESVERT DIF -------------------------------------------------------------
 
-load("results/sdmTMB_models/amphibians_sesvert.RData")
-amph.sesvert = pred.f
+amph = readRDS("results/sdmTMB_models2/predictions/amphibians_sesvert.rds")
+amph.sesvert = amph$pred.f
+pred = amph$pred
+
+# load("results/sdmTMB_models/amphibians_sesvert.RData")
+# amph.sesvert = pred.f
 amph.sesvert$vertvar = "SES Mean Verticality"
-amph.sesvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean.ses) - min(pred$vert.mean.ses))*100
-amph.sesvert$biome = pred$biome
+# amph.sesvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean.ses) - min(pred$vert.mean.ses))*100
+#amph.sesvert$biome = pred$biome
 
 amph.pdat = amph.sesvert %>% 
   mutate(x = x*1e5, y = y*1e5) %>% 
   rast(type = "xyz", crs = "+proj=cea +datum=WGS84") %>% 
   project("epsg:4326")
 
-amph.sesvert.dif.plt = plt_rast_margin(r = amph.pdat, v = wd, margin.title = "avg. relative\ndifference by latitude",
-                                       plot.title = "Amphibians", var = "est.reldif", rd = 1, 
-                                       scale_limits = c(-5.2,5.2))
+amph.sesvert.dif.plt = plt_rast_margin(r = amph.pdat, v = wd, margin.title = "Avg. Difference by Latitude",
+                                       plot.title = "Amphibians", var = "est.dif", rd = 1, 
+                                       scale_limits = c(-1.5,1.5))
 amph.sesvert.dif.plt
 
 
@@ -144,105 +148,109 @@ amph.sesvert.dif.plt
 # * amphibians Meanvert DIF -------------------------------------------------------------
 
 load("results/sdmTMB_models/amphibians_meanvert.RData")
-amph.meanvert = pred.f
-amph.meanvert = pred.f
-amph.meanvert$vertvar = "Mean Verticality"
-amph.meanvert$est.reldif = pred.f$est.dif/(max(pred$vert.mean) - min(pred$vert.mean))*100
-
-
-amph.pdat = amph.meanvert %>% 
[…]

### 2025-02-10 scripts/03_analyses/01_glmms/01_sdmTMB_amphibians.R
fix typo

@@ -213,12 +213,10 @@ compMods_aic = compareMods_AIC(f = forms, dat, mesh, taxon = taxon, response_var
 saveRDS(list(mesh = mesh, compMods_aic = compMods_aic),  "results/sdmTMB_models2/amph_sesvert.rds")
 
 # refit best model with REML estimation
+compMods_aic = out$compMods_aic
 compMods_aic$modsel
 sanity(compMods_aic$modlist[[2]])
 summary(compMods_aic$modlist[[2]])
-out = readRDS("results/sdmTMB_models2/amph_sesvert.rds")
-mesh = out$mesh
-compMods_aic = out$compMods_aic
 
 # refit model using REML
 bestmod = compareMods_AIC(f = list(forms[[2]]), dat, mesh, taxon = taxon, response_var = response_var, family = gaussian(), reml = T)
@@ -260,6 +258,7 @@ saveRDS(list(mesh = mesh, compMods_aic = compMods_aic, bestmod = bestmod), "resu
 out = readRDS("results/sdmTMB_models2/amphibians_sesvert.rds")
 bestmod = out$bestmod
 
+
 plot_resids(mod = bestmod, response_var = "vert.mean.ses", 
             fpath = paste0("figures/residual_checks/sdmTMB2/amphibians/",fname_end), integer_response = F)
 
@@ -279,7 +278,7 @@ plot_resids(mod = bestmod, response_var = "vert.mean.ses",
 # * - predict svc + realm model to the future ---------------------------------------
 
 predict_future(mod = bestmod, newdata = dat.f, type = "response",
-               fpath = paste0("results/sdmTMB_models2/predictions/",fname_end,".RData"))
+               fpath = paste0("results/sdmTMB_models2/predictions/",fname_end,".rds"))
 
 
 

### 2025-02-10 verticality_future_mkd.Rmd
update figs

@@
 ```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap.pre="Fig.
-", fig.cap=" Coefficient estimates for models of mean and SES verticality with 95% confidence intervals.
+", fig.cap=" Coefficient estimates for the most parsimonious models of SES verticality with 95% confidence intervals.
 Red points indicate coefficients with 95% CIs that do not overlap zero."}
 
-include_graphics("figures/main_figs/fig3_sdmTMB_coefs.png", error = F)
-```
-
-\newpage
-
-```{r fig.id="svc", fig.width=180, fig.height=120, fig.cap.pre="Fig.
-", fig.cap="Spatially varying effect of canopy height on the mean and standard effect size (SES) of verticality for birds, mammals, reptiles, and amphibians.
-The spatially varying effect is the sum of the main effect of canopy height and the spatially varying random effect at any given point.
-Larger positive numbers indicate a stronger positive effect of canopy height on verticality and larger negative numbers indicate a stronger negative effect of canopy height on verticality.
-Dark gray areas indicate locations where the effect of canopy height was not significant.
-Light gray areas indicate areas that are not wooded or that have fewer than five species."}
-
-include_graphics("figures/main_figs/fig4_canopy_height_coefs.png", error = F)
+include_graphics("figures/main_figs/fig3/fig3_sdmTMB_coefs.png", error = F)
 ```
@@
 
-```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap.pre="Fig.
-", fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians.
-Bar plots indicate average relative percent change per latitude.
+```{r fig.id="vert-dif-map", fig.width=180, fig.height=76.94, fig.cap.pre="Fig.
+", fig.cap="Difference in predicted vertical between future and present climate conditions for a) birds, b) mammals, c) reptiles, and d) amphibians.
+Bar plots indicate average difference per latitude.
 Gray areas indicate areas that are not wooded or that have fewer than five species."}
 
-include_graphics("figures/main_figs/fig6_vert_reldifs.png", error = F)
+include_graphics("figures/main_figs/fig5/fig5_vert_difs.png", error = F)
 ```

### 2025-02-25 verticality_future_mkd.Rmd
figures and methods updates

@@
 To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool.
-For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness.
+For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell based on 100 simulations.
+For each simulation, we randomized verticality within the grid cell by randomly drawing the same number of verticality scores from the respective species pool as the observed richness in the cell.
 Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.
@@
 
-We used 10 variables to represent variation in environmental conditions.
-Climate data were represented with six predictors: (i) mean daily maximum air temperature of the warmest month, (ii) mean daily minimum air temperature of the coldest month, (iii) temperature seasonality, (iv) precipitation of the wettest month, (v) precipitation of the driest month, and (vi)  precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
+We used 11 variables to represent variation in environmental conditions.
+Climate data were represented with seven predictors: (i) mean daily maximum air temperature of the warmest month (Tmax warm), (ii) mean daily minimum air temperature of the coldest month (Tmin cold), (iii) temperature seasonality (Temp sea), (iv) precipitation of the wettest month (Precip wet), (v) precipitation of the driest month (Precip dry), (vi)  precipitation seasonality, and (vii) mean monthly precipitation of the warmest quarter [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
 All climatic variables were extracted from CHELSA v2.1 (30 arc-second resolution) for both current (1981-2010) and future (2071-2100) periods, the later represented by an ensemble of the five general circulation models in CHELSA CMIP6 ISIMIP3 data projected under the Shared Socioeconomic Pathway (SSP) 585.
-Vegetation data included (vii) canopy height [@langHighresolutionCanopyHeight2023], (viii) vegetation density [@crowtherMappingTreeDensity2015], and (ix) vegetation complexity (the product of canopy height and vegetation density).
-Additionally, we included (x) historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
+Vegetation data included (viii) canopy height [@langHighresolutionCanopyHeight2023], (ix) vegetation density [@crowtherMappingTreeDensity2015], and (x) vegetation complexity (the product of canopy height and vegetation density).
+Additionally, we included (xi) historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019].
 All environmental data were resampled to the spatial resolution and projection of the species data.
@@
 
-```{r fig.id="richness-mods", fig.width = 160, fig.height=115.6, fig.cap.pre="Fig.
-", fig.cap="Slopes for linear models of richness as a function of SES or mean verticality across all biomes and for individual biomes.
-Hollow circles indicate non-significant relationships (p ≥ 0.05) and filled circles indicate significant relationships.
-R^2^ values are represented by the size of the circle.
-Purple = all biomes, gray = boreal biomes, orange = temperate biomes, green = tropical biomes.
-Flooded GS = Flooded grassland and savannas, Mediterranean FWS = Mediterranean forests, woodlands, and scrub, Montane GS = Montane grasslands and shrublands, Temperate BMF = Temperate broadleaf and mixed forests, Temperate CF = Temperate conifer forests, Temperate GSS = Temperate grasslands, savannas, and shrublands, Tropical CF = Tropical and subtropical conifer forests, Tropical DBF = Tropical and subtropical dry broadleaf forests, Tropical GSS = Tropical and subtropical grasslands, savannas, and shrublands, Tropical MBF = Tropical and subtropical moist broadleaf forests"}
+```{r fig.id="cor-plot", fig.width = 180, fig.height=91.35, fig.cap.pre="Fig.
+", fig.cap=" Pearson correlation coefficients per biome for relationships between richness and SES verticality (x-axis) and SES verticality and canopy height (y-axis) for a) birds, b) mammals, c) reptiles, and d) amphibians.
+Colors indicated biome.
+Black rings indicate significance of the correlation.
+The outer ring indicates significance of the correlation between verticality and canopy height and the inner ring indicates significance of the correlation between richness and verticality."}
 
-include_graphics("figures/main_figs/fig5_vert_rich_lm_slopes.png", error = F)
+include_graphics("figures/main_figs/fig4/fig4_correlations_annotated.png", error = F)
 ```

### 2025-06-26 manuscript.Rmd
add manuscript

---
date: "`r Sys.Date()`"
author: "Your Name"
title: "Forest canopies are better generators of biodiversity in the tropics"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE)
library(officedown)
library(officer)

fp <- fp_par(
  text.align = "center", 
  padding.bottom = 20, padding.top = 120, 
  border.bottom = fp_border())

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
```

\newpage

"The observer who leaves the center of the earth by an infinitely small amount compared to the radius can reach a new world, so to speak, and he can observe more variations in the aspect of the soil and more modifications in the atmosphere than he would if he were to pass from one latitude to another" - Humbolt pg 77 Essay on the Geography of Plants

"The result is that on mountains in temperate zones, nature can develop the
variety of organized beings and meteorological phenomena on only half the
surface offered by tropical regions, where vegetation ceases to exist only at
4,793 meters (2,460 toises)" - pg. 77 Essay on the Geography of Plants

"Current hypotheses for large-scale geographical patterns of species diversity fail to explain the extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5). We call this challenge “Humboldt’s enigma.”" - Rahbek

For centuries, ecologists have been intrigued by the role of vertical landscape structures in biogeographic patterns from local to global scales (cite). In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. As vertical features on the landscape, mountains host an extraordinarly high biodiversity by increasing the land surface within a given area and producing steep climate gradients that promote high turnover in species composition relative to adjacent climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and more variable climates that increase range size and reduce elevational turnover in composition relative to climatically stable tropical mountain slopes (cite Rahbek, Janzen, other mountain paper, and Humbolt essay on geo of plants). Over a century later, MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as critical to our understanding of biogeographic patterns as our ability to measure this structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure vegetation structure.

Tall and structurally complex forests have been proposed to generate biodiversity through a similar mechanism to mountains. Tall, multi-layered forests increase habitat volume and contain vertical microclimate and resource profiles, which allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, as forest height increases and provides more habitable space, additional species may join the assemblage, leading to greater richness within a given area compared to shorter forests (cite). This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex (cite). Yet relationships between canopy height or volume and species richness are often weak in temperate and boreal forests, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015] - a pattern resembling Humbolt's Enigma in forests.

Climate may play a key role in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles. Vertical microclimate gradients produced by micrometeorological processes involving interactions between macroclimate, solar radiation, wind patterns, and forest structure are characterized by decreasing humidity and increasing thermal variability from the ground to the canopy and may help or hinder the evolution of diverse arboreal assemblages. In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in tropical regions could thus support higher animal verticality and richness. In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite). Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity, the mechanisms that allow tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and assemblage verticality in wooded biomes across the globe. We first examining the vegetative dimension to Humbolt's enigma by modeling relationships between species richness and canopy height in forests across latitudes and hypothesize that tropical forests will have greater richness than temperate or boreal forests for a given canopy height. We then examine the mechanistic processes driving this pattern. If vertical stratification influences species richness patterns, we expect the use or arboreal habitats by a disproportionately large number of species to be associated with high richness. We therefore begin by examining correlations between assemblage verticality and species richness. If verticality is positively associated with species richness, then mechanisms limiting verticality will limit richness. To examine these mechanisms, we model verticality of the four vertebrate groups by macroclimate and vegetative structure and investigate the microclimatic processes that underlie these macroclimatic patterns. While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we model vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.

But only within the last decade have we been able to measure forest structure and acquire data on vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive understanding of the impact of vertical habitat structure on species richness patterns.

In forests, species assemblages stratify vertically along resource and microclimate gradients that extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition habitat across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].

# Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics

# Use of vertical habitat space is associated with high species richness

# Vegetative structure and multi-scale climatic processes moderate the capacity for tall forests to support arboreal communities and high species richness

This document presents most of the features of the package `r ftext("officedown", ft)`. 
`r fp`

## Table of content

<!---BLOCK_TOC--->

## List of figures

<!---BLOCK_TOC{seq_id: 'fig'}--->

## List of tables

<!---BLOCK_TOC{seq_id: 'tab'}--->

\newpage

## Sections

### A two columns section

<!---BLOCK_MULTICOL_START--->

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat. Maecenas facilisis dapibus egestas. Curabitur dignissim pharetra pulvinar. Nunc bibendum elit sed cursus congue.

`r run_columnbreak()`Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus. Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor. Donec consequat congue erat in iaculis. Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat. Maecenas facilisis dapibus egestas. Curabitur dignissim pharetra pulvinar. Nunc bibendum elit sed cursus congue. Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus. Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor. Donec consequat congue erat in iaculis. Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.

<!---BLOCK_MULTICOL_STOP{widths: [3,3], space: 0.2, sep: true}--->

### End of two columns section

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat. Maecenas facilisis dapibus egestas. Curabitur dignissim pharetra pulvinar. Nunc bibendum elit sed cursus congue. Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus. Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor. Donec consequat congue erat in iaculis. Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.

<!---BLOCK_LANDSCAPE_START--->

### Landscape section

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat.

```{r, echo=FALSE}
barplot(1:5, col=1:5)
```

<!---BLOCK_LANDSCAPE_STOP--->

### Normal section

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat. Maecenas facilisis dapibus egestas. Curabitur dignissim pharetra pulvinar. Nunc bibendum elit sed cursus congue. Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus. Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor. Donec consequat congue erat in iaculis. Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nulla eu pulvinar arcu, quis aliquam dui. In at cursus ante. Vestibulum non sagittis lacus. Duis vitae iaculis dui. Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna. Sed facilisis dapibus diam nec volutpat. Maecenas facilisis dapibus egestas. Curabitur dignissim pharetra pulvinar. Nunc bibendum elit sed cursus congue. Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus. Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor. Donec consequat congue erat in iaculis. Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.

## Tables

### Table 1

```{r tab.cap="caption 1", tab.id="mtcars"}
head(mtcars)
```

### Table 2

```{r tab.cap="iris"}
head(iris)
```

### Table 3

```{r tab.cap="cars", tab.id="cars"}
head(cars)
```

## figures

### A boxplot

```{r fig.cap="A boxplot", fig.id = "boxplot"}
boxplot(1:8)
```

### A barplot

```{r fig.cap="What a barplot", fig.id = "barplot"}
barplot(1:8)
```

## Lists

Amet nunc eros curabitur tellus massa, eros maximus porttitor sociosqu, pellentesque.

* Erat mauris egestas finibus tincidunt sed in rhoncus a tellus etiam. 
    - A adipiscing per ultricies justo tellus lorem. 
        - Imperdiet ut dui primis, sed gravida, at sed nulla. 
        - Sem posuere lacus consequat inceptos dapibus duis malesuada finibus. 
    - Urna sed dui, ornare, eu turpis mus pellentesque amet amet bibendum. 
* Himenaeos tincidunt, auctor dapibus scelerisque, montes nunc faucibus 
sodales malesuada ridiculus sed cubilia ligula.

1. Erat mauris egestas finibus tincidunt sed in rhoncus a tellus etiam. 
    1. A adipiscing per ultricies justo tellus lorem. 
        1. Imperdiet ut dui primis, sed gravida, at sed nulla. 
        2. Sem posuere lacus consequat inceptos dapibus duis malesuada finibus. 
    2. Urna sed dui, ornare, eu turpis mus pellentesque amet amet bibendum. 
2. Himenaeos tincidunt, auctor dapibus scelerisque, montes nunc faucibus 
sodales malesuada ridiculus sed cubilia ligula.

Lorem dolor interdum orci eros pellentesque semper tristique, sodales, et sed
ut! Porta mattis natoque et. Ac facilisi ipsum viverra elementum vestibulum
ligula amet enim magnis luctus ullamcorper. Rhoncus rhoncus elit in at nisl.
Tincidunt habitant sit.

* Aptent conubia quam montes id sagittis.
    - Mattis nisi nascetur, aliquam duis ex, tristique.
        - Imperdiet ut dui primis, sed gravida, at sed nulla. 
        - Donec ligula nulla ac. Nisl ac at accumsan sagittis eros felis lobortis amet 
        nec phasellus urna bibendum sapien.
    - Eu dui ac id, dictum proin consectetur convallis.
* Facilisi eu lectus mauris lorem. Et sed sapien pellentesque sed etiam vehicula. 
* In porttitor id lorem eu efficitur, nisl dis!

## Reference

see figure \@ref(fig:boxplot) and table \@ref(tab:mtcars)!

### 2025-06-27 manuscript.Rmd
ms revisions

@@
 
-For centuries, ecologists have been intrigued by the role of vertical landscape structures in biogeographic patterns from local to global scales (cite).
+For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
 In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
-As vertical features on the landscape, mountains host an extraordinarly high biodiversity by increasing the land surface within a given area and producing steep climate gradients that promote high turnover in species composition relative to adjacent climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
-Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and more variable climates that increase range size and reduce elevational turnover in composition relative to climatically stable tropical mountain slopes (cite Rahbek, Janzen, other mountain paper, and Humbolt essay on geo of plants).
-Over a century later, MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation.
-Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as critical to our understanding of biogeographic patterns as our ability to measure this structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure vegetation structure.
+As vertical, topographically complex landscape features, mountains provide a larger volume of geographic space and produce steep climate gradients over short distance that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
+Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
+Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961].
+Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
 
@@
 # Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics
+- general richness patterns
+  - increase with area, productivity, and topographic complexity
+- mountain biodiversity:
+  - tropical regions contain more species than expected for a given area, while temperate and boreal regions contain fewer species than expected for a given area
+- our results
+  - high latitude forests have lower richness than expected for a given forest height based on a global model of richness by canopy height
 
@@
 # Vegetative structure and multi-scale climatic processes moderate the capacity for tall forests to support arboreal communities and high species richness
+
+- comparison to mountains
+  - climate bands are narrower in the tropics -> higher turnover rates
+  - we don't see this pattern for vertical climate profiles (similar shapes in profiles across latitudes and less variation in the troipcs)
+- results
+  - the vertical difference in rhmin and tmax is greater than the latitudinal difference from tropical to boreal forests
+  - stable microclimates in tropical canopies -> specialization and speciation
 

### 2025-06-27 scripts/01_get_data/01-1_data_canopy_height.R
mask and aggregate canopy height

@@ -1,0 +1,177 @@
+# Aggregate and average Lang et al. 2023 canopy height raster
+# excluding non-forested areas/human-modified (e.g., oil palm)
+# Non-forested areas are masked by Hansen et al. 2022 Global land use and land cover
+# Canopy height data available from https://langnico.github.io/globalcanopyheight/
+# Landcover data available from https://glad.umd.edu/dataset/global-land-cover-land-use-v1
+
+# This script must be run on hipergator as canopy height and land cover data is stored there
+
+library(terra)
+library(numform)
+library(doParallel)
+library(foreach)
+
+# canopy height indexed by lower left corner (ymin, xmin) (NXXEXXX_Map.tif)
+ch = list.files("/orange/scheffers/dklinges/soiltemp/SoilTemp-big-data/vegetation/canopy_height/lang_et_al_2023_10m/", full.names = T)
+ch = grep("Map.tif", ch, value = T)
+
+# lc indexed by upper left corner (ymax, xmin) (XXNXXXE.tif)
+lc = list.files("/orange/scheffers/lydia.soifer/big_data/land_use/hansen2022/tiles_strata/", full.names = T)
+lctest = rast(lc[1])
+
+# path to landcover tiles
+lcpath = "/orange/scheffers/lydia.soifer/big_data/land_use/hansen2022/tiles_strata/GLCLU_2019_strata_"
+
+# split list of canopy height tiles into 10 to parallelize
+breaksNE = paste0("N", f_pad_zero(seq(0,87,3), 2), "E")
+breaksSE = paste0("S", f_pad_zero(seq(0,87,3), 2), "E")
+breaksNW = paste0("N", f_pad_zero(seq(0,87,3), 2), "W")
+breaksSW = paste0("S", f_pad_zero(seq(0,87,3), 2), "W")
+
+tilesubsNE = lapply(breaksNE, grep, ch, value = T)
+tilesubsSE = lapply(breaksSE, grep, ch, value = T)
+tilesubsNW = lapply(breaksNW, grep, ch, value = T)
+tilesubsSW = lapply(breaksSW, grep, ch, value = T)
+
+tilesubs = c(tilesubsNE, tilesubsNW, tilesubsSE, tilesubsSW)
+lengthsubs = sapply(tilesubs, length)
+tilesubs = tilesubs[lengthsubs>0] # remove groups with no tiles
+
[…]

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/01_get_data.R
remove old file

@@ -1,173 +1,0 @@
-# Get data for modelling vertical microclimates in boreal, temperate, and tropical forests
-# last updated June 12th, 2025
-# run on hipergator
-
-library(terra)
-library(microclim3D) # installed from lysoifer/microclim3D
-library(data.table)
-library(sf)
-library(gfcanalysis)
-library(dplyr)
-library(tidyterra)
-library(fasterize)
-library(ggplot2)
-
-
-# FUNCTIONS ---------------------------------------------------------------
-
-# find name of Hansen tile
-find_gfw_tile = function(tiles, output_folder, 
-                         images = c("treecover2000", "lossyear", "gain", "datamask"),
-                         dataset = "GFC-2022-v1.10") 
-{
-  stopifnot(all(images %in% c("treecover2000", "lossyear", 
-                              "gain", "datamask", "first", "last")))
-  if (!file_test("-d", output_folder)) {
-    stop("output_folder does not exist")
-  }
-  message(paste(dim(tiles)[1], "tiles to download/check."))
-  successes <- 0
-  failures <- 0
-  skips <- 0
-  for (n in 1:dim(tiles)[1]) {
-    gfc_tile <- tiles[n, ]
-    min_x <- st_bbox(gfc_tile)[1]
-    max_y <- st_bbox(gfc_tile)[4]
-    if (min_x < 0) {
-      min_x <- paste0(sprintf("%03i", abs(min_x)), "W")
-    }
-    else {
-      min_x <- paste0(sprintf("%03i", min_x), "E")
-    }
-    if (max_y < 0) {
-      max_y <- paste0(sprintf("%02i", abs(max_y)), "S")
-    }
-    else {
-      max_y <- paste0(sprintf("%02i", max_y), "N")
-    }
-    file_root <- paste0("Hansen_", dataset, "_")
-    file_suffix <- paste0("_", max_y, "_", min_x, ".tif")
-    filenames <- paste0(file_root, images, file_suffix)
-    filenames = paste0(output_folder, filenames)
-    return(filenames)
-  }}
-
-
-# Copernicus closed forest raster -----------------------------------------
-
-# run on hipergator
-# Select forested land cover
-# slow to run
-
[…]

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/01_get_gedi_data.R
download gedi tiles and select points

@@ -1,0 +1,187 @@
+# Get data for modelling vertical microclimates in boreal, temperate, and tropical forests
+# This file must be run on hipergator
+
+library(terra)
+library(microclim3D) # installed from lysoifer/microclim3D
+library(data.table)
+library(sf)
+library(gfcanalysis)
+library(dplyr)
+library(tidyterra)
+library(fasterize)
+library(ggplot2)
+
+
+# FUNCTIONS ---------------------------------------------------------------
+
+# find name of Hansen tile
+find_gfw_tile = function(tiles, output_folder, 
+                         images = c("treecover2000", "lossyear", "gain", "datamask"),
+                         dataset = "GFC-2022-v1.10") 
+{
+  stopifnot(all(images %in% c("treecover2000", "lossyear", 
+                              "gain", "datamask", "first", "last")))
+  if (!file_test("-d", output_folder)) {
+    stop("output_folder does not exist")
+  }
+  message(paste(dim(tiles)[1], "tiles to download/check."))
+  successes <- 0
+  failures <- 0
+  skips <- 0
+  for (n in 1:dim(tiles)[1]) {
+    gfc_tile <- tiles[n, ]
+    min_x <- st_bbox(gfc_tile)[1]
+    max_y <- st_bbox(gfc_tile)[4]
+    if (min_x < 0) {
+      min_x <- paste0(sprintf("%03i", abs(min_x)), "W")
+    }
+    else {
+      min_x <- paste0(sprintf("%03i", min_x), "E")
+    }
+    if (max_y < 0) {
+      max_y <- paste0(sprintf("%02i", abs(max_y)), "S")
+    }
+    else {
+      max_y <- paste0(sprintf("%02i", max_y), "N")
+    }
+    file_root <- paste0("Hansen_", dataset, "_")
+    file_suffix <- paste0("_", max_y, "_", min_x, ".tif")
+    filenames <- paste0(file_root, images, file_suffix)
+    filenames = paste0(output_folder, filenames)
+    return(filenames)
+  }}
+
+
+# Copernicus closed forest raster -----------------------------------------
+
+# run on hipergator
+# Select forested land cover
+# slow to run
+
[…]

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/01.2_get_landcover_and_soil_data.R
download lc and soil data

@@ -1,0 +1,48 @@
+# Download MODIS landcover data and soil data for regional extents
+# The extents cover 1x1 degree gedi tiles
+# RUN ON HIPERGATOR
+
+library(microclim3D) # on github lysoifer/microclim3D
+library(data.table)
+library(terra)
+
+start = "2019-01-01"
+end = "2022-12-31"
+outpath = "/orange/scheffers/lydia.soifer/big_data/land_use/landcover_modis/regions2/"
+usr = "lysoifer"
+pwd = "Timp4mEDa."
+
+# These are the processed gedi data, filtering out points with forest loss, etc.
+gedi = list.files("data/derivative_data/gedi_filtered/", full.names = T)
+
+# Get extents of gedi data
+e = list()
+for(i in 1:length(gedi)) {
+  g = fread(gedi[i], select = c("lat_lm_a0", "lon_lm_a0"))
+  e[[i]] = ext(min(g$lon_lm_a0), max(g$lon_lm_a0), min(g$lat_lm_a0), max(g$lat_lm_a0))
+}
+regions = gsub(".csv", "", basename(gedi))
+regions = gsub("gedi_", "", regions)
+names(e) = regions
+
+# Download modis landcover using modisfast wrapper
+microclim3D::get_modis_lc(e, start, end, usr, pwd, collection = "MCD12Q1.061",
+                          variables = c("LC_Type1", "QC"), regions = names(e),
+                          outpath = outpath)
+
+
+
+# Download soil data ------------------------------------------------------
+
+for(i in 1:length(e)) {
+  out = paste0("/orange/scheffers/lydia.soifer/big_data/soil/soilgrid/regions/", names(e)[i], "/")
+  if(!dir.exists(out)) dir.create(out)
+  e2 = extend(e[[i]], 1) # using this for danum because I am getting NA soil values at a couple of points
+  roi = vect(e2, crs = "epsg:4326")
+  get_soil(roi, outdir = paste0(out, "soilgrids.tif"), overwrite = T)
+}
+
+
+
+
+

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/02_prep_micropoint.R
prepare micropoint inputs

@@ -1,0 +1,352 @@
+# run micropoint models for 100 points in each forested area
+library(terra)
+library(mcera5)
+library(ncdf4)
+library(data.table)
+library(microclim3D)
+library(microclimf)
+library(foreach)
+library(numform)
+
+# read in gedi data
+# danum gedi points are all labeled as mangroves - resselect point
+
+### HARD CODES ###
+#gedi = fread("data/derivative_data/gedi_filtered/gedi_pacaya_samiria.csv")
+
+era = paste0("/orange/scheffers/dklinges/soiltemp/SoilTemp-big-data/era5/mcera5_",2019:2022, "_time.nc")
+yrs = 2019:2022
+land_sea_mask = rast("/blue/scheffers/dklinges/soiltemp/SoilTemp-daily-offset/data/enviro_parameters/land_cover/landprop_30km.tif")
+
+# For all the regions to run in a loop
+# landcover for each region
+
+gedifiles = list.files("data/derivative_data/gedi_filtered/", full.names = T)
+
+lcpath = list.dirs("/orange/scheffers/lydia.soifer/big_data/land_use/landcover_modis/regions2/data/", full.names = T, recursive = T)
+lcpath = grep("MCD", lcpath, value = T)
+
+# soil for each region
+spath = list.files("/orange/scheffers/lydia.soifer/big_data/soil/soilgrid/regions/", recursive = T, full.names = T)
+
+# global elevation and elevation of era5
+elev_path = "/orange/scheffers/lydia.soifer/big_data/topography/elevation_geodataR/elevation/wc2.1_30s/wc2.1_30s_elev.tif"
+dtmc_path = "/orange/scheffers/lydia.soifer/big_data/topography/dtmc_era5/demera5.tif"
+
+# file paths to save output for each region
+regions = sapply(spath, function(x)stringr::str_split_i(x, "/", 10))
+op = paste0("/blue/scheffers/lydia.soifer/verticality_future/data/derivative_data/micropoint_prep/prep_micropoint_", regions, ".rds")
+
+
+
+#########
+
+## Functions ##
+
+# prep micropoint is from the microclim3D package that I am working on 
+# Here is the version used in this analysis
+# As the package is under active development, the function may change overtime
+prep_micropoint = function(e, start, end, era_path = NA, clima = NA, landcover_path, soilpath, elev_path, dtmc_path) {
+  # extend requested extent (so distance weighting works)
[…]

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/03_run_micropoint.R
run micropoint profiles

@@ -1,0 +1,1062 @@
+# RUN MICROPOINT MODELS
+# MUST RUN ON DESKTOP FOR NOW
+# micropoint prep run on hipergator because large data is stored there
+# I exported the inputed needed to run micropoint, including
+# the env variables (climate, veg, soil params), and sampled gedi
+# points for min temp of coldest month and max temp of warmest month.
+# Months with min and max temps were determined using era5 climate grids
+# For example, I calculated min temp of each day, then averaged min temps
+# by month, then found the month with the coldest min temp. This was done
+# for each year.
+# In general months are more consistent in temperate/boreal than in tropics
+
+library(microclim3D)
+library(foreach)
+library(data.table)
+library(soiltexture)
+library(dplyr)
+library(tidyr)
+soilparams = micropoint::soilparams
+
+# Function to run micropoint - part of microclim3D package, but package currently isn't functioning
+run_micropoint = function(tme, gedi, climr, vegp, soil, elev, asp, slp, dtmc,
+                          method, plotout, n, maxiter, fout, modis_path, reqhgts = NA, vertpai_method = "pai") {
+  # if(method == "vertprof") {
+  #   vertmat = matrix(ncol = 7, nrow = 0)
+  #   colnames(vertmat) = c("z", "tair", "canopy_height", "elev", "lon", "lat", "doy")
+  #   write.table(vertmat, fout)
+  # }
+  
+  # Unpack spatrasters
+  # if we have multiple years of climate data
+  if(is(climr[[1]], "list")) {
+    for(a in 1:length(climr)) {
+      climr[[a]] = lapply(climr[[a]], unwrap)
+    }
+  } else { # if there is only one year of climate data
+    climr = unwrap(climr)
+  }
+  
+  # unpack vegp
+  vegp = lapply(vegp, unwrap)
+  soil = unwrap(soil)
+  elev = unwrap(elev)
+  asp = unwrap(asp)
+  slp = unwrap(slp)
+  dtmc = unwrap(dtmc)
+  
+  # Loop through GEDI points and calculate vertical microclimate gradient for each
+  v = foreach(i = 1:nrow(gedi), .combine = rbind) %do% {
+    print(i)
+    
+    # If we are modelling for only the month in which the GEDI point was observed
+    # Then we need to pull out the year and month of the observation
[…]

### 2025-06-27 scripts/03_analyses/02_vertical_microclim_models/04_plot_profiles.R
plot vertical climate profiles

@@ -1,0 +1,249 @@
+# Plot boxplots of vertical microclimate profiles
+# Then consider running an NMDS to look at bioclim at different latitudes and iin canopy and ground
+
+library(ggthemes)
+library(dplyr)
+library(ggplot2)
+library(foreach)
+library(data.table)
+library(tidyr)
+
+# Load in modeling data
+micromods = list.files("results/microclim_models/", recursive = T, full.names = T)
+micromods.tmin = grep("mintemp", micromods, value = T) # points for the coldest month of the year
+micromods.tmax = grep("tmax", micromods, value = T) # points for the warmest month of the year
+indat = list.files("D:/projects/verticality_future/data/derivative_data/micropoint_prep/", full.names = T) # gedi points used for models
+
+# dataframe for month with coldest min temp
+mm.tmin = foreach(f = 1:length(micromods.tmin), .combine = rbind) %do% {
+  fi = fread(micromods.tmin[f])
+  nm = stringr::str_split_i(dirname(micromods.tmin[f]), "/", 3)
+  fi$region = nm
+  fi
+}
+
+# dataframe for month with hottest max temp
+mm.tmax = foreach(f = 1:length(micromods.tmax), .combine = rbind) %do% {
+  fi = fread(micromods.tmax[f])
+  nm = stringr::str_split_i(dirname(micromods.tmax[f]), "/", 3)
+  fi$region = nm
+  fi
+}
+
+# dataframe for PAD
+pad = foreach(f = 1:length(indat), .combine = rbind) %do% {
+  dat = readRDS(indat[f])
+  gedi.tmin = dat$gedisamp.mcold
+  gedi.tmax = dat$gedisamp.mhot
+  gedi = rbind(gedi.tmin, gedi.tmax, fill = T)
+  cnames = grep("^pavd", names(gedi), value = T)
+  cnames = c("shot_num", "rh_100_a0", "lat_lm_a0", "lon_lm_a0", grep("[0-9]$", cnames, value = T))
+  gedi = gedi[, ..cnames]
+  gedi[gedi==0] <- NA # convert zeros to NAs
+  gedi = gedi %>% 
+    pivot_longer(cols = 5:ncol(gedi), names_to = "height", values_to = "pavd") %>% 
+    drop_na()
+  
+  nm = basename(indat[f])
+  nm = gsub("prep_micropoint_", "", nm)
+  nm = gsub(".rds", "", nm)
+  gedi$region = nm
+  gedi
[…]

### 2025-06-27 scripts/03_analyses/00_tools/tools.R
project tools

@@ -1,0 +1,9 @@
+reproj_df = function(df, from, to, method, torast) {
+  d = df %>% 
+    rast(crs = from) %>%
+    project(to, method = method)
+  
+  if(!torast) {d = as.data.frame(d, xy = T)}
+  
+  return(d)
+}

### 2025-06-27 manuscript.Rmd
revisions

@@
 For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
-As vertical, topographically complex landscape features, mountains provide a larger volume of geographic space and produce steep climate gradients over short distance that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
+In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains as vertical and topographically complex landscape features in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
+The vertical nature of mountains provides a larger volume of geographic space and produces steep climate gradients over short distances that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
 Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
 Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961].
-Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
+Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
 
-Tall and structurally complex forests have been proposed to generate biodiversity through a similar mechanism to mountains.
-Tall, multi-layered forests increase habitat volume and contain vertical microclimate and resource profiles, which allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-Thus, as forest height increases and provides more habitable space, additional species may join the assemblage, leading to greater richness within a given area compared to shorter forests (cite).
-This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex (cite).
-Yet relationships between canopy height or volume and species richness are often weak in temperate and boreal forests, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015] - a pattern resembling Humbolt's Enigma in forests.
+Tall, multi-layered forests have been proposed to generate biodiversity through a similar mechanism to mountains.
+By increasing habitat volume and producing vertical profiles of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to shorter forests [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967].
+This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+Yet impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020] - a pattern resembling that of mountains.
 
-Climate may play a key role in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles.
+Climate and vegetative complexity may play key roles in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles and impacting the availability of vertical niche opportunities.
 Vertical microclimate gradients produced by micrometeorological processes involving interactions between macroclimate, solar radiation, wind patterns, and forest structure are characterized by decreasing humidity and increasing thermal variability from the ground to the canopy and may help or hinder the evolution of diverse arboreal assemblages.
@@
   - stable microclimates in tropical canopies -> specialization and speciation
+
+# Methods
+## Data
+The canopy height layer was obtained from Lang et al.
+[-@langHighresolutionCanopyHeight2023], which represents canopy height in the year 2020 at a 10 m resolution.
+To obtain canopy height in forests, we aggregated the data to match the resolution of Hansen et al.
+[-@hansenGlobalLandUse2022] global land cover and land use in 2019 (~ 30 m resolution) and used this layer to mask all grid cells with woody vegetation < 3 m, surface water and permanent ice, built-up land, cropland, recent forest loss, and land use associated with forestry and shifting cultivation (e.g., oil palm).
+We then aggregated and averaged the remaining pixels to the resolution of the verticality data (50 km).
 

### 2025-06-30 manuscript.Rmd
canopy height richness models and plots

@@
 ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
+```
+
+```{r}
+# analysis scripts
+
+"scripts/03_analyses/01_canopyheight_richness_relationships.R" # canopy height richness models and plots
 ```
@@
 We then aggregated and averaged the remaining pixels to the resolution of the verticality data (50 km).
+
+We classified biomes into larger categories based on the 2017 biome categories.
+For biomes that occurred across latitudes (e.g., Montane Grasslands and Shrublands), we classified them as tropical if they occurred within the latitudes covered by biomes classified as tropical and temperate if they occurred within the latitudes classified as temperate.
+Additionally, Mediterranean forests, woodlands, and scrub were partitioned into temperate forests if the ecoregion contained "forest" and temperate woodland, savanna, and shrubland if the ecoregion name did not contain the word "forest".
+
+glmmTMB was used to produce models of canopy height by richness. sdmTMB was used to produce models of verticality. DHARMa was used to check model fit and assumptions.
+
+```{r}
+# Fig. 1
+
+# Legend: a) canopy height, b) vertebrate richness (all vertebrates combined)
+# c) Deviance residuals for a model of total vertebrate richness by canopy height
+# d) Relationships between species richness and canopy height for each taxon in each
+# biome category. Dots indicate observed values and lines indicate negative binomial
+# model fits of richness by canopy height for each combination of taxon and biome.
+
+"figures/ms_figures/fig1_canopy_richness.png"
+
+```
+
+# Data availability statement
+canopy height:
 

### 2025-06-30 scripts/03_analyses/01_canopyheight_richness_relationships.R

@@ -1,0 +1,429 @@
+# Model relationships between canopy height and species richness
+
+library(DHARMa)
+library(dplyr)
+library(foreach)
+library(terra)
+library(tidyterra)
+library(glmmTMB)
+library(ggplot2)
+library(patchwork)
+
+amph = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+birds = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
+rept = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv")
+mammals = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
+
+amph$taxa = "Amphibians"
+birds$taxa = "Birds"
+rept$taxa = "Reptiles"
+mammals$taxa = "Mammals"
+
+df = rbind(amph, birds, rept, mammals)
+
+# identify tropical, temperate, and boreal latitudes
+tropical = range(abs(df$y[grepl("Tropical", df$biome)]))
+temperate = range(abs(df$y[grepl("Temperate", df$biome)]))
+boreal = range(abs(df$y[grepl("Boreal|Tundra", df$biome)]))
+
+# Classify biomes more coarsely
+# if not forest and not temperate, then tropical woodland, savanna, shrubland
+# if not forest and not tropical or boreal, then temperate woodland, savanna, shrubland
+# if tropical and forest then tropical forest
+# if temperate and forest then temperate and forest
+# if boreal/tundra, then boreal
+# if Mediterranean and forest, then temperate forest
+# if Mediterranean and not forest then temperate woodland, savanna, shrubland
+df = df %>% 
+  mutate(biome2 = case_when((!grepl("Forest|Temperate", biome)) & (abs(df$y) >= tropical[1]) & (abs(df$y) < tropical[2]) ~ "Tropical woodland, savanna, and shrubland",
+                            (!grepl("Forest|Tropical|Tundra", biome)) & (abs(df$y) >= temperate[1]) & (abs(df$y) < temperate[2]) ~ "Temperate woodland, savanna, and shrubland",
+                            grepl("(?=.*Forest)(?=.*Tropical)", biome, perl=T) ~ "Tropical forest",
+                            grepl("(?=.*Forest)(?=.*Temperate)", biome, perl=T) ~ "Temperate forest",
[…]

### 2025-07-04 manuscript.Rmd
ms revisions

@@
 # analysis scripts
+# storing all figures in figures/ms_figures/
+# other figures are from older versions
 
@@
 # Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics
-- general richness patterns
-  - increase with area, productivity, and topographic complexity
-- mountain biodiversity:
-  - tropical regions contain more species than expected for a given area, while temperate and boreal regions contain fewer species than expected for a given area
-- our results
-  - high latitude forests have lower richness than expected for a given forest height based on a global model of richness by canopy height
 
-# Use of vertical habitat space is associated with high species richness
+```{r}
+source("scripts/03_analyses/01_canopyheight_richness_relationships.R")
+# ratedifs provides table of differences in richness between forests that are 15 and 25 m tall for each biome/taxa combo
+```
 
-# Vegetative structure and multi-scale climatic processes moderate the capacity for tall forests to support arboreal communities and high species richness
+To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally and mapped the residuals.
+We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.
+
+Globally, taller trees, which increase habitat volume, support more species (Fig.
+S\@ref(fig:global-chrich)).
+However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Fig.
+\@ref(fig:fig1), S\@ref(fig:resids-taxa)).
+Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions.
+Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig.
+\ref(fig:fig1)).
+For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions.
+Within the tropics, increasing canopy height generally yields greater gains in richness in forests than in drier woodlands, savannas, and shrublands, though baseline richness in areas with short vegetation are similar (Fig.
+\@ref(fig:fig1)).
+For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species.
+The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Fig.
+\@ref(fig:fig1)).
+
+Minor deviations from these patterns may indicate variations in evolutionary drivers of species richness patterns.
+  [- This pattern is particularly stark in the western hemisphere.
+In the eastern hemisphere, the pattern is weaker where evolutionary forces may drive overprediction of richness in parts of the lowland African tropics and Indo-Malaysian.
+In Indo-Malaysia, island size may reduce vertebrate richness relative to expectations by canopy height.]
+
+ [ - While this pattern could be driven by the latitudinal richness gradient]
+
+However, the described patterns are broadly consistent among taxa and mirror findings that for a given area, mountains have higher richness in the tropics [@rahbekHumboldtsEnigmaWhat2019].
+Thus, increases in habitat volume via either mountains or forests have greater impacts on richness in the tropics.
+In other words, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the tropics.
+
+  - TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE REGION
+ - on second thought, I don't think this is necessary because we want to know why more species can't exist in trees in boreal regions
+
+# The vertical structure of vertebrate assemblages
+The vertical layering of species within forests may play a key role in the capacity for tall forests to support rich assemblages.
+By vertically stratifying across canopy layers, arboreal species can fill the larger volume of habitat offered by tall forests, increasing the richness of the assemblage [cite].
+Within wooded ecosystems, we thus expect species rich assemblages to have a higher proportion of arboreal species and thus assemblages with vertical distributions of richness skewed toward the canopy [@oliveiraVerticalStratificationInfluences2019].
+To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@scheffersVerticalArborealityHorizontal2017].
+For species occurring in multiple habitats, we averaged scores across habitats.
+We then calculated verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).
+
+Verticality is strongly correlated with richness patterns, peaking in the tropics and declining toward higher latitudes.
+- At its highest, x % of species exhibited arboreal strategies, corresponding to some of the richest assemblages in tropical regions.
+- In these regions, expansion up the canopy combined with fine-scale niche partitioning, and thus species packing, within the canopy produce species rich assemblages.
+
+Though temperate and boreal regions have tall trees, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, consisting of greater proportions of fossorial and terrestrial species.
+This pattern suggests a reduction in the fine-scale partitioning of canopies relative to tropical regions, and thus a constraint on the capacity for taller canopies to generate biodiversity via the vertical layering of species.
+
+Environmental factors constraining verticality or the fine-scale partitioning of niche space will limit the capacity for tall forests to generate biodiversity.
+
+Dry (and cold) habitats have lower vertical complexity. Less layering = fewer opportunities for niche speciailizes = lower richness
+Though climate may still limit arobreality at higher latitudes, higher structural complexity and vegetation density across forest profiles in select extra-tropical regions increases verticality and consequently species richness.
+
+where verticality declines, species may use terrestrial habitats during the dry season to maintain moisture requirements assemblage verticality declines because species exhibit semi-arboreal life strategies, which reduces assemblage verticality.
+
+"short tropical forests show higher complexity relative to other forests of similar heights" - de Conto et al. 2023
+
+# Canopy Climates
+To examine the mechanisms contributing to variation in the capacity for tall forests to support highly arboreal, and thus species rich, assemblages, we use spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with climatic factors that may constrain or support arboreality, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month (Supplementary methods).
+We additionally include vegetation density in the models, as this will impact the availability of habitat volume, and did not include foliage height diversity due to its strong correlation with canopy height, as has previously been observed [@decontoCharacterizingStructuralComplexity2024; @atkinsPowerLawScaling2022].
+The climate variables included in these models represent macroclimate conditions in open habitats rather than the climate conditions experienced in forest canopies.
+We therefore additionally modeled vertical microclimate profiles in six lowland forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [] with ERA5 macroclimate data at a high temporal resolution [] in a process-based microclimate model that resolves within-canopy processes generating vertical climate gradients [].
+These models provide an estimate of climate conditions proximal to those experienced by forest-dwelling organisms and thus better represent the biogeographic and ecophysiological processes that allow tall forests to support vertical and rich assemblages in the tropics more so than at higher latitudes [@klingesProximalMicroclimateMoving2024].
+Modeling microclimates in conjuction with modeling verticality based on macroclimate conditions thus allows us to infer the microclimate processes that underlie macroclimate patterns in the absence of global 3D climate data.
+
+Our models indicate that taller canopies support highly vertical vertebrate assemblages, while dry, hot, and cold climates limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying extents across vertebrate classes (Fig.
+\@ref(fig:fig3).
+These patterns arise because the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.
+
+These patterns arise because macroclimate indirectly constrain verticality via its effect on vegetation structure and niche opportunities, while microclimate conditions produced by the interaction between macroclimate and vegetation structure place physiological constraints on arboreal lifestyles.
+the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.
+
+This occurs because climate indirectly constrains verticality via its effects on vegetation structure and niche opportunities for specialization across vertical strata and the interaction between macro- and micro-climate gradients directly constrain verticality....
+
+Climate mediates that capacity for forests to support vertical assemblages indirectly, through its effect on vegetation structure, and directly, via impacts on animal physiology.
+
+Interactions between macroclimate, vegetation structure, and microclimate impact verticality by affecting niche opportunities and imposing physiological constraints.
+
+## Climate produces structural constraints to verticality
+Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014].
+High annual precipitation and low precipitation seasonality enable development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
+Wet climates additionally allow development of diverse epiphyte communities, which provide additional microhabitats in the canopy [CITE].
+Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding extraordinarily rich communities.
+In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig.
+\@ref(fig:fig1)).
+
+At higher latitudes, lower precipitation and light hinder the development of densely layered forests by restricting the functional and structural configurations of trees [@ehbrechtGlobalPatternsClimatic2021a; @hakkenbergClimateMediatesRelationship2021; @atkinsPowerLawScaling2022].
+In evergreen needleleaf forests that dominate boreal regions, these environmental constraints dampen the rate of increase in structural complexity with canopy height [@atkinsPowerLawScaling2022].
+Tall canopies therefore provide fewer niche opportunities in boreal regions, which may constrain verticality and limit species richness relative to canopies of equivalent height in lower latitude forests.
+Yet, hotspots of structural complexity in temperate forests, including the North Pacific Alaskan coastal forest, Valdivian forest, and Tasmanian forest, that nearly match the complexity of tropical forests [@ehbrechtGlobalPatternsClimatic2021a] also contain assemblages with lower verticality than the tropics and fewer species than expected based upon their canopy height (Figs.
+\@ref(fig:fig1), \@ref(fig:fig2)).
+MacArthur & MacArthur [-@macarthurBirdSpeciesDiversity1961] observed this pattern over 50 years ago, noting that "tropical bird species diversity is much greater than the temperate one for habitats of comparable profiles." The effects of the interaction between macroclimate and vegetation profiles on vertical microclimate profiles provide an explanation [MIRRORING MECHANISMS THAT ALLOW MOUNTAINS TO BE BETTER GENERATORS OF BIODIVERSITY IN THE TROPICS]
+
+## Physiological constraints to arboreality truncate suitable habitat 
+## Canopy climates truncate suitable habitat in cold and dry forests
+
+physiological tolerance hypothesis
+
+In regions that experience low dry season precipitation, drier conditions in the canopy relative to the understory truncate the availability of vertical habitat space.
+
+Across vertebrate classes, low dry season precipitation limits verticality.
+While vertical gradients in humidity are steep in forests across latitudes, high dry season precipitation and thus lower precipitation seasonality in tropical moist forests provides a consistent source of water.
+Arboreal animals are therefore able to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014].
+In tropical dry regions the volume of habitat that can support a high diversity of arboreal lifestyles may shift seasonally.
+For example, birds and frogs that use primarily arboreal habitats during the wet season increase their use of more humid terrestrial habitats during the dry season to avoid desiccation (Basham and Scheffers 2020; Rajaonarivelo et al.
+2021).
+In many temperate and/or boreal biomes, low precipitation combined with dry canopy microclimates may similarly constrains amphibian arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses (Sunday et al.
+2014; Tracy, Christian, and Tracy 2010).
+
+We also observed consistently significant interactions between canopy height and extreme temperature variables.
+Canopies experience lower minimum temperatures and higher maximum temperatures than understories that are buffered from thermal extremes (De Frenne et al.
+2019).
+Particularly when combined with windier conditions in the canopy, cold temperatures can cause ectotherm body temperature to drop below critical thermal minima and impose energetic constraints to endotherms (Sunday et al.
+2014) (Byman, Hay, and Bakken 1988).
+Furthermore, high temperatures may cause body temperatures to exceed critical thermal maxima or cause desiccation due to evapotranspirative cooling when water is scarce (McKechnie et al.
+2017; Sunday et al.
+2014).
+When minimum temperatures are warmer and maximum temperatures are cooler, the effect of canopy height on verticality generally becomes more positive, likely because more benign thermal microclimate conditions reduce physiological constraints on arboreal life strategies.
+Therefore, canopy microclimates that impose physiological challenges to arboreal life strategies within prevailing macroclimate, truncate the volume of suitable habitat and prevent taller canopies from support a highly vertical and species rich assemblage.
+
+- Bird richness weighted toward the ground in higher latitude forests (Basham et al. 2023)
+
+Yet temperate regions that equal the height and complexity of tropical areas contain lower verticality, which may contribute to lower richness.
+Direct constraints of climate on animal physiology may contribute to this pattern.
+
+Niche specilization
+
+The canopy supports fewer arboreal species at high latitudes because
+1) physiological limitations and physiological tolerance hypothesis
+2) exaggerated spatial and temporal variability of canopy microclimates reduces opportunities for specialization that would otherwise yield high richness
+
+[temperature seasonality exaggerated in the canopy relative to the ground] - would increase range size of arboreal species and therefore reduce opportunities for specialization and high richness
+- climate stratification - more offset in more complex forests in the tropics? - not consistently supported by our microclim models
+
+- Structurally complex forests capture a remarkable volume of climate space given their geographic footprint, bringing distinct climates within close proximity of each other.
+But the climate space is truncated at higher latitudes and in drier forests due to physiological constraints.
+
+- Low spatial and temporal variation across vertical strata in the tropics supports specialization and stratification, while greater spatial and temporal variability at higher latitudes brings animals into contact with more climates, thereby increases range size and reducing opportunities for specialization that promote speciation.
+
+# Conclusion - bring back to Humbolt's engima
+- Trees, just as mountains, bring different climate zones within close proximity of each [@rahbekHumboldtsEnigmaWhat2019].
+While the climatic difference from the ground to the canopy is much less than the climatic difference across thousands of meters in elevation, the climatic rate of change is often faster from the ground to the canopy.
+This climatic layering allows distinct assemblages of species to exist across forest strata.
+
+- climate stability, range size, and niche width - greater climate stability in the tropics allows for more specialization within the canopy, while higher climate variability in the canopy in high latitudes reduces specialization and turnover.
+Compare to the ground - see Dave's mountaintops paper?
+
+- The capacity for tall forests to support rich arboreal assemblages depends on the density of vegetation across the vertical profile and the presence of canopy microclimates that limit physiological constraints on arboreal life strategies.
+
+- In high latitude forests, tall canopies may help generate biodiversity on the ground by buffering microclimate and providing small increases in structural complexity that can be inhabited by few arboreal animals with specialized physiologies.
+However, harsh microclimates in the upper canopy largely truncate niche availability, preventing fine-scale partitioning of canopy habitats by arboreal species and thus limiting the extent to which taller canopies increase species richness.
+
+- The biodiversity-structure hypothesis suggests xyz...
+  - Greater veg density across vertical profile = higher productivity
+
+[INTEGRATE VERTICAL COMPONENT INTO THIS PARAGRAPH]
 
@@
   - stable microclimates in tropical canopies -> specialization and speciation
+
+In climate conditions that do not constrain verticality, shifts up the canopy are associated with radiations within the canopy that produce positive relationships between canopy verticality and species richness
+In climate conditions that constrain verticality, arboreal species occur, but arboreal radiations do not, causing neutral or negative relationships between verticality and richness (although this mechanism/relationship is noisy)
 
@@
 
-glmmTMB was used to produce models of canopy height by richness. sdmTMB was used to produce models of verticality. DHARMa was used to check model fit and assumptions.
+MASS was used to produce models of canopy height by richness.
+sdmTMB was used to produce models of verticality.
+DHARMa was used to check model fit and assumptions.
+ggeffects was used to generate model predictions and confidence intervals.
+Figures were generating using tidyterra, patchwork, ggplot2
 
-```{r}
+# Data availability statement
+canopy height:
+
+# Figures
+
+```{r fig1, fig.id="fig1"}
 # Fig. 1
 
-# Legend: a) canopy height, b) vertebrate richness (all vertebrates combined)
-# c) Deviance residuals for a model of total vertebrate richness by canopy height
+# Legend: Spatial distribution of a) canopy height, b) vertebrate richness (all vertebrates combined), and
+# c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height
 # d) Relationships between species richness and canopy height for each taxon in each
 # biome category. Dots indicate observed values and lines indicate negative binomial
-# model fits of richness by canopy height for each combination of taxon and biome.
+# model fits of richness by canopy height.
 
@@
 
-# Data availability statement
-canopy height:
+# Supplementary figures
+
+```{r splot-global-chrich, fig.id="global-chrich", width=90, height=90, fig.cap="Global relationship between total vertebrate richness and canopy height.
+Line indicates predictions from a negative binomial model of richness by canopy height ± 95% CI.
+Points indicate observed richness by canopy height.}
+
+"figures/ms_figures/supp_figs/total_richness~canopy.png"
+```
+
+```{r splot-resids-taxa, fig.id="resids-taxa", width=180, height=100, fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians."}
+
+"figures/ms_figures/supp_figs/deviance_resids_taxa.png"
+```
 

### 2025-07-07 manuscript.Rmd
ms revisions

@@
 
-"The result is that on mountains in temperate zones, nature can develop the
-variety of organized beings and meteorological phenomena on only half the
-surface offered by tropical regions, where vegetation ceases to exist only at
-4,793 meters (2,460 toises)" - pg. 77 Essay on the Geography of Plants
-
 "Current hypotheses for large-scale geographical patterns of species diversity fail to explain the extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5).
@@
 For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains as vertical and topographically complex landscape features in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
-The vertical nature of mountains provides a larger volume of geographic space and produces steep climate gradients over short distances that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
-Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
+In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
+The vertical nature of mountains provides more land surface per unit land area than flat regions and produces steep climate gradients over short distances that allow rapid rates of turnover.
+Combined, these factors generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
+Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctiveness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
 Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961].
@@
 
-Tall, multi-layered forests have been proposed to generate biodiversity through a similar mechanism to mountains.
-By increasing habitat volume and producing vertical profiles of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to shorter forests [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967].
+Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains.
+By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967].
 This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
@@
 
-Climate and vegetative complexity may play key roles in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles and impacting the availability of vertical niche opportunities.
-Vertical microclimate gradients produced by micrometeorological processes involving interactions between macroclimate, solar radiation, wind patterns, and forest structure are characterized by decreasing humidity and increasing thermal variability from the ground to the canopy and may help or hinder the evolution of diverse arboreal assemblages.
-In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019].
-Taller canopies in tropical regions could thus support higher animal verticality and richness.
-In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
-Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.
+Vegetative complexity and climate may play key roles in mediating the relationships between canopy height and species richness by impacting the availability of vertical niche opportunities physiologically and constraining arboreal lifestyles [CITE].
+Densely layered forest structures not only provide habitat across the forest profile, but produce vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed.
+From the humid and thermally stable understory to the dry and thermally variable canopy that experiences hot and cold temperature extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or hinder the capacity for tall forests to generate high levels of biodiversity.
+For example, increased exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat that can support arboreal species and thus limit richness despite tall canopies [@oliveiraVerticalStratificationInfluences2019].
+Moreover, relatively low thermal variability in the canopies of tropical forests may increase the distinctness of vertical climate bands that support high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics [CITE].
 
-Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity, the mechanisms that allow tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
+Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [CITE], the mechanisms that allow tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
 To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and assemblage verticality in wooded biomes across the globe.
-We first examining the vegetative dimension to Humbolt's enigma by modeling relationships between species richness and canopy height in forests across latitudes and hypothesize that tropical forests will have greater richness than temperate or boreal forests for a given canopy height.
-We then examine the mechanistic processes driving this pattern.
-If vertical stratification influences species richness patterns, we expect the use or arboreal habitats by a disproportionately large number of species to be associated with high richness.
-We therefore begin by examining correlations between assemblage verticality and species richness.
-If verticality is positively associated with species richness, then mechanisms limiting verticality will limit richness.
-To examine these mechanisms, we model verticality of the four vertebrate groups by macroclimate and vegetative structure and investigate the microclimatic processes that underlie these macroclimatic patterns.
-While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we model vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.
-
-But only within the last decade have we been able to measure forest structure and acquire data on vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive understanding of the impact of vertical habitat structure on species richness patterns.
-
-In forests, species assemblages stratify vertically along resource and microclimate gradients that extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition habitat across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height.
+We then examined the mechanistic processes driving this pattern.
+If vertical stratification contributes to the capacity for tall forests to support diverse assemblages, we expect that species rich assemblages will exhibit high verticality (i.e., a high average vertical niche position) and that environmental factors limiting verticality will limit species richness.
+To test this hypothesis, we examined correlations between assemblage verticality and species richness.
+We then modeled assemblage verticality of the four vertebrate classes by macroclimate and vegetative structure and investigated the microclimatic processes that underlie these macroclimatic patterns.
+While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we modeled vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.
 
@@
 
- [ - While this pattern could be driven by the latitudinal richness gradient]
-
 However, the described patterns are broadly consistent among taxa and mirror findings that for a given area, mountains have higher richness in the tropics [@rahbekHumboldtsEnigmaWhat2019].
@@
 
-  - TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE REGION
- - on second thought, I don't think this is necessary because we want to know why more species can't exist in trees in boreal regions
-
-# The vertical structure of vertebrate assemblages
-The vertical layering of species within forests may play a key role in the capacity for tall forests to support rich assemblages.
+# The mediating effect of verticality
+The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages.
 By vertically stratifying across canopy layers, arboreal species can fill the larger volume of habitat offered by tall forests, increasing the richness of the assemblage [cite].
-Within wooded ecosystems, we thus expect species rich assemblages to have a higher proportion of arboreal species and thus assemblages with vertical distributions of richness skewed toward the canopy [@oliveiraVerticalStratificationInfluences2019].
+Within wooded ecosystems, we therefore expect species rich assemblages to exhibit high verticality, or a skew in the vertical distribution of species toward occurring in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
 To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@scheffersVerticalArborealityHorizontal2017].
@@
 Though temperate and boreal regions have tall trees, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, consisting of greater proportions of fossorial and terrestrial species.
-This pattern suggests a reduction in the fine-scale partitioning of canopies relative to tropical regions, and thus a constraint on the capacity for taller canopies to generate biodiversity via the vertical layering of species.
+This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
 
-Environmental factors constraining verticality or the fine-scale partitioning of niche space will limit the capacity for tall forests to generate biodiversity.
+Environmental factors that constrain the evolution of diverse arboreal assemblages are therefore key to unraveling the vertical dimension to Humbolt's enigma.
 
-Dry (and cold) habitats have lower vertical complexity. Less layering = fewer opportunities for niche speciailizes = lower richness
-Though climate may still limit arobreality at higher latitudes, higher structural complexity and vegetation density across forest profiles in select extra-tropical regions increases verticality and consequently species richness.
-
-where verticality declines, species may use terrestrial habitats during the dry season to maintain moisture requirements assemblage verticality declines because species exhibit semi-arboreal life strategies, which reduces assemblage verticality.
-
-"short tropical forests show higher complexity relative to other forests of similar heights" - de Conto et al. 2023
+constraining verticality or the fine-scale partitioning of niche space will limit the capacity for tall forests to generate biodiversity.
 
@@
 
+Depends on the distribution of vegetation as well as the spatial and temporal variation of climate in the canopy.
+
 ## Climate produces structural constraints to verticality
@@
 
-physiological tolerance hypothesis
+Spatial and temporal climate variability from macro to micro scales strongly impact macroecological patterns (Klinges and Scheffers 2021; Rahbek)  [CITE].
+Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019].
+By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce steep vertical microclimate gradients [THAT CAN CHANGE AT RATES OF UP TO XXX PER METER] and therefore capture a remarkable volume of climatic variation given their geographic footprint.
+For example, canopies experience drier conditions and hotter maximum temperatures than understories across latitudes, while temperate and boreal canopies additionally experience colder minimum temperatures [FIGURE AND SUPP FIGURES WITH ADDED DETAIL].
+When macroclimate conditions producing canopy microclimates favorable for arboreal life, species may partition climate space across the entire forest volume.
+For example, high dry season precipitation and thus low precipitation seasonality in tropical moist forests support arboreal life by limiting seasonal water stress in the canopy and allow arboreal animals to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014] (Fig.
+\@ref(fig:fig3)).
+Low physiological stress allows high rates of turnover across the vertical gradient that yields a dense packing of species in the canopy that increases assemblage verticality and richness [CITE].
 
-In regions that experience low dry season precipitation, drier conditions in the canopy relative to the understory truncate the availability of vertical habitat space.
+In contrast, unfavourable canopy climates may reduce the number of viable physiological strategies thereby limiting the extent to which taller canopies can support increasing richness.
+For example, low dry season precipitation, which inhibits verticality (Fig.
+\@ref(fig:fig3)), exaggerates water stress in canopies and may prevent fully arboreal species from occurring.
+In tropical dry forests, animals seasonally adjust habitat use, shifting from using primarily arboreal habitats during the wet season to more humid terrestrial habitats during the dry season to avoid desiccation (Basham and Scheffers 2020; Rajaonarivelo et al.
+2021).
+In temperate and boreal biomes, low precipitation combined with dry canopy microclimates may similarly constrain ectotherm arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses (Sunday et al.
+2014; Tracy, Christian, and Tracy 2010).
+Harsh macroclimate conditions therefore seasonally truncate the volume of habitat that can support arboreal species, which collapses vertically stratified assemblages and reduces opportunities for fine-scale partitioning of the entire vertical climate gradient.
+[And flattens rates of turnover as height increases??] This prevents taller canopies from supporting the addition of arboreal species to assemblages.
+
+PHYSIOLOGICAL TOLERANCE HYPOTHESIS - REDUCE NUMBER OF FUNCTIONAL TYPES THAT CAN INHABIT A CLIMATE SPACE, LIMITING THE EXTENT TO WHICH AN INCREASE IN CANOPY HEIGHT INCREASES SPECIES RICHNESS
+ENVIRONMENTAL FILTERING IN ARBOREAL COMMUNITIES LIMITS ARBOREAL DIVERSITY IN HARSHER MACROCLIMATE CONDITIONS
+
+Harsh macroclimate conditions place environmental filters on viable physiologies for arboreal species, effectively truncating the viable functional space and therefore the amount of habitat volume that can be occupied.
+[https://nsojournals.onlinelibrary.wiley.com/doi/full/10.1111/ecog.07122]
+
+Temporal climate variability also plays a key role in the capacity for verticality to mediate biogeographic patterns by impacting climatic distinctness.
+[relationships between species richness and canopy height by impacting climatic distinctness].
+In montane regions, climatic distinctness across elevations declines with latitude due to increasing seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019; @janzenWhyMountainPasses1967].
+Narrow and distinct climate bands in the tropics produce high rates of turnover that allow tropical mountains to generate extraordinary levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
+In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and high levels of biodiversity relative to their tropical counterparts [@rahbekHumboldtsEnigmaWhat2019].
+
+Our models suggest that a similar mechanism may contribute to the capacity for tropical canopies to support assemblages with higher verticality and species richness than temperate canopies of equivalent heights.
+At the macroclimate scale, spatial models showed mixed conclusions regarding the role of temperature seasonality.
+However, at the microclimate scale, daily thermal variation in the canopy is greater in high latitude forests.
+Moreover, the increase in thermal variability from the ground to the canopy was steeper in temperate and boreal regions.
+Therefore, temperate and boreal canopies canopies are less likely to contain distinct vertical zones that support unique communities even in regions with climate conditions that physiologically support arboreal life.
+
+- climate variation increases vertical range size of species - Leahy
+
+- Low spatial and temporal variation across vertical strata in the tropics supports specialization and stratification, while greater spatial and temporal variability at higher latitudes brings animals into contact with more climates, thereby increases range size and reducing opportunities for specialization that promote speciation.
+
+indicated that temperature seasonality, which mirrored impacts of minimum temperatures when supplemented in the models given their strong negative correlation (Supplementary material), had mixed effects on verticality.
+[HOWEVER, WHEN RESTRICTING THE ANALYSIS TO FORESTED BIOMES, EXCLUDING MORE SPARSELY VEGETATED REGIONS, NEGATIVE IMPACTS OF TEMPERATURE SEASONALITY BECAME APPARENT].
+
+- Exposure to climate variation interacts with the steepness of the vertical gradient to influence how finely species partition niche space. 
+- Higher thermal variability reduces vertical zonation of temperate and boreal canopies
+- greater temporal climate variation increases the range of conditions that animals are exposed to
+- being exposed to a large range of conditions increases range size and climate niche of a species
+- larger range size and climate niche reduce the number of species that can coexist
+- Higher temporal thermal variation reduces species richness by increasing niche size
 
@@
 2014; Tracy, Christian, and Tracy 2010).
+
+physiological tolerance hypothesis
 
@@
 
-- Bird richness weighted toward the ground in higher latitude forests (Basham et al. 2023)
-
 Yet temperate regions that equal the height and complexity of tropical areas contain lower verticality, which may contribute to lower richness.
 Direct constraints of climate on animal physiology may contribute to this pattern.
-
-Niche specilization
 
@@
 2) exaggerated spatial and temporal variability of canopy microclimates reduces opportunities for specialization that would otherwise yield high richness
-
-[temperature seasonality exaggerated in the canopy relative to the ground] - would increase range size of arboreal species and therefore reduce opportunities for specialization and high richness
-- climate stratification - more offset in more complex forests in the tropics? - not consistently supported by our microclim models
-
-- Structurally complex forests capture a remarkable volume of climate space given their geographic footprint, bringing distinct climates within close proximity of each other.
-But the climate space is truncated at higher latitudes and in drier forests due to physiological constraints.
-
-- Low spatial and temporal variation across vertical strata in the tropics supports specialization and stratification, while greater spatial and temporal variability at higher latitudes brings animals into contact with more climates, thereby increases range size and reducing opportunities for specialization that promote speciation.
 
@@
 
+```{r fig3, fig.id="fig3", fig.cap="Coefficients of standardized predictors for spatial LMMs of verticality as a function of environmental factors.
+Models were conducted for each vertebrate class.
+Error bars represent 95% confidence intervals.
+Filled circles with solid error bars have confidence intervals that do not overlap zero.
+Open circles with dashed error bars have confidence intervals that overlap zero.
+Tmin~cold~ = minimum temperature of the coldest month; Tmax~warm~ = maximum temperature of the warmest month; Temp~diu~ = diurnal temperature range; Precip~dry~ = precipitation of the driest quarter; Precip~warm~ = precipitation of the warmest quarter.", fig.width=140, fig.height=120}
+
+# Fig. 3
+
+"figures/main_figs/fig3_sdmTMB_coefs.png"
+
+```
+
+THINGS TO FIX IN FIG4
+- diurnal temp range is for warmest month in strata plot and coldest month in profile plot - make consistent
+- replace map with concept figure similar to Rahbek
+- should we add strata plots for other variables? Or add to supplement - probable supplement
+- text sizing
+- point size in map if we keep the map
+```{r fig4, fig.id="fig4", fig.cap=", fig.width=140, fig.height=120}
+
+# Fig. 4
+
+"figures/ms_figures/fig4_profiles.png"
+
+```
 # Supplementary figures
@@
 ```
+
+# SCRATCH SPACE
+  - TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE REGION
+ - on second thought, I don't think this is necessary because we want to know why more species can't exist in trees in boreal regions
+
+in wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019].
+Taller canopies in tropical regions could thus support higher animal verticality and richness.
+In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
+Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.
+
+Integrating new technologies can shed light on the mechanisms driving decades-old ecological patterns.
+
+But only within the last decade have we been able to measure forest structure and acquire data on vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive understanding of the impact of vertical habitat structure on species richness patterns.
+
+In forests, species assemblages stratify vertically along resource and microclimate gradients that extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
+Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition habitat across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+
+where verticality declines, species may use terrestrial habitats during the dry season to maintain moisture requirements assemblage verticality declines because species exhibit semi-arboreal life strategies, which reduces assemblage verticality.
+
+"short tropical forests show higher complexity relative to other forests of similar heights" - de Conto et al. 2023
 

### 2025-07-07 scripts/01_get_data/01_data_download.R
model updates with canopy height and diurnal temp range. Figure updates

@@ -8,10 +8,10 @@ library(rerddap)
 library(tidyverse)
 
 # REFERENCE PROJECTION MAP
-mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
-             crs = "+proj=cea +datum=WGS84")
-res(mapa) <- 111000
-values(mapa) = 1
+# mapa <- rast(xmin = -20592508, xmax = 20588492, ymin = -5743602, ymax = 6573398,
+#              crs = "+proj=cea +datum=WGS84")
+# res(mapa) <- 111000
+# values(mapa) = 1
 
 ## CHELSA V2.1 bioclimate data
 bio_layers = c(1,2,4,5,6,12:18)
@@ -142,9 +142,9 @@ curl::curl_download(f, dest)
 ## Nat Ecol Evol 7, 1778–1789 (2023). https://doi.org/10.1038/s41559-023-02206-6
 ## Downloaded from google earth engine where I aggregated data to 1km resolution
 
-canopy_rr1 = rast("data/original/canopy_height/coarsened/canopy_height_reduceRes_1km_1.tif")
-canopy_rr2 = rast("data/original/canopy_height/coarsened/canopy_height_reduceRes_1km_2.tif")
-canopy_height = mosaic(canopy_rr1, canopy_rr2, filename = "data/original/canopy_height/coarsened/canopy_height_reduceRes_mosaic.tif")
+# canopy_rr1 = rast("data/original/canopy_height/coarsened/canopy_height_reduceRes_1km_1.tif")
+# canopy_rr2 = rast("data/original/canopy_height/coarsened/canopy_height_reduceRes_1km_2.tif")
+# canopy_height = mosaic(canopy_rr1, canopy_rr2, filename = "data/original/canopy_height/coarsened/canopy_height_reduceRes_mosaic.tif")
 
 # elevation
 browse("srtm30plus_v11_land")

### 2025-07-07 scripts/01_get_data/02_get_env_data.R

@@ -169,13 +169,15 @@ writeRaster(precip_warm.f, "data/derivative_data/resampled_env_rasters_50km/chel
 
 # canopy height
 # canopy height from Lang et al. 2023 https://doi.org/10.1038/s41559-023-02206-6
-canopy_height = rast("data/original/canopy_height/coarsened/canopy_height_reduceRes_mosaic.tif")
-crs(canopy_height) = "+init=epsg:4326" # set crs because it was not being read correctly leading to warnings when projecting to cea
+# This layer represents canopy height after masking out non-woody and human-altered habitats
+# based on Hansen et al. 2022 land cover.
+# then aggregating and averaging to coarser resolution
+canopy_height = rast("data/derivative_data/resampled_env_rasters_50km/canopy_height_lang2023/canopy_height_agg_ForestOnly.tif")
 canopy_height = crop(canopy_height, ext(-180, 180, -64.7937, 90))
 canopy_height = project(canopy_height, "+proj=cea +datum=WGS84")
 canopy_height = resample(canopy_height, mapa)
 names(canopy_height) = "canopy_height"
-writeRaster(canopy_height, "data/derivative_data/resampled_env_rasters_50km/canopy_height.tif", overwrite = T)
+writeRaster(canopy_height, "data/derivative_data/resampled_env_rasters_50km/canopy_height_lang2023/canopy_height_agg_ForestOnly_cea.tif", overwrite = T)
 
 # vegetation density
 # Crowther et al. 2015 https://doi.org/10.1038/nature14967
@@ -206,27 +208,42 @@ clim_velocity = project(clim_velocity, mapa)
 names(clim_velocity) = "clim_velocity"
 writeRaster(clim_velocity, "data/derivative_data/resampled_env_rasters_50km/clim_velocity.tif", overwrite = T)
 
+
+
+# add biorealms -----------------------------------------------------------
+
 # realms
 # from Dinnerstein et al. 2017 https://doi.org/10.1093/biosci/bix014
 biorealm = vect("data/original/env_data/ecoregions/Ecoregions2017.shp")
+  
 biorealm = project(biorealm, "+proj=cea datum=WGS84")
 realm = rasterize(biorealm, mapa, field = "REALM")
 realm = subst(realm, "N/A", NA)
 names(realm) = "realm"
[…]

### 2025-07-07 scripts/01_get_data/04_get_sp_and_cell_data_forestsOnly.R

@@ -41,6 +41,9 @@ world.rast = mask(world.rast, world.cea)
 world.df = as.data.frame(world.rast, xy = T) %>% 
   dplyr::select(x,y)
 
+wd = vect("data/original/World_Continents/World_Continents.shp")
+wd = project(wd, "+proj=cea +datum=WGS84")
+
 # load env data
 env = read.csv("data/derivative_data/env_data_50km.csv")
 
@@ -184,6 +187,16 @@ env = env %>% dplyr::select(x,y, precip_warm)
 griddat = left_join(griddat, env, by = c("x", "y"))
 fwrite(griddat, "data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv", row.names = F)
 
+griddat = fread("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+envsub = env %>% dplyr::select(x,y, canopy_height2)
+griddat = left_join(griddat, envsub, by = c("x", "y"))
+fwrite(griddat, "data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv", row.names = F)
+
+envsub = env %>% dplyr::select(x,y,fhd)
+griddat = left_join(griddat, envsub, by = c("x", "y"))
+fwrite(griddat, "data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv", row.names = F)
+
+
 # # error here because I didn't use forest env
 # tic()
 # amph_comdat = get_gridcell_vert_parallel(trait_dat = amph_spdat, occ = occr, env = env, rich_min = 5, ncore = 7, nsim = 100)
@@ -298,6 +311,17 @@ env = env %>% dplyr::select(x,y, precip_warm)
 griddat = left_join(griddat, env, by = c("x", "y"))
 fwrite(griddat,  "data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv", row.names = F)
 
+
+griddat = fread("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
+envsub = env %>% dplyr::select(x,y, canopy_height2)
+griddat = left_join(griddat, envsub, by = c("x", "y"))
+fwrite(griddat, "data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv", row.names = F)
+
+envsub = env %>% dplyr::select(x,y,fhd)
[…]

### 2025-07-07 scripts/01_get_data/sandbox.R

@@ -1,0 +1,239 @@
+library(data.table)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(terra)
+library(tidyterra)
+library(colorspace)
+
+env = read.csv("data/derivative_data/env_data_50km.csv")
+head(env)
+wd = vect("data/original/rnaturalearth_world.shp") %>% 
+  project("epsg:4326") %>% 
+  crop(ext(-180,180,-60,83.64))
+
+env = env %>% drop_na(-fhd.mean)
+hist(env$veg_den)
+ggplot(env, aes(veg_den, fhd.mean)) +
+  geom_point(pch = ".") +
+  geom_smooth(method = "lm")
+cor.test(env$fhd.mean, env$veg_den)
+
+# forested biomes only
+env.for = env %>% 
+  filter(grepl("Forest", biome)) %>% 
+  filter(!grepl("Med", biome))
+env.for.r = rast(env.for, crs = "+proj=cea +datum=WGS84")
+
+ggplot() +
+  geom_spatvector(data = wd, fill = "gray", color = NA) +
+  geom_spatraster(data = env.for.r, aes(fill = veg_den)) +
+  scale_fill_continuous_sequential(palette = "viridis", na.value = NA) +
+  theme_void()
+
+# calculate quantiles of vegetation density in forests per realm
+forden = env.for %>% 
+  group_by(realm) %>% 
+  summarise(vegden_min = min(veg_den),
+            vegden05 = quantile(veg_den, prob = 0.05),
+            vegden10 = quantile(veg_den, prob = 0.1))
+
+
+
+# Min vegden --------------------------------------------------------------
+
+env.denmin = env %>% 
+  left_join(forden, by = "realm") %>% 
+  filter(veg_den >= vegden_min)
+
+env.denmin.r = rast(env.denmin, crs = "+proj=cea +datum=WGS84")
+
+ggplot() +
+  geom_spatvector(data = wd, fill = "gray", color = NA) +
+  geom_spatraster(data = env.denmin.r, aes(fill = veg_den)) +
+  scale_fill_continuous_sequential(palette = "viridis", na.value = NA) +
+  theme_void()
+
+
+# Veg den05 ---------------------------------------------------------------
+
+# filter out veg den below 5th quantile of forest per realm
+
+env.den05 = env %>% 
[…]

### 2025-07-07 scripts/03_analyses/01_canopyheight_richness_relationships.R

@@ -585,3 +585,52 @@ maxrichcell = df %>%
 ratedifs = left_join(ratedifs, maxrichcell, by = c("taxa", "biome2")) %>% 
   mutate(relative_increase = dif/maxrich * 100,
          percent_increase = dif/h15*100)
+
+
+
+
+
+# SCRATCH SPACE -----------------------------------------------------------
+
+df2 = df %>% filter(rich >= 5)
+
+amph2 = df2 %>% filter(taxa == "Amphibians")
+
+m1 = lm(rich ~ vert.mean.ses, data = amph2)
+m1.resid = resid(m1)
+plot(amph2$canopy_height2, m1.resid)
+plot(amph2$precip_dry, m1.resid)
+ggplot(amph2, aes(log(precip_dry +1), m1.resid)) + geom_point() + geom_smooth(method = "lm")
+ggplot(amph2, aes(canopy_height2, m1.resid)) + geom_point() + geom_smooth(method = "lm")
+ggplot(amph2, aes(tmax_warm, m1.resid)) + geom_point() + geom_smooth(method = "lm")
+ggplot(amph2, aes(tmin_cold, m1.resid)) + geom_point() + geom_smooth(method = "lm")
+ggplot(amph2, aes(vert.mean.ses, rich)) + 
+  geom_point(alpha = 0.05) + 
+  geom_smooth(method = "glm",  method.args = list(family = "poisson"))
+
+# so it looks like precip and canopy height drive verticality, which influences richness
+# therefore, verticality mediates relationships between canopy height and richness
+# and factors limiting verticality will limit richness
+# but temperature influences richness independently of verticality
+
+m2 = lm(rich ~ vert.mean.ses + canopy_height2, data = amph2)
+summary(m2)
+
+m3 = lm(rich ~ canopy_height2, data = amph2)
+summary(m3)
+
+m1 = MASS::glm.nb(rich ~ vert.mean.ses, data = amph2)
+m1.resid = residuals(m1, type = "deviance")
+amph2$m1.resid = m1.resid
+ggplot(amph2, aes(canopy_height2, m1.resid)) +
+  geom_point(alpha = 0.05) +
+  geom_smooth(method = "lm")
+
+mod.resid = lm(m1.resid ~ canopy_height2, data = amph2)
+summary(mod.resid)
+
+cor.test(m1.resid, amph2$canopy_height2)
+
+
+
+

### 2025-07-07 scripts/03_analyses/02_vertical_microclim_models/04_plot_profiles.R

@@ -1,249 +1,0 @@
-# Plot boxplots of vertical microclimate profiles
-# Then consider running an NMDS to look at bioclim at different latitudes and iin canopy and ground
-
-library(ggthemes)
-library(dplyr)
-library(ggplot2)
-library(foreach)
-library(data.table)
-library(tidyr)
-
-# Load in modeling data
-micromods = list.files("results/microclim_models/", recursive = T, full.names = T)
-micromods.tmin = grep("mintemp", micromods, value = T) # points for the coldest month of the year
-micromods.tmax = grep("tmax", micromods, value = T) # points for the warmest month of the year
-indat = list.files("D:/projects/verticality_future/data/derivative_data/micropoint_prep/", full.names = T) # gedi points used for models
-
-# dataframe for month with coldest min temp
-mm.tmin = foreach(f = 1:length(micromods.tmin), .combine = rbind) %do% {
-  fi = fread(micromods.tmin[f])
-  nm = stringr::str_split_i(dirname(micromods.tmin[f]), "/", 3)
-  fi$region = nm
-  fi
-}
-
-# dataframe for month with hottest max temp
-mm.tmax = foreach(f = 1:length(micromods.tmax), .combine = rbind) %do% {
-  fi = fread(micromods.tmax[f])
-  nm = stringr::str_split_i(dirname(micromods.tmax[f]), "/", 3)
-  fi$region = nm
-  fi
-}
-
-# dataframe for PAD
-pad = foreach(f = 1:length(indat), .combine = rbind) %do% {
-  dat = readRDS(indat[f])
-  gedi.tmin = dat$gedisamp.mcold
-  gedi.tmax = dat$gedisamp.mhot
-  gedi = rbind(gedi.tmin, gedi.tmax, fill = T)
-  cnames = grep("^pavd", names(gedi), value = T)
-  cnames = c("shot_num", "rh_100_a0", "lat_lm_a0", "lon_lm_a0", grep("[0-9]$", cnames, value = T))
-  gedi = gedi[, ..cnames]
-  gedi[gedi==0] <- NA # convert zeros to NAs
-  gedi = gedi %>% 
-    pivot_longer(cols = 5:ncol(gedi), names_to = "height", values_to = "pavd") %>% 
-    drop_na()
-  
-  nm = basename(indat[f])
-  nm = gsub("prep_micropoint_", "", nm)
-  nm = gsub(".rds", "", nm)
-  gedi$region = nm
-  gedi
[…]

### 2025-07-07 scripts/03_analyses/02_vertmaps_and_vertRichness_correlations.R

@@ -1,0 +1,270 @@
+library(tidyverse)
+library(png)
+library(ggplot2)
+library(patchwork)
+library(terra)
+library(data.table)
+library(tidyterra)
+library(colorspace)
+library(grid)
+#library(tricolore)
+library(lemon)
+
+
+# load data ----------------------------------------------------
+
+wd = vect("data/original/rnaturalearth_world.shp") %>% 
+  crop(ext(-180,180,-60,83.64)) %>% 
+  project("+proj=robin +datum=WGS84") 
+
+wd = vect("data/original/rnaturalearth_world.shp") %>% 
+  crop(ext(-180,180,-60,83.64)) %>% 
+  project("+proj=cea +datum=WGS84") 
+
+ymin = ext(wd)[3]
+ymax = ext(wd)[4]
+
+amph = fread("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
+  dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na() %>% 
+  rast(crs = "+proj=cea +datum=WGS84") %>% 
+  project("+proj=robin +datum=WGS84") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(taxa = "Amphibians")
+
+mammals = fread("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
+  dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na() %>% 
+  rast(crs = "+proj=cea +datum=WGS84") %>% 
+  project("+proj=robin +datum=WGS84") %>% 
+  as.data.frame(xy = T) %>% 
+  mutate(taxa = "Mammals")
+
+birds = fread("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
+  dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
[…]

### 2025-07-07 scripts/03_analyses/03_glmms/01-2_sdmTMB_amphibians.R

@@ -1,0 +1,1177 @@
+# Fit models of verticality using sdmTMB (SECOND ROUND OF MODELS USING FHD AND REVISED CANOPY HEIGHT)
+# Step 1: compare random effects structure using cross validation, reml = T, include all fixed effects
+# Step 2: compare fixed effects structure for different hypotheses regarding environmental predictors using cross validation, reml = F
+# Step 3: refit the best model using all data
+# Step 4: predict the best model to future env data
+
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+library(car)
+library(data.table)
+library(DHARMa)
+source("scripts/00_functions/manuscript_functions.R")
+source("scripts/00_functions/00_plot_functions.R")
+
+# amphibian forest only 50km resolution
+raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
+
+
+# remove cells with fewer than five species
+raw = raw %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na(!fhd) %>% 
+  dplyr::select(!canopy_height)
+
+# plot relationship between SES verticality and env predictors
+raw %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm) %>%
+  pivot_longer(cols = 3:17, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
+dat.t = raw %>% 
+  mutate(log_clim_velocity = log10(clim_velocity),
+         log_tmin_cold = log10(tmin_cold + 50),
+         log_precip_dry = log10(precip_dry + 1),
+         log_precip_wet = log10(precip_wet))
+
+dat.t %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm, log_clim_velocity, log_precip_dry, 
+                log_tmin_cold, log_precip_wet) %>%
[…]

### 2025-07-07 scripts/03_analyses/03_glmms/01-2_sdmTMB_birds.R

@@ -1,0 +1,332 @@
+# Fit models of verticality using sdmTMB (SECOND ROUND OF MODELS USING FHD AND REVISED CANOPY HEIGHT)
+# Step 1: compare random effects structure using cross validation, reml = T, include all fixed effects
+# Step 2: compare fixed effects structure for different hypotheses regarding environmental predictors using cross validation, reml = F
+# Step 3: refit the best model using all data
+# Step 4: predict the best model to future env data
+
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+library(car)
+library(data.table)
+library(DHARMa)
+source("scripts/00_functions/manuscript_functions.R")
+source("scripts/00_functions/00_plot_functions.R")
+
+# amphibian forest only 50km resolution
+raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
+
+
+# remove cells with fewer than five species
+raw = raw %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na(!fhd) %>% 
+  dplyr::select(!canopy_height)
+
+# plot relationship between SES verticality and env predictors
+raw %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm) %>%
+  pivot_longer(cols = 3:17, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
+dat.t = raw %>% 
+  mutate(log_clim_velocity = log10(clim_velocity),
+         log_tmin_cold = log10(tmin_cold + 50),
+         log_precip_dry = log10(precip_dry + 1),
+         log_precip_wet = log10(precip_wet))
+
+dat.t %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm, log_clim_velocity, log_precip_dry, 
[…]

### 2025-07-07 scripts/03_analyses/03_glmms/01-2_sdmTMB_mammals.R

@@ -1,0 +1,246 @@
+# Fit models of verticality using sdmTMB (SECOND ROUND OF MODELS USING FHD AND REVISED CANOPY HEIGHT)
+# Step 1: compare random effects structure using cross validation, reml = T, include all fixed effects
+# Step 2: compare fixed effects structure for different hypotheses regarding environmental predictors using cross validation, reml = F
+# Step 3: refit the best model using all data
+# Step 4: predict the best model to future env data
+
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+library(car)
+library(data.table)
+library(DHARMa)
+source("scripts/00_functions/manuscript_functions.R")
+source("scripts/00_functions/00_plot_functions.R")
+
+# amphibian forest only 50km resolution
+raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
+
+
+# remove cells with fewer than five species
+raw = raw %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na(!fhd) %>% 
+  dplyr::select(!canopy_height)
+
+# plot relationship between SES verticality and env predictors
+raw %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm) %>%
+  pivot_longer(cols = 3:17, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
+dat.t = raw %>% 
+  mutate(log_clim_velocity = log10(clim_velocity),
+         log_tmin_cold = log10(tmin_cold + 50),
+         log_precip_dry = log10(precip_dry + 1),
+         log_precip_wet = log10(precip_wet))
+
+dat.t %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
[…]

### 2025-07-07 scripts/03_analyses/03_glmms/01-2_sdmTMB_reptiles.R

@@ -1,0 +1,138 @@
+library(sdmTMB)
+library(lme4)
+library(tidyverse)
+library(terra)
+library(car)
+library(data.table)
+library(DHARMa)
+source("scripts/00_functions/manuscript_functions.R")
+source("scripts/00_functions/00_plot_functions.R")
+
+# reptiles forest only 50km resolution
+raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/reptiles_comdat.csv")
+v = vect("data/original/rnaturalearth_world.shp")
+v = project(v, "+proj=cea +datum=WGS84")
+
+
+# remove cells with fewer than five species
+raw = raw %>% 
+  # subset to richness >= 5
+  filter(rich >= 5) %>% 
+  filter(!is.na(vert.mean.ses)) %>% 
+  drop_na(!fhd) %>% 
+  dplyr::select(!canopy_height)
+
+# plot relationship between SES verticality and env predictors
+raw %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm) %>%
+  pivot_longer(cols = 3:17, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
+dat.t = raw %>% 
+  mutate(log_clim_velocity = log10(clim_velocity),
+         log_tmin_cold = log10(tmin_cold + 50),
+         log_precip_dry = log10(precip_dry + 1),
+         log_precip_wet = log10(precip_wet))
+
+dat.t %>% 
+  dplyr::select(vert.mean.ses, biome:tmin_cold, elev, veg_den, veg_complexity,
+                canopy_height2, fhd, precip_warm, log_clim_velocity, log_precip_dry, 
+                log_tmin_cold, log_precip_wet) %>%
+  pivot_longer(cols = 3:21, names_to = "var", values_to = "val") %>% 
+  ggplot(aes(x = val, y = vert.mean.ses, color = biome)) +
+  geom_point(pch = ".") +
+  facet_wrap(~var, scales = "free") +
+  theme_classic()
+
+# scale the data
+dat = dat.t %>%
+  mutate_at(.vars = vars(mat:tmin_cold, elev, veg_den, veg_complexity, 
[…]

### 2025-07-07 scripts/03_analyses/03_glmms/02_fig3_coef_plot.R

@@ -9,12 +9,12 @@ library(sdmTMB)
 
 # SES Verticality Model Coefficients --------------------------------------
 
-amph = readRDS("results/sdmTMB_models2/amphibians_sesvert.rds")
-birds = readRDS("results/sdmTMB_models2/birds_sesvert.rds")
-mammals = readRDS("results/sdmTMB_models2/mammals_sesvert.rds")
-repts = readRDS("results/sdmTMB_models2/reptiles_sesvert.rds")
+amph = readRDS("results/sdmTMB_models2.3/amphibians_sesvert_tempdiu.rds")
+birds = readRDS("results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
+mammals = readRDS("results/sdmTMB_models2.3/mammals_sesvert_tempdiu.rds")
+repts = readRDS("results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
 
-mods = list(amph$bestmod, birds$bestmod, mammals$bestmod, repts$bestmod)
+mods = list(amph$fullmod$modlist[[1]], birds$fullmod, mammals$fullmod, repts$fullmod)
 mods = lapply(mods, tidy, effects = "fixed", conf.int = T)
 mods[[1]]$taxa = "Amphibians"
 mods[[2]]$taxa = "Birds"
@@ -25,24 +25,26 @@ mods = bind_rows(mods)
 mods = mods %>% 
   filter(term != "(Intercept)") %>%
   mutate(taxa = factor(taxa, levels = c("Birds", "Mammals", "Reptiles", "Amphibians")),
-         term = case_when(term == "tmin_cold:canopy_height" ~ "Tmin[cold]:canopy height",
-                          term == "tmax_warm:canopy_height" ~ "Tmax[warm]:canopy height",
-                          term == "I(tmax_warm^2):canopy_height" ~ "Tmax[warm]\u00b2:canopy height",
-                          term == "log_precip_dry:canopy_height" ~ "log(Precip[dry]):canopy height",
-                          term == "precip_warm:canopy_height" ~ "Precip[warm]:canopy height",
-                          term == "log_clim_velocity" ~ "log(climate velocity)", 
+         term = case_when(term == "tmin_cold:canopy_height2" ~ "Tmin[cold]:canopy height",
+                          term == "tmax_warm:canopy_height2" ~ "Tmax[warm]:canopy height",
[…]

### 2025-07-07 scripts/03_analyses/04_vertical_microclim_models/03_run_micropoint.R

@@ -16,6 +16,7 @@ library(data.table)
 library(soiltexture)
 library(dplyr)
 library(tidyr)
+library(terra)
 soilparams = micropoint::soilparams
 
 # Function to run micropoint - part of microclim3D package, but package currently isn't functioning
@@ -107,7 +108,7 @@ run_micropoint = function(tme, gedi, climr, vegp, soil, elev, asp, slp, dtmc,
       paii = pai_z_season[[2]]
     }
     
-    # test if height is greater than 5m (errors when there is only one paiz value - not sure why)
+    # test if height is greater than 5m (errors when there are fewer than 10 nodes to estimate the vert profile)
     #print(pai_z)
     print(h)
     
@@ -228,7 +229,7 @@ run_micropoint = function(tme, gedi, climr, vegp, soil, elev, asp, slp, dtmc,
 # Run vertical microclimate profiles --------------------------------------
 
 # Run mintemp vertical profiles
-indat = list.files("D:/projects/verticality_future/data/derivative_data/micropoint_prep/", full.names = T)
+indat = list.files("D:/projects/verticality_future/data/derivative_data/micropoint_prep2/", full.names = T)
 # run vertical profiles in the coldest month
 for(i in 1:6) {
   dat = readRDS(indat[i])
@@ -258,7 +259,7 @@ for(i in 1:6) {
     method = "temporal_month",
     reqhgts = NA,
     vertpai_method = "pai",
-    fout = paste0("results/microclim_models/",nm,"/mintemp_cold.csv"))
+    fout = paste0("results/microclim_models2/",nm,"/mintemp_cold.csv"))
   
 }
 
@@ -291,7 +292,7 @@ for(i in 1:6) {
     method = "temporal_month",
     reqhgts = NA,
     vertpai_method = "pai",
-    fout = paste0("results/microclim_models/",nm,"/tmax_warm.csv"))
+    fout = paste0("results/microclim_models2/",nm,"/tmax_warm.csv"))
   
 }
 
@@ -300,7 +301,8 @@ for(i in 1:6) {
 # VISUALIZATION -----------------------------------------------------------
 
 # Load in modeling data
-micromods = list.files("results/microclim_models/", recursive = T, full.names = T)
+micromods = list.files("results/microclim_models2/", pattern = ".csv",
+                       recursive = T, full.names = T)
 micromods.tmin = grep("mintemp", micromods, value = T) # points for the coldest month of the year
 micromods.tmax = grep("tmax", micromods, value = T) # points for the warmest month of the year
 

### 2025-07-07 scripts/03_analyses/04_vertical_microclim_models/04_plot_profiles.R

@@ -1,0 +1,452 @@
+# Plot boxplots of vertical microclimate profiles
+# Then consider running an NMDS to look at bioclim at different latitudes and iin canopy and ground
+
+library(ggthemes)
+library(dplyr)
+library(ggplot2)
+library(foreach)
+library(data.table)
+library(tidyr)
+
+thm = theme(legend.text = element_text(size = 4),
+            legend.title = element_text(size = 6),
+            axis.text = element_text(size = 4),
+            axis.title = element_text(size = 6),
+            strip.text = element_text(size = 6))
+
+# Load in modeling data
+micromods = list.files("results/microclim_models2/", recursive = T, full.names = T)
+micromods.tmin = grep("mintemp", micromods, value = T) # points for the coldest month of the year
+micromods.tmax = grep("tmax", micromods, value = T) # points for the warmest month of the year
+indat = list.files("D:/projects/verticality_future/data/derivative_data/micropoint_prep/", full.names = T) # gedi points used for models
+
+# dataframe for month with coldest min temp
+mm.tmin = foreach(f = 1:length(micromods.tmin), .combine = rbind) %do% {
+  fi = fread(micromods.tmin[f])
+  nm = stringr::str_split_i(dirname(micromods.tmin[f]), "/", 3)
+  fi$region = nm
+  fi
+}
+
+# dataframe for month with hottest max temp
+mm.tmax = foreach(f = 1:length(micromods.tmax), .combine = rbind) %do% {
+  fi = fread(micromods.tmax[f])
+  nm = stringr::str_split_i(dirname(micromods.tmax[f]), "/", 3)
+  fi$region = nm
+  fi
+}
+
+# dataframe for PAD
+pad = foreach(f = 1:length(indat), .combine = rbind) %do% {
+  dat = readRDS(indat[f])
+  gedi.tmin = dat$gedisamp.mcold
+  gedi.tmax = dat$gedisamp.mhot
+  gedi = rbind(gedi.tmin, gedi.tmax, fill = T)
+  cnames = grep("^pavd", names(gedi), value = T)
+  cnames = c("shot_num", "rh_100_a0", "lat_lm_a0", "lon_lm_a0", grep("[0-9]$", cnames, value = T))
+  gedi = gedi[, ..cnames]
+  gedi[gedi==0] <- NA # convert zeros to NAs
[…]

### 2025-07-07 scripts/03_analyses/zz_retired/05_humbolts_enigma.R

@@ -1,3 +1,5 @@
+# Model richness as a function of canopy height
+
 library(tidyverse)
 library(Data4Ecologists)
 library(terra)
@@ -5,6 +7,11 @@ library(foreach)
 library(colorspace)
 library(tidyterra)
 library(MASS)
+source("scripts/03_analyses/00_tools/tools.R")
+
+wd = vect("data/original/rnaturalearth_world.shp") %>% 
+  project("epsg:4326") %>% 
+  crop(ext(-180,180,-60,83.64))
 
 amph = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
 birds = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
@@ -30,7 +37,14 @@ df = df %>%
  # drop_na(biome2) %>% 
   filter(rich > 5)
 
+# Explore some preliminary plots
 ggplot(df, aes(canopy_height, rich)) +
+  geom_point(pch = ".", alpha = 0.2) +
+  geom_smooth(method = "lm") +
+  facet_wrap(~taxa, scales = "free_y") +
+  theme_classic()
+
+ggplot(df, aes(log(canopy_height), log(rich))) +
   geom_point(pch = ".", alpha = 0.2) +
   geom_smooth(method = "lm") +
   facet_wrap(~taxa, scales = "free_y") +
@@ -48,93 +62,79 @@ ggplot(df, aes(canopy_height, vert.mean.ses, color = biome)) +
   facet_wrap(~taxa, scales = "free_y") +
   theme_classic()
 
-richness.ch.resids = foreach(i = unique(df$taxa)) %do% {
+
+
+# Model Richness by canopy height for each taxon globally -----------------
+# I initially try a poisson regression. These does not fit the data well for any
+# taxonomic group. Then I try a negabive binomial. This works much better.
+# and residuals look ok.
+
+richness.ch.resids.global = foreach(i = unique(df$taxa)) %do% {
   d = df %>% filter(taxa == i) %>% 
       drop_na(vert.mean.ses, rich, canopy_height) %>% 
       filter(rich > 5 & canopy_height > 0) %>% 
       mutate(log_ch = log(canopy_height))
   
   # model richness by canopy height
-  # m1 = glm(rich ~ log_ch,
-  #         data = d,
-  #         family = poisson())
+  # Poisson distribution does not fit data well
+  m1 = glm(rich ~ log(canopy_height),
[…]

### 2025-07-07 scripts/03_analyses/zz_retired/06_phylodistance/phylodistance.R

@@ -1,0 +1,327 @@
+library(picante)
+library(foreach)
+
+data(phylocom)
+names(phylocom)
+
+phy = phylocom$phylo
+commtest = phylocom$sample
+traits = phylocom$traits
+
+phy
+comm
+traits
+
+prunedphy = prune.sample(comm, phy)
+prunedphy
+
+par(mfrow = c(2,3))
+for (i in row.names(comm)) {
+  plot(prunedphy, show.tip.label = F, main = i)
+  tiplabels(tip = which(prunedphy$tip.label %in% names(which(comm[i,]>0))), pch = 19, cex = 2)
+}
+
+par(mfrow = c(2,2))
+for (i in row.names(comm)) {
+  plot(phy, show.tip.label = F, main = i)
+  tiplabels(pch=22, col = traits[,i], bg = traits[,i], cex = 1.5)
+}
+
+pd.result = pd(comm, phy, include.root = T)
+pd.result
+
+phydist = cophenetic(phy)
+ses.mpd.result = ses.mpd(comm, phydist, null.model = "taxa.labels", abundance.weighted = F, runs = 99)
+ses.mpd.result
+
+
+# Amphibians --------------------------------------------------------------
+tree = read.tree("data/original/phylotrees/amph_shl_new_Consensus_7238.tre")
+tree
+
+sp = read.csv("data/derivative_data/species_data/amph_spdat_moura_forestsOnly.csv")
+tree.sp = tree$tip.label
+head(sort(tree.sp))
+
+# community data matrix for amphibians
+comm = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat_occr.csv")
+
+# fix hyphens in column names
+colnames(comm)[which(colnames(comm) == "Pristimantis_w.nigrum")] = "Pristimantis_w-nigrum"
+colnames(comm)[which(colnames(comm) == "Scinax_x.signatus")] = "Scinax_x-signatus"
+
+#rownames(comm) = paste0(1:nrow(comm))
+commdat = as.matrix(comm)
+rownames(commdat) = paste0(commdat[,"x"], "_", commdat[,"y"])
+commdat = commdat[,-c(1,2)]
+griddat = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
+
+# moura and vertlife tree use same taxonomic backbone, so no need for phylomatcher
[…]

### 2025-07-07 scripts/04_plots/sdmTMB_marginaleffects.R

@@ -9,36 +9,34 @@ library(doParallel)
 library(tidyverse)
 library(data.table)
 
-amph = readRDS("results/sdmTMB_models2/predictions/amphibians_sesvert.rds")
-amphmod = amph$mod
-amphmod = sdmTMB:::reload_model(amphmod)
+amph = readRDS("results/sdmTMB_models2.3/amphibians_sesvert_tempdiu.rds")
+amphmod = amph$fullmod$modlist[[1]]
 
-birds = readRDS("results/sdmTMB_models2/predictions/birds_sesvert.rds")
-birdsmod = birds$mod
+birds = readRDS("results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
+birdsmod = birds$fullmod
 
-mammals = readRDS("results/sdmTMB_models2/predictions/mammals_sesvert.rds")
-mammalsmod = mammals$mod
+mammals = readRDS("results/sdmTMB_models2.3/mammals_sesvert_tempdiu.rds")
+mammalsmod = mammals$fullmod
 
-repts = readRDS("results/sdmTMB_models2/predictions/reptiles_sesvert.rds")
-reptsmod = repts$mod
+repts = readRDS("results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
+reptsmod = repts$fullmod
 
 range(birdsmod$data$log_precip_dry)
 range(birdsmod$data$tmax_warm)
 range(birdsmod$data$tmin_cold)
 range(birdsmod$data$canopy_height)
 range(birdsmod$data$veg_den)
-range(birdsmod$data$precip_wet)
 range(birdsmod$data$precip_warm)
-range(birdsmod$data$log_clim_velocity)
+range(birdsmod$data$temp_diu)
 
 var = list(
-  "log_precip_dry [-4.6:2.3 by=0.1]",
-  "tmax_warm [-3.6:2.7 by=0.1]",
-  "tmin_cold [-2.54:1.37 by=0.1]",
-  "canopy_height [-1.3:3.5 by=0.1]",
+  "log_precip_dry [-2.1:2.7 by=0.1]",
+  "tmax_warm [-3.8:2.9 by=0.1]",
+  "tmin_cold [-2.54:1.34 by=0.1]",
+  "canopy_height [-1.8:3.1 by=0.1]",
   "veg_den [-1.3:9.0 by=0.1]",
[…]

### 2025-07-09 manuscript.Rmd
text and fig2 revisions

@@
 In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
-The vertical nature of mountains provides more land surface per unit land area than flat regions and produces steep climate gradients over short distances that allow rapid rates of turnover.
-Combined, these factors generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
-Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctiveness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
-Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961].
-Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
+[Yet until recently, hypotheses failed to explain the extraordinary richness of mountains particularly in the tropics.] The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover.
+Combined, these factors allow mountains to generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
+Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
 
+NEED TO MENTION HUMBOLT'S ENIGMA SOMEWHERE IN FIRST PARAGRAPH AND THAT THE MAIN EXPLANATIONS ARE TOPOGRAPHY AND CLIMATE
+
+Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogegoraphic patterns [@macarthurBirdSpeciesDiversity1961].
 Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains.
@@
 This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
-Yet impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020] - a pattern resembling that of mountains.
+Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+- a vertical dimension to Humbolt's enigma
 
-Vegetative complexity and climate may play key roles in mediating the relationships between canopy height and species richness by impacting the availability of vertical niche opportunities physiologically and constraining arboreal lifestyles [CITE].
-Densely layered forest structures not only provide habitat across the forest profile, but produce vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed.
-From the humid and thermally stable understory to the dry and thermally variable canopy that experiences hot and cold temperature extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or hinder the capacity for tall forests to generate high levels of biodiversity.
-For example, increased exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat that can support arboreal species and thus limit richness despite tall canopies [@oliveiraVerticalStratificationInfluences2019].
-Moreover, relatively low thermal variability in the canopies of tropical forests may increase the distinctness of vertical climate bands that support high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics [CITE].
+Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
+Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages.
+For example, the denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [CITE or CONCEPT FIG].
+Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes.
+Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020] [CITE].
+Yet only recently have we been able to address these hypotheses, as our ability to measure vegetation structure and quantify its use by animals globally [AND MODEL MICROCLIMATES GLOBALLY] increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
 
-Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [CITE], the mechanisms that allow tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
-To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and assemblage verticality in wooded biomes across the globe.
+In this study we combine range maps of nearly 24,000 terrestrial vertebrate species and comprehensive databases of vertical habitat use, global macroclimate and forest structure data, and process-based microclimate models to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities.
+We first investigated the capacity for taller forests to support increasing species richness in forests and less densely vegetated wooded biomes across latitudes.
+If the vertical structure of species assemblages mediates the relationship between canopy height and species richness, then we expect positive correlations between the verticality of assemblages (i.e., their average vertical niche position) and species richness.
+[AND THAT CANOPY HEIGHT WILL EXPLAIN LITTLE VARIATION IN RICHNESS ONCE VERTICALITY HAS BEEN ACCOUNTED FOR]
+
+and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height.
+We then examined the mechanistic processes driving this pattern.
+EXPLANATION OF TETRAPODTRAITS AND ELTONTRAITS
+
+, including TetrapodTraits and EltonTraits, to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.
+
+To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.
 We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height.
@@
 
-To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally and mapped the residuals.
+To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally.
+We then mapped the residuals of these models to determine where canopy height over- or under-predicted species richness.
 We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.
@@
 The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages.
-By vertically stratifying across canopy layers, arboreal species can fill the larger volume of habitat offered by tall forests, increasing the richness of the assemblage [cite].
-Within wooded ecosystems, we therefore expect species rich assemblages to exhibit high verticality, or a skew in the vertical distribution of species toward occurring in arboreal habitats [@oliveiraVerticalStratificationInfluences2019].
-To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@scheffersVerticalArborealityHorizontal2017].
+As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].
+An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017].
+Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019].
+To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017].
 For species occurring in multiple habitats, we averaged scores across habitats.
@@
 
-Verticality is strongly correlated with richness patterns, peaking in the tropics and declining toward higher latitudes.
-- At its highest, x % of species exhibited arboreal strategies, corresponding to some of the richest assemblages in tropical regions.
-- In these regions, expansion up the canopy combined with fine-scale niche partitioning, and thus species packing, within the canopy produce species rich assemblages.
+Verticality peaked in the tropics and declined toward the poles, exhibiting strong correlations with verticality across taxa (Fig.
+\@ref(fig:fig2)).
+Moist tropical forests contained assemblages comprised of x% to x% species exhibiting arboreal strategies (SUPP FIG), corresponding to some of the most species rich assemblages.
+Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies.
+This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
+Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.
 
-Though temperate and boreal regions have tall trees, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, consisting of greater proportions of fossorial and terrestrial species.
-This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
+# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities
+To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may constrain or support arboreal life strategies, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month, as well as with macroclimatic factors that may influence the distinctness of vertical climate banding, including diurnal temperature range (Supplementary methods).
+We additionally included vegetation density in the models, which impacts the availability of habitat volume.
+While foliage height diversity and temperature seasonality also impact niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height and minimum temperatures, respectively, and were therefore excluded from the models.
+The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018].
+While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact community structure [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020].
+We therefore additionally modeled vertical microclimate profiles in six forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy processes to predict vertical climate gradients [@macleanMicroclimfFastCanopy2024].
+To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within a 1° x 1° grid cell.
+By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.
 
-Environmental factors that constrain the evolution of diverse arboreal assemblages are therefore key to unraveling the vertical dimension to Humbolt's enigma.
-
-constraining verticality or the fine-scale partitioning of niche space will limit the capacity for tall forests to generate biodiversity.
-
-# Canopy Climates
-To examine the mechanisms contributing to variation in the capacity for tall forests to support highly arboreal, and thus species rich, assemblages, we use spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with climatic factors that may constrain or support arboreality, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month (Supplementary methods).
-We additionally include vegetation density in the models, as this will impact the availability of habitat volume, and did not include foliage height diversity due to its strong correlation with canopy height, as has previously been observed [@decontoCharacterizingStructuralComplexity2024; @atkinsPowerLawScaling2022].
-The climate variables included in these models represent macroclimate conditions in open habitats rather than the climate conditions experienced in forest canopies.
-We therefore additionally modeled vertical microclimate profiles in six lowland forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [] with ERA5 macroclimate data at a high temporal resolution [] in a process-based microclimate model that resolves within-canopy processes generating vertical climate gradients [].
-These models provide an estimate of climate conditions proximal to those experienced by forest-dwelling organisms and thus better represent the biogeographic and ecophysiological processes that allow tall forests to support vertical and rich assemblages in the tropics more so than at higher latitudes [@klingesProximalMicroclimateMoving2024].
-Modeling microclimates in conjuction with modeling verticality based on macroclimate conditions thus allows us to infer the microclimate processes that underlie macroclimate patterns in the absence of global 3D climate data.
-
-Our models indicate that taller canopies support highly vertical vertebrate assemblages, while dry, hot, and cold climates limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying extents across vertebrate classes (Fig.
-\@ref(fig:fig3).
-These patterns arise because the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.
-
-These patterns arise because macroclimate indirectly constrain verticality via its effect on vegetation structure and niche opportunities, while microclimate conditions produced by the interaction between macroclimate and vegetation structure place physiological constraints on arboreal lifestyles.
-the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.
-
-This occurs because climate indirectly constrains verticality via its effects on vegetation structure and niche opportunities for specialization across vertical strata and the interaction between macro- and micro-climate gradients directly constrain verticality....
-
-Climate mediates that capacity for forests to support vertical assemblages indirectly, through its effect on vegetation structure, and directly, via impacts on animal physiology.
-
-Interactions between macroclimate, vegetation structure, and microclimate impact verticality by affecting niche opportunities and imposing physiological constraints.
-
-Depends on the distribution of vegetation as well as the spatial and temporal variation of climate in the canopy.
+Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry, and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying degrees across vertebrate classes (Fig.
+\@ref(fig:fig3)).
+These patterns arise due to impacts of spatial climate variation on vegetation structure and physiological constraints to arboreal lifestyles as well as impacts of temporal climate variation on the distinctness of vertical climate bands.
 
@@
 Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014].
-High annual precipitation and low precipitation seasonality enable development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
-Wet climates additionally allow development of diverse epiphyte communities, which provide additional microhabitats in the canopy [CITE].
-Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding extraordinarily rich communities.
+High dry season precipitation and thus low precipitation seasonality may support highly vertical assemblages by enabling development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
+Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004].
+Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].
 In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig.
@@
 In evergreen needleleaf forests that dominate boreal regions, these environmental constraints dampen the rate of increase in structural complexity with canopy height [@atkinsPowerLawScaling2022].
-Tall canopies therefore provide fewer niche opportunities in boreal regions, which may constrain verticality and limit species richness relative to canopies of equivalent height in lower latitude forests.
-Yet, hotspots of structural complexity in temperate forests, including the North Pacific Alaskan coastal forest, Valdivian forest, and Tasmanian forest, that nearly match the complexity of tropical forests [@ehbrechtGlobalPatternsClimatic2021a] also contain assemblages with lower verticality than the tropics and fewer species than expected based upon their canopy height (Figs.
+Tall canopies therefore provide fewer niche opportunities in boreal regions, which may constrain assemblage verticality and limit species richness relative to canopies of equivalent height at lower latitudes.
+Yet, hotspots of structural complexity in temperate forests, including North Pacific Alaskan, Valdivian, and Tasmanian forests, that nearly match the complexity of tropical forests [@ehbrechtGlobalPatternsClimatic2021a] also contain assemblages with lower verticality than the tropics and fewer species than expected based upon their canopy height (Figs.
 \@ref(fig:fig1), \@ref(fig:fig2)).
-MacArthur & MacArthur [-@macarthurBirdSpeciesDiversity1961] observed this pattern over 50 years ago, noting that "tropical bird species diversity is much greater than the temperate one for habitats of comparable profiles." The effects of the interaction between macroclimate and vegetation profiles on vertical microclimate profiles provide an explanation [MIRRORING MECHANISMS THAT ALLOW MOUNTAINS TO BE BETTER GENERATORS OF BIODIVERSITY IN THE TROPICS]
+MacArthur & MacArthur [-@macarthurBirdSpeciesDiversity1961] observed this pattern over 50 years ago, noting that "tropical bird species diversity is much greater than the temperate one for habitats of comparable profiles." The interaction between macroclimate and vertical microclimate profiles provides an explanation that mirrors the climatic mechanisms allowing mountains to be better generators of biodiversity in the tropics [@rahbekHumboldtsEnigmaWhat2019].
 
-## Physiological constraints to arboreality truncate suitable habitat 
-## Canopy climates truncate suitable habitat in cold and dry forests
+## The influence of canopy climates
 
-Spatial and temporal climate variability from macro to micro scales strongly impact macroecological patterns (Klinges and Scheffers 2021; Rahbek)  [CITE].
-Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019].
-By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce steep vertical microclimate gradients [THAT CAN CHANGE AT RATES OF UP TO XXX PER METER] and therefore capture a remarkable volume of climatic variation given their geographic footprint.
-For example, canopies experience drier conditions and hotter maximum temperatures than understories across latitudes, while temperate and boreal canopies additionally experience colder minimum temperatures [FIGURE AND SUPP FIGURES WITH ADDED DETAIL].
-When macroclimate conditions producing canopy microclimates favorable for arboreal life, species may partition climate space across the entire forest volume.
-For example, high dry season precipitation and thus low precipitation seasonality in tropical moist forests support arboreal life by limiting seasonal water stress in the canopy and allow arboreal animals to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014] (Fig.
+Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019; @leahyVerticalNicheElevation2021; @vinodThermalSensitivityForest2023].
+By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a remarkable volume of climatic variation given their geographic footprint [@defrenneForestMicroclimatesClimate2021].
+These steep climate gradients can support high rates of turnover from the ground to the canopy that yield vertically stratified and species rich assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018].
+However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness, just as lower elevation snow lines limit the capacity for tall mountains to generate biodiversity at higher latitudes [@oliveiraVerticalStratificationInfluences2019].
+
+For example, our models indicate that low dry season precipitation constrains verticality across vertebrate classes (Fig.
 \@ref(fig:fig3)).
-Low physiological stress allows high rates of turnover across the vertical gradient that yields a dense packing of species in the canopy that increases assemblage verticality and richness [CITE].
+These macroclimate patterns were accompanied by steep gradients in minimum relative humidity, which declined x% from the ground to the top of the canopy (Fig.
+\@ref(fig:fig4)).
+Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot climates, where arboreal species depend on evapotranspiration to maintain body temperatures below critical thermal maxima, or in cold climates, where basking to increase body temperature can quickly lead to desiccation without adequate moisture availability or physiological defenses.
+By imposing environmental filters on the functional morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage verticality and limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [https://onlinelibrary.wiley.com/doi/full/10.1111/j.1461-0248.2004.00671.x].
+This pattern is evident for frog communities in the seasonally dry forests of Panama, where collapse of vertical stratification toward the ground during the dry season due to water limitation in the canopy is accompanied by richness declines and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020].
+Simply put, extreme canopy microclimates that physiologicall constrain arboreal life truncate the volume of available habitat, which prevents tall trees from supporting species rich assemblages.
+In contrast, high dry season precipitation in tropical moist forests supports highly vertical assemblages by reducing physiological constraints on arboreal life strategies in the upper canopy.
+Species are therefore able to finely partition the entire volume of forested habitat, leading to a dense packing of species in the canopy and allowing increases in canopy height to yield rapid increases in species richness (Fig.
+\@ref(fig:fig1)).
 
-In contrast, unfavourable canopy climates may reduce the number of viable physiological strategies thereby limiting the extent to which taller canopies can support increasing richness.
-For example, low dry season precipitation, which inhibits verticality (Fig.
-\@ref(fig:fig3)), exaggerates water stress in canopies and may prevent fully arboreal species from occurring.
-In tropical dry forests, animals seasonally adjust habitat use, shifting from using primarily arboreal habitats during the wet season to more humid terrestrial habitats during the dry season to avoid desiccation (Basham and Scheffers 2020; Rajaonarivelo et al.
-2021).
-In temperate and boreal biomes, low precipitation combined with dry canopy microclimates may similarly constrain ectotherm arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses (Sunday et al.
-2014; Tracy, Christian, and Tracy 2010).
-Harsh macroclimate conditions therefore seasonally truncate the volume of habitat that can support arboreal species, which collapses vertically stratified assemblages and reduces opportunities for fine-scale partitioning of the entire vertical climate gradient.
-[And flattens rates of turnover as height increases??] This prevents taller canopies from supporting the addition of arboreal species to assemblages.
+Hot and cold temperatures had less consistent effects across taxa, but may still play critical roles in physiological constraints, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures.
+For example, high maximum temperatures reduced verticality and/or [REDUCED THE POSITIVE IMPACT OF CANOPY HEIGHT ON VERTICALITY FOR REPTILES, MAMMALS, AND BIRDS] across vertebrate classes, while low minimum temperatures reduced positive effects of canopy height for birds and reptiles (Fig.
+\@ref(fig:fig3)).
+High maximum temperatures may cause ectotherm body temperatures to exceed critical thermal maxima (McKechnie et al.
+2017; Sunday et al.
+2014), particularly in drier regions where maximum temperatures peaked.
+Microclimate models additionally indicated that vertical maximum temperature gradients were steepest in the temperate forests, though canopy temperatures exceeding 40C may limit the capacity for species to stratify across the entire vertical climate gradient (Fig.
+\@ref(fig:fig4)).
+Similarly, low minimum temperatures can cause body temperatures to exceed critical thermal minima for ectotherms or impose energetic constraints to endotherms, particularly in the windier conditions that characterize the canopy  (Sunday et al.
+2014) (Byman, Hay, and Bakken 1988).
+Shallow vertical minimum temperature gradients (Fig.
+\@ref(fig:fig4)), particularly during leaf off conditions [CITE], cause environmental filters to act across the entire canopy, posing severe limitations on the capacity for taller canopies to increase species richness.
 
-PHYSIOLOGICAL TOLERANCE HYPOTHESIS - REDUCE NUMBER OF FUNCTIONAL TYPES THAT CAN INHABIT A CLIMATE SPACE, LIMITING THE EXTENT TO WHICH AN INCREASE IN CANOPY HEIGHT INCREASES SPECIES RICHNESS
-ENVIRONMENTAL FILTERING IN ARBOREAL COMMUNITIES LIMITS ARBOREAL DIVERSITY IN HARSHER MACROCLIMATE CONDITIONS
-
-Harsh macroclimate conditions place environmental filters on viable physiologies for arboreal species, effectively truncating the viable functional space and therefore the amount of habitat volume that can be occupied.
 [https://nsojournals.onlinelibrary.wiley.com/doi/full/10.1111/ecog.07122]
@@
 In montane regions, climatic distinctness across elevations declines with latitude due to increasing seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019; @janzenWhyMountainPasses1967].
-Narrow and distinct climate bands in the tropics produce high rates of turnover that allow tropical mountains to generate extraordinary levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
-In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and high levels of biodiversity relative to their tropical counterparts [@rahbekHumboldtsEnigmaWhat2019].
+Low seasonal variation in the tropics produces narrow and distinct climate bands that exhibit little overlap across elevations, yielding high rates of turnover that allow tropical mountains to generate extraordinary levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
+In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
 
 Our models suggest that a similar mechanism may contribute to the capacity for tropical canopies to support assemblages with higher verticality and species richness than temperate canopies of equivalent heights.
-At the macroclimate scale, spatial models showed mixed conclusions regarding the role of temperature seasonality.
-However, at the microclimate scale, daily thermal variation in the canopy is greater in high latitude forests.
+At the macroclimate scale, models of verticality generally indicated no significant impact of diurnal temperature variation, though low diurnal temperature variation increased verticality of mammals (Fig.
+\@ref(fig:fig3)).
+However, microclimate models indicated clear increases in daily temperature variation in higher latitude forests (Fig.
+\@ref(fig:fig4)).
 Moreover, the increase in thermal variability from the ground to the canopy was steeper in temperate and boreal regions.
-Therefore, temperate and boreal canopies canopies are less likely to contain distinct vertical zones that support unique communities even in regions with climate conditions that physiologically support arboreal life.
+Temperate and boreal canopies canopies therefore exhibited less distinct climate bands, with vertical zonation collapsing nearly completely during morning hours.
+Without unique climatic niches across the forest profile, these regions are unlikely to support rapid vertical turnover, limiting the number of species that may be added to the community given an increase in canopy height even in regions with climate conditions that physiologically support arboreal life.
+In contrast, tropical canopies maintained distinct climate bands throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
+Narrow and distinct vertical climate niches may therefore support rapid vertical turnover and increases in richness with canopy height.
+that contributes the capacity for tropical  unique communities
 
@@
 
-- Low spatial and temporal variation across vertical strata in the tropics supports specialization and stratification, while greater spatial and temporal variability at higher latitudes brings animals into contact with more climates, thereby increases range size and reducing opportunities for specialization that promote speciation.
+# Conclusion - bring back to Humbolt's engima
 
-indicated that temperature seasonality, which mirrored impacts of minimum temperatures when supplemented in the models given their strong negative correlation (Supplementary material), had mixed effects on verticality.
-[HOWEVER, WHEN RESTRICTING THE ANALYSIS TO FORESTED BIOMES, EXCLUDING MORE SPARSELY VEGETATED REGIONS, NEGATIVE IMPACTS OF TEMPERATURE SEASONALITY BECAME APPARENT].
+Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
 
-- Exposure to climate variation interacts with the steepness of the vertical gradient to influence how finely species partition niche space. 
-- Higher thermal variability reduces vertical zonation of temperate and boreal canopies
-- greater temporal climate variation increases the range of conditions that animals are exposed to
-- being exposed to a large range of conditions increases range size and climate niche of a species
-- larger range size and climate niche reduce the number of species that can coexist
-- Higher temporal thermal variation reduces species richness by increasing niche size
-
-Across vertebrate classes, low dry season precipitation limits verticality.
-While vertical gradients in humidity are steep in forests across latitudes, high dry season precipitation and thus lower precipitation seasonality in tropical moist forests provides a consistent source of water.
-Arboreal animals are therefore able to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014].
-In tropical dry regions the volume of habitat that can support a high diversity of arboreal lifestyles may shift seasonally.
-For example, birds and frogs that use primarily arboreal habitats during the wet season increase their use of more humid terrestrial habitats during the dry season to avoid desiccation (Basham and Scheffers 2020; Rajaonarivelo et al.
-2021).
-In many temperate and/or boreal biomes, low precipitation combined with dry canopy microclimates may similarly constrains amphibian arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses (Sunday et al.
-2014; Tracy, Christian, and Tracy 2010).
-
-physiological tolerance hypothesis
-
-We also observed consistently significant interactions between canopy height and extreme temperature variables.
-Canopies experience lower minimum temperatures and higher maximum temperatures than understories that are buffered from thermal extremes (De Frenne et al.
-2019).
-Particularly when combined with windier conditions in the canopy, cold temperatures can cause ectotherm body temperature to drop below critical thermal minima and impose energetic constraints to endotherms (Sunday et al.
-2014) (Byman, Hay, and Bakken 1988).
-Furthermore, high temperatures may cause body temperatures to exceed critical thermal maxima or cause desiccation due to evapotranspirative cooling when water is scarce (McKechnie et al.
-2017; Sunday et al.
-2014).
-When minimum temperatures are warmer and maximum temperatures are cooler, the effect of canopy height on verticality generally becomes more positive, likely because more benign thermal microclimate conditions reduce physiological constraints on arboreal life strategies.
-Therefore, canopy microclimates that impose physiological challenges to arboreal life strategies within prevailing macroclimate, truncate the volume of suitable habitat and prevent taller canopies from support a highly vertical and species rich assemblage.
-
-Yet temperate regions that equal the height and complexity of tropical areas contain lower verticality, which may contribute to lower richness.
-Direct constraints of climate on animal physiology may contribute to this pattern.
+Since verticality research was started, most research has been on impacts of vegetation structure.
+Yet no hypotheses had been proposed regarding inconsistent conclusions of the impacts of vegetation structure on richness.
 
@@
 1) physiological limitations and physiological tolerance hypothesis
-2) exaggerated spatial and temporal variability of canopy microclimates reduces opportunities for specialization that would otherwise yield high richness
+2) high temporal climate variabiltiy in high latitude canopies reduces vertical zonation, limiting vertical turnover and therefore richness
 
-# Conclusion - bring back to Humbolt's engima
 - Trees, just as mountains, bring different climate zones within close proximity of each [@rahbekHumboldtsEnigmaWhat2019].
@@
 
+# Supplementary material
+
+## Microclimate models
+
+We selected 6 1 x 1 degree grid cells, with 2 each in tropical, temperate, and boreal forests.
+We downloaded GEDI points that intersected with these cells.
+All GEDI points were leaf on conditions due to low accuracy of vertical profiles during leaf-off conditions.
+When modeling climate in the coldest month, this resulted in only few points in temperate regions, as these were restricted to mixed forests and did not include deciduous forests.
+
 # SCRATCH SPACE
@@
  - on second thought, I don't think this is necessary because we want to know why more species can't exist in trees in boreal regions
+
+Just as topographic complexity and climate explain Humbolt's enigma in mountains, vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the availability of vertical niche opportunities and physiologically constraining arboreal lifestyles [CITE].
+Densely layered forest structures not only provide habitat across the forest profile, but produce vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed.
+From the humid and thermally stable understory to the dry and thermally variable canopy that experiences both hot and cold temperature extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or hinder the capacity for tall forests to generate high levels of biodiversity.
+For example, increased exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat that can support arboreal species and thus limit richness despite tall canopies [@oliveiraVerticalStratificationInfluences2019].
+Moreover, relatively low thermal variability in the canopies of tropical forests may increase the distinctness of vertical climate bands that support high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics [CITE].
 

### 2025-07-09 scripts/03_analyses/02_vertmaps_and_vertRichness_correlations.R

@@ -9,6 +9,7 @@ library(colorspace)
 library(grid)
 #library(tricolore)
 library(lemon)
+library(cowplot)
 
 
 # load data ----------------------------------------------------
@@ -16,10 +17,6 @@ library(lemon)
 wd = vect("data/original/rnaturalearth_world.shp") %>% 
   crop(ext(-180,180,-60,83.64)) %>% 
   project("+proj=robin +datum=WGS84") 
-
-wd = vect("data/original/rnaturalearth_world.shp") %>% 
-  crop(ext(-180,180,-60,83.64)) %>% 
-  project("+proj=cea +datum=WGS84") 
 
 ymin = ext(wd)[3]
 ymax = ext(wd)[4]
@@ -82,14 +79,14 @@ for(i in levels(all$taxa)) {
   
   map = ggplot() +
     geom_spatvector(data = wd, fill = "black", color = NA) +
-    geom_raster(data = d, aes(x,y,fill = vert.mean.ses)) +
+    geom_tile(data = d, aes(x,y,fill = vert.mean.ses)) +
     #geom_spatvector(data = wd, fill = NA, color = "black") +
     scale_x_continuous("SES Verticality", expand = c(0,0)) +
     scale_fill_continuous_divergingx("spectral", na.value = NA, limits = col.lims,
                                      rev = T, guide = guide_colorbar(title = "Verticality")) +
     #facet_wrap(vars(taxa), ncol = 1, strip.position = "left") +
     coord_sf(crs = "+proj=robin +datum=WGS84") +
-    theme_classic() +
+    theme_void() +
     theme(strip.text = element_blank(),
           legend.position = "bottom",
           legend.title.position = "top",
@@ -99,7 +96,8 @@ for(i in levels(all$taxa)) {
           axis.title = element_blank(),
           axis.line = element_line(color = "white"),
           axis.text = element_blank(),
-          axis.ticks = element_line(color = "white"))
+          axis.ticks = element_line(color = "white"),
+          panel.background = element_blank())
   
   vert.legend = ggpubr::get_legend(map)
   #vert.legend = as_ggplot(vert.legend)
@@ -140,9 +138,9 @@ for(i in levels(all$taxa)) {
 
 pal = divergingx_hcl(palette = "zissou", n = 3)
 
-cola = pal[1]
+cola = pal[3]
 colt = pal[2]
-colf = pal[3]
+colf = pal[1]
 
 
[…]

### 2025-07-09 scripts/04_plots/sdmTMB_marginaleffects.R

@@ -33,7 +33,7 @@ var = list(
   "log_precip_dry [-2.1:2.7 by=0.1]",
   "tmax_warm [-3.8:2.9 by=0.1]",
   "tmin_cold [-2.54:1.34 by=0.1]",
-  "canopy_height [-1.8:3.1 by=0.1]",
+  "canopy_height2 [-1.8:3.1 by=0.1]",
   "veg_den [-1.3:9.0 by=0.1]",
   "precip_warm [-1.3:15.4 by=0.1]",
   "temp_diu [-3.7:3.2 by=0.1]"
@@ -41,7 +41,8 @@ var = list(
 
 
 amphplots = foreach(v = 1:length(var)) %do% {
-  g = predict_response(amphmod, var[[v]])
+  g = ggeffect(amphmod, var[[v]])
+  #g = predict_response(amphmod, var[[v]])
   g = cbind(g)
   g$term = var[[v]]
   g$taxa = "amphibians"
@@ -49,7 +50,7 @@ amphplots = foreach(v = 1:length(var)) %do% {
   g
 }
 
-saveRDS(amphplots, "results/sdmTMB_models2/marginal_effects/amph.rds")
+saveRDS(amphplots, "results/sdmTMB_models2.3/marginal_effects/amph.rds")
 
 # mammals
 
@@ -58,11 +59,11 @@ mammalsplots = foreach(v = 1:length(var)) %do% {
   g = cbind(g)
   g$term = var[[v]]
   g$taxa = "mammals"
-  saveRDS(g, paste0("results/sdmTMB_models2/marginal_effects/mammals/mammals_", v, ".rds"))
+  saveRDS(g, paste0("results/sdmTMB_models2.3/marginal_effects/mammals/mammals_", v, ".rds"))
   g
 }
 
-saveRDS(mammalsplots, "results/sdmTMB_models2/marginal_effects/mammals.rds")
+saveRDS(mammalsplots, "results/sdmTMB_models2.3/marginal_effects/mammals.rds")
 
 # reptiles
 
@@ -71,12 +72,12 @@ reptsplots = foreach(v = 1:length(var)) %do% {
   g = cbind(g)
   g$term = var[[v]]
   g$taxa = "reptiles"
-  saveRDS(g, paste0("results/sdmTMB_models2/marginal_effects/reptiles/reptiles_", v, ".rds"))
[…]

### 2025-07-14 manuscript.Rmd
ms conclusion

@@
 ---
-date: "`r Sys.Date()`"
-author: "Your Name"
-title: "Forest canopies are better generators of biodiversity in the tropics"
+bibliography: "`r rbbt::bbt_write_bib('r-refs.bib', overwrite = TRUE)`"
+csl: science.csl
 output: 
   officedown::rdocx_document:
+    reference_docx: "style-guide.docx"
     mapstyles:
       Normal: ['First Paragraph']
+plots:
+  caption:
+    style: Image Caption
+    pre: 'Figure '
+    sep: ': '
+header-includes:
+- \usepackage{fancyhdr}
+- \pagestyle{fancy}
+- \rhead{Global Verticality}
+- \begin{document}
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE)
+knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
 library(officedown)
 library(officer)
+library(tidyverse)
+library(spatialreg)
+library(data.table)
+library(flextable)
+library(knitr)
+library(chunkhooks)
 
@@
 # other figures are from older versions
-
-"scripts/03_analyses/01_canopyheight_richness_relationships.R" # canopy height richness models and plots
 ```
@@
 
-"The observer who leaves the center of the earth by an infinitely small amount compared to the radius can reach a new world, so to speak, and he can observe more variations in the aspect of the soil and more modifications in the atmosphere than he would if he were to pass from one latitude to another" - Humbolt pg 77 Essay on the Geography of Plants
+**Forest canopies are better generators of biodiversity in the tropics**
 
-"Current hypotheses for large-scale geographical patterns of species diversity fail to explain the extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5).
-We call this challenge “Humboldt’s enigma.”" - Rahbek
+Lydia G. Soifer^1\*^, David P. Edwards^2,3^, Brunno F Oliveira^4,5^, Mario R Moura^6^, Brett R. Scheffers^4^
+
+1 School of Natural Resource and Environment, University of Florida, USA
+
+2 Department of Plant Sciences and Centre for Global Wood Security, University of Cambridge, Cambridge, UK
+
+3 Conservation Research Institue, University of Cambridge, Cambridge, UK
+
+4 Centre de Synthèse et d’Analyse sur la Biodiversité, Montpellier, France
+
+5 Department of Wildlife Ecology and Conservation, University of Florida, USA
+
+6 Departamento de Sistemática e Ecologia, Universidade Federal da Paraíba, João Pessoa, Brazil
+
+Email addresses:
+
+^\*^ lysoifer@gmail.com (corresponding author)
+
+dpe29@cam.ac.uk
+
+brunno.oliveira@me.com
+
+mariormoura@gmail.com
+
+brett.scheffers@ufl.edu
+
+\newpage
+
+# Abstract
+
+\newpage
 
 For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019].
-[Yet until recently, hypotheses failed to explain the extraordinary richness of mountains particularly in the tropics.] The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover.
-Combined, these factors allow mountains to generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019].
-Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].
+In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global biodiversity patterns as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019].
+The extraordinarily high biodiversity of mountains and their enhanced capacity to generate species richness in the tropics has been termed "Humbolt's enigma", and can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019].
+The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover.
+Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].
 
-NEED TO MENTION HUMBOLT'S ENIGMA SOMEWHERE IN FIRST PARAGRAPH AND THAT THE MAIN EXPLANATIONS ARE TOPOGRAPHY AND CLIMATE
-
-Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogegoraphic patterns [@macarthurBirdSpeciesDiversity1961].
+Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogeographic patterns [@macarthurBirdSpeciesDiversity1961; @fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
 Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains.
@@
 Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967].
-This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024].
 Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
-- a vertical dimension to Humbolt's enigma
+The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical dimension to Humbolt's enigma.
 
-Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
+Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
 Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages.
-For example, the denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [CITE or CONCEPT FIG].
+For example, denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019].
 Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes.
-Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020] [CITE].
-Yet only recently have we been able to address these hypotheses, as our ability to measure vegetation structure and quantify its use by animals globally [AND MODEL MICROCLIMATES GLOBALLY] increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].
+Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019].
+Yet only recently have we been able to address these hypotheses, as our ability to quantify variation in vegetation structure and microclimates globally and examine their impacts on biogeographic patterns in three-dimensional space has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates any where on earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].
 
-In this study we combine range maps of nearly 24,000 terrestrial vertebrate species and comprehensive databases of vertical habitat use, global macroclimate and forest structure data, and process-based microclimate models to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities.
-We first investigated the capacity for taller forests to support increasing species richness in forests and less densely vegetated wooded biomes across latitudes.
-If the vertical structure of species assemblages mediates the relationship between canopy height and species richness, then we expect positive correlations between the verticality of assemblages (i.e., their average vertical niche position) and species richness.
-[AND THAT CANOPY HEIGHT WILL EXPLAIN LITTLE VARIATION IN RICHNESS ONCE VERTICALITY HAS BEEN ACCOUNTED FOR]
-
-and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height.
-We then examined the mechanistic processes driving this pattern.
-EXPLANATION OF TETRAPODTRAITS AND ELTONTRAITS
-
-, including TetrapodTraits and EltonTraits, to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.
-
-To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.
-We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height.
-We then examined the mechanistic processes driving this pattern.
-If vertical stratification contributes to the capacity for tall forests to support diverse assemblages, we expect that species rich assemblages will exhibit high verticality (i.e., a high average vertical niche position) and that environmental factors limiting verticality will limit species richness.
-To test this hypothesis, we examined correlations between assemblage verticality and species richness.
-We then modeled assemblage verticality of the four vertebrate classes by macroclimate and vegetative structure and investigated the microclimatic processes that underlie these macroclimatic patterns.
-While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we modeled vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.
+In this study we combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities.
+We first investigated the capacity for taller canopies to support increasing species richness in forests and less densely vegetated woody biomes across latitudes.
+We find that tall canopies are better generators of biodiversity in the tropics, particularly in moist forests (Fig.
+\ref(fig:fig1)).
+We then examined the processes driving this pattern using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and process-based microclimate models [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] for smaller regions that integrate forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023].
+While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.
 
@@
 
-To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally.
-We then mapped the residuals of these models to determine where canopy height over- or under-predicted species richness.
-We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.
+To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of vertebrate richness as a function of canopy height globally for all vertebrate combined and each vertebrate class (birds, mammals, reptiles, and amphibians).
+We then mapped the deviance residuals of these models to determine where canopy height over- or under-predicted species richness.
+To examine variation in the capacity for tall canopies to support rich assemblages across latitudes, we produces separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands.
 
@@
 S\@ref(fig:global-chrich)).
-However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Fig.
-\@ref(fig:fig1), S\@ref(fig:resids-taxa)).
+However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs.
+\@ref(fig:fig1)c, S\@ref(fig:resids-taxa)).
 Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions.
 Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig.
-\ref(fig:fig1)).
+\ref(fig:fig1)d).
 For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions.
@@
 For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species.
-The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Fig.
-\@ref(fig:fig1)).
+The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs.
+\@ref(fig:fig1)c, S\@ref(fig:resids-taxa)).
+Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier or higher latitude regions, mirroring impacts of mountains [@rahbekHumboldtsEnigmaWhat2019].
+Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics.
 
-Minor deviations from these patterns may indicate variations in evolutionary drivers of species richness patterns.
-  [- This pattern is particularly stark in the western hemisphere.
-In the eastern hemisphere, the pattern is weaker where evolutionary forces may drive overprediction of richness in parts of the lowland African tropics and Indo-Malaysian.
-In Indo-Malaysia, island size may reduce vertebrate richness relative to expectations by canopy height.]
-
-However, the described patterns are broadly consistent among taxa and mirror findings that for a given area, mountains have higher richness in the tropics [@rahbekHumboldtsEnigmaWhat2019].
-Thus, increases in habitat volume via either mountains or forests have greater impacts on richness in the tropics.
-In other words, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the tropics.
+[ADD MODEL TABLES TO SUPPLEMENT]
 
 # The mediating effect of verticality
-The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages.
+
+```{r}
+source("scripts/03_analyses/02_vertmaps_and_vertRichness_correlations.R")
+
+# mean proportion of arboreal species
+parb.wettropics = round(as.numeric(parb.mean[parb.mean$biome=="Tropical & Subtropical Moist Broadleaf Forests", 2]),2)
+
+```
+
+The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a].
 As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].
@@
 For species occurring in multiple habitats, we averaged scores across habitats.
-We then calculated verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).
+We then calculated assemblage verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).
 
-Verticality peaked in the tropics and declined toward the poles, exhibiting strong correlations with verticality across taxa (Fig.
+Verticality peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig.
+\@ref(fig:fig2), S\@ref(fig:vertmean-biome)).
+Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig.
+S\@(fig:meanprop-biome)).
+Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig.
 \@ref(fig:fig2)).
-Moist tropical forests contained assemblages comprised of x% to x% species exhibiting arboreal strategies (SUPP FIG), corresponding to some of the most species rich assemblages.
-Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies.
 This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
@@
 # Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities
-To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may constrain or support arboreal life strategies, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month, as well as with macroclimatic factors that may influence the distinctness of vertical climate banding, including diurnal temperature range (Supplementary methods).
+
+To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages.
+Macroclimate variables included those that may induce physiological constraints - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods).
 We additionally included vegetation density in the models, which impacts the availability of habitat volume.
-While foliage height diversity and temperature seasonality also impact niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height and minimum temperatures, respectively, and were therefore excluded from the models.
-The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018].
-While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact community structure [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020].
-We therefore additionally modeled vertical microclimate profiles in six forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy processes to predict vertical climate gradients [@macleanMicroclimfFastCanopy2024].
-To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within a 1° x 1° grid cell.
+While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models.
+The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across vertical forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018].
+While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact vertical assemblage structure and richness patterns [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020].
+We therefore additionally modeled vertical microclimate profiles in six 1° x 1° grid cells within forested regions from tropical to boreal latitudes.
+To do so, we integrated information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025].
+To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within the modeled region (Supplementary methods).
 By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.
 
-Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry, and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying degrees across vertebrate classes (Fig.
-\@ref(fig:fig3)).
-These patterns arise due to impacts of spatial climate variation on vegetation structure and physiological constraints to arboreal lifestyles as well as impacts of temporal climate variation on the distinctness of vertical climate bands.
+Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig.
+\@ref(fig:fig3), Table S\@ref(tab:vert-model-results)).
+These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal lifestyles, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.
 
 ## Climate produces structural constraints to verticality
+
 Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014].
-High dry season precipitation and thus low precipitation seasonality may support highly vertical assemblages by enabling development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
+Our models indicated that high dry season precipitation positively impacts assemblage verticality across taxa (Fig.
+\@ref(fig:fig3)).
+Wet climates with low precipitation seasonality enable the development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
 Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004].
-Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].
-In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig.
-\@ref(fig:fig1)).
+Hotspots of structural complexity in wet regions that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].
 
-At higher latitudes, lower precipitation and light hinder the development of densely layered forests by restricting the functional and structural configurations of trees [@ehbrechtGlobalPatternsClimatic2021a; @hakkenbergClimateMediatesRelationship2021; @atkinsPowerLawScaling2022].
+In drier forests and higher latitudes, lower precipitation and light hinder the development of densely layered forests by restricting the functional and structural configurations of trees [@ehbrechtGlobalPatternsClimatic2021a; @hakkenbergClimateMediatesRelationship2021; @atkinsPowerLawScaling2022].
 In evergreen needleleaf forests that dominate boreal regions, these environmental constraints dampen the rate of increase in structural complexity with canopy height [@atkinsPowerLawScaling2022].
@@
 
+[ADDITIONAL CITATIONS] - <https://doi.org/10.1016/j.rse.2006.11.016>
+-<https://doi.org/10.1111/1365-2745.13944>
+
 ## The influence of canopy climates
@@
 These steep climate gradients can support high rates of turnover from the ground to the canopy that yield vertically stratified and species rich assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018].
-However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness, just as lower elevation snow lines limit the capacity for tall mountains to generate biodiversity at higher latitudes [@oliveiraVerticalStratificationInfluences2019].
+However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness [@oliveiraVerticalStratificationInfluences2019].
 
-For example, our models indicate that low dry season precipitation constrains verticality across vertebrate classes (Fig.
-\@ref(fig:fig3)).
-These macroclimate patterns were accompanied by steep gradients in minimum relative humidity, which declined x% from the ground to the top of the canopy (Fig.
+Constraints of low dry season precipitation on verticality (Fig.
+\@ref(fig:fig3)) were accompanied by steep gradients in minimum relative humidity, which microclimate models estimated declined x% from the ground to the top of the canopy (Fig.
 \@ref(fig:fig4)).
-Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot climates, where arboreal species depend on evapotranspiration to maintain body temperatures below critical thermal maxima, or in cold climates, where basking to increase body temperature can quickly lead to desiccation without adequate moisture availability or physiological defenses.
-By imposing environmental filters on the functional morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage verticality and limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [https://onlinelibrary.wiley.com/doi/full/10.1111/j.1461-0248.2004.00671.x].
-This pattern is evident for frog communities in the seasonally dry forests of Panama, where collapse of vertical stratification toward the ground during the dry season due to water limitation in the canopy is accompanied by richness declines and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020].
-Simply put, extreme canopy microclimates that physiologicall constrain arboreal life truncate the volume of available habitat, which prevents tall trees from supporting species rich assemblages.
-In contrast, high dry season precipitation in tropical moist forests supports highly vertical assemblages by reducing physiological constraints on arboreal life strategies in the upper canopy.
-Species are therefore able to finely partition the entire volume of forested habitat, leading to a dense packing of species in the canopy and allowing increases in canopy height to yield rapid increases in species richness (Fig.
-\@ref(fig:fig1)).
+Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot or cold climates, where evapotranspiration to maintain body temperatures below thermal maxima or basking to maintain body temperatures above thermal minima can quickly lead to desiccation without adequate physiological defenses [@scheffersIncreasingArborealityAltitude2013; @tracyNotJustSmall2010; @mckechnieAvianThermoregulationHeat2017].
+By imposing environmental filters on the functional
+morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage
+verticality and limit the number of species that an increase in canopy height can support, in line
+with the physiological tolerance hypothesis [@curriePredictionsTestsClimatebased2004]. For example,
+in the seasonally dry forests of of Panama, water limitation in the canopy causes the vertical
+stratification of frog communities to collapse toward the ground during the dry season - a shift
+coincident with declines in richness and homogenization of arboreal diversity
+[@bashamVerticalStratificationCollapses2020]. Dry climates can therefore reduce the capacity for
+increasingly tall canopies to support greater biodiversity (Fig. \@ref(fig:fig1)), just as cold
+climates limit the capacity for tall mountains to generate biodiversity by reducing the elevation of
+the snow line [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013;
+@janzenWhyMountainPasses1967]. In contrast, high dry season precipitation in tropical moist forests
+supports highly vertical assemblages by reducing physiological constraints on arboreal life
+strategies in the upper canopy. Species are therefore able to finely partition the entire volume of
+forested habitat, leading to a dense packing of species in the canopy and allowing increases in
+canopy height to yield rapid increases in species richness
+[@oliveiraVerticalStratificationInfluences2019] (Fig. \@ref(fig:fig1)).
 
-Hot and cold temperatures had less consistent effects across taxa, but may still play critical roles in physiological constraints, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures.
-For example, high maximum temperatures reduced verticality and/or [REDUCED THE POSITIVE IMPACT OF CANOPY HEIGHT ON VERTICALITY FOR REPTILES, MAMMALS, AND BIRDS] across vertebrate classes, while low minimum temperatures reduced positive effects of canopy height for birds and reptiles (Fig.
-\@ref(fig:fig3)).
-High maximum temperatures may cause ectotherm body temperatures to exceed critical thermal maxima (McKechnie et al.
-2017; Sunday et al.
-2014), particularly in drier regions where maximum temperatures peaked.
-Microclimate models additionally indicated that vertical maximum temperature gradients were steepest in the temperate forests, though canopy temperatures exceeding 40C may limit the capacity for species to stratify across the entire vertical climate gradient (Fig.
-\@ref(fig:fig4)).
-Similarly, low minimum temperatures can cause body temperatures to exceed critical thermal minima for ectotherms or impose energetic constraints to endotherms, particularly in the windier conditions that characterize the canopy  (Sunday et al.
-2014) (Byman, Hay, and Bakken 1988).
-Shallow vertical minimum temperature gradients (Fig.
-\@ref(fig:fig4)), particularly during leaf off conditions [CITE], cause environmental filters to act across the entire canopy, posing severe limitations on the capacity for taller canopies to increase species richness.
+Maximum and minimum temperatures had less consistent effects across taxa (Fig. \@ref(fig:fig3)), but
+may still place physiological constraints on arboreal life strategies that influence the capacity
+for taller canopies to support richer species assemblages, particularly in regions exhibiting
+steeper vertical gradients characterized by hotter maximum and colder minimum temperatures. For
+example, high maximum temperatures generally reduced verticality and/or reduced the positive impact
+of canopy height on verticality, barring mammals for which verticality responded positively to
+maximum temperatures. High temperatures may cause ectotherm body temperatures to exceed critical
+thermal limits [@sundayThermalsafetyMarginsNecessity2014]. While consistent water availability in
+the canopies of tropical wet forests may allow ectotherms to maintain body temperature below
+critical limits, the drier conditions of temperate forests combined with steeper vertical gradients
+in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal
+habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats
+[@sundayThermalsafetyMarginsNecessity2014] (Fig. \@ref(fig:fig4)). Moreover, low minimum
+temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig.
+\@ref(fig:fig3)). Cold temperatures at higher latitudes can cause ectotherm body temperatures to
+exceed critical thermal minima outside fossorial habitats or impose energetic constraints to
+endotherms, particularly in the windier conditions that characterize the canopy
+[@sundayThermalsafetyMarginsNecessity2014] [REPLACE MAMMAL EXAMPLE WITH BIRD PHYSIOLOGY]. The
+physiological constraints imposed by minimum temperatures may act across the entire canopy given
+given their shallow vertical gradient, (Fig. \@ref(fig:fig4)), posing severe limitations on the
+capacity for taller canopies to increase species richness in boreal forests.
 
-[https://nsojournals.onlinelibrary.wiley.com/doi/full/10.1111/ecog.07122]
+Temporal climate variability may also play a role in the capacity for verticality to mediate
+relationships between species richness and canopy height by impacting climatic distinctness. In
+montane regions, climatic distinctness across elevations declines with latitude due to increasing
+seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019;
+@janzenWhyMountainPasses1967]. Low seasonal variation in the tropics produces narrow and distinct
+climate bands that exhibit little overlap across elevations. Unique climates across elevations
+restrict species' range size, leading to high rates of turnover across elevations that allow
+tropical mountains to generate extraordinary levels of biodiversity [@janzenWhyMountainPasses1967;
+@sheldonFiftyYearsMountain2018; @ghalamborAreMountainPasses2006; @rahbekHumboldtsEnigmaWhat2019]. In
+contrast, high thermal variation in temperate regions increases the similarity between climates
+experienced at high and low elevations, reducing the capacity for temperate mountains to support
+high rates of turnover and thus high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019;
+@ghalamborAreMountainPasses2006].
 
-Temporal climate variability also plays a key role in the capacity for verticality to mediate biogeographic patterns by impacting climatic distinctness.
-[relationships between species richness and canopy height by impacting climatic distinctness].
-In montane regions, climatic distinctness across elevations declines with latitude due to increasing seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019; @janzenWhyMountainPasses1967].
-Low seasonal variation in the tropics produces narrow and distinct climate bands that exhibit little overlap across elevations, yielding high rates of turnover that allow tropical mountains to generate extraordinary levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
-In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019].
+Within wooded habitats, diurnal temperature range may influence the distinctness of vertical climate
+bands from the ground to the canopy, and thus the vertical turnover of species. Our models of
+verticality indicate that greater diurnal temperature variation reduces verticality of mammals,
+though effects for other classes were insignificant (Fig. \@ref(fig:fig3)). Furthermore,
+microclimate models indicated that increases in daily temperature variation, as well as the rate of
+increase in temperature variation from the ground to the canopy, were greater in higher latitude
+forests (Fig. \@ref(fig:fig4)c). High daily temperature variation reduced the distinctness of
+climate bands across the vertical profile, with vertical zonation collapsing nearly completely
+during morning hours in temperate and boreal forests (Fig. \@ref(fig:fig4)B). Without unique
+climatic niches across the forest profile, these regions are unlikely to support rapid vertical
+turnover, limiting the number of species that may be added to the assemblage for a given increase in
+canopy height, even in wetter temperate regions with climate conditions that physiologically support
+arboreal strategies. In contrast, tropical canopies maintained more distinct vertical climate bands
+throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
+Narrow and distinct vertical climate niches may therefore support rapid vertical turnover, leading
+to greater increases in richness with canopy height and enhancing the capacity for tropical canopies
+generate biodiversity.
 
-Our models suggest that a similar mechanism may contribute to the capacity for tropical canopies to support assemblages with higher verticality and species richness than temperate canopies of equivalent heights.
-At the macroclimate scale, models of verticality generally indicated no significant impact of diurnal temperature variation, though low diurnal temperature variation increased verticality of mammals (Fig.
-\@ref(fig:fig3)).
-However, microclimate models indicated clear increases in daily temperature variation in higher latitude forests (Fig.
-\@ref(fig:fig4)).
-Moreover, the increase in thermal variability from the ground to the canopy was steeper in temperate and boreal regions.
-Temperate and boreal canopies canopies therefore exhibited less distinct climate bands, with vertical zonation collapsing nearly completely during morning hours.
-Without unique climatic niches across the forest profile, these regions are unlikely to support rapid vertical turnover, limiting the number of species that may be added to the community given an increase in canopy height even in regions with climate conditions that physiologically support arboreal life.
-In contrast, tropical canopies maintained distinct climate bands throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
-Narrow and distinct vertical climate niches may therefore support rapid vertical turnover and increases in richness with canopy height.
-that contributes the capacity for tropical  unique communities
+[TRY LABELLING MICROCLIM OVERLAP FIG AS BOREAL, TEMPERATE, TROPICAL COLUMNS AND EASTERN AND WETERN
+HEMISPHERE ROWS TO SAVE SPACE ALSO REORDE PROFILE PANELS INTO THE ORDER THEY ARE DISCUSSED]
 
-- climate variation increases vertical range size of species - Leahy
+# Canopy height in a changing world
 
-# Conclusion - bring back to Humbolt's engima
+Decades of research have elucidated the mechanisms that allow mountains to generate high levels of biodiversity, particularly in the tropics [CITE].
+Yet only recently has the three-dimensional nature of terrestrial landscapes been widely recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
+We bring the vertical dimension of habitats to the forefront of global biogeography by showing that the climatic characteristics of wooded ecosystems from macro- to micro-scales mediate the capacity for tall forests to support high levels of biodiversity via impacts on the vertical structure of vertebrate assemblages.
+Tropical rainforests, with wet and thermally stable climates that increase the structural complexity of vegetation, reduce physiological stress in the canopy, and produce distinct vertical climate bands, are able to support fine-scale partitioning of resources and climatic niches throughout the entire forest volume.
+Consequently, taller trees in tropical forests support higher biodiversity and greater increases in species richness than temperate and boreal ecosystems of equivalent heights where climatic conditions reduce the structural complexity of vegetation, pose physiological constraints on arboreal lifestyles, and reduce vertical climate zonation from the ground to the canopy.
 
-Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
-
-Since verticality research was started, most research has been on impacts of vegetation structure.
-Yet no hypotheses had been proposed regarding inconsistent conclusions of the impacts of vegetation structure on richness.
-
-The canopy supports fewer arboreal species at high latitudes because
-1) physiological limitations and physiological tolerance hypothesis
-2) high temporal climate variabiltiy in high latitude canopies reduces vertical zonation, limiting vertical turnover and therefore richness
-
-- Trees, just as mountains, bring different climate zones within close proximity of each [@rahbekHumboldtsEnigmaWhat2019].
-While the climatic difference from the ground to the canopy is much less than the climatic difference across thousands of meters in elevation, the climatic rate of change is often faster from the ground to the canopy.
-This climatic layering allows distinct assemblages of species to exist across forest strata.
-
-- climate stability, range size, and niche width - greater climate stability in the tropics allows for more specialization within the canopy, while higher climate variability in the canopy in high latitudes reduces specialization and turnover.
-Compare to the ground - see Dave's mountaintops paper?
-
-- The capacity for tall forests to support rich arboreal assemblages depends on the density of vegetation across the vertical profile and the presence of canopy microclimates that limit physiological constraints on arboreal life strategies.
-
-- In high latitude forests, tall canopies may help generate biodiversity on the ground by buffering microclimate and providing small increases in structural complexity that can be inhabited by few arboreal animals with specialized physiologies.
-However, harsh microclimates in the upper canopy largely truncate niche availability, preventing fine-scale partitioning of canopy habitats by arboreal species and thus limiting the extent to which taller canopies increase species richness.
-
-- The biodiversity-structure hypothesis suggests xyz...
-  - Greater veg density across vertical profile = higher productivity
-
-[INTEGRATE VERTICAL COMPONENT INTO THIS PARAGRAPH]
-
-- comparison to mountains
-  - climate bands are narrower in the tropics -> higher turnover rates
-  - we don't see this pattern for vertical climate profiles (similar shapes in profiles across latitudes and less variation in the troipcs)
-- results
-  - the vertical difference in rhmin and tmax is greater than the latitudinal difference from tropical to boreal forests
-  - stable microclimates in tropical canopies -> specialization and speciation
-
-In climate conditions that do not constrain verticality, shifts up the canopy are associated with radiations within the canopy that produce positive relationships between canopy verticality and species richness
-In climate conditions that constrain verticality, arboreal species occur, but arboreal radiations do not, causing neutral or negative relationships between verticality and richness (although this mechanism/relationship is noisy)
-
-# Methods
-## Data
-The canopy height layer was obtained from Lang et al.
-[-@langHighresolutionCanopyHeight2023], which represents canopy height in the year 2020 at a 10 m resolution.
-To obtain canopy height in forests, we aggregated the data to match the resolution of Hansen et al.
-[-@hansenGlobalLandUse2022] global land cover and land use in 2019 (~ 30 m resolution) and used this layer to mask all grid cells with woody vegetation < 3 m, surface water and permanent ice, built-up land, cropland, recent forest loss, and land use associated with forestry and shifting cultivation (e.g., oil palm).
-We then aggregated and averaged the remaining pixels to the resolution of the verticality data (50 km).
-
-We classified biomes into larger categories based on the 2017 biome categories.
-For biomes that occurred across latitudes (e.g., Montane Grasslands and Shrublands), we classified them as tropical if they occurred within the latitudes covered by biomes classified as tropical and temperate if they occurred within the latitudes classified as temperate.
-Additionally, Mediterranean forests, woodlands, and scrub were partitioned into temperate forests if the ecoregion contained "forest" and temperate woodland, savanna, and shrubland if the ecoregion name did not contain the word "forest".
-
-MASS was used to produce models of canopy height by richness.
-sdmTMB was used to produce models of verticality.
-DHARMa was used to check model fit and assumptions.
-ggeffects was used to generate model predictions and confidence intervals.
-Figures were generating using tidyterra, patchwork, ggplot2
+While tall forests, particularly in the tropics, are crucial for supporting species rich assemblages, regions containing tall forests are globally rare, under-protected, and threatened by anthropogenic disturbance and climate change [@huangPrioritizingGlobalTall2023; @bourgoinHumanDegradationTropical2024; @bennettLargerTreesSuffer2015; @rowlandDeathDroughtTropical2015].
+For example, disturbances from fire and logging produce persistent reductions of canopy height in the tropics [@bourgoinHumanDegradationTropical2024], with potential negative effects on the capacity for these forests to support high levels of biodiversity.
+Furthermore, climate change is increasing the frequency of extreme heat and drought [@seneviratneWeatherClimateExtreme2021].
+In addition to directly reducing forest height [@bennettLargerTreesSuffer2015; @rowlandDeathDroughtTropical2015], these trends could truncate the volume of physiological suitability habitat for arboreal species.
+Climate change may therefore flatten the vertical structure of species assemblages toward the ground [@oliveiraVerticalStratificationInfluences2019; @scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], just as warming temperatures are causing upslope shifts in species' distributions [@lenoirSpeciesBetterTrack2020; @chanClimateVelocitiesSpecies2024].
+Improving our understanding of the vertical structure of species assemblages as canopies heat up and dry out will be critical for evaluating the continued capacity of tropical forests to generate extraordinarily high levels of biodiversity.
 
 # Data availability statement
+
 canopy height:
+
+# Acknowledgements
+LGS was supported by the NSF GRFP...
+We thank Patrick Burns for access to the processed GEDI L2b points.
+
+\newpage
 
@@
 
-"figures/ms_figures/fig1_canopy_richness.png"
+include_graphics("figures/ms_figures/fig1_canopy_richness.png", error = F)
+
+```
+
+```{r fig2, fig.id="fig2", fig.cap="Global variation in the vertical structure of vertebrate assemblages and its association with species richness for birds, mammals, reptiles, and amphibians.
+a) The average ± SD proportion of species exhibiting arboreal (red), terrestrial (yellow), and fossorial (blue) strategies across latitudes.
+Proportions can exceed one because species can exhibit multiple vertical strategies.
+b) The verticality of vertebrate communities in wooded biomes across the globe.
+High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance." c) Correlations between verticality and species richness globally.
+Lines indicate linear best fit line with 95% CIs.
+}
+# Fig. 1
+
+include_graphics("figures/ms_figures/fig2_vertmaps.png", error = F)
 
@@
 
-"figures/main_figs/fig3_sdmTMB_coefs.png"
+include_graphics("figures/ms_figures/fig3_sdmTMB_coefs.png", error = F)
 
@@
 
-THINGS TO FIX IN FIG4
-- diurnal temp range is for warmest month in strata plot and coldest month in profile plot - make consistent
-- replace map with concept figure similar to Rahbek
-- should we add strata plots for other variables? Or add to supplement - probable supplement
-- text sizing
-- point size in map if we keep the map
+THINGS TO FIX IN FIG4 - diurnal temp range is for warmest month in strata plot and coldest month in
+profile plot - make consistent - replace map with concept figure similar to Rahbek - should we add
+strata plots for other variables? Or add to supplement - probable supplement - text sizing - point
+size in map if we keep the map
+
 ```{r fig4, fig.id="fig4", fig.cap=", fig.width=140, fig.height=120}
@@
 
-"figures/ms_figures/fig4_profiles.png"
+include_graphics("figures/ms_figures/fig4_profiles.png", error = F)
 
-```
-# Supplementary figures
-
-```{r splot-global-chrich, fig.id="global-chrich", width=90, height=90, fig.cap="Global relationship between total vertebrate richness and canopy height.
-Line indicates predictions from a negative binomial model of richness by canopy height ± 95% CI.
-Points indicate observed richness by canopy height.}
-
-"figures/ms_figures/supp_figs/total_richness~canopy.png"
-```
-
-```{r splot-resids-taxa, fig.id="resids-taxa", width=180, height=100, fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians."}
-
-"figures/ms_figures/supp_figs/deviance_resids_taxa.png"
 ```
@@
 
+# Materials and Methods
+
+## Assmeblage richness and verticality
+
+```{r}
+amph_spdat = read.csv("data/derivative_data/species_data/amph_spdat_moura_forestsOnly.csv")
+rept_spdat = read.csv("data/derivative_data/species_data/reptiles_spdat_moura_forestsOnly.csv")
+mammals_spdat = read.csv("data/derivative_data/species_data/mammals_spdat_moura_forestsOnly.csv")
+bird_spdat = read.csv("data/derivative_data/species_data/birds_spdat_elton_forestsOnly.csv")
+
+namph = nrow(amph_spdat)
+nrept = nrow(rept_spdat)
+nmammals = nrow(mammals_spdat)
+nbirds = nrow(bird_spdat)
+
+```
+
+We mapped global richness and verticality of terrestrial vertebrate assemblages in wooded habitats
+[@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells using a cylindrical equal
+area projection. Range maps were obtained from the IUCN Red List for mammals and amphibians
+[@iucnIUCNRedList2023], BirdLife International for bird breeding and resident ranges
+[@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global
+Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017;
+@caetanoAutomatedAssessmentReveals2022]. Overall, we obtained range maps for `r nbirds` birds,
+`r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians.
+
+To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a
+verticality score from zero to one and then took the average verticality of all species whose ranges
+overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based
+on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the
+TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024a]. For birds, we
+calculated verticality scores based on their use of different foraging strata as identified in the
+EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale
+vertical classification than the dataset from Moura et al.
+[-@mouraPhylogenyinformedCharacterisationGlobal2024a]. For birds, verticality scores were as
+follows: foraging below the water surface, around the water surface, and on the ground = 0.5,
+foraging in the understory = 0.667; foraging in the mid-high canopy = 0.8337; foraging in or above
+the upper canopy = 1. For all taxa, if species used multiple vertical habitats, vertical scores were
+averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it
+would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that
+did not occur in terrestrial habitats (e.g., marine mammals) were excluded from the analysis.
+
+To examine spatial patterns of assemblage verticality independently of species richness and accounting for biogeographic history, we calculated the standard effect size (SES) of mean verticality (hereafter ‘verticality’): $SES=(x_i-\bar{x})/sd(x)$ where x is a vector of expected mean verticality scores given random selection of species from a species pool.
+For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm (Dinerstein et al.
+2017) and generated a null distribution of expected verticality for each grid cell based on 100 simulations.
+Realms included Palearctic, Nearctic, Indomalayan, Neotropic, Oceania, Afrotropic, and Australasia.
+For each simulation, we randomized verticality within the grid cell by randomly drawing the same number of verticality scores from the respective species pool as the observed richness in the cell.
+Highly positive SES verticality scores represent assemblages having higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.
+
+## Environmental data
+
+We obtained macroclimate data from CHELSA v2.1 (30 arc-second resolution), representing average conditions for the time period 1981-2010 [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017].
+Macroclimate variables represented climatic conditions that could pose physiological constraints to arboreal life styles or impact the distinctness of vertical climate banding, and included mean daily maximum air temperature of the warmest month (Tmax~warm~; °C), mean daily minimum air temperature of the coldest month (Tmin~cold~; °C), precipitation amount of the driest month (Precip~dry~; kg m^-2^ month^-1^), mean monthly precipitation of the warmest quarter (Precip~warm~; kg m^-2^ month^-1^), temperature seasonality (°C), mean diurnal temperature range (Temp~diu~; °C), and precipitation seasonality (Precip~sea~; kg m^-2^).
+
+We obtained vegetation data, including canopy height [@langHighresolutionCanopyHeight2023], foliage height diversity [@burnsMultiresolutionGriddedMaps2024] vegetation density [@crowtherMappingTreeDensity2015].
+The canopy height layer represents canopy height at a 10 m resolution for the year 2020 and was developed using a fusion of height data from the Global Ecosystem Dynamics Investigation (GEDI) and optical satellite images from Sentinel-2.
+To obtain canopy height in in-tact wooded habitat, excluding human modified land covers, such as plantations and shifting agriculture, we applied a mask using Hansen et al.
+[-@hansenGlobalLandUse2022] global land cover and land use in 2019.
+To do so, we aggregated the data to match the resolution of the land cover layer (~ 30 m), and then masked all grid cells with woody vegetation \< 3 m, surface water and permanent ice, built-up land, cropland, recent forest loss, and land use associated with forestry and shifting cultivation (e.g., oil palm).
+Foliage height diversity (FHD) was obtained from gridded GEDI data at a 1 km resolution and represents the vertical heterogeneity of vegetation.
+FHD was masked using the same method as canopy height.
+The vegetation density map represents stems per hectare and was generated using regression models linking stem density in forest plots to climate, topography, vegetation characteristics, and the proportion of developed land cover.
+Because the vegetation density estimates were generated using plots in forests rather than spatially continuous observations including human modified land use types, we did not mask vegetation density with the land cover map.
+
+All environmental data were resampled to the spatial resolution and projection of the species data.
+
+## Vertebrate richness models
+
+Vertebrate richness is expected to increase with increasing canopy height.
+We first modeled richness as a function of canopy height using a generalized linear model with a Poisson distribution and a log-link.
+Models were produced for global richness across all vertebrate class, global richness for each vertebrate class, and biome richness for each vertebrate class.
+Initial model evaluation indicated over-dispersion in model residuals.
+We therefore fit a negative binomial model with a log-link function.
+Since richness often exhibits a log-log relationship with covariates, we additionally fit a model of richness as a function of log(canopy height), and compared this model to the original using AIC and pseudo-R^2^ values.
+We report results of richness as a function of canopy height, as it consistently performed better except for the model for reptiles in temperate forests.
+
+We classified biomes into larger categories based on the [@dinersteinEcoregionBasedApproachProtecting2017] 2017 biome categories.
+For biomes that occurred across latitudes (e.g., Montane Grasslands and Shrublands), we classified them as tropical if they occurred within the latitudes covered by biomes classified as tropical and temperate if they occurred within the latitudes classified as temperate.
+Additionally, Mediterranean forests, woodlands, and scrub were partitioned into temperate forests if the ecoregion contained "forest" and temperate woodland, savanna, and shrubland if the ecoregion name did not contain the word "forest".
+
+We also examined relationships between richness and verticality using Pearson correlations.
+
+## Verticality models
+
+We examined relationships between verticality and environmental covariates using spatial linear mixed models (LMMs).
+Grid cells with fewer than five species were excluded, because they may provide unrealiable estimates of the SES.
+To linearize the relationships between predictor and response variables, we log10-transformed precipitation of the driest month.
+We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using Pearson correlations.
+We excluded precipitation seasonality, temperature seasonality, and foliage height diversity due to strong correlations (|r| > 0.8) with dry season precipitation, minimum temperature of the coldest month, and canopy height, respectively.
+Additionally, we included a quadratic effect for maximum temperature of the warmest month, since extremely high temperatures may inhibit arboreal life strategies.
+
+Spatial LMMs were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which approximates a spatial random effect using a finite element analysis mesh as implemented in R-INLA [@thorsonSpatioTemporalModelsEcologists2024].
+The full model can be described as:
+
+$$
+\begin{align}
+y_s ∼ 𝜇_{s}, \\
+𝜇_s = 𝜲_s𝛽 + ⍵_s, \\
+⍵_s ∼ MVNormal(0,𝚺_⍵)
+\end{align}
+$$
+
+where $y_s$ is verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected verticality.
+$Χ_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients, including additive effects of Tmax~warm~, Tmax~warm~^2^, Tmin~cold~, Temp~diu~, Precip~warm~, log10(Precip~dry~), vegetation density, and canopy height, as well as interactions between canopy height and the five climatic variables.
+Interactions were included because we expect the impact of canopy height on verticality to decline when harsh macroclimate conditions, including extreme cold, hot, or dry conditions, reduce the suitability of canopy microhabitats for arboreal life strategies, or when high climate variability reduces distinctness of vertical climate bands.
+$⍵_s$ represents a spatial random effect at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates.
+The spatial random effect was modeled as a Gaussian random field with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
+Gaussian random fields were approximated for models of verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].
+
+We validated models first using the ‘sanity’ function from the sdmTMB package.
+We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to assess residual normality and lack of patterning across space.
+
 ## Microclimate models
 
-We selected 6 1 x 1 degree grid cells, with 2 each in tropical, temperate, and boreal forests.
-We downloaded GEDI points that intersected with these cells.
-All GEDI points were leaf on conditions due to low accuracy of vertical profiles during leaf-off conditions.
-When modeling climate in the coldest month, this resulted in only few points in temperate regions, as these were restricted to mixed forests and did not include deciduous forests.
+To examine vertical microclimate gradients, we mechanistically modeled microclimates from the ground to the canopy using the ‘micropoint’ R package [@macleanMicropointCanopyPoint2025].
+Models were produced for boreal forests, temperate forests, and tropical forests from the Americas and from Eurasia.
 
-# SCRATCH SPACE
-  - TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE REGION
- - on second thought, I don't think this is necessary because we want to know why more species can't exist in trees in boreal regions
+### Vegetation data
 
-Just as topographic complexity and climate explain Humbolt's enigma in mountains, vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the availability of vertical niche opportunities and physiologically constraining arboreal lifestyles [CITE].
-Densely layered forest structures not only provide habitat across the forest profile, but produce vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed.
-From the humid and thermally stable understory to the dry and thermally variable canopy that experiences both hot and cold temperature extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or hinder the capacity for tall forests to generate high levels of biodiversity.
-For example, increased exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat that can support arboreal species and thus limit richness despite tall canopies [@oliveiraVerticalStratificationInfluences2019].
-Moreover, relatively low thermal variability in the canopies of tropical forests may increase the distinctness of vertical climate bands that support high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics [CITE].
+Modeling microclimates within the canopy requires detailed information on the vertical distribution of vegetation, which we obtained from spaceborne lidar data acquired through NASA's GEDI.
+The footprint level L2b data was preprocessed by Burns et al.
+[-@burnsGriddedGEDIVegetation2024; @burnsMultiresolutionGriddedMaps2024] to filter out low quality points and gridded into 1° x 1° tiles.
+We selected six tiles in which we modeled microclimates, representing boreal, temperate, and tropical forests in the Americas and Eurasia in non-mountainous regions that contained relatively large areas of tree cover based on Google satellite imagery and that were within the latitudinal bounds of the GEDI data.
+As the highest latitude is 51.6°N, we were only able to produce microclimate models for southern boreal forests.
+Boreal forests included regions in Canada (50.75°,-89.54°) and Russia (50.8°,137.77°); temperate forests included regions in the midwestern United States (37.69°,-91.08°) and Europe (48.33°,10.50°); tropical forests included regions in northern Peru (-5.14°,-75.57°) and Malaysia (4.95°,117.69°).
 
-in wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019].
-Taller canopies in tropical regions could thus support higher animal verticality and richness.
-In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
-Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.
+Points that fell in areas outside closed forest, as identified by Copernicus Global Dynamic Land Cover 2019 v3 [@buchhornCopernicusGlobalLand2020], areas that experienced deforestation between 2001 and 2024, as identified by Hansen global forest watch [@hansenHighResolutionGlobalMaps2013], and areas above 500 m in elevation were excluded from analysis.
+We restricted analysis to low elevation forests to control for any elevational impacts on vertical climate gradients.
+From the remaining points, we identified those that were recorded during the months with the average coldest minimum temperature and average warmest maximum temperature for the year in which they were obtained.
+These months were identified using ERA5 climate data [@hersbachERA5HourlyData2023].
+For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile.
+However, there were not 100 points that matched our selection criteria in temperate forests during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite) (26 points in temperate USA and 14 points in temperate Europe).
+All temperate forest points therefore fell in mixed rather than deciduous forests.
+There were also only 65 points that met our selection criteria in temperate European forests during the warmest month.
 
-Integrating new technologies can shed light on the mechanisms driving decades-old ecological patterns.
+For each point, GEDI data provided information on canopy height, plant area index (PAI), and vertical distribution of the PAI.
+The GEDI data describes the vertical distribution of PAI as the cumulative PAI from the top of the canopy to the stratum representing the vertical space from 0-5 m above the ground.
+For example, the PAI 0-5 m includes nearly all of the canopy space and is therefore nearly equal to the total PAI in most cases, while PAI 10-15 m describes the PAI from the top of the canopy to 10-15 m above the ground.
+However, the microclimate models require that the vertical PAI distribution represent the PAI within individual height bins and that the sum of the vertical PAI profile equal the total PAI.
+We therefore subtracted total PAI – PAI 0-5 m to represent the PAI in the 0-5 m bin, PAI 0-5 m – PAI 5-10 m to represent the PAI in the 5-10 m bin, and so forth.
 
-But only within the last decade have we been able to measure forest structure and acquire data on vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive understanding of the impact of vertical habitat structure on species richness patterns.
+### Climate data
 
-In forests, species assemblages stratify vertically along resource and microclimate gradients that extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition habitat across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+We used ERA5 for macroclimate forcing data.
+To reduce gridding effects of the data, we averaged the four closest grid cells to the location of the GEDI point using inverse distance weighting.
+ERA5 data was processed into the format required by the model using the 'mcera5' R package [@klingesMcera5DrivingMicroclimate].
+ [DETAILS ON DIURNAL TEMPERATURE CORRECTION AND LAND SEA MASK]
 
-where verticality declines, species may use terrestrial habitats during the dry season to maintain moisture requirements assemblage verticality declines because species exhibit semi-arboreal life strategies, which reduces assemblage verticality.
+### Soil data
 
-"short tropical forests show higher complexity relative to other forests of similar heights" - de Conto et al. 2023
+Soil data was obtained from the soilgrids database at a 250 m resolution using the soilDB R package, and included sand, silt, and clay content at the top soil layer (0-5 cm below the ground).
+[DESCRIBE METHODS USED TO CONVERT TO SOIL DATA USED BY MICROCLIMATE MODEL].
 
-This document presents most of the features of the package `r ftext("officedown", ft)`. 
-`r fp`
+MASS was used to produce models of canopy height by richness. sdmTMB was used to produce models of
+verticality. DHARMa was used to check model fit and assumptions. ggeffects was used to generate
+model predictions and confidence intervals. Figures were generating using tidyterra, patchwork,
+ggplot2
 
-## Table of content
+# Supplementary tables
 
-<!---BLOCK_TOC--->
+```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab"}
+# table from: "scripts/03_analyses/01_canopyheight_richness_relationships.R"
 
-## List of figures
+fit.global.tidy %>% 
+  mutate(term = ifelse(term == "canopy_height2", "canopy height", term)) %>% 
+  mutate(across(.cols = estimate:p.value, .fns = function(x) {round(x,2)})) %>% 
+  rename(SE = std.error, "p-value" = p.value) %>% 
+  flextable()
 
-<!---BLOCK_TOC{seq_id: 'fig'}--->
+```
 
-## List of tables
+```{r vert-model-results, tab.id="vert-model-results", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}
 
-<!---BLOCK_TOC{seq_id: 'tab'}--->
+amphmod = readRDS("results/sdmTMB_models2.3/amphibians_sesvert_tempdiu.rds")
+reptmod = readRDS("results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
+birdsmod = readRDS("results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
+mammalsmod = readRDS("results/sdmTMB_models2.3/mammals_sesvert_tempdiu.rds")
+
+amphmod = amphmod$fullmod$modlist[[1]]
+amphtab = tidy(amphmod) %>% 
+  mutate(Amphibians = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
+  dplyr::select(term, Amphibians)
+
+birdtab = birdsmod$fullmod %>% 
+  tidy() %>% 
+  mutate(Birds = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
+  dplyr::select(term, Birds)
+
+mammalstab = mammalsmod$fullmod %>% 
+  tidy() %>% 
+  mutate(Mammals = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
+  dplyr::select(term, Mammals)
+
+repttab = reptmod$fullmod %>% 
+  tidy() %>% 
+  mutate(Reptiles = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
+  dplyr::select(term, Reptiles)
+
+tab = plyr::join_all(list(birdtab, mammalstab, repttab, amphtab), by = "term", type = "left")
+
+tab %>%   flextable() %>% 
+  compose(i = ~term == "tmax_warm", j = "term", value = as_paragraph("Tmax", as_sub("warm"))) %>% 
+  compose(i = ~term == "I(tmax_warm^2)", j = "term", value = as_paragraph("Tmax", as_sub("warm"), as_sup("2"))) %>% 
+  compose(i = ~term == "tmin_cold", j = "term", value = as_paragraph("Tmin", as_sub("cold"))) %>% 
+  compose(i = ~term == "temp_diu", j = "term", value = as_paragraph("Temp", as_sub("diu"))) %>% 
+  compose(i = ~term == "precip_warm", j = "term", value = as_paragraph("Precip", as_sub("warm"))) %>% 
+  compose(i = ~term == "log_precip_dry", j = "term", value = as_paragraph("log10(precip", as_sub("dry"), ")")) %>% 
+  compose(i = ~term == "canopy_height2", j = "term", value = as_paragraph("Canopy height")) %>% 
+  compose(i = ~term == "veg_den", j = "term", value = as_paragraph("Vegetation density")) %>% 
+  compose(i = ~term == "precip_warm:canopy_height2", j = "term", value = as_paragraph("Precip", as_sub("warm"), ":Canopy height")) %>% 
+  compose(i = ~term == "tmin_cold:canopy_height2", j = "term", value = as_paragraph("Tmin", as_sub("cold"), ":Canopy height")) %>% 
+  compose(i = ~term == "tmax_warm:canopy_height2", j = "term", value = as_paragraph("Tmax", as_sub("warm"), ":Canopy height")) %>% 
+  compose(i = ~term == "I(tmax_warm^2):canopy_height2", j = "term", value = as_paragraph("Tmax", as_sub("warm"), as_sup("2"), ":Canopy height")) %>% 
+  compose(i = ~term == "log_precip_dry:canopy_height2", j = "term", value = as_paragraph("log10(precip", as_sub("dry"), "):Canopy height")) %>% 
+  compose(i = ~term == "temp_diu:canopy_height2", j = "term", value = as_paragraph("Temp", as_sub("diu"), ":Canopy height")) %>% 
+  autofit()
+
+```
 
@@
 
-## Sections
+# Supplementary figures
 
-### A two columns section
+```{r splot-global-chrich, fig.id="global-chrich",  fig.lp="suppfig", fig.autonum.start_at=1, width=90, height=90, fig.cap="Global relationship between total vertebrate richness and canopy height.
+Line indicates predictions from a negative binomial model of richness by canopy height ± 95% CI.
+Points indicate observed richness by canopy height.}
 
-<!---BLOCK_MULTICOL_START--->
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-Maecenas facilisis dapibus egestas.
-Curabitur dignissim pharetra pulvinar.
-Nunc bibendum elit sed cursus congue.
-
-`r run_columnbreak()`Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus.
-Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor.
-Donec consequat congue erat in iaculis.
-Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-Maecenas facilisis dapibus egestas.
-Curabitur dignissim pharetra pulvinar.
-Nunc bibendum elit sed cursus congue.
-Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus.
-Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor.
-Donec consequat congue erat in iaculis.
-Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.
-
-<!---BLOCK_MULTICOL_STOP{widths: [3,3], space: 0.2, sep: true}--->
-
-### End of two columns section
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-Maecenas facilisis dapibus egestas.
-Curabitur dignissim pharetra pulvinar.
-Nunc bibendum elit sed cursus congue.
-Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus.
-Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor.
-Donec consequat congue erat in iaculis.
-Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.
-
-<!---BLOCK_LANDSCAPE_START--->
-
-### Landscape section
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-
-```{r, echo=FALSE}
-barplot(1:5, col=1:5)
+include_graphics("figures/ms_figures/supp_figs/total_richness~canopy.png", error = F)
 ```
 
-<!---BLOCK_LANDSCAPE_STOP--->
+```{r splot-resids-taxa, fig.id="resids-taxa", fig.lp="suppfig", width=180, height=100, fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians."}
 
-### Normal section
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-Maecenas facilisis dapibus egestas.
-Curabitur dignissim pharetra pulvinar.
-Nunc bibendum elit sed cursus congue.
-Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus.
-Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor.
-Donec consequat congue erat in iaculis.
-Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Nulla eu pulvinar arcu, quis aliquam dui.
-In at cursus ante.
-Vestibulum non sagittis lacus.
-Duis vitae iaculis dui.
-Vivamus tempor, nibh ut pretium tempus, enim lorem dignissim quam, at euismod massa magna at magna.
-Sed facilisis dapibus diam nec volutpat.
-Maecenas facilisis dapibus egestas.
-Curabitur dignissim pharetra pulvinar.
-Nunc bibendum elit sed cursus congue.
-Curabitur ligula quam, iaculis faucibus orci quis, vestibulum lobortis lectus.
-Suspendisse fringilla nisl pulvinar, laoreet tellus sed, sollicitudin tortor.
-Donec consequat congue erat in iaculis.
-Curabitur luctus tellus ut turpis iaculis, nec laoreet ligula scelerisque.
-
-## Tables
-
-### Table 1
-
-```{r tab.cap="caption 1", tab.id="mtcars"}
-head(mtcars)
+include_graphics("figures/ms_figures/supp_figs/deviance_resids_taxa.png", error = F)
 ```
 
-### Table 2
+```{r splot-meanvert-biome, fig.id = "meanvert-biome", fig.lp="suppfig", width = 180, height = 130, fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
+(2017) and b) ecoregions aggregated to coarser classifications.
+Red diamonds indicate mean assemblage verticality.
+Ecoregions that do not contain woody vegetation were excluded."}
 
-```{r tab.cap="iris"}
-head(iris)
+include_graphics("figures/ms_figures/supp_figs/biome_mean_verticality.png", error = F)
 ```
 
-### Table 3
+```{r splot-meanprop-biome, fig.id="meanprop-biome", fig.lp="suppfig", width = 160, height = 160, fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
+(2017) and b) ecoregions aggregated to coarser classifications.
+Red diamonds indicate mean assemblage verticality.
+Ecoregions that do not contain woody vegetation were excluded."}
 
-```{r tab.cap="cars", tab.id="cars"}
-head(cars)
+include_graphics("figures/ms_figures/supp_figs/biome_mean_proportion_strategies.png", error = F)
 ```
 
-## figures
-
-### A boxplot
-
-```{r fig.cap="A boxplot", fig.id = "boxplot"}
-boxplot(1:8)
-```
-
-### A barplot
-
-```{r fig.cap="What a barplot", fig.id = "barplot"}
-barplot(1:8)
-```
-
-## Lists
-
-Amet nunc eros curabitur tellus massa, eros maximus porttitor sociosqu, pellentesque.
-
-* Erat mauris egestas finibus tincidunt sed in rhoncus a tellus etiam. 
-    - A adipiscing per ultricies justo tellus lorem. 
-        - Imperdiet ut dui primis, sed gravida, at sed nulla. 
-        - Sem posuere lacus consequat inceptos dapibus duis malesuada finibus. 
-    - Urna sed dui, ornare, eu turpis mus pellentesque amet amet bibendum. 
-* Himenaeos tincidunt, auctor dapibus scelerisque, montes nunc faucibus 
-sodales malesuada ridiculus sed cubilia ligula.
-
-1. Erat mauris egestas finibus tincidunt sed in rhoncus a tellus etiam. 
-    1. A adipiscing per ultricies justo tellus lorem. 
-        1. Imperdiet ut dui primis, sed gravida, at sed nulla. 
-        2. Sem posuere lacus consequat inceptos dapibus duis malesuada finibus. 
-    2. Urna sed dui, ornare, eu turpis mus pellentesque amet amet bibendum. 
-2. Himenaeos tincidunt, auctor dapibus scelerisque, montes nunc faucibus 
-sodales malesuada ridiculus sed cubilia ligula.
-
-Lorem dolor interdum orci eros pellentesque semper tristique, sodales, et sed
-ut! Porta mattis natoque et. Ac facilisi ipsum viverra elementum vestibulum
-ligula amet enim magnis luctus ullamcorper. Rhoncus rhoncus elit in at nisl.
-Tincidunt habitant sit.
-
-* Aptent conubia quam montes id sagittis.
-    - Mattis nisi nascetur, aliquam duis ex, tristique.
-        - Imperdiet ut dui primis, sed gravida, at sed nulla. 
-        - Donec ligula nulla ac. Nisl ac at accumsan sagittis eros felis lobortis amet 
-        nec phasellus urna bibendum sapien.
-    - Eu dui ac id, dictum proin consectetur convallis.
-* Facilisi eu lectus mauris lorem. Et sed sapien pellentesque sed etiam vehicula. 
-* In porttitor id lorem eu efficitur, nisl dis!
+\newpage
 
@@
 
-see figure \@ref(fig:boxplot) and table \@ref(tab:mtcars)!
+# SCRATCH SPACE
+
+"The observer who leaves the center of the earth by an infinitely small amount compared to the
+radius can reach a new world, so to speak, and he can observe more variations in the aspect of the
+soil and more modifications in the atmosphere than he would if he were to pass from one latitude to
+another" - Humbolt pg 77 Essay on the Geography of Plants
+
+"Current hypotheses for large-scale geographical patterns of species diversity fail to explain the
+extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5). We call
+this challenge “Humboldt’s enigma.”" - Rahbek
+
+-   TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE
+    IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE
+    REGION
+-   on second thought, I don't think this is necessary because we want to know why more species
+    can't exist in trees in boreal regions
+
+Just as topographic complexity and climate explain Humbolt's enigma in mountains, vegetative
+complexity and climate may play key roles in mediating the relationship between canopy height and
+species richness by impacting the availability of vertical niche opportunities and physiologically
+constraining arboreal lifestyles [CITE]. Densely layered forest structures not only provide habitat
+across the forest profile, but produce vertical microclimate gradients via interactions with
+macroclimate conditions, solar radiation, and wind speed. From the humid and thermally stable
+understory to the dry and thermally variable canopy that experiences both hot and cold temperature
+extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or
+hinder the capacity for tall forests to generate high levels of biodiversity. For example, increased
+exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat
+that can support arboreal species and thus limit richness despite tall canopies
+[@oliveiraVerticalStratificationInfluences2019]. Moreover, relatively low thermal variability in the
+canopies of tropical forests may increase the distinctness of vertical climate bands that support
+high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics
+[CITE].
+
+in wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates,
+which reduce thermal and hydric stress and may allow species to finely partition the entire
+three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in
+tropical regions could thus support higher animal verticality and richness. In contrast, extreme
+cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles,
+constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
+Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space,
+thereby limiting vertical stratification and species richness despite the presence of vegetation.
+
+Integrating new technologies can shed light on the mechanisms driving decades-old ecological
+patterns.
+
+But only within the last decade have we been able to measure forest structure and acquire data on
+vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive
+understanding of the impact of vertical habitat structure on species richness patterns.
+
+In forests, species assemblages stratify vertically along resource and microclimate gradients that
+extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a;
+@xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests increase both habitat volume and
+niche opportunities, enabling species to partition habitat across vertical strata akin to
+multi-floored apartments [@macarthurLimitingSimilarityConvergence1967;
+@macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
+
+In contrast, lower structural complexity in parts the Afrotropics
+[@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding
+less vertical assemblages and lower richness than expected based on the height of the canopy (Fig.
+\@ref(fig:fig1)).
+
+where verticality declines, species may use terrestrial habitats during the dry season to maintain
+moisture requirements assemblage verticality declines because species exhibit semi-arboreal life
+strategies, which reduces assemblage verticality.
+
+"short tropical forests show higher complexity relative to other forests of similar heights" - de
+Conto et al. 2023
+
+-   highest risk for to exceeding critical maxima in tropical lowlands and mid-latitudes (Sunday et
+    al. 2014)
+-   tolerance of max temps remains stable across latitudes - suggests advantage for temperate
+    ectotherms (Sunday et al. 2011)
+-   increasing tolerance of low min temps with latitude (Sunday et al.
+    2011)
+
+          -   Declines in tolerance of low minimum temperatures toward the equator suggest that the
+              cooler temperatures in the canopy may limit arboreality
+-   in terrestrial ectotherms, upper thermal limits decline marginally with latitude, while lower
+    thermal limits decline steeply with latitude (Sunday et al. 2019) - so upper thermal limits most
+    problematic in hot environments where species have lower thermal safety margins - safety margin
+    will be even lower in the canopy
+    -   steep vertical gradients in maximum temperatures reduce thermal safety margins, which may
+        pose challenges to fully arboreal life strategies, particularly in temperate regions where
+        maximum temperature gradients were greatest (fig) and where ectotherms are at the highest
+        ris for exceeding critical thermal maxima (Sunday et al. 2014)
+-   operative temps are typically higher than critical thermal maxima, suggesting that terrestrial
+    ectotherms must rely on behavioral adjustments to survive extreme heat (Sunday et al. 2014) -
+    this would force terrestriality, particularly in temperate latitudes where the vertical climate
+    gradient is very steep
+-   max air temps highest at mid-latitudes (Sunday et al. 2014) - I can show this with a map in the
+    supp figures
+-   heat tolerance varied little with latitude (Sundary et al. 2014)
+-   upper thermal tolerance limits were generally lower than maximum operative temps for ectotherms
+    (Sunday et al. 2014)
+-   cold tolerance limits generally lower than lowest operative temps (Sunday et al. 2014)
+-   reptiles usually have operative temps higher than critical maxima when in the sun, so must use
+    behavioral thermoregulation - particularly at lower to mid latitudes (Sunday et al. 2014) -
+    Therefore, in temperate forests where vertical max temp gradients are steep and canopies are
+    drier than the tropics, with canopies exceeding xC, high temps may prevent verticality
+-   amphibians can have operative temps that remain below critical maxima when water is available
+    (Sunday et al. 2014)
+-   burrowing or evapotranspiration can reduce body temps (Sunday et al.
+    2014) 
+-   when water is not available, amphibians are forced to seek shade under high temps to maintain
+    temps below thermal safety margin (Sunday et al. 2014)
+-   at high latitudes, reptile operative temperatue declines below CTmin, so they need to burrow to
+    stay warm enough (Sunday et al.
+    2014)

### 2025-07-17 scripts/03_analyses/03_glmms/01-2_sdmTMB_amphibians.R
update with null models and nonspatial models to calculate percent deviance

@@ -87,15 +87,6 @@ vif(lm(vert.mean.ses ~ canopy_height2 + veg_den + tmax_warm +
 
 # VIF all under 3
 
-f1 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold +
-               precip_warm + log_precip_dry +
-               canopy_height2 + veg_den +
-               precip_warm:canopy_height2 + tmin_cold:canopy_height2 + 
-               tmax_warm:canopy_height2 + I(tmax_warm^2):canopy_height2 +
-               log_precip_dry:canopy_height2)
-
-
-
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
@@ -114,199 +105,6 @@ ggplot(datfor, aes(tmin_cold, vert.mean.ses)) +
 
 
 
-# SES VERT ----------------------------------------------------------------
-
-# * - Set up mesh for analysis ------------------------------------------------
-
-# set up spatial mesh
-
-# first estimate range of spatial autocorrelation
-samp = dat %>% sample_n(1000)
-samp.cor = ncf::correlog(x = samp$x, y = samp$y, z = samp$vert.mean.ses, increment = 50000/1e5, resamp = 99)
-ncf:::plot.correlog(samp.cor)
-ncf:::plot.correlog(samp.cor, xlim = c(0,100))
-
-# set initial range as 30 and max.edge as range/5
-# https://haakonbakkagit.github.io/btopic104.html
-# Bakka, H., J. Vanhatalo, J. Illian, D. Simpson, and H. Rue. 2016. “Accounting for Physical Barriers in Species Distribution Modeling with Non-Stationary Spatial Random Effects.” arXiv preprint arXiv:1608.03787. Norwegian University of Science; Technology, Trondheim, Norway. 
-
-# removed SVC for canopy height
-fitmesh = fit_mesh(f1, dat, range = samp.cor$x.intercept, v = v, family = gaussian())
-
-mesh = fitmesh$meshes[[length(fitmesh$meshes)]]
-saveRDS(list(mesh = mesh), "results/sdmTMB_models2.3/amph_sesvert.rds")
-out = readRDS("results/sdmTMB_models2.3/amph_sesvert.rds")
-mesh = out$mesh
-
-
-# * - Compare models with AIC -------------------------------------------------
-
-taxon = "Amphibians"
-response_var = "SES verticality"
-fname_end = "amphibians_sesvert"
-
[…]

### 2025-07-17 scripts/03_analyses/03_glmms/01-2_sdmTMB_birds.R

@@ -107,6 +107,15 @@ f1 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold + temp_diu +
                tmax_warm:canopy_height2 + I(tmax_warm^2):canopy_height2 +
                log_precip_dry:canopy_height2 + temp_diu:canopy_height2)
 
+f2 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold + temp_diu +
+               precip_warm + log_precip_dry + 
+               canopy_height2 + veg_den)
+
+f3 = formula(vert.mean.ses ~ 1)
+
+forms = list(f1, f2, f3)
+
+
 samp = dat %>% sample_n(1000)
 samp.cor = ncf::correlog(x = samp$x, y = samp$y, z = samp$vert.mean.ses, increment = 50000/1e5, resamp = 99)
 ncf:::plot.correlog(samp.cor)
@@ -128,205 +137,20 @@ taxon = "Birds"
 response_var = "SES verticality"
 fname_end = "birds_sesvert"
 
-fullmod = compareMods_AIC(f = list(f1), dat, mesh, taxon = taxon, response_var = response_var, family = gaussian(), reml = T)
-fullmod = fullmod$modlist[[1]]
+mods.spatial = compareMods_AIC(f = forms, dat, mesh, taxon = taxon,
+                               response_var = response_var, family = gaussian(), reml = F, spatial = "on")
+mods.nonspatial = compareMods_AIC(f = forms, dat, mesh, taxon = taxon,
+                                  response_var = response_var, family = gaussian(), reml = F, spatial = "off")
+
+
+# fullmod = compareMods_AIC(f = list(f1), dat, mesh, taxon = taxon, response_var = response_var, family = gaussian(), reml = T)
+# fullmod = fullmod$modlist[[1]]
 #bestmod = run_extra_optimization(bestmod)
-saveRDS(list(mesh = mesh, fullmod = fullmod), "results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
+saveRDS(list(mesh = mesh, mods.spatial = mods.spatial, mods.nonspatial = mods.nonspatial), "results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
 
 
 
-plot_resids(mod = fullmod, response_var = "vert.mean.ses", 
+plot_resids(mod = mods.spatial$modlist[[1]], response_var = "vert.mean.ses", 
[…]

### 2025-07-17 scripts/03_analyses/03_glmms/01-2_sdmTMB_mammals.R

@@ -14,7 +14,7 @@ library(DHARMa)
 source("scripts/00_functions/manuscript_functions.R")
 source("scripts/00_functions/00_plot_functions.R")
 
-# amphibian forest only 50km resolution
+# mammals forest only 50km resolution
 raw = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv")
 v = vect("data/original/rnaturalearth_world.shp")
 v = project(v, "+proj=cea +datum=WGS84")
@@ -87,15 +87,6 @@ vif(lm(vert.mean.ses ~ canopy_height2 + veg_den + tmax_warm + I(tmax_warm^2)  + 
 
 # VIF all under 5
 
-# f1 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold +
-#                precip_warm + log_precip_dry +
-#                canopy_height2 + veg_den +
-#                precip_warm:canopy_height2 + tmin_cold:canopy_height2 + 
-#                tmax_warm:canopy_height2 + I(tmax_warm^2):canopy_height2 +
-#                log_precip_dry:canopy_height2)
-# 
-
-
 dat$x = dat$x/1e5
 dat$y = dat$y/1e5
 dat$ecoregion = factor(dat$ecoregion)
@@ -104,97 +95,6 @@ dat$biorealm = factor(paste(dat$biome, dat$realm, sep = "_"))
 dat$realm = factor(dat$realm)
 
 # SES VERT ----------------------------------------------------------------
-
-# * - Set up mesh for analysis ------------------------------------------------
-
-# set up spatial mesh
-
-# first estimate range of spatial autocorrelation
-samp = dat %>% sample_n(1000)
-samp.cor = ncf::correlog(x = samp$x, y = samp$y, z = samp$vert.mean.ses, increment = 50000/1e5, resamp = 99)
-ncf:::plot.correlog(samp.cor)
-ncf:::plot.correlog(samp.cor, xlim = c(0,100))
-
-# set initial range as 30 and max.edge as range/5
-# https://haakonbakkagit.github.io/btopic104.html
[…]

### 2025-07-17 scripts/03_analyses/03_glmms/01-2_sdmTMB_reptiles.R

@@ -99,6 +99,14 @@ f1 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold + temp_diu +
                tmax_warm:canopy_height2 + I(tmax_warm^2):canopy_height2 +
                log_precip_dry:canopy_height2 + temp_diu:canopy_height2)
 
+f2 = formula(vert.mean.ses ~ tmax_warm + I(tmax_warm^2) + tmin_cold + temp_diu +
+               precip_warm + log_precip_dry + 
+               canopy_height2 + veg_den)
+
+f3 = formula(vert.mean.ses ~ 1)
+
+forms = list(f1, f2, f3)
+
 samp = dat %>% sample_n(1000)
 samp.cor = ncf::correlog(x = samp$x, y = samp$y, z = samp$vert.mean.ses, increment = 50000/1e5, resamp = 99)
 ncf:::plot.correlog(samp.cor)
@@ -120,11 +128,13 @@ taxon = "Reptiles"
 response_var = "SES verticality"
 fname_end = "reptiles_sesvert"
 
-fullmod = compareMods_AIC(f = list(f1), dat, mesh, taxon = taxon, response_var = response_var, family = gaussian(), reml = T)
+mods.spatial = compareMods_AIC(f = forms, dat, mesh, taxon = taxon,
+                               response_var = response_var, family = gaussian(), reml = F, spatial = "on")
+mods.nonspatial = compareMods_AIC(f = forms, dat, mesh, taxon = taxon,
+                                  response_var = response_var, family = gaussian(), reml = F, spatial = "off")
 
-fullmod = fullmod$modlist[[1]]
-#bestmod = run_extra_optimization(bestmod)
-saveRDS(list(mesh = mesh, fullmod = fullmod), "results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
+
+saveRDS(list(mesh = mesh, mods.spatial = mods.spatial, mods.nonspatial = mods.nonspatial), "results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
 
 
 

### 2025-07-17 scripts/03_analyses/03_glmms/03_conditional_effect_plots.R

@@ -1,0 +1,132 @@
+# Goal: Make response plots to visualize predictions of SES verticality across predictor variables
+# These are conditional plots with all variables held at the mean, while the variable of interest is allowed to vary
+# from min to max value
+
+library(sdmTMB)
+library(ggeffects)
+library(foreach)
+library(doParallel)
+library(tidyverse)
+library(data.table)
+
+amph = readRDS("results/sdmTMB_models2.3/amphibians_sesvert_tempdiu.rds")
+amphmod = amph$fullmod$modlist[[1]]
+
+birds = readRDS("results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
+birdsmod = birds$fullmod
+
+mammals = readRDS("results/sdmTMB_models2.3/mammals_sesvert_tempdiu.rds")
+mammalsmod = mammals$fullmod
+
+repts = readRDS("results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
+reptsmod = repts$fullmod
+
+range(birdsmod$data$log_precip_dry)
+range(birdsmod$data$tmax_warm)
+range(birdsmod$data$tmin_cold)
+range(birdsmod$data$canopy_height)
+range(birdsmod$data$veg_den)
+range(birdsmod$data$precip_warm)
+range(birdsmod$data$temp_diu)
+
+var = c(
+  "log_precip_dry",
+  "tmax_warm",
+  "tmin_cold",
+  "canopy_height2",
+  "veg_den",
+  "precip_warm",
+  "temp_diu",
+  "log_precip_dry,canopy_height2",
+  "tmax_warm,canopy_height2",
+  "tmin_cold,canopy_height2",
+  "precip_warm,canopy_height2",
+  "temp_diu,canopy_height2",
+  "canopy_height2,tmin_cold"
+  )
+
+
+mods = list(amphmod, birdsmod, mammalsmod, reptsmod) # list of model fits
+names(mods) = c("Amphibians", "Birds", "Mammals", "Reptiles")
+predfit = list() # store predictions for each taxon
+
+# for each taxon,
+# predict varying one variable and holding others constant at their mean
+for(m in 1:length(mods)) {
+  modi = mods[[m]] # select model
+  pred.df = data.frame() # dataframe to store results
+  for(i in var) {
+    
+    # if varying 2 variables (canopy height and a climate var)
+    if(grepl(",", i)) {
+      v = str_split_1(i, ",")
+      v1 = c(-2,0,2)
[…]

### 2025-07-17 manuscript.Rmd
revise ms methods - microclimate methods

@@
 ---
-bibliography: "`r rbbt::bbt_write_bib('r-refs.bib', overwrite = TRUE)`"
+bibliography: "r-refs.bib"
 csl: science.csl
@@
 Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot or cold climates, where evapotranspiration to maintain body temperatures below thermal maxima or basking to maintain body temperatures above thermal minima can quickly lead to desiccation without adequate physiological defenses [@scheffersIncreasingArborealityAltitude2013; @tracyNotJustSmall2010; @mckechnieAvianThermoregulationHeat2017].
-By imposing environmental filters on the functional
-morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage
-verticality and limit the number of species that an increase in canopy height can support, in line
-with the physiological tolerance hypothesis [@curriePredictionsTestsClimatebased2004]. For example,
-in the seasonally dry forests of of Panama, water limitation in the canopy causes the vertical
-stratification of frog communities to collapse toward the ground during the dry season - a shift
-coincident with declines in richness and homogenization of arboreal diversity
-[@bashamVerticalStratificationCollapses2020]. Dry climates can therefore reduce the capacity for
-increasingly tall canopies to support greater biodiversity (Fig. \@ref(fig:fig1)), just as cold
-climates limit the capacity for tall mountains to generate biodiversity by reducing the elevation of
-the snow line [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013;
-@janzenWhyMountainPasses1967]. In contrast, high dry season precipitation in tropical moist forests
-supports highly vertical assemblages by reducing physiological constraints on arboreal life
-strategies in the upper canopy. Species are therefore able to finely partition the entire volume of
-forested habitat, leading to a dense packing of species in the canopy and allowing increases in
-canopy height to yield rapid increases in species richness
-[@oliveiraVerticalStratificationInfluences2019] (Fig. \@ref(fig:fig1)).
+By imposing environmental filters on the functional morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage verticality and limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [@curriePredictionsTestsClimatebased2004].
+For example, in the seasonally dry forests of of Panama, water limitation in the canopy causes the vertical stratification of frog communities to collapse toward the ground during the dry season - a shift coincident with declines in richness and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020].
+Dry climates can therefore reduce the capacity for increasingly tall canopies to support greater biodiversity (Fig.
+\@ref(fig:fig1)), just as cold climates limit the capacity for tall mountains to generate biodiversity by reducing the elevation of the snow line [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].
+In contrast, high dry season precipitation in tropical moist forests supports highly vertical assemblages by reducing physiological constraints on arboreal life strategies in the upper canopy.
+Species are therefore able to finely partition the entire volume of forested habitat, leading to a dense packing of species in the canopy and allowing increases in canopy height to yield rapid increases in species richness [@oliveiraVerticalStratificationInfluences2019] (Fig.
+\@ref(fig:fig1)).
 
-Maximum and minimum temperatures had less consistent effects across taxa (Fig. \@ref(fig:fig3)), but
-may still place physiological constraints on arboreal life strategies that influence the capacity
-for taller canopies to support richer species assemblages, particularly in regions exhibiting
-steeper vertical gradients characterized by hotter maximum and colder minimum temperatures. For
-example, high maximum temperatures generally reduced verticality and/or reduced the positive impact
-of canopy height on verticality, barring mammals for which verticality responded positively to
-maximum temperatures. High temperatures may cause ectotherm body temperatures to exceed critical
-thermal limits [@sundayThermalsafetyMarginsNecessity2014]. While consistent water availability in
-the canopies of tropical wet forests may allow ectotherms to maintain body temperature below
-critical limits, the drier conditions of temperate forests combined with steeper vertical gradients
-in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal
-habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats
-[@sundayThermalsafetyMarginsNecessity2014] (Fig. \@ref(fig:fig4)). Moreover, low minimum
-temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig.
-\@ref(fig:fig3)). Cold temperatures at higher latitudes can cause ectotherm body temperatures to
-exceed critical thermal minima outside fossorial habitats or impose energetic constraints to
-endotherms, particularly in the windier conditions that characterize the canopy
-[@sundayThermalsafetyMarginsNecessity2014] [REPLACE MAMMAL EXAMPLE WITH BIRD PHYSIOLOGY]. The
-physiological constraints imposed by minimum temperatures may act across the entire canopy given
-given their shallow vertical gradient, (Fig. \@ref(fig:fig4)), posing severe limitations on the
-capacity for taller canopies to increase species richness in boreal forests.
+Maximum and minimum temperatures had less consistent effects across taxa (Fig.
+\@ref(fig:fig3)), but may still place physiological constraints on arboreal life strategies that influence the capacity for taller canopies to support richer species assemblages, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures.
+For example, high maximum temperatures generally reduced verticality and/or reduced the positive impact of canopy height on verticality, barring mammals for which verticality responded positively to maximum temperatures.
+High temperatures may cause ectotherm body temperatures to exceed critical thermal limits [@sundayThermalsafetyMarginsNecessity2014].
+While consistent water availability in the canopies of tropical wet forests may allow ectotherms to maintain body temperature below critical limits, the drier conditions of temperate forests combined with steeper vertical gradients in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats [@sundayThermalsafetyMarginsNecessity2014] (Fig.
+\@ref(fig:fig4)).
+Moreover, while models indicated that higher minimum temperatures were associated with lower verticality for birds and amphibians because verticality is low in hot but drier tropical regions while remaining high in cooler but wet tropical montane regions, low minimum temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig.
+\@ref(fig:fig3)).
+The interaction with canopy height reflects the physiological constraints that cold temperatures can impose on arboreal life strategies.
+Cold temperatures at higher latitudes can cause ectotherm body temperatures to exceed critical thermal minima outside fossorial habitats or impose energetic constraints on endotherms, particularly in the windier conditions of the upper canopy [@sundayThermalsafetyMarginsNecessity2014; @villen-perezWoodlandMediterraneanBirds2014; @wolfThermalEffectsRadiation1996].
+For example, birds that typically forage in the upper canopy shift to lower canopy strata during periods of cold winter temperatures [@villen-perezWoodlandMediterraneanBirds2014].
+Cold temperatures therefore limit the degree to which animals can specialization to upper canopy strata and consequently prevent taller canopies from increasing species richness.
 
-Temporal climate variability may also play a role in the capacity for verticality to mediate
-relationships between species richness and canopy height by impacting climatic distinctness. In
-montane regions, climatic distinctness across elevations declines with latitude due to increasing
-seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019;
-@janzenWhyMountainPasses1967]. Low seasonal variation in the tropics produces narrow and distinct
-climate bands that exhibit little overlap across elevations. Unique climates across elevations
-restrict species' range size, leading to high rates of turnover across elevations that allow
-tropical mountains to generate extraordinary levels of biodiversity [@janzenWhyMountainPasses1967;
-@sheldonFiftyYearsMountain2018; @ghalamborAreMountainPasses2006; @rahbekHumboldtsEnigmaWhat2019]. In
-contrast, high thermal variation in temperate regions increases the similarity between climates
-experienced at high and low elevations, reducing the capacity for temperate mountains to support
-high rates of turnover and thus high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019;
-@ghalamborAreMountainPasses2006].
+Temporal climate variability may also play a role in the capacity for verticality to mediate relationships between species richness and canopy height by impacting climatic distinctness.
+In montane regions, climatic distinctness across elevations declines with latitude due to increasing seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019; @janzenWhyMountainPasses1967].
+Low seasonal variation in the tropics produces narrow and distinct climate bands that exhibit little overlap across elevations.
+Unique climates across elevations restrict species' range size, leading to high rates of turnover across elevations that allow tropical mountains to generate extraordinary levels of biodiversity [@janzenWhyMountainPasses1967; @sheldonFiftyYearsMountain2018; @ghalamborAreMountainPasses2006; @rahbekHumboldtsEnigmaWhat2019].
+In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and thus high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019; @ghalamborAreMountainPasses2006].
 
-Within wooded habitats, diurnal temperature range may influence the distinctness of vertical climate
-bands from the ground to the canopy, and thus the vertical turnover of species. Our models of
-verticality indicate that greater diurnal temperature variation reduces verticality of mammals,
-though effects for other classes were insignificant (Fig. \@ref(fig:fig3)). Furthermore,
-microclimate models indicated that increases in daily temperature variation, as well as the rate of
-increase in temperature variation from the ground to the canopy, were greater in higher latitude
-forests (Fig. \@ref(fig:fig4)c). High daily temperature variation reduced the distinctness of
-climate bands across the vertical profile, with vertical zonation collapsing nearly completely
-during morning hours in temperate and boreal forests (Fig. \@ref(fig:fig4)B). Without unique
-climatic niches across the forest profile, these regions are unlikely to support rapid vertical
-turnover, limiting the number of species that may be added to the assemblage for a given increase in
-canopy height, even in wetter temperate regions with climate conditions that physiologically support
-arboreal strategies. In contrast, tropical canopies maintained more distinct vertical climate bands
-throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
-Narrow and distinct vertical climate niches may therefore support rapid vertical turnover, leading
-to greater increases in richness with canopy height and enhancing the capacity for tropical canopies
-generate biodiversity.
+Within wooded habitats, diurnal temperature range may influence the distinctness of vertical climate bands from the ground to the canopy, and thus the vertical turnover of species.
+Our models of verticality indicate that greater diurnal temperature variation reduces verticality of mammals, though effects for other classes were insignificant (Fig.
+\@ref(fig:fig3)).
+Furthermore, microclimate models indicated that increases in daily temperature variation, as well as the rate of increase in temperature variation from the ground to the canopy, were greater in higher latitude forests (Fig.
+\@ref(fig:fig4)c).
+High daily temperature variation reduced the distinctness of climate bands across the vertical profile, with vertical zonation collapsing nearly completely during morning hours in temperate and boreal forests (Fig.
+\@ref(fig:fig4)B).
+Without unique climatic niches across the forest profile, these regions are unlikely to support rapid vertical turnover, limiting the number of species that may be added to the assemblage for a given increase in canopy height, even in wetter temperate regions with climate conditions that physiologically support arboreal strategies.
+In contrast, tropical canopies maintained more distinct vertical climate bands throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
+Narrow and distinct vertical climate niches may therefore support rapid vertical turnover, leading to greater increases in richness with canopy height and enhancing the capacity for tropical canopies generate biodiversity.
 
@@
 
-Decades of research have elucidated the mechanisms that allow mountains to generate high levels of biodiversity, particularly in the tropics [CITE].
+Centuries of research have elucidated the mechanisms that allow mountains to generate high levels of biodiversity, particularly in the tropics [@vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967; @rahbekHumboldtsEnigmaWhat2019].
 Yet only recently has the three-dimensional nature of terrestrial landscapes been widely recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
@@
 
-```{r fig1, fig.id="fig1"}
+```{r fig1, fig.id="fig1", fig.cap="Spatial distribution of a) canopy height, b) vertebrate richness (for all vertebrates), and c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height.
+d) Relationships between species richness and canopy height for each taxon in different biomes.
+Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=180}
+
 # Fig. 1
-
-# Legend: Spatial distribution of a) canopy height, b) vertebrate richness (all vertebrates combined), and
-# c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height
-# d) Relationships between species richness and canopy height for each taxon in each
-# biome category. Dots indicate observed values and lines indicate negative binomial
-# model fits of richness by canopy height.
 
@@
 High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance." c) Correlations between verticality and species richness globally.
-Lines indicate linear best fit line with 95% CIs.
-}
-# Fig. 1
+Lines indicate linear best fit line with 95% CIs., fit.width = 150, fig.height=150}
+
+# Fig. 2
 
@@
 
-```{r fig4, fig.id="fig4", fig.cap=", fig.width=140, fig.height=120}
+```{r fig4, fig.id="fig4", fig.cap="a) Locations of modeled vertical microclimate profiles.
+b) Average hourly microclimate across relative heights in the canopy.
+Relative height is calculated as absolute height above ground divided height of the canopy.
+Shading represents the range of temperatures experienced in the uppermost vertical band (0.9-1 relative height) and the lowermost vertical band (0-0.1 relative height).
+c) Vertical profiles of microclimate variables and plant area volume density (PAVD) (median +/- SD) in tropical, temperate, and boreal forest sites.
+PAVD was obtained from GEDI shots.", fig.width=140, fig.height=120}
 
@@
 
-We mapped global richness and verticality of terrestrial vertebrate assemblages in wooded habitats
-[@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells using a cylindrical equal
-area projection. Range maps were obtained from the IUCN Red List for mammals and amphibians
-[@iucnIUCNRedList2023], BirdLife International for bird breeding and resident ranges
-[@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global
-Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017;
-@caetanoAutomatedAssessmentReveals2022]. Overall, we obtained range maps for `r nbirds` birds,
-`r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians.
+We mapped global richness and verticality of terrestrial vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells using a cylindrical equal area projection.
+Range maps were obtained from the IUCN Red List for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for bird breeding and resident ranges [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].
+Overall, we obtained range maps for `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians.
 
-To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a
-verticality score from zero to one and then took the average verticality of all species whose ranges
-overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based
-on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the
-TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024a]. For birds, we
-calculated verticality scores based on their use of different foraging strata as identified in the
-EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale
-vertical classification than the dataset from Moura et al.
-[-@mouraPhylogenyinformedCharacterisationGlobal2024a]. For birds, verticality scores were as
-follows: foraging below the water surface, around the water surface, and on the ground = 0.5,
-foraging in the understory = 0.667; foraging in the mid-high canopy = 0.8337; foraging in or above
-the upper canopy = 1. For all taxa, if species used multiple vertical habitats, vertical scores were
-averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it
-would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that
-did not occur in terrestrial habitats (e.g., marine mammals) were excluded from the analysis.
+To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score from zero to one and then took the average verticality of all species whose ranges overlapped with the grid cell.
+For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024a].
+For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale vertical classification than the dataset from Moura et al.
+[-@mouraPhylogenyinformedCharacterisationGlobal2024a].
+For birds, verticality scores were as follows: foraging below the water surface, around the water surface, and on the ground = 0.5, foraging in the understory = 0.667; foraging in the mid-high canopy = 0.8337; foraging in or above the upper canopy = 1.
+For all taxa, if species used multiple vertical habitats, vertical scores were averaged.
+For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75.
+Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats (e.g., marine mammals) were excluded from the analysis.
 
 To examine spatial patterns of assemblage verticality independently of species richness and accounting for biogeographic history, we calculated the standard effect size (SES) of mean verticality (hereafter ‘verticality’): $SES=(x_i-\bar{x})/sd(x)$ where x is a vector of expected mean verticality scores given random selection of species from a species pool.
-For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm (Dinerstein et al.
-2017) and generated a null distribution of expected verticality for each grid cell based on 100 simulations.
+For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell based on 100 simulations.
 Realms included Palearctic, Nearctic, Indomalayan, Neotropic, Oceania, Afrotropic, and Australasia.
@@
 We first modeled richness as a function of canopy height using a generalized linear model with a Poisson distribution and a log-link.
-Models were produced for global richness across all vertebrate class, global richness for each vertebrate class, and biome richness for each vertebrate class.
+Models were produced for global richness across all vertebrate classes and global richness for each vertebrate class.
+For each vertebrate class, we then modeled richness as a function of the interactive effect between canopy height and biome using a means parameterization.
 Initial model evaluation indicated over-dispersion in model residuals.
-We therefore fit a negative binomial model with a log-link function.
-Since richness often exhibits a log-log relationship with covariates, we additionally fit a model of richness as a function of log(canopy height), and compared this model to the original using AIC and pseudo-R^2^ values.
-We report results of richness as a function of canopy height, as it consistently performed better except for the model for reptiles in temperate forests.
+We therefore fit negative binomial models with a log-link function.
+Since richness often exhibits a log-log relationship with covariates, we additionally fit models of richness as a function of log(canopy height) or the interactive effect of log(canopy height) and biome, and compared this model to the original using AIC and conditional percent deviance explained (CPDE).
+Models fit with canopy height rather than log(canopy height) consistently performed better, and we report these results.
 
@@
 
-We also examined relationships between richness and verticality using Pearson correlations.
+We additionally examined relationships between richness and verticality using Pearson correlations.
 
@@
 
-We examined relationships between verticality and environmental covariates using spatial linear mixed models (LMMs).
-Grid cells with fewer than five species were excluded, because they may provide unrealiable estimates of the SES.
+We examined relationships between the verticality of each taxon and environmental covariates described above using spatial linear mixed models (LMMs).
+Grid cells with fewer than five species were excluded, because they may provide unreliable estimates of the SES.
 To linearize the relationships between predictor and response variables, we log10-transformed precipitation of the driest month.
-We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using Pearson correlations.
+We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using Pearson correlations and variance inflation factors.
 We excluded precipitation seasonality, temperature seasonality, and foliage height diversity due to strong correlations (|r| > 0.8) with dry season precipitation, minimum temperature of the coldest month, and canopy height, respectively.
-Additionally, we included a quadratic effect for maximum temperature of the warmest month, since extremely high temperatures may inhibit arboreal life strategies.
+The remaining variables all had VIF < 5.
+Additionally, we included a quadratic effect for maximum temperature of the warmest month, because extremely high temperatures are expected to inhibit arboreal life strategies as well as interaction effects between the climatic covariates and canopy height.
+Interactions were included because we expect the impact of canopy height on verticality to decline when harsh macroclimate conditions, including extreme cold, hot, or dry conditions, reduce the suitability of canopy microhabitats for arboreal life strategies, or when high climate variability reduces distinctness of vertical climate bands.
+The fixed effects therefore included Tmax~warm~, Tmax~warm~^2^, Tmin~cold~, Temp~diu~, Precip~warm~, log10(Precip~dry~), vegetation density, and canopy height, as well as interactions between canopy height and the climatic covariates.
 
-Spatial LMMs were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which approximates a spatial random effect using a finite element analysis mesh as implemented in R-INLA [@thorsonSpatioTemporalModelsEcologists2024].
+Spatial LMMs were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which accounts for spatially autocorrelated data using stochastic partial differential equations (SPDE) to approximate spatial Gaussian random fields as Gaussian Markov Random Fields (GMRF) with a Matérn covariance function [@lindgrenExplicitLinkGaussian2011].
 The full model can be described as:
@@
 \begin{align}
-y_s ∼ 𝜇_{s}, \\
+y_s ∼ N(𝜇_{s}, \sigma^2), \\
 𝜇_s = 𝜲_s𝛽 + ⍵_s, \\
-⍵_s ∼ MVNormal(0,𝚺_⍵)
+⍵_s ∼ MVNormal(0,𝚺_⍵),
 \end{align}
@@
 
-where $y_s$ is verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected verticality.
-$Χ_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients, including additive effects of Tmax~warm~, Tmax~warm~^2^, Tmin~cold~, Temp~diu~, Precip~warm~, log10(Precip~dry~), vegetation density, and canopy height, as well as interactions between canopy height and the five climatic variables.
-Interactions were included because we expect the impact of canopy height on verticality to decline when harsh macroclimate conditions, including extreme cold, hot, or dry conditions, reduce the suitability of canopy microhabitats for arboreal life strategies, or when high climate variability reduces distinctness of vertical climate bands.
-$⍵_s$ represents a spatial random effect at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates.
-The spatial random effect was modeled as a Gaussian random field with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance.
-Gaussian random fields were approximated for models of verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].
+where the response variable at location $s$ is normally distributed with mean $\mu_s$ and variance $\sigma^2$; $𝜇_s$ is the expected verticality.
+$Χ_s$ is a design matrix at location $s$; 𝛽 is a vector of fixed-effect coefficients; $⍵_s$ represents a spatial random effect at location $s$; and $\Sigma_{\omega}$ represents the covariance matrix of the spatial random field.
+The SPDE method requires a finite element analysis mesh, which consists of a set of triangles covering the spatial domain, and allows for the approximation of the spatial random effect at mesh vertices [@thorsonSpatioTemporalModelsEcologists2024].
+We created a mesh for each vertebrate class using *fmesher* and tuned each mesh, such that the maximum edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].
 
-We validated models first using the ‘sanity’ function from the sdmTMB package.
-We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to assess residual normality and lack of patterning across space.
+We evaluated models using the sanity check function in sdmTMB and by examining quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022].
+To determine the degree to which environmental factors improve model performance, we compare this model to a spatial random effect model with  model with We also report conditional percent deviance explained (CPDE).
 
@@
 
-To examine vertical microclimate gradients, we mechanistically modeled microclimates from the ground to the canopy using the ‘micropoint’ R package [@macleanMicropointCanopyPoint2025].
-Models were produced for boreal forests, temperate forests, and tropical forests from the Americas and from Eurasia.
+To examine vertical microclimate gradients, we mechanistically modeled microclimates from the ground to the canopy using the ‘micropoint’ R package [@macleanMicropointCanopyPoint2025; @macleanMicroclimfFastModelling2025].
+Point models of vertical climate profiles were produced for boreal, temperate, and tropical forests from the Americas and Eurasia within six 1° x 1° tiles located in regions with relatively flat topography that contained relatively large areas of tree cover based on Google satellite imagery and that were within the latitudinal bounds of the GEDI data, which provided information on vegetation structure (see below).
+Boreal forests included regions in Canada (50°,-90°) and Russia (50°,137°); temperate forests included regions in the midwestern United States (37°,-92°) and Europe (48°,10°); tropical forests included regions in northern Peru (-6°,-76°) and Malaysia (4°,117°), where coordinates indicate the lower left corners of tiles.
+
+Points within these tiles were selected as locations of GEDI footprint level L2b data, which was preprocessed by Burns et al.
+[-@burnsGriddedGEDIVegetation2024; @burnsMultiresolutionGriddedMaps2024] to filter out low quality points.
+From the high quality points, we excluded points that 1) fell outside closed forest, as identified by Copernicus Global Dynamic Land Cover 2019 v3 [@buchhornCopernicusGlobalLand2020], 2) were located in areas that experienced deforestation between 2001 and 2024, as identified by Hansen global forest watch [@hansenHighResolutionGlobalMaps2013], 3) were obtained from coverage rather than power beams, 4) represented vegetation less than 10 m tall, and 4) occurred above 500 m in elevation.
+We restricted analysis to low elevation forests to control for any elevational impacts on vertical climate gradients.
+From the remaining points, we identified those that were recorded during the months with the average coldest minimum temperature and average warmest maximum temperature for the year in which they were obtained.
+These months were identified using the ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023].
+For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile.
+However, there were only 26 points in temperate USA and 14 points in temperate Europe during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite).
+The points retained therefore represented mixed rather than deciduous forest types.
+Furthermore, only 65 points met our selection criteria in temperate European forests during the warmest month.
+For each point, the microclimate model was run for the month and year in which the point was obtained at 0.15 cm above ground, 2 m above ground, and then from 5 m to the top of the canopy at 5 m intervals (equivalent to the vertical resolution of GEDI data) to produce hourly estimates of air temperature and relative humidity.
+
+### Climate data
+
+The ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023] was used for meteorological inputs to the model, which provided near surface, open-air climate conditions at a 0.25$^\circ$ spatial resolution and hourly temporal resolution.
+ERA5 climate data was processed using the *mcera5* R package, which converts ERA5 variables into those required by the model and applies a coastal diurnal temperature correction to terrestrial locations located within predominanrly marine grid cells [@klingesMcera5DrivingMicroclimate2022].
+Meteorological conditions at the GEDI point of interest were then extracted from the climate grid by taking the weighted average of climate conditions across the four nearest neighboring cells, where the weight was defined as the inverse distance to the point.
+Finally, we applied an elevational correction that adjusts temperature and pressure at the location of interest by accounting for the difference in elevation between the center of the ERA5 grid cell and the point of interest using a humidity-dependent lapse rate.
 
@@
 
-Modeling microclimates within the canopy requires detailed information on the vertical distribution of vegetation, which we obtained from spaceborne lidar data acquired through NASA's GEDI.
-The footprint level L2b data was preprocessed by Burns et al.
-[-@burnsGriddedGEDIVegetation2024; @burnsMultiresolutionGriddedMaps2024] to filter out low quality points and gridded into 1° x 1° tiles.
-We selected six tiles in which we modeled microclimates, representing boreal, temperate, and tropical forests in the Americas and Eurasia in non-mountainous regions that contained relatively large areas of tree cover based on Google satellite imagery and that were within the latitudinal bounds of the GEDI data.
-As the highest latitude is 51.6°N, we were only able to produce microclimate models for southern boreal forests.
-Boreal forests included regions in Canada (50.75°,-89.54°) and Russia (50.8°,137.77°); temperate forests included regions in the midwestern United States (37.69°,-91.08°) and Europe (48.33°,10.50°); tropical forests included regions in northern Peru (-5.14°,-75.57°) and Malaysia (4.95°,117.69°).
-
-Points that fell in areas outside closed forest, as identified by Copernicus Global Dynamic Land Cover 2019 v3 [@buchhornCopernicusGlobalLand2020], areas that experienced deforestation between 2001 and 2024, as identified by Hansen global forest watch [@hansenHighResolutionGlobalMaps2013], and areas above 500 m in elevation were excluded from analysis.
-We restricted analysis to low elevation forests to control for any elevational impacts on vertical climate gradients.
-From the remaining points, we identified those that were recorded during the months with the average coldest minimum temperature and average warmest maximum temperature for the year in which they were obtained.
-These months were identified using ERA5 climate data [@hersbachERA5HourlyData2023].
-For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile.
-However, there were not 100 points that matched our selection criteria in temperate forests during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite) (26 points in temperate USA and 14 points in temperate Europe).
-All temperate forest points therefore fell in mixed rather than deciduous forests.
-There were also only 65 points that met our selection criteria in temperate European forests during the warmest month.
-
-For each point, GEDI data provided information on canopy height, plant area index (PAI), and vertical distribution of the PAI.
+Vegetation structure influences the amount of radiation and vertical air mixing. We obtained information on vegetation from the GEDI footprint level L2b data.
+For each point, GEDI data provided information on canopy height (RH100; the height at which 100% of the waveform was returned), plant area index (PAI), and the vertical distribution of the PAI.
 The GEDI data describes the vertical distribution of PAI as the cumulative PAI from the top of the canopy to the stratum representing the vertical space from 0-5 m above the ground.
@@
 
-### Climate data
+*Micropoint* requires a number of additional parameters describing vegetation structure that are not available from remote sensing products, such leaf angle distribution and stomatal conductance, among others.
+We use *microclimf's* 'vegpfromhab()' function to estimate these parameters based on land cover type.
+Land cover was determined based on MODIS land cover type (MCD12Q1) version 6.1 for the year 2022.
 
-We used ERA5 for macroclimate forcing data.
-To reduce gridding effects of the data, we averaged the four closest grid cells to the location of the GEDI point using inverse distance weighting.
-ERA5 data was processed into the format required by the model using the 'mcera5' R package [@klingesMcera5DrivingMicroclimate].
- [DETAILS ON DIURNAL TEMPERATURE CORRECTION AND LAND SEA MASK]
+Vegetation parameters were assumed to remain constant for the month and year in which the GEDI shot was obtained.
 
-### Soil data
+[WE SHOULD PRODUCE MODELS AT DRIEST MONTH OF THE YEAR]
 
-Soil data was obtained from the soilgrids database at a 250 m resolution using the soilDB R package, and included sand, silt, and clay content at the top soil layer (0-5 cm below the ground).
-[DESCRIBE METHODS USED TO CONVERT TO SOIL DATA USED BY MICROCLIMATE MODEL].
+### Topography and soil composition
+
+To characterize topography, we extracted elevation from the lowest mode return of the GEDI shot.
+Slope and aspect were calculated using *terrain()* based on SRTM elevation data at a 1 km resolution, which was obtained using the *geodata* R package.
+Sand, silt, and clay content in the top soil layer (0-5 cm below the ground) were obtained from the SoilGrids 2.0 at a 250 m resolution using the soilDB R package, and included sand, silt, and clay content at the top soil layer (0-5 cm below the ground).
+Soil type was determined based on the relative percentage of sand, silt, and clay based on USDA soil classifications using the *soiltexture* R package.
+Soil properties were then defined for the given soil class as default values provided by *micropoint*.
+
+### Microclimate analysis
+
+To examine global patterns of vertical microclimate profiles, we calculated the average minimum temperature of the coldest month, maximum temperature of the warmest month, minimum relative humidity of the warmest month, and diurnal temperature range of the coldest month.
+For each forest type (boreal, temperate, and tropical), we graphed the median climate value within vertical intervals, each representing 10% of the total canopy height.
+Therefore, values at the top of the canopy always occur in the 0.9-1 relative height interval regardless of the absolute height of the canopy.
+We additionally examined diurnal temperature fluctuations across forest types by taking the mean value across points for a given relative height and hour of the day.
+
+### Software
+All analyses were conducted using R Statistical Software (v4.4.1; @rcite).
 
@@
 
-```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab"}
-# table from: "scripts/03_analyses/01_canopyheight_richness_relationships.R"
+```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab".
+tab.cap="Coefficient estimates (with 95% confidence intervals) and conditional percent deviance explained (CPDE) for global models of richness as a function of canopy height for all taxa and each vertebrate class and for biome models of richness as a fucntion of canopy height x biome.
+Biome models used a means parameterization such that the effect of biome represents the mean estimated richness at a canopy height of zero and the interaction effect represents the estimated impact of canopy height on richness for each biome."}
 
-fit.global.tidy %>% 
-  mutate(term = ifelse(term == "canopy_height2", "canopy height", term)) %>% 
-  mutate(across(.cols = estimate:p.value, .fns = function(x) {round(x,2)})) %>% 
-  rename(SE = std.error, "p-value" = p.value) %>% 
-  flextable()
+# tables from: "scripts/03_analyses/01_canopyheight_richness_relationships.R"
+
+tidy.global = fit.global.tidy %>% 
+  mutate(term = ifelse(term == "canopy_height2", "Canopy height", term),
+         taxa = "All vertebrates") %>% 
+  mutate(across(.cols = estimate:conf.high, .fns = function(x) {round(x,2)})) %>% 
+  mutate(estimate = ifelse(term != "CPDE (%)",
+                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
+                           paste0(estimate))) %>% 
+  dplyr::select(taxa, term, estimate) %>% 
+  pivot_wider(names_from = "taxa", values_from = "estimate")
+
+tidy.global.taxa = global.taxa.tidy %>% 
+  dplyr::select(taxa, term, estimate, conf.low, conf.high) %>% 
+  mutate(term = ifelse(term == "canopy_height2", "Canopy height", term)) %>%
+  mutate(across(.cols = estimate:conf.high, .fns = function(x){round(x,3)})) %>% 
+  arrange(taxa) %>% 
+  mutate(estimate = ifelse(term != "CPDE (%)",
+                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
+                           paste0(estimate)),
+         taxa = factor(taxa, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
+  dplyr::select(taxa, term, estimate) %>% 
+  pivot_wider(names_from = "taxa", values_from = "estimate")
+
+tidy.global = left_join(tidy.global, tidy.global.taxa, by = "term") %>% 
+  mutate(model = "Global model") %>% 
+  relocate(model)
+
+tidy.biome = biome.mod.tidy %>% 
+  dplyr::select(taxa, term, estimate, conf.low, conf.high) %>% 
+  mutate(term = gsub("e2", "e = ", term),
+         term = gsub(":", " x ", term),
+         term = gsub("2", "", term)) %>% 
+  mutate(across(.cols = estimate:conf.high, .fns = function(x){round(x,3)})) %>% 
+  mutate(estimate = ifelse(term != "CPDE (%)",
+                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
+                           paste0(estimate)),
+         taxa = factor(taxa, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
+  dplyr::select(taxa, term, estimate) %>% 
+  pivot_wider(names_from = "taxa", values_from = "estimate") %>% 
+  mutate(model = "Biome model") %>% 
+  relocate(model)
+
+tidymods = bind_rows(tidy.global, tidy.biome)
+
+tidymods %>% 
+  flextable() %>% 
+  hline(i = 3, part = "body", border = fp_border(color = "black", width = 3)) %>% 
+  merge_v(j = c("model")) %>% 
+  autofit()
 
@@
 
-```{r vert-model-results, tab.id="vert-model-results", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}
+```{r vert-model-results, tab.id="vert-model-results", tab.lp="stab", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}
 
@@
 
+```{r splot-microtempdiu, fig.id="microtempdiu", fig.lp="suppfig", width = 120, height = 70, fig.cap="Average hourly microclimate across relative heights in the canopy.
+Relative height is calculated as absolute height above ground divided height of the canopy."}
+
+include_graphics("figures/ms_figures/supp_figs/microclim_temp_diu_mintempMonth.png", error = F)
+
+```
+
 \newpage
@@
 ## Reference
-
-# SCRATCH SPACE
-
-"The observer who leaves the center of the earth by an infinitely small amount compared to the
-radius can reach a new world, so to speak, and he can observe more variations in the aspect of the
-soil and more modifications in the atmosphere than he would if he were to pass from one latitude to
-another" - Humbolt pg 77 Essay on the Geography of Plants
-
-"Current hypotheses for large-scale geographical patterns of species diversity fail to explain the
-extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5). We call
-this challenge “Humboldt’s enigma.”" - Rahbek
-
--   TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE
-    IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE
-    REGION
--   on second thought, I don't think this is necessary because we want to know why more species
-    can't exist in trees in boreal regions
-
-Just as topographic complexity and climate explain Humbolt's enigma in mountains, vegetative
-complexity and climate may play key roles in mediating the relationship between canopy height and
-species richness by impacting the availability of vertical niche opportunities and physiologically
-constraining arboreal lifestyles [CITE]. Densely layered forest structures not only provide habitat
-across the forest profile, but produce vertical microclimate gradients via interactions with
-macroclimate conditions, solar radiation, and wind speed. From the humid and thermally stable
-understory to the dry and thermally variable canopy that experiences both hot and cold temperature
-extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or
-hinder the capacity for tall forests to generate high levels of biodiversity. For example, increased
-exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat
-that can support arboreal species and thus limit richness despite tall canopies
-[@oliveiraVerticalStratificationInfluences2019]. Moreover, relatively low thermal variability in the
-canopies of tropical forests may increase the distinctness of vertical climate bands that support
-high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics
-[CITE].
-
-in wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates,
-which reduce thermal and hydric stress and may allow species to finely partition the entire
-three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in
-tropical regions could thus support higher animal verticality and richness. In contrast, extreme
-cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles,
-constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
-Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space,
-thereby limiting vertical stratification and species richness despite the presence of vegetation.
-
-Integrating new technologies can shed light on the mechanisms driving decades-old ecological
-patterns.
-
-But only within the last decade have we been able to measure forest structure and acquire data on
-vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive
-understanding of the impact of vertical habitat structure on species richness patterns.
-
-In forests, species assemblages stratify vertically along resource and microclimate gradients that
-extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a;
-@xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests increase both habitat volume and
-niche opportunities, enabling species to partition habitat across vertical strata akin to
-multi-floored apartments [@macarthurLimitingSimilarityConvergence1967;
-@macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-
-In contrast, lower structural complexity in parts the Afrotropics
-[@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding
-less vertical assemblages and lower richness than expected based on the height of the canopy (Fig.
-\@ref(fig:fig1)).
-
-where verticality declines, species may use terrestrial habitats during the dry season to maintain
-moisture requirements assemblage verticality declines because species exhibit semi-arboreal life
-strategies, which reduces assemblage verticality.
-
-"short tropical forests show higher complexity relative to other forests of similar heights" - de
-Conto et al. 2023
-
--   highest risk for to exceeding critical maxima in tropical lowlands and mid-latitudes (Sunday et
-    al. 2014)
--   tolerance of max temps remains stable across latitudes - suggests advantage for temperate
-    ectotherms (Sunday et al. 2011)
--   increasing tolerance of low min temps with latitude (Sunday et al.
-    2011)
-
-          -   Declines in tolerance of low minimum temperatures toward the equator suggest that the
-              cooler temperatures in the canopy may limit arboreality
--   in terrestrial ectotherms, upper thermal limits decline marginally with latitude, while lower
-    thermal limits decline steeply with latitude (Sunday et al. 2019) - so upper thermal limits most
-    problematic in hot environments where species have lower thermal safety margins - safety margin
-    will be even lower in the canopy
-    -   steep vertical gradients in maximum temperatures reduce thermal safety margins, which may
-        pose challenges to fully arboreal life strategies, particularly in temperate regions where
-        maximum temperature gradients were greatest (fig) and where ectotherms are at the highest
-        ris for exceeding critical thermal maxima (Sunday et al. 2014)
--   operative temps are typically higher than critical thermal maxima, suggesting that terrestrial
-    ectotherms must rely on behavioral adjustments to survive extreme heat (Sunday et al. 2014) -
-    this would force terrestriality, particularly in temperate latitudes where the vertical climate
-    gradient is very steep
--   max air temps highest at mid-latitudes (Sunday et al. 2014) - I can show this with a map in the
-    supp figures
--   heat tolerance varied little with latitude (Sundary et al. 2014)
--   upper thermal tolerance limits were generally lower than maximum operative temps for ectotherms
-    (Sunday et al. 2014)
--   cold tolerance limits generally lower than lowest operative temps (Sunday et al. 2014)
--   reptiles usually have operative temps higher than critical maxima when in the sun, so must use
-    behavioral thermoregulation - particularly at lower to mid latitudes (Sunday et al. 2014) -
-    Therefore, in temperate forests where vertical max temp gradients are steep and canopies are
-    drier than the tropics, with canopies exceeding xC, high temps may prevent verticality
--   amphibians can have operative temps that remain below critical maxima when water is available
-    (Sunday et al. 2014)
--   burrowing or evapotranspiration can reduce body temps (Sunday et al.
-    2014) 
--   when water is not available, amphibians are forced to seek shade under high temps to maintain
-    temps below thermal safety margin (Sunday et al. 2014)
--   at high latitudes, reptile operative temperatue declines below CTmin, so they need to burrow to
-    stay warm enough (Sunday et al.
-    2014)

### 2025-07-28 manuscript.Rmd
adjust figure 1 to show all taxa for scatter plot

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
+knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
 library(officedown)
@@
 
-hook_figure_unit("mm")
+hook_figure_unit("mm") # sets figure units to mm
 ```
@@
 
-\newpage
+::: {custom-style="Author"}
 
@@
 
-3 Conservation Research Institue, University of Cambridge, Cambridge, UK
+3 Conservation Research Institute, University of Cambridge, Cambridge, UK
 
@@
 
+:::
+
 \newpage
@@
 
+::: {custom-style="Normal"}
+
+Three-dimensional landscape structure, characterized by mountains and trees, plays a key role in shaping species richness patterns.
+Taller forests have been proposed to support high levels of biodiversity by increasing habitat volume and producing vertical zones of microclimate and resource availability that allow species to partition niche opportunities from the ground to the canopy.
+Yet the capacity for taller forests to support greater biodiversity is inconsistent globally.
+Here, we examine this pattern and the climatic mechanisms from macro to micro scales may mediate the capacity for tall canopies to increase species richness via impacts on the vertical structure of vertebrate communities.
+Overall, we find that tall canopies support greater biodiversity globally, but high precipitation allows them to yield more rapid increases in richness in the wet tropics by supporting structurally complex vegetation and reducing physiological constraints imposed by dry canopy microclimates.
+
+:::
+
 \newpage
 
-For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
-In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global biodiversity patterns as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019].
-The extraordinarily high biodiversity of mountains and their enhanced capacity to generate species richness in the tropics has been termed "Humbolt's enigma", and can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019].
-The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover.
+::: {custom-style="Normal"}
+
+For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a].
+In 1799, Alexander von Humboldt began an expedition that would ultimately canonize the role of mountains as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019].
+This pattern—'Humboldt's enigma'—can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019].
+The vertical nature of mountains provides more land surface per unit area than flat regions and brings distinct climates within close proximity of each, which produces rapid rates of species turnover.
 Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].
 
-Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogeographic patterns [@macarthurBirdSpeciesDiversity1961; @fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a].
+Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961].
 Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains.
 By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].
-Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967].
-This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024].
-Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
-The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical dimension to Humbolt's enigma.
+This vertical stacking of species with increasing arboreal niche opportunities has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024].
+Yet taller canopies are not consistently associated with higher biodiversity across the globe [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents 'MacArthur's enigma'.
 
-Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests.
-Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages.
-For example, denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019].
-Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes.
-Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019].
-Yet only recently have we been able to address these hypotheses, as our ability to quantify variation in vegetation structure and microclimates globally and examine their impacts on biogeographic patterns in three-dimensional space has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates any where on earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].
+Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to produce species rich assemblages remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and land-use change on biodiversity, and the conservation value of forests.
+Following explanations of Humboldt's enigma [@rahbekHumboldtsEnigmaWhat2019], we examine how climate from macro to micro scales mediates the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages.
+For example, climatic conditions that promote structurally complex canopies may provide more niche opportunities, allowing tall canopies to support more biodiversity than structurally simpler canopies of equivalent height [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019].
+Moreover, physiological constraints imposed on arboreal life strategies by dry, hot, or cold canopy microclimates combined with the distinctness of vertical climate bands may impact species turnover from the ground to canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019].
+Only recently have we been able to address these hypotheses, as our ability to quantify global variation in 3D vegetation structure, microclimates, and animal community structure has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates anywhere on Earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].
 
-In this study we combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities.
-We first investigated the capacity for taller canopies to support increasing species richness in forests and less densely vegetated woody biomes across latitudes.
-We find that tall canopies are better generators of biodiversity in the tropics, particularly in moist forests (Fig.
-\ref(fig:fig1)).
-We then examined the processes driving this pattern using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and process-based microclimate models [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] for smaller regions that integrate forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023].
-While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.
+We combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how climate from macro to micro scales may mediate the capacity for tall forests to support diverse assemblages via impacts on the vertical structure of vertebrate communities.
+We first investigate the capacity for taller canopies to support increasing species richness in woody biomes across latitudes.
+We then examine the processes driving observed patterns using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and regional models of vertical microclimate profiles [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025], which integrate forest structure data from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023] to estimate temperature and relative humidity from the ground to the canopy.
 
-# Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics
+:::
+
+# Tall canopies generate biodiversity, but more so in the tropics
 
@@
 
-To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of vertebrate richness as a function of canopy height globally for all vertebrate combined and each vertebrate class (birds, mammals, reptiles, and amphibians).
-We then mapped the deviance residuals of these models to determine where canopy height over- or under-predicted species richness.
-To examine variation in the capacity for tall canopies to support rich assemblages across latitudes, we produces separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands.
+::: {custom-style="Normal"}
 
-Globally, taller trees, which increase habitat volume, support more species (Fig.
-S\@ref(fig:global-chrich)).
-However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs.
+To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of global vertebrate richness as a function of canopy height for all vertebrates combined and each vertebrate class (birds, mammals, reptiles, and amphibians), mapping the deviance residuals to determine where canopy height over- or under-predicted species richness.
+Globally, taller canopies support more species (Fig.
+S\@ref(fig:global-chrich); Table S\@ref(tab:richness-mod-results)), because they increase habitat volume and complexity, which allows fine-scale niche partitioning from the ground to the canopy [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a].
+Yet global models of vertebrate richness by canopy height generally overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs.
 \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)).
 Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions.
-Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig.
-\ref(fig:fig1)d).
-For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions.
-Within the tropics, increasing canopy height generally yields greater gains in richness in forests than in drier woodlands, savannas, and shrublands, though baseline richness in areas with short vegetation are similar (Fig.
-\@ref(fig:fig1)).
-For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species.
-The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs.
+
+Models that allow impacts of canopy height to vary across biomes indicate that increasing canopy height yields greater gains in richness in the wet tropics than at higher latitudes or drier tropical biomes (Fig.
+\@ref(fig:fig1)d).
+For example, an increase in canopy height from 15 m to 25 m corresponds to an increase of `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` bird and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibian species in tropical forests, but only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` bird and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibian species in boreal regions.
+The difference in impacts of taller canopies on biodiversity between wet and dry tropical regions is particularly evident in Madagascar, where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs.
 \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)).
-Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier or higher latitude regions, mirroring impacts of mountains [@rahbekHumboldtsEnigmaWhat2019].
-Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics.
+Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier regions or higher latitudes.
+Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics - a pattern mirroring that of mountains [@rahbekHumboldtsEnigmaWhat2019].
 
-[ADD MODEL TABLES TO SUPPLEMENT]
+:::
 
@@
 
-The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a].
+::: {custom-style="Normal"}
+
+Verticality—the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]—peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig.
+\@ref(fig:fig2), S\@ref(fig:meanvert-biome)).
+Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig.
+S\@ref(fig:meanprop-biome)).
+Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig.
+\@ref(fig:fig2)).
+This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
+Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling MacArthur's enigma.
+
+The vertical layering of species within forests may mediate the capacity for tall forests to support diverse assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a].
 As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].
 An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017].
-Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019].
-To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017].
+Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support diverse assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019].
+To quantify the verticality of an assemblage, we used the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017].
 For species occurring in multiple habitats, we averaged scores across habitats.
@@
 
-Verticality peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig.
-\@ref(fig:fig2), S\@ref(fig:vertmean-biome)).
-Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig.
-S\@(fig:meanprop-biome)).
-Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig.
-\@ref(fig:fig2)).
-This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.
-Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.
+:::
 
-# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities
+# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical assemblages
 
-To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages.
-Macroclimate variables included those that may induce physiological constraints - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods).
-We additionally included vegetation density in the models, which impacts the availability of habitat volume.
-While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models.
-The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across vertical forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018].
-While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact vertical assemblage structure and richness patterns [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020].
-We therefore additionally modeled vertical microclimate profiles in six 1° x 1° grid cells within forested regions from tropical to boreal latitudes.
-To do so, we integrated information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025].
-To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within the modeled region (Supplementary methods).
-By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.
+::: {custom-style="Normal"}
 
-Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig.
-\@ref(fig:fig3), Table S\@ref(tab:vert-model-results)).
-These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal lifestyles, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.
+The microclimate models integrate forest structure data derived  While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.
+
+To examine environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages.
+Macroclimate variables included those that may impact vegetation structural complexity or place physiological constraints on arboreal life strategies - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods).
+We additionally included vegetation density in the models, which impacts habitat volume.
+While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables were strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models.
+
+While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, macroclimate maps do not represent the microclimatic conditions experienced across vertical forest profiles that directly influence the vertical structure of animal assemblages [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018; @scheffersIncreasingArborealityAltitude2013].
+We therefore additionally modeled vertical microclimate profiles in six 1° x 1° regions from tropical to boreal forests.
+To do so, we integrated information on the vertical distribution of vegetation derived from the GEDI spaceborne lidar [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] (Supplementary methods).
+By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and diverse animal assemblages.
+
+Overall, our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig.
+\@ref(fig:fig3), S\@ref(fig:preds-univariate)-\@ref(fig:preds-interaction); Table S\@ref(tab:vert-model-results)).
+These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal strategies, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.
+
+:::
 
@@
 
-Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014].
-Our models indicated that high dry season precipitation positively impacts assemblage verticality across taxa (Fig.
+::: {custom-style="Normal"}
+
+Climate impacts the structural complexity of vegetation, which influences the availability of arboreal niche opportunities and is typically positively associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014; @goetzLaserRemoteSensing2007].
+Our models indicated that high dry season precipitation positively affects assemblage verticality across taxa (Fig.
 \@ref(fig:fig3)).
-Wet climates with low precipitation seasonality enable the development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021].
-Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004].
+Wet climates with low precipitation seasonality support tall and structurally complex forests with densely layered vegetation [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], as well as diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004].
 Hotspots of structural complexity in wet regions that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].
@@
 
-[ADDITIONAL CITATIONS] - <https://doi.org/10.1016/j.rse.2006.11.016>
--<https://doi.org/10.1111/1365-2745.13944>
+:::
 
@@
 
+::: {custom-style="Normal"}
+
 Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019; @leahyVerticalNicheElevation2021; @vinodThermalSensitivityForest2023].
-By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a remarkable volume of climatic variation given their geographic footprint [@defrenneForestMicroclimatesClimate2021].
-These steep climate gradients can support high rates of turnover from the ground to the canopy that yield vertically stratified and species rich assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018].
+By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a substantial volume of climatic variation from the cool and wet understory to the climatically variable canopy, which experiences drier, hotter, and colder conditions [@defrenneForestMicroclimatesClimate2021].
+These steep climate gradients can support high rates of species turnover from the ground to the canopy that yield vertically stratified, diverse assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018].
 However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness [@oliveiraVerticalStratificationInfluences2019].
@@
 Constraints of low dry season precipitation on verticality (Fig.
-\@ref(fig:fig3)) were accompanied by steep gradients in minimum relative humidity, which microclimate models estimated declined x% from the ground to the top of the canopy (Fig.
+\@ref(fig:fig3)) were accompanied by steep gradients in minimum relative humidity (Fig.
 \@ref(fig:fig4)).
-Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot or cold climates, where evapotranspiration to maintain body temperatures below thermal maxima or basking to maintain body temperatures above thermal minima can quickly lead to desiccation without adequate physiological defenses [@scheffersIncreasingArborealityAltitude2013; @tracyNotJustSmall2010; @mckechnieAvianThermoregulationHeat2017].
-By imposing environmental filters on the functional morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage verticality and limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [@curriePredictionsTestsClimatebased2004].
-For example, in the seasonally dry forests of of Panama, water limitation in the canopy causes the vertical stratification of frog communities to collapse toward the ground during the dry season - a shift coincident with declines in richness and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020].
+Water scarcity risks desiccation in hot or cold climates, where evapotranspiration to maintain body temperatures below thermal maxima or basking to maintain body temperatures above thermal minima can quickly lead to desiccation without adequate physiological defenses [@scheffersIncreasingArborealityAltitude2013; @tracyNotJustSmall2010; @mckechnieAvianThermoregulationHeat2017].
+Dry canopy microclimates combined with low precipitation inputs therefore impose environmental filters on the functional morphologies that can inhabit canopy environments, which may limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [@curriePredictionsTestsClimatebased2004].
+For example, in the seasonally dry forests of Panama, water limitation in the canopy causes the vertical stratification of frog communities to collapse toward the ground during the dry season - a shift coincident with declines in richness and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020].
 Dry climates can therefore reduce the capacity for increasingly tall canopies to support greater biodiversity (Fig.
@@
 Maximum and minimum temperatures had less consistent effects across taxa (Fig.
-\@ref(fig:fig3)), but may still place physiological constraints on arboreal life strategies that influence the capacity for taller canopies to support richer species assemblages, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures.
-For example, high maximum temperatures generally reduced verticality and/or reduced the positive impact of canopy height on verticality, barring mammals for which verticality responded positively to maximum temperatures.
+\@ref(fig:fig3)), but may still place physiological constraints on arboreal life strategies that influence the capacity for taller canopies to support richer species assemblages, particularly in regions exhibiting steeper vertical temperature gradients.
+For example, high maximum temperatures generally reduced verticality and/or reduced the positive impact of canopy height on verticality, barring mammals which exhibited the opposite response (Figs.
+\@ref(fig:fig3), S).
 High temperatures may cause ectotherm body temperatures to exceed critical thermal limits [@sundayThermalsafetyMarginsNecessity2014].
-While consistent water availability in the canopies of tropical wet forests may allow ectotherms to maintain body temperature below critical limits, the drier conditions of temperate forests combined with steeper vertical gradients in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats [@sundayThermalsafetyMarginsNecessity2014] (Fig.
+While consistent water availability in the canopies of tropical wet forests may allow amphibians to maintain body temperature below critical limits [@sundayThermalsafetyMarginsNecessity2014], the drier conditions of temperate forests combined with steeper vertical gradients in maximum temperatures may impede use of the canopy and require that animals thermoregulate by moving to terrestrial or fossorial habitats [@sundayThermalsafetyMarginsNecessity2014; @scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020] (Fig.
 \@ref(fig:fig4)).
-Moreover, while models indicated that higher minimum temperatures were associated with lower verticality for birds and amphibians because verticality is low in hot but drier tropical regions while remaining high in cooler but wet tropical montane regions, low minimum temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig.
+
+Moreover, low minimum temperatures reduced positive effects of canopy height on verticality for birds and reptiles despite being associated with lower verticality for mammals and amphibians due to high minimum temperatures in hot and dry regions (Fig.
 \@ref(fig:fig3)).
-The interaction with canopy height reflects the physiological constraints that cold temperatures can impose on arboreal life strategies.
-Cold temperatures at higher latitudes can cause ectotherm body temperatures to exceed critical thermal minima outside fossorial habitats or impose energetic constraints on endotherms, particularly in the windier conditions of the upper canopy [@sundayThermalsafetyMarginsNecessity2014; @villen-perezWoodlandMediterraneanBirds2014; @wolfThermalEffectsRadiation1996].
-For example, birds that typically forage in the upper canopy shift to lower canopy strata during periods of cold winter temperatures [@villen-perezWoodlandMediterraneanBirds2014].
-Cold temperatures therefore limit the degree to which animals can specialization to upper canopy strata and consequently prevent taller canopies from increasing species richness.
+The interaction between minimum temperatures and canopy height reflects physiological constraints on arboreality.
+Extreme cold at higher latitudes may reduce ectotherm body temperatures below critical thermal minima or impose energetic constraints on endotherms, particularly in the colder and windier conditions of the upper canopy [@sundayThermalsafetyMarginsNecessity2014; @villen-perezWoodlandMediterraneanBirds2014; @wolfThermalEffectsRadiation1996] (Fig.
+\@ref(fig:fig4)).
+To thermoregulate, animals may forage in lower canopy positions or use fossorial habitats [@sundayThermalsafetyMarginsNecessity2014; @villen-perezWoodlandMediterraneanBirds2014].
+Cold temperatures therefore limit the degree to which animals can specialize to upper canopy strata and consequently prevent taller canopies from increasing species richness.
 
-Temporal climate variability may also play a role in the capacity for verticality to mediate relationships between species richness and canopy height by impacting climatic distinctness.
+Temporal climate variability may also influence the capacity for verticality to mediate relationships between species richness and canopy height by impacting climatic distinctness.
 In montane regions, climatic distinctness across elevations declines with latitude due to increasing seasonal temperature variation toward the poles [@rahbekHumboldtsEnigmaWhat2019; @janzenWhyMountainPasses1967].
 Low seasonal variation in the tropics produces narrow and distinct climate bands that exhibit little overlap across elevations.
-Unique climates across elevations restrict species' range size, leading to high rates of turnover across elevations that allow tropical mountains to generate extraordinary levels of biodiversity [@janzenWhyMountainPasses1967; @sheldonFiftyYearsMountain2018; @ghalamborAreMountainPasses2006; @rahbekHumboldtsEnigmaWhat2019].
-In contrast, high thermal variation in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and thus high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019; @ghalamborAreMountainPasses2006].
+Unique climates across elevations restrict species' range size, leading to high rates of turnover over short distances that yield extraordinary levels of biodiversity in tropical montane regions [@janzenWhyMountainPasses1967; @sheldonFiftyYearsMountain2018; @ghalamborAreMountainPasses2006; @rahbekHumboldtsEnigmaWhat2019].
+In contrast, greater temperature seasonality in temperate regions increases the similarity between climates experienced at high and low elevations, reducing the capacity for temperate mountains to support high rates of turnover and thus high levels of biodiversity [@rahbekHumboldtsEnigmaWhat2019; @ghalamborAreMountainPasses2006].
 
-Within wooded habitats, diurnal temperature range may influence the distinctness of vertical climate bands from the ground to the canopy, and thus the vertical turnover of species.
+In forests, diurnal temperature range may influence the distinctness of vertical climate bands from the ground to the canopy, and thus the vertical turnover of species.
 Our models of verticality indicate that greater diurnal temperature variation reduces verticality of mammals, though effects for other classes were insignificant (Fig.
 \@ref(fig:fig3)).
-Furthermore, microclimate models indicated that increases in daily temperature variation, as well as the rate of increase in temperature variation from the ground to the canopy, were greater in higher latitude forests (Fig.
-\@ref(fig:fig4)c).
-High daily temperature variation reduced the distinctness of climate bands across the vertical profile, with vertical zonation collapsing nearly completely during morning hours in temperate and boreal forests (Fig.
-\@ref(fig:fig4)B).
-Without unique climatic niches across the forest profile, these regions are unlikely to support rapid vertical turnover, limiting the number of species that may be added to the assemblage for a given increase in canopy height, even in wetter temperate regions with climate conditions that physiologically support arboreal strategies.
-In contrast, tropical canopies maintained more distinct vertical climate bands throughout the day, with the lowest and highest bands exhibiting little to no temperature overlap.
-Narrow and distinct vertical climate niches may therefore support rapid vertical turnover, leading to greater increases in richness with canopy height and enhancing the capacity for tropical canopies generate biodiversity.
+Furthermore, microclimate models indicated that lower daily temperature variation and greater decoupling between canopy and understory temperatures generally reduces thermal overlap across vertical strata in tropical forests relative to temperate and boreal forests (Fig.
+\@ref(fig:fig4)b, S\@ref(fig:microtempdiu)).
+Without unique climatic niches across the forest profile, high latitude forests are unlikely to support rapid vertical turnover, limiting the number of species that may be added to the assemblage for a given increase in canopy height, even in wetter temperate regions with climate conditions that physiologically support arboreal strategies.
+In contrast, tropical canopies maintained more distinct vertical climate bands throughout the day, which may support rapid vertical turnover, leading to greater increases in richness with canopy height and enhancing the capacity for tropical canopies generate biodiversity.
 
@@
 
+:::
+
 # Canopy height in a changing world
+
+::: {custom-style="Normal"}
 
@@
 We bring the vertical dimension of habitats to the forefront of global biogeography by showing that the climatic characteristics of wooded ecosystems from macro- to micro-scales mediate the capacity for tall forests to support high levels of biodiversity via impacts on the vertical structure of vertebrate assemblages.
-Tropical rainforests, with wet and thermally stable climates that increase the structural complexity of vegetation, reduce physiological stress in the canopy, and produce distinct vertical climate bands, are able to support fine-scale partitioning of resources and climatic niches throughout the entire forest volume.
+Tropical rainforests, with wet and thermally stable climates that increase the structural complexity of vegetation, reduce physiological stress in the canopy, and produce more distinct vertical climate bands, can support fine-scale partitioning of resources and climatic niches throughout the entire forest volume.
 Consequently, taller trees in tropical forests support higher biodiversity and greater increases in species richness than temperate and boreal ecosystems of equivalent heights where climatic conditions reduce the structural complexity of vegetation, pose physiological constraints on arboreal lifestyles, and reduce vertical climate zonation from the ground to the canopy.
@@
 While tall forests, particularly in the tropics, are crucial for supporting species rich assemblages, regions containing tall forests are globally rare, under-protected, and threatened by anthropogenic disturbance and climate change [@huangPrioritizingGlobalTall2023; @bourgoinHumanDegradationTropical2024; @bennettLargerTreesSuffer2015; @rowlandDeathDroughtTropical2015].
-For example, disturbances from fire and logging produce persistent reductions of canopy height in the tropics [@bourgoinHumanDegradationTropical2024], with potential negative effects on the capacity for these forests to support high levels of biodiversity.
+Disturbances from fire and logging produce persistent reductions in canopy height in the tropics [@bourgoinHumanDegradationTropical2024], with potential negative effects on the capacity for these forests to support high levels of biodiversity.
 Furthermore, climate change is increasing the frequency of extreme heat and drought [@seneviratneWeatherClimateExtreme2021].
-In addition to directly reducing forest height [@bennettLargerTreesSuffer2015; @rowlandDeathDroughtTropical2015], these trends could truncate the volume of physiological suitability habitat for arboreal species.
-Climate change may therefore flatten the vertical structure of species assemblages toward the ground [@oliveiraVerticalStratificationInfluences2019; @scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], just as warming temperatures are causing upslope shifts in species' distributions [@lenoirSpeciesBetterTrack2020; @chanClimateVelocitiesSpecies2024].
+In addition to directly reducing forest height [@bennettLargerTreesSuffer2015; @rowlandDeathDroughtTropical2015], these trends could truncate the volume of physiologically suitable habitat for arboreal species.
+Climate change may therefore flatten the vertical structure of species assemblages toward the ground [@oliveiraVerticalStratificationInfluences2019; @scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020; @soiferMultidimensionalMicroclimateVelocities2025], just as warming temperatures are causing upslope shifts in species' distributions [@lenoirSpeciesBetterTrack2020; @chanClimateVelocitiesSpecies2024].
 Improving our understanding of the vertical structure of species assemblages as canopies heat up and dry out will be critical for evaluating the continued capacity of tropical forests to generate extraordinarily high levels of biodiversity.
+
+:::
 
@@
 LGS was supported by the NSF GRFP...
-We thank Patrick Burns for access to the processed GEDI L2b points.
+We thank Patrick Burns for access to the processed GEDI L2b points and Alex Baecher for assistance with the spatial statistics.
 
@@
 ```{r fig1, fig.id="fig1", fig.cap="Spatial distribution of a) canopy height, b) vertebrate richness (for all vertebrates), and c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height.
-d) Relationships between species richness and canopy height for each taxon in different biomes.
-Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=180}
+d) Relationships between species richness of all vertebrates and canopy height across biomes.
+Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=140}
 
@@
 
-include_graphics("figures/ms_figures/fig1_canopy_richness.png", error = F)
+include_graphics("figures/ms_figures/fig1-2_canopy_richness.png", error = F)
 
@@
 b) The verticality of vertebrate communities in wooded biomes across the globe.
-High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance." c) Correlations between verticality and species richness globally.
-Lines indicate linear best fit line with 95% CIs., fit.width = 150, fig.height=150}
+High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance.
+c) Correlations between verticality and species richness globally.
+Lines indicate linear best fit line with 95% CIs.", fig.width = 150, fig.height=150}
 
@@
 c) Vertical profiles of microclimate variables and plant area volume density (PAVD) (median +/- SD) in tropical, temperate, and boreal forest sites.
-PAVD was obtained from GEDI shots.", fig.width=140, fig.height=120}
+PAVD was obtained from GEDI shots.", fig.width=180, fig.height=110}
 
@@
 
-## Assmeblage richness and verticality
+## Assemblage richness and verticality
 
@@
 Overall, we obtained range maps for `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians.
+Wooded habitats included  all forested biomes, as well as ecoregions identified as forests, pine,  woodlands, woods, taiga, espinal, várzea, yungas, savanna, thicket, matorral,  chaparral, mezquital, pineyards, pantanal, the cuban wetlands, cerrado, bushveld, chaco, lowveld, and mallee as defined by Dinnerstein et al.
+[-@dinersteinEcoregionBasedApproachProtecting2017].
+We did not include mangroves given their lack of terrestrial and fossorial habitat.
 
@@
 For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024a].
+The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature.
+For species where imputation was necessary, we considered a species to use a given vertical stratum if over 70% of the 10,000 imputations predicted use of the stratum.
 For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale vertical classification than the dataset from Moura et al.
@@
 where the response variable at location $s$ is normally distributed with mean $\mu_s$ and variance $\sigma^2$; $𝜇_s$ is the expected verticality.
-$Χ_s$ is a design matrix at location $s$; 𝛽 is a vector of fixed-effect coefficients; $⍵_s$ represents a spatial random effect at location $s$; and $\Sigma_{\omega}$ represents the covariance matrix of the spatial random field.
+$Χ_s$ is a design matrix at location $s$; 𝛽 is a vector of fixed-effect coefficients; $⍵_s$ represents a gaussian spatial random field at location $s$; and $\Sigma_{\omega}$ represents the covariance matrix of the spatial random field.
 The SPDE method requires a finite element analysis mesh, which consists of a set of triangles covering the spatial domain, and allows for the approximation of the spatial random effect at mesh vertices [@thorsonSpatioTemporalModelsEcologists2024].
-We created a mesh for each vertebrate class using *fmesher* and tuned each mesh, such that the maximum edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].
+The mesh used to approximate the spatial random field impacts model results as well as computation time [@andersonSdmTMBPackageFast2024].
+To produce an optimal mesh that would avoid over-smoothing the spatial random field and avoid over-fitting the model and unnecessarily increasing computational demands, we used an iterative procedure to update minimum and maximum edge lengths of the triangles in the mesh based on the spatial range of the data (i.e., the distance at which points exhibit a correlation of approximately 0.1) [@thorsonSpatioTemporalModelsEcologists2024].
+The tuning procedure for the mesh was conducted for each taxon using *fmesher* and consisted of the following steps.
+First, we identified the spatial range of the verticality using a correlogram on a subset of 1000 points.
+Next, we constructed an initial mesh with a maximum edge length of the inner domain as 1/5 the spatial range, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range.
+To avoid small triangles, we also specified the minimum edge lenght as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019].
+We then fit a model using the resulting mesh.
+If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model.
+We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of the mesh.
 
 We evaluated models using the sanity check function in sdmTMB and by examining quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022].
-To determine the degree to which environmental factors improve model performance, we compare this model to a spatial random effect model with  model with We also report conditional percent deviance explained (CPDE).
+To determine the degree to which environmental factors improve model performance, we compare the models to spatial random effects models without environmental predictors and report AIC and conditional percent deviance explained (CPDE).
 
@@
 These months were identified using the ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023].
-For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile.
+For minimum and maximum temperature subsets, we then selected up to 100 GEDI points within each 1° x 1° tile from the top 75th to 98th quantile of canopy height.
 However, there were only 26 points in temperate USA and 14 points in temperate Europe during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite).
@@
 The ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023] was used for meteorological inputs to the model, which provided near surface, open-air climate conditions at a 0.25$^\circ$ spatial resolution and hourly temporal resolution.
-ERA5 climate data was processed using the *mcera5* R package, which converts ERA5 variables into those required by the model and applies a coastal diurnal temperature correction to terrestrial locations located within predominanrly marine grid cells [@klingesMcera5DrivingMicroclimate2022].
+ERA5 climate data was processed using the *mcera5* R package, which converts ERA5 variables into those required by the model and applies a coastal diurnal temperature correction to terrestrial locations located within predominantly marine grid cells [@klingesMcera5DrivingMicroclimate2022].
 Meteorological conditions at the GEDI point of interest were then extracted from the climate grid by taking the weighted average of climate conditions across the four nearest neighboring cells, where the weight was defined as the inverse distance to the point.
@@
 
-### Software
-All analyses were conducted using R Statistical Software (v4.4.1; @rcite).
-
-MASS was used to produce models of canopy height by richness. sdmTMB was used to produce models of
-verticality. DHARMa was used to check model fit and assumptions. ggeffects was used to generate
-model predictions and confidence intervals. Figures were generating using tidyterra, patchwork,
-ggplot2
+<!---BLOCK_LANDSCAPE_START--->
 
@@
 
-```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab".
-tab.cap="Coefficient estimates (with 95% confidence intervals) and conditional percent deviance explained (CPDE) for global models of richness as a function of canopy height for all taxa and each vertebrate class and for biome models of richness as a fucntion of canopy height x biome.
+```{r richness-mod-results, tab.id="richness-mod-results", tab.lp="stab", tab.cap="Coefficient estimates (with 95% confidence intervals) and conditional percent deviance explained (CPDE) for global models of richness as a function of canopy height for all taxa and each vertebrate class and for biome models of richness as a fucntion of canopy height x biome.
 Biome models used a means parameterization such that the effect of biome represents the mean estimated richness at a canopy height of zero and the interaction effect represents the estimated impact of canopy height on richness for each biome."}
@@
   merge_v(j = c("model")) %>%
+  set_table_properties(width = 1, layout = "autofit")
+
+```
+
+```{r vert-model-eval, tab.id="vert-model-eval", tab.lp="stab", tab.cap="Degrees of freedom, AIC, and conditional percent deviance explained (CPDE) for models of verticality by environmental predictors and spatial random effects and models of verticality by spatial random effects only.
+In the models, 'Fixed' refers to fixed environmental predictors and 'Spatial' refers to spatial random effects."}
+
+tab = read.csv("scripts/03_analyses/03_glmms/result_tables/compareMods.csv")
+
+tab %>% 
+  mutate(across(.cols = AIC:CPDE, function(x) {round(x,2)})) %>% 
+  rename(Model = name, "CPDE (%)" = "CPDE", "\u0394AIC" = "deltaAIC") %>% 
+  flextable() %>% 
+  merge_v(j = c("taxon")) %>% 
+  bg(i = c(1:3,7:9), j = 1:6, bg="grey80") %>% 
+  fix_border_issues() %>% 
   autofit()
@@
 
-```{r vert-model-results, tab.id="vert-model-results", tab.lp="stab", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}
+```{r vert-model-results, tab.id="vert-model-results", tab.lp="stab", tab.cap="Coefficient estimates (with 95% confidence intervals) for models of verticality as a function of standardized predictor variables."}
 
-amphmod = readRDS("results/sdmTMB_models2.3/amphibians_sesvert_tempdiu.rds")
-reptmod = readRDS("results/sdmTMB_models2.3/reptiles_sesvert_tempdiu.rds")
-birdsmod = readRDS("results/sdmTMB_models2.3/birds_sesvert_tempdiu.rds")
-mammalsmod = readRDS("results/sdmTMB_models2.3/mammals_sesvert_tempdiu.rds")
-
-amphmod = amphmod$fullmod$modlist[[1]]
-amphtab = tidy(amphmod) %>% 
-  mutate(Amphibians = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
-  dplyr::select(term, Amphibians)
-
-birdtab = birdsmod$fullmod %>% 
-  tidy() %>% 
-  mutate(Birds = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
-  dplyr::select(term, Birds)
-
-mammalstab = mammalsmod$fullmod %>% 
-  tidy() %>% 
-  mutate(Mammals = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
-  dplyr::select(term, Mammals)
-
-repttab = reptmod$fullmod %>% 
-  tidy() %>% 
-  mutate(Reptiles = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")) %>% 
-  dplyr::select(term, Reptiles)
-
-tab = plyr::join_all(list(birdtab, mammalstab, repttab, amphtab), by = "term", type = "left")
+# tab produced in "scripts/03_analyses/03_glmms/04_resultsTab.R"
+tab = read.csv("scripts/03_analyses/03_glmms/result_tables/modelCoefs.csv")
 
@@
 
+<!---BLOCK_LANDSCAPE_STOP--->
+
 \newpage
@@
 
-```{r splot-global-chrich, fig.id="global-chrich",  fig.lp="suppfig", fig.autonum.start_at=1, width=90, height=90, fig.cap="Global relationship between total vertebrate richness and canopy height.
+```{r splot-global-chrich, fig.id="global-chrich",  fig.lp="suppfig", fig.autonum.start_at=1, fig.cap="Global relationship between total vertebrate richness and canopy height.
 Line indicates predictions from a negative binomial model of richness by canopy height ± 95% CI.
-Points indicate observed richness by canopy height.}
+Points indicate observed richness by canopy height.", , fig.width=90, fig.height=90}
 
@@
 
-```{r splot-resids-taxa, fig.id="resids-taxa", fig.lp="suppfig", width=180, height=100, fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians."}
+```{r splot-resids-taxa, fig.id="resids-taxa", fig.lp="suppfig", fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians.", fig.width=180, fig.height=100}
 
@@
 
-```{r splot-meanvert-biome, fig.id = "meanvert-biome", fig.lp="suppfig", width = 180, height = 130, fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
+```{r splot-meanvert-biome, fig.id = "meanvert-biome", fig.lp="suppfig", fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
 (2017) and b) ecoregions aggregated to coarser classifications.
 Red diamonds indicate mean assemblage verticality.
-Ecoregions that do not contain woody vegetation were excluded."}
+Ecoregions that do not contain woody vegetation were excluded.", fig.width = 180, fig.height = 130}
 
@@
 
-```{r splot-meanprop-biome, fig.id="meanprop-biome", fig.lp="suppfig", width = 160, height = 160, fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
+```{r splot-meanprop-biome, fig.id="meanprop-biome", fig.lp="suppfig", fig.cap="Distribution of verticality within a) biomes as classified by Dinerstein et al.
 (2017) and b) ecoregions aggregated to coarser classifications.
 Red diamonds indicate mean assemblage verticality.
-Ecoregions that do not contain woody vegetation were excluded."}
+Ecoregions that do not contain woody vegetation were excluded.", fig.width = 160, fig.height = 160}
 
@@
 
-```{r splot-microtempdiu, fig.id="microtempdiu", fig.lp="suppfig", width = 120, height = 70, fig.cap="Average hourly microclimate across relative heights in the canopy.
-Relative height is calculated as absolute height above ground divided height of the canopy."}
+```{r splot-preds-univariate, fig.id="preds-univariate", fig.lp="suppfig", fig.cap="Conditional plots of model predictions for assemblage verticality of each vertebrate class in response to different environmental predictors.
+All variables except for that displayed on the x-axis were held constant at their mean.
+Error bars represent 95% confidence intervals.", fig.width=180, fig.height=120}
+
+include_graphics("figures/ms_figures/supp_figs/preds_univariate.png", error = F)
+
+```
+
+```{r splot-preds-interaction, fig.id="preds-interaction", fig.lp="suppfig", fig.cap="Conditional plots of model predictions for assemblage verticality of each vertebrate class across canopy heights and at different values of climatic interaction terms.
+Interaction terms were allowed to vary across their mean, -2 standard deviations, and +2 standard deviations, while all other predictions were held constant at their mean.
+Error bars represent 95% confidence intervals.", fig.width=180, fig.height=120}
+
+include_graphics("figures/ms_figures/supp_figs/preds_interaction.png", error = F)
+
+```
+
+```{r splot-microtempdiu, fig.id="microtempdiu", fig.lp="suppfig", fig.cap="Average hourly microclimate across relative heights in the canopy.
+Relative height is calculated as absolute height above ground divided height of the canopy.", fig.width = 120, fig.height = 70}
 
@@
 
-## Reference
+# References

### 2025-07-28 scripts/03_analyses/01_canopyheight_richness_relationships.R

@@ -5,12 +5,14 @@ library(dplyr)
 #library(foreach)
 library(terra)
 library(tidyterra)
+library(broom.mixed)
 #library(glmmTMB)
 library(MASS)
 library(ggplot2)
 library(patchwork)
 library(performance)
 library(ggeffects)
+library(colorspace)
 
 amph = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv")
 birds = read.csv("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv")
@@ -211,7 +213,8 @@ AIC(fit.global.all1, fit.global.all2) # rich~canopyheight better (fit.global.all
 mnull = MASS::glm.nb(total_richness~1, data = df_combined)
 fit.global.all.pd = (1-(fit.global.all1$deviance/mnull$deviance))*100
 
-nd = data.frame(canopy_height2 = c(min(df_combined$canopy_height2), seq(1,floor(max(df_combined$canopy_height2)),1)))
+nd = data.frame(canopy_height2 = c(min(df_combined$canopy_height2), 
+                                   seq(1,floor(max(df_combined$canopy_height2)),1)))
 fit.global.all.pred = predict_response(fit.global.all1, nd)
 
 # fit.resid = simulateResiduals(fit.global.all1, plot = T)
@@ -223,128 +226,40 @@ fit.global.tidy = tidy(fit.global.all1, conf.int = T, conf.level = 0.95)
 global.pd = data.frame(term = "CPDE (%)", estimate = fit.global.all.pd)
 fit.global.tidy = bind_rows(fit.global.tidy, global.pd)
 
-# Models by taxon and biome2 ----------------------------------------------
-# model richness by canopy height for each taxon in each of the coarser biome/lat classifications
 
-# biome.taxa = list() # data used in models with deviance residuals for m3 and m4 models
-# biome.taxa.pred.m3mod = list() # rich~canopy height predictions
-# biome.taxa.mod.m3mod = list() # rich~canopy height models
-# biome.taxa.pred.m4logmod = list() # rich~log(canopy height) predictions
-# biome.taxa.mod.m4logmod = list() # rich~log(canopy height) models
-# biome.taxa.mod.aicdif = list() # store difference in AIC values between model with canopy height and model with log(canopy height)
[…]

### 2025-08-26 scripts/03_analyses/01_canopyheight_richness_relationships.R
update global richness by canopy height plot to include global regression line

@@ -352,17 +352,26 @@ biome.mod.tidy = bind_rows(biome.mod.tidy, biome.pd)
 # Fig 1. REVISED
 
 # *- richness by canopy height for all taxa by biome ----------------------
+
+# combine global model df with biome model df for plotting
+global.all.pred = fit.global.all.pred %>% 
+  as.data.frame() %>% 
+  rename(canopy_height = x, richness = predicted, biome2 = group) %>% 
+  mutate(biome2 = "Global")
+
 biome.all.pred = fit.biome.all.pred %>% 
   as.data.frame() %>% 
+  bind_rows(global.all.pred) %>% 
   mutate(biome2 = factor(biome2, levels = c("Tropical forest", "Tropical woodland, savanna, and shrubland",
                                             "Temperate forest", "Temperate woodland, savanna, and shrubland",
-                                            "Boreal")),
+                                            "Boreal", "Global")),
          col = case_when(biome2 == "Tropical forest" ~ "#008234",
                          biome2 == "Tropical woodland, savanna, and shrubland" ~ "#50DE00",
                          biome2 == "Temperate forest" ~ "#E08A00",
                          biome2 == "Temperate woodland, savanna, and shrubland" ~ "#FFCB76",
-                         biome2 == "Boreal" ~ "#56B4E9"),
-         col = factor(col, levels = c("#008234", "#50DE00","#E08A00", "#FFCB76", "#56B4E9")))
+                         biome2 == "Boreal" ~ "#56B4E9",
+                         biome2 == "Global" ~ "black"),
+         col = factor(col, levels = c("#008234", "#50DE00","#E08A00", "#FFCB76", "#56B4E9", "black")))
 
 df_combined = df_combined %>% 
   mutate(biome2 = factor(biome2, levels = c("Tropical forest", "Tropical woodland, savanna, and shrubland",
@@ -464,14 +473,15 @@ total_resid.plt = ggplot() +
 
 
 total.global.map = canopy_height.plt / total_rich.plt / total_resid.plt +
-  plot_layout(nrow = 3)
+  plot_layout(nrow = 3) & theme(plot.tag.position = c(0.1, 0.9),
+                                plot.tag = element_text(size = 6))
 
+ch_rich.plt  = ch_rich.plt + theme(plot.tag.position = c(0.03, 0.95),
+                                   plot.tag = element_text(size = 6))
 
[…]

### 2025-08-26 manuscript.Rmd
Figure 1 caption update

@@
 The vertical nature of mountains provides more land surface per unit area than flat regions and brings distinct climates within close proximity of each, which produces rapid rates of species turnover.
-Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].
+Yet mountains are better generators of biodiversity in the tropics because they have higher elevation snow lines than high-latitude mountains, which increase the area of land suitable for diverse communities, longer climate gradients due to warmer lowlands, and more distinct elevational bands due to less seasonal temperature variation [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].
 
@@
 This vertical stacking of species with increasing arboreal niche opportunities has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024].
-Yet taller canopies are not consistently associated with higher biodiversity across the globe [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
-The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents 'MacArthur's enigma'.
+Yet MacArthur & MacArthur [ -@macarthurBirdSpeciesDiversity1961] noted over 50 years ago that tropical forests support greater bird diversity than temperate forests with comparable profiles, and more recent studies have revealed inconsistent effects of canopy height on biodiversity across the globe [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020].
+The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical component to Humboldt's enigma.
 
@@
 
-```{r fig1, fig.id="fig1", fig.cap="Spatial distribution of a) canopy height, b) vertebrate richness (for all vertebrates), and c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height.
-d) Relationships between species richness of all vertebrates and canopy height across biomes.
-Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=140}
+```{r fig1, fig.id="fig1", fig.cap="a) Relationships between species richness and canopy height across all vertebrate taxa globally and in different biomes.
+Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of canopy height (black) or richness as a function of the interaction between canopy height and biome.
+b-d) Spatial distribution of canopy height, vertebrate richness (for all vertebrates), and deviance residuals for the global model of vertebrate richness.", fig.height=100, fig.width=140}
 
@@
 We use *microclimf's* 'vegpfromhab()' function to estimate these parameters based on land cover type.
-Land cover was determined based on MODIS land cover type (MCD12Q1) version 6.1 for the year 2022.
+Land cover was determined based on Copernicus Global Dynamic Land Cover version 3 for the year 2019 [@buchhornCopernicusGlobalLand2020].
 
 Vegetation parameters were assumed to remain constant for the month and year in which the GEDI shot was obtained.
-
-[WE SHOULD PRODUCE MODELS AT DRIEST MONTH OF THE YEAR]
 

### 2025-08-26 scripts/03_analyses/01_canopyheight_richness_relationships.R
update supplementary figure 1 to combine taxon specific richness by canopy height maps with deviance residuals

@@ -486,92 +486,43 @@ ch_rich.plt  = ch_rich.plt + theme(plot.tag.position = c(0.03, 0.95),
 ggsave('figures/ms_figures/fig1-2_canopy_richness.png', width = 140, height = 100,
        units = "mm", dpi = 300)
 
-
-# Fig. 1 OLD ------------------------------------------------------------------
-
-# Legend: a) canopy height, b) vertebrate richness (all vertebrates combined)
-# c) Deviance residuals for a model of total vertebrate richness by canopy height
-# d) Relationships between species richness and canopy height for each taxon in each
-# biome category. Dots indicate observed values and lines indicate negative binomial
-# model fits of richness by canopy height for each combination of taxon and biome.
-
-# *- Total canopy, richness, and resid maps -------------------------------
-
-wd = vect("data/original/rnaturalearth_world.shp") %>% 
-  crop(ext(-180,180,-60,90)) %>% 
-  project("+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m")
-
-# Map of canopy richness, canopy height, and residuals for total_richness model
-df_combined.r = df_combined %>% 
-  dplyr::select(x,y,canopy_height2, total_richness, resid.deviance) %>% 
-  rast(crs = "+proj=cea +datum=WGS84") %>% 
-  project("+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m")
-
-total_rich.plt = ggplot() + 
-  geom_spatvector(data = wd, fill = "black", color = NA) +
-  geom_spatraster(data = df_combined.r, aes(fill = total_richness)) +
-  scale_fill_continuous_sequential(palette = "Sunset", na.value = NA,
-                                   guide = guide_colorbar(title = "")) +
-  theme_void() +
-  ggtitle("Vertebrate Richness") +
-  theme(legend.position = "inside",
-        legend.justification = c(0.1,0),
-        legend.key.size = unit(2, "mm"),
-        legend.text = element_text(size = 4),
-        legend.title = element_text(size = 4),
-        plot.title = element_text(hjust = 0.5, size = 6))
-  
-canopy_height.plt = ggplot() + 
-  geom_spatvector(data = wd, fill = "black", color = NA) +
-  geom_spatraster(data = df_combined.r, aes(fill = canopy_height2)) +
[…]

### 2025-08-26 manuscript.Rmd
Update supplementary figure 1 and caption

@@
 ```{r fig1, fig.id="fig1", fig.cap="a) Relationships between species richness and canopy height across all vertebrate taxa globally and in different biomes.
-Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of canopy height (black) or richness as a function of the interaction between canopy height and biome.
+Dots indicate observed values and lines indicate negative binomial model fits ± 95% CI of richness as a function of canopy height (black) or richness as a function of the interaction between canopy height and biome.
 b-d) Spatial distribution of canopy height, vertebrate richness (for all vertebrates), and deviance residuals for the global model of vertebrate richness.", fig.height=100, fig.width=140}
@@
 
-```{r splot-global-chrich, fig.id="global-chrich",  fig.lp="suppfig", fig.autonum.start_at=1, fig.cap="Global relationship between total vertebrate richness and canopy height.
-Line indicates predictions from a negative binomial model of richness by canopy height ± 95% CI.
-Points indicate observed richness by canopy height.", , fig.width=90, fig.height=90}
+```{r taxa-rich-ch, fig.id="taxa-rich-ch",  fig.lp="suppfig", fig.autonum.start_at=1, fig.cap=" a) Relationships between species richness and canopy height for birds, mammals, reptiles, and amphibians globally and in different biomes.
+Dots indicate observed values and lines indicate negative binomial model fits ± 95% CI of richness as a function of canopy height (black) or richness as a function of the interaction between canopy height and biome.
+b)  Deviance residuals from negative binomial models of global richness by canopy height for each vertebrate class.", fig.width=180, fig.height=200}
 
-include_graphics("figures/ms_figures/supp_figs/total_richness~canopy.png", error = F)
-```
-
-```{r splot-resids-taxa, fig.id="resids-taxa", fig.lp="suppfig", fig.cap="Deviance residuals from negative binomial models of richness by canopy height for each vertebrate class: a) birds, b) mammals, c) reptiles, d) amphibians.", fig.width=180, fig.height=100}
-
-include_graphics("figures/ms_figures/supp_figs/deviance_resids_taxa.png", error = F)
+include_graphics("figures/ms_figures/supp_figs/richness~canopyHeight_taxa.png", error = F)
 ```

### 2025-08-26 scripts/03_analyses/02_vertmaps_and_vertRichness_correlations.R
move richness axis to left

@@ -22,41 +22,46 @@ wd = vect("data/original/rnaturalearth_world.shp") %>%
 ymin = ext(wd)[3]
 ymax = ext(wd)[4]
 
-amph = fread("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
+amph1 = fread("data/derivative_data/gridcell_data/env_forest/50_km/amph_comdat.csv") %>% 
   dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
   filter(rich >= 5) %>% 
   filter(!is.na(vert.mean.ses)) %>% 
-  drop_na() %>% 
+  drop_na()
+  
+amph = amph1 %>% 
   rast(crs = "+proj=cea +datum=WGS84") %>% 
   project("+proj=robin +datum=WGS84") %>% 
   as.data.frame(xy = T) %>% 
   mutate(taxa = "Amphibians")
 
-mammals = fread("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
+mammals1 = fread("data/derivative_data/gridcell_data/env_forest/50_km/mammals_comdat.csv") %>% 
   dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
   filter(rich >= 5) %>% 
   filter(!is.na(vert.mean.ses)) %>% 
-  drop_na() %>% 
+  drop_na()
+mammals = mammals1 %>% 
   rast(crs = "+proj=cea +datum=WGS84") %>% 
   project("+proj=robin +datum=WGS84") %>% 
   as.data.frame(xy = T) %>% 
   mutate(taxa = "Mammals")
 
-birds = fread("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
+birds1 = fread("data/derivative_data/gridcell_data/env_forest/50_km/birds_comdat.csv") %>% 
   dplyr::select(x, y, rich, p.arb, p.fos, p.ter, vert.mean, vert.mean.ses, biome) %>% 
   filter(rich >= 5) %>% 
   filter(!is.na(vert.mean.ses)) %>% 
-  drop_na()%>% 
+  drop_na()
+birds = birds1 %>% 
   rast(crs = "+proj=cea +datum=WGS84") %>% 
   project("+proj=robin +datum=WGS84") %>% 
[…]
